---
title: "Prediction analyses - run prediction and outputs"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

**Language / Idioma:** [English](analysis_prediction_run_outputs.html) | [Português](analysis_prediction_run_outputs_pt.html)

**Module navigation / Navegação do módulo:** [Part I: setup, models and CV design](analysis_prediction.html)

## 1. Introduction

This second submodule concentrates the operational part of the prediction workflow. The goal here is to document how the prediction jobs are organized, where the outputs are written, and how those outputs are prepared for the downstream summaries consumed later by `visualization.Rmd` and `visualization.html`.

The workflow is intentionally divided into two layers:

- a **heavy execution layer**, which fits the 18 models under `CV1`, `CV2`, `CV0`, and `CV00`;
- a **light output-review layer**, which inventories the files already written by the prediction stage and summarizes their structure.

In the tutorial version, the heavy execution chunks remain disabled. This keeps the document readable and avoids rerunning computationally expensive model fitting during rendering.

## 2. Packages and project paths

This chunk loads only the packages required for the light review of prediction outputs and standardizes the paths used later in the file. The objective is to make the document robust whether it is rendered from the project root or from the `analysis/` folder.

```{r setup}
rm(list = ls())
options(scipen = 999)

library(readr)
library(dplyr)
library(stringr)
library(knitr)

project_root <- getwd()

if (!dir.exists(file.path(project_root, "output")) &&
    dir.exists(file.path(project_root, "..", "output"))) {
  project_root <- normalizePath(file.path(project_root, ".."))
}

if (!dir.exists(file.path(project_root, "output"))) {
  stop("Could not find the folder 'output/'. Render this file from the project root or from analysis/.")
}

results_dir <- file.path(project_root, "output", "results")
bglr_runs_dir <- file.path(results_dir, "bglr_runs")
analysis_tables_dir <- file.path(project_root, "output", "tables", "analysis_prediction")

dir.create(analysis_tables_dir, recursive = TRUE, showWarnings = FALSE)
```

## 3. Prepare the output folders used by the prediction workflow

This chunk shows the first operational step of the prediction stage: creating the folders where the raw CSV files and the auxiliary `BGLR` artifacts are stored. Because these folders are part of the heavy execution layer, the chunk remains disabled in the tutorial.

```{r prepare-output-folders, eval=FALSE}
dir.create("output/results", recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("output", "results", "bglr_runs"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("output", "tables", "analysis_prediction"), recursive = TRUE, showWarnings = FALSE)
```

## 4. Define the objects that control the run

This chunk defines the traits, environment labels, and pedigree labels that control the loops used later in the execution. It is kept disabled here because it belongs to the same heavy execution layer and depends on the matrices and phenotypic table prepared earlier in the pipeline.

```{r define-run-objects, eval=FALSE}
traits_to_run <- c("GY", "KW")
env_levels <- unique(Pheno$Env)
pedigrees <- unique(Pheno$Pedigree)
```

## 5. Start the parallel backend

This chunk documents the parallel backend used by `foreach`. The cluster is started, registered, and configured so that each worker can fit models independently. The chunk remains disabled in the tutorial because it is part of the expensive execution stage.

```{r start-parallel-backend, eval=FALSE}
cl <- parallel::makeCluster(n_cores)
doParallel::registerDoParallel(cl)

on.exit({
  try(parallel::stopCluster(cl), silent = TRUE)
  foreach::registerDoSEQ()
}, add = TRUE)

parallel::clusterEvalQ(cl, {
  library(BGLR)
  NULL
})
```

## 6. Run prediction analyses for CV1 and CV2

This chunk contains the operational block used for `CV1` and `CV2`. It creates genotype folds, masks the appropriate observations, fits the 18 models in parallel, and stores one CSV per result object. The chunk is disabled because it is the computational core of the prediction pipeline.

```{r run-predictions-cv12, eval=FALSE}
for (trait_name in traits_to_run) {

  y_full <- Pheno[[trait_name]]

  set.seed(123)

  for (rep_id in 1:num_rep) {

    fold_id <- sample(rep(1:5, length.out = length(pedigrees)))
    fold_map <- data.frame(Pedigree = pedigrees, Fold = fold_id, stringsAsFactors = FALSE)

    for (fold in 1:5) {

      pedigrees_in_fold <- fold_map$Pedigree[fold_map$Fold == fold]

      # CV1
      y_cv1 <- y_full
      y_cv1[Pheno$Pedigree %in% pedigrees_in_fold] <- NA

      cv1_results <- foreach(
        m = 1:18,
        .packages = c("BGLR"),
        .export = c("Pheno", "Eta_list", "model_catalog", "y_cv1", "y_full", "env_levels",
                    "pedigrees_in_fold", "nIter", "burnIn", "thin", "trait_name", "rep_id", "fold")
      ) %dopar% {
        model_label <- model_catalog$Model[m]

        run_dir <- file.path(
          "output", "results", "bglr_runs",
          "CV1", trait_name, model_label,
          paste0("rep", rep_id),
          paste0("fold", fold)
        )

        dir.create(run_dir, recursive = TRUE, showWarnings = FALSE)

        existing_files <- list.files(run_dir, full.names = TRUE, all.files = TRUE, no.. = TRUE)
        if (length(existing_files) > 0) {
          unlink(existing_files, recursive = TRUE, force = TRUE)
        }

        save_prefix <- file.path(
          run_dir,
          paste0("CV1_", trait_name, "_", model_label, "_rep", rep_id, "_fold", fold, "_")
        )

        fit_cv1 <- BGLR(
          y = y_cv1,
          ETA = Eta_list[[m]],
          nIter = nIter,
          burnIn = burnIn,
          thin = thin,
          saveAt = save_prefix,
          verbose = FALSE
        )

        pred_cv1 <- fit_cv1$yHat

        out_cv1 <- data.frame(
          ObsID = Pheno$ObsID,
          Pedigree = Pheno$Pedigree,
          Env = Pheno$Env,
          observed = y_full,
          predicted = pred_cv1,
          cor = NA_real_,
          stringsAsFactors = FALSE
        )

        for (env_now in env_levels) {
          idx_env <- which(out_cv1$Env == env_now & out_cv1$Pedigree %in% pedigrees_in_fold)

          if (length(idx_env) > 1) {
            out_cv1$cor[idx_env] <- cor(
              out_cv1$observed[idx_env],
              out_cv1$predicted[idx_env],
              use = "complete.obs"
            )
          }
        }

        list(model_number = m, model_label = model_label, result_table = out_cv1)
      }

      for (k in seq_along(cv1_results)) {
        model_label <- cv1_results[[k]]$model_label
        out_cv1 <- cv1_results[[k]]$result_table

        write.csv(
          out_cv1,
          file = paste0(
            "output/results/CV1_",
            trait_name,
            "_", model_label,
            "_rep", rep_id,
            "_fold", fold,
            ".csv"
          ),
          row.names = FALSE
        )
      }

      # CV2
      y_cv2 <- y_full

      for (ped_now in pedigrees_in_fold) {
        idx_ped <- which(Pheno$Pedigree == ped_now)
        envs_ped <- unique(Pheno$Env[idx_ped])

        if (length(envs_ped) > 1) {
          env_to_mask <- sample(envs_ped, size = floor(length(envs_ped) / 2))
          idx_mask <- which(Pheno$Pedigree == ped_now & Pheno$Env %in% env_to_mask)
          y_cv2[idx_mask] <- NA
        }
      }

      cv2_results <- foreach(
        m = 1:18,
        .packages = c("BGLR"),
        .export = c("Pheno", "Eta_list", "model_catalog", "y_cv2", "y_full", "env_levels",
                    "nIter", "burnIn", "thin", "trait_name", "rep_id", "fold")
      ) %dopar% {
        model_label <- model_catalog$Model[m]

        run_dir <- file.path(
          "output", "results", "bglr_runs",
          "CV2", trait_name, model_label,
          paste0("rep", rep_id),
          paste0("fold", fold)
        )

        dir.create(run_dir, recursive = TRUE, showWarnings = FALSE)

        existing_files <- list.files(run_dir, full.names = TRUE, all.files = TRUE, no.. = TRUE)
        if (length(existing_files) > 0) {
          unlink(existing_files, recursive = TRUE, force = TRUE)
        }

        save_prefix <- file.path(
          run_dir,
          paste0("CV2_", trait_name, "_", model_label, "_rep", rep_id, "_fold", fold, "_")
        )

        fit_cv2 <- BGLR(
          y = y_cv2,
          ETA = Eta_list[[m]],
          nIter = nIter,
          burnIn = burnIn,
          thin = thin,
          saveAt = save_prefix,
          verbose = FALSE
        )

        pred_cv2 <- fit_cv2$yHat

        out_cv2 <- data.frame(
          ObsID = Pheno$ObsID,
          Pedigree = Pheno$Pedigree,
          Env = Pheno$Env,
          observed = y_full,
          predicted = pred_cv2,
          cor = NA_real_,
          stringsAsFactors = FALSE
        )

        idx_na <- which(is.na(y_cv2))

        for (env_now in env_levels) {
          idx_env <- which(out_cv2$Env == env_now)
          idx_eval <- intersect(idx_env, idx_na)

          if (length(idx_eval) > 1) {
            out_cv2$cor[idx_eval] <- cor(
              out_cv2$observed[idx_eval],
              out_cv2$predicted[idx_eval],
              use = "complete.obs"
            )
          }
        }

        list(model_number = m, model_label = model_label, result_table = out_cv2)
      }

      for (k in seq_along(cv2_results)) {
        model_label <- cv2_results[[k]]$model_label
        out_cv2 <- cv2_results[[k]]$result_table

        write.csv(
          out_cv2,
          file = paste0(
            "output/results/CV2_",
            trait_name,
            "_", model_label,
            "_rep", rep_id,
            "_fold", fold,
            ".csv"
          ),
          row.names = FALSE
        )
      }
    }
  }
}
```

## 7. Run prediction analyses for CV0 and CV00

This chunk documents the operational block for the leave-one-environment-out scenarios. The logic is similar to the previous chunk, but now the masking strategy is defined at the environment level. The chunk remains disabled in the tutorial version.

```{r run-predictions-cv0-cv00, eval=FALSE}
for (trait_name in traits_to_run) {

  y_full <- Pheno[[trait_name]]

  for (env_leave in env_levels) {

    env_parts <- strsplit(env_leave, "_")[[1]]

    # CV0
    y_cv0 <- y_full
    idx_env_leave <- which(Pheno$Env == env_leave)
    y_cv0[idx_env_leave] <- NA

    cv0_results <- foreach(
      m = 1:18,
      .packages = c("BGLR"),
      .export = c("Pheno", "Eta_list", "model_catalog", "y_cv0", "y_full", "idx_env_leave",
                  "nIter", "burnIn", "thin", "trait_name", "env_leave")
    ) %dopar% {
      model_label <- model_catalog$Model[m]

      run_dir <- file.path(
        "output", "results", "bglr_runs",
        "CV0", trait_name, model_label,
        "rep1", "fold1", env_leave
      )

      dir.create(run_dir, recursive = TRUE, showWarnings = FALSE)

      existing_files <- list.files(run_dir, full.names = TRUE, all.files = TRUE, no.. = TRUE)
      if (length(existing_files) > 0) {
        unlink(existing_files, recursive = TRUE, force = TRUE)
      }

      save_prefix <- file.path(
        run_dir,
        paste0("CV0_", trait_name, "_", model_label, "_rep1_fold1_", env_leave, "_")
      )

      fit_cv0 <- BGLR(
        y = y_cv0,
        ETA = Eta_list[[m]],
        nIter = nIter,
        burnIn = burnIn,
        thin = thin,
        saveAt = save_prefix,
        verbose = FALSE
      )

      pred_cv0 <- fit_cv0$yHat
      cor_cv0 <- cor(y_full[idx_env_leave], pred_cv0[idx_env_leave], use = "complete.obs")

      out_cv0 <- data.frame(
        ObsID = Pheno$ObsID[idx_env_leave],
        Pedigree = Pheno$Pedigree[idx_env_leave],
        Env = Pheno$Env[idx_env_leave],
        observed = y_full[idx_env_leave],
        predicted = pred_cv0[idx_env_leave],
        cor = cor_cv0,
        stringsAsFactors = FALSE
      )

      list(model_number = m, model_label = model_label, result_table = out_cv0)
    }

    for (k in seq_along(cv0_results)) {
      model_label <- cv0_results[[k]]$model_label
      out_cv0 <- cv0_results[[k]]$result_table

      write.csv(
        out_cv0,
        file = paste0(
          "output/results/CV0_",
          trait_name,
          "_", env_parts[1], "_", env_parts[2],
          "_", model_label,
          "_rep1_fold1.csv"
        ),
        row.names = FALSE
      )
    }

    # CV00
    pedigrees_test <- unique(Pheno$Pedigree[Pheno$Env == env_leave])
    pedigrees_train <- unique(Pheno$Pedigree[Pheno$Env != env_leave])
    pedigrees_new <- setdiff(pedigrees_test, pedigrees_train)

    idx_cv00 <- which(Pheno$Env == env_leave & Pheno$Pedigree %in% pedigrees_new)

    if (length(idx_cv00) > 1) {

      y_cv00 <- y_full
      y_cv00[idx_cv00] <- NA

      cv00_results <- foreach(
        m = 1:18,
        .packages = c("BGLR"),
        .export = c("Pheno", "Eta_list", "model_catalog", "y_cv00", "y_full", "idx_cv00",
                    "nIter", "burnIn", "thin", "trait_name", "env_leave")
      ) %dopar% {
        model_label <- model_catalog$Model[m]

        run_dir <- file.path(
          "output", "results", "bglr_runs",
          "CV00", trait_name, model_label,
          "rep1", "fold1", env_leave
        )

        dir.create(run_dir, recursive = TRUE, showWarnings = FALSE)

        existing_files <- list.files(run_dir, full.names = TRUE, all.files = TRUE, no.. = TRUE)
        if (length(existing_files) > 0) {
          unlink(existing_files, recursive = TRUE, force = TRUE)
        }

        save_prefix <- file.path(
          run_dir,
          paste0("CV00_", trait_name, "_", model_label, "_rep1_fold1_", env_leave, "_")
        )

        fit_cv00 <- BGLR(
          y = y_cv00,
          ETA = Eta_list[[m]],
          nIter = nIter,
          burnIn = burnIn,
          thin = thin,
          saveAt = save_prefix,
          verbose = FALSE
        )

        pred_cv00 <- fit_cv00$yHat
        cor_cv00 <- cor(y_full[idx_cv00], pred_cv00[idx_cv00], use = "complete.obs")

        out_cv00 <- data.frame(
          ObsID = Pheno$ObsID[idx_cv00],
          Pedigree = Pheno$Pedigree[idx_cv00],
          Env = Pheno$Env[idx_cv00],
          observed = y_full[idx_cv00],
          predicted = pred_cv00[idx_cv00],
          cor = cor_cv00,
          stringsAsFactors = FALSE
        )

        list(model_number = m, model_label = model_label, result_table = out_cv00)
      }

      for (k in seq_along(cv00_results)) {
        model_label <- cv00_results[[k]]$model_label
        out_cv00 <- cv00_results[[k]]$result_table

        write.csv(
          out_cv00,
          file = paste0(
            "output/results/CV00_",
            trait_name,
            "_", env_parts[1], "_", env_parts[2],
            "_", model_label,
            "_rep1_fold1.csv"
          ),
          row.names = FALSE
        )
      }
    }
  }
}
```

## 8. Inventory of the prediction output files

This chunk creates lightweight inventory tables that summarize the CSV files already produced by the prediction stage. These inventories are useful for auditing the step, checking naming consistency, and feeding the Portuguese tutorial version without rerunning the heavy loops.

```{r prediction-output-inventory}
prediction_files <- character(0)

if (dir.exists(results_dir)) {
  prediction_files <- list.files(results_dir, pattern = "\\.csv$", full.names = TRUE)
}

inventory_cv12 <- data.frame()
inventory_cvloo <- data.frame()

if (length(prediction_files) > 0) {
  for (current_file in prediction_files) {
    current_name <- basename(current_file)
    current_name_no_ext <- sub("\\.csv$", "", current_name)
    current_parts <- strsplit(current_name_no_ext, "_")[[1]]

    if (length(current_parts) == 5 && current_parts[1] %in% c("CV1", "CV2")) {
      inventory_cv12 <- bind_rows(
        inventory_cv12,
        data.frame(
          file_name = current_name,
          CV = current_parts[1],
          Trait = current_parts[2],
          Model = current_parts[3],
          Rep = current_parts[4],
          Fold = current_parts[5],
          stringsAsFactors = FALSE
        )
      )
    }

    if (length(current_parts) == 6 && current_parts[1] %in% c("CV0", "CV00")) {
      inventory_cvloo <- bind_rows(
        inventory_cvloo,
        data.frame(
          file_name = current_name,
          CV = current_parts[1],
          Trait = current_parts[2],
          Env_leave = paste(current_parts[3], current_parts[4], sep = "_"),
          Model = current_parts[5],
          RepFold = current_parts[6],
          stringsAsFactors = FALSE
        )
      )
    }
  }
}

write_csv(inventory_cv12, file.path(analysis_tables_dir, "analysis_prediction_inventory_cv12.csv"))
write_csv(inventory_cvloo, file.path(analysis_tables_dir, "analysis_prediction_inventory_cvloo.csv"))

if (nrow(inventory_cv12) > 0) {
  knitr::kable(
    head(inventory_cv12, 10),
    caption = "Preview of the inventory of CV1 and CV2 prediction files.",
    align = "c"
  )
} else {
  cat("No CV1 or CV2 prediction CSV files were found in output/results/.\n")
}
```

## 9. Counts of files by validation scheme and trait

This chunk summarizes the number of prediction files already available by validation scenario and by trait. The objective is to provide a quick operational check before moving to `visualization.Rmd`.

```{r prediction-output-counts}
counts_by_cv <- data.frame()
counts_by_trait <- data.frame()

if (nrow(inventory_cv12) > 0 || nrow(inventory_cvloo) > 0) {
  combined_inventory <- bind_rows(
    inventory_cv12 %>% mutate(Env_leave = NA_character_),
    inventory_cvloo %>% mutate(Fold = NA_character_)
  )

  counts_by_cv <- combined_inventory %>%
    count(CV, name = "n_files") %>%
    arrange(CV)

  counts_by_trait <- combined_inventory %>%
    count(CV, Trait, name = "n_files") %>%
    arrange(CV, Trait)
}

write_csv(counts_by_cv, file.path(analysis_tables_dir, "analysis_prediction_counts_by_cv.csv"))
write_csv(counts_by_trait, file.path(analysis_tables_dir, "analysis_prediction_counts_by_trait.csv"))

if (nrow(counts_by_cv) > 0) {
  knitr::kable(
    counts_by_cv,
    caption = "Number of prediction files by validation scheme.",
    align = "cc"
  )
} else {
  cat("Prediction counts by validation scheme are not available yet.\n")
}
```

## 10. Examples of the output naming pattern

This chunk organizes a few concrete examples of file names and their meaning. It is a compact operational reference that helps connect the prediction step to the downstream summaries.

```{r prediction-output-examples}
output_examples <- data.frame()

if (nrow(inventory_cv12) > 0) {
  output_examples <- bind_rows(
    output_examples,
    inventory_cv12 %>%
      mutate(ExampleType = "CV1/CV2") %>%
      select(ExampleType, everything()) %>%
      head(5)
  )
}

if (nrow(inventory_cvloo) > 0) {
  output_examples <- bind_rows(
    output_examples,
    inventory_cvloo %>%
      mutate(ExampleType = "CV0/CV00") %>%
      select(ExampleType, everything()) %>%
      head(5)
  )
}

write_csv(output_examples, file.path(analysis_tables_dir, "analysis_prediction_output_examples.csv"))

if (nrow(output_examples) > 0) {
  knitr::kable(
    output_examples,
    caption = "Examples of file names produced by the prediction workflow.",
    align = "c"
  )
} else {
  cat("No prediction output examples are available yet.\n")
}
```

## 11. Summary of the auxiliary `BGLR` run folders

This chunk summarizes the directory structure created by `saveAt` under `output/results/bglr_runs/`. These folders are important because they store the auxiliary outputs of each fitted model, separated by validation scenario, trait, model, and fold or held-out environment.

```{r bglr-run-folder-summary}
bglr_inventory <- data.frame()

if (dir.exists(bglr_runs_dir)) {
  bglr_files <- list.files(bglr_runs_dir, recursive = TRUE, full.names = TRUE)

  if (length(bglr_files) > 0) {
    for (current_file in bglr_files) {
      relative_path <- gsub(
        paste0("^", normalizePath(bglr_runs_dir, winslash = "/"), "/?"),
        "",
        normalizePath(current_file, winslash = "/")
      )

      path_parts <- strsplit(relative_path, "/")[[1]]

      current_cv <- ifelse(length(path_parts) >= 1, path_parts[1], NA)
      current_trait <- ifelse(length(path_parts) >= 2, path_parts[2], NA)
      current_model <- ifelse(length(path_parts) >= 3, path_parts[3], NA)

      bglr_inventory <- bind_rows(
        bglr_inventory,
        data.frame(
          relative_path = relative_path,
          CV = current_cv,
          Trait = current_trait,
          Model = current_model,
          file_name = basename(current_file),
          stringsAsFactors = FALSE
        )
      )
    }
  }
}

write_csv(bglr_inventory, file.path(analysis_tables_dir, "analysis_prediction_bglr_run_inventory.csv"))

if (nrow(bglr_inventory) > 0) {
  bglr_summary <- bglr_inventory %>%
    count(CV, Trait, Model, name = "n_files") %>%
    arrange(CV, Trait, Model)

  knitr::kable(
    head(bglr_summary, 12),
    caption = "Preview of the auxiliary BGLR run folders.",
    align = "c"
  )
} else {
  cat("No auxiliary BGLR run files were found in output/results/bglr_runs/.\n")
}
```

## 12. Final note

This submodule closes the prediction stage by documenting both the heavy execution blocks and the structure of the files that feed the next step of the pipeline. The downstream summaries, weighted averages, manuscript-ready tables, and final figures are generated later by `visualization.Rmd` and `visualization.html`.
