---
title: "Analysis of NIR and Genomic Kernel Integration"
author: 
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: TRUE
    toc_depth: 3
    toc_float: TRUE
  highlight: github
  code_download: yes
editor_options:
  chunk_output_type: console
github-repo: wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, out.width = "100%")
```

# Analysis of NIR and Genomic Kernel Integration

This tutorial documents the analytical pipeline used to integrate near-infrared reflectance spectroscopy (NIR), genomic, and climate data using kernel-based methods to improve predictive performance. The workflow includes preprocessing, model fitting, and predictive evaluation.

This script focuses on the foundational stages of **data loading, quality control, and alignment** across data modalities.

## 1) Load Required Libraries and Data

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r load-libraries, message=FALSE}
# Core packages
library(tidyverse)
library(data.table)
library(BGLR)

# Parallel processing
library(parallel)
library(doParallel)

# Statistical and plotting utilities
library(MASS)
library(gridExtra)
library(grid)

# Modular tutorial functions
source("code/tutorial_functions.R")
```

## 2) Data Loading and Preprocessing

This section describes the corresponding methodological step in the workflow.

### 2.1) Data Loading

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
    1.  IDs (`Pedigree`, `Env`).
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r load-core-data}
core_data <- load_core_data()
NIR <- core_data$NIR
Geno <- core_data$Geno
Pheno <- core_data$Pheno

# Quick inspection
head(NIR[, 1:7])
head(Geno[, 1:5])
head(Pheno)
unique(NIR$Env)
```


### 2.2) Genotype Alignment

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r filter-shared-genotypes}
filtered <- filter_shared_genotypes(NIR, Geno, Pheno)
NIR <- filtered$NIR
Geno <- filtered$Geno
Pheno <- filtered$Pheno
Pedigree <- filtered$Pedigree

cat("Genotypes in NIR:", length(unique(NIR$Pedigree)), "\n")
cat("Genotypes in Geno:", length(Geno$taxa), "\n")
cat("Genotypes in Pheno:", length(unique(Pheno$Pedigree)), "\n")
```

## 3) Splitting NIR data by environment

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r preprocess-data}
env_split <- split_nir_by_environment(NIR)
a <- env_split$a
b <- env_split$b
c <- env_split$c
d <- env_split$d
```

### 3.1) Identification of genotypes common to all environments

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r shared-genotypes}
# Tutorial step
# Tutorial step
# Tutorial step
# Tutorial step
Pedigree_communs <- Reduce(intersect, list(Geno$taxa, a$Pedigree, b$Pedigree, c$Pedigree, d$Pedigree))
length(Pedigree_communs)  # Number of consistent genotypes
```

## 4) Building the genomic relationship matrix (ZG)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r zg}
Geno.matrix <- prepare_geno_matrix(Geno)
kernel_geno <- build_genomic_kernel(Geno.matrix, env_split)
Geno.all <- kernel_geno$Geno.all
dim(Geno.all)
```

```{r zg2, eval = FALSE}
# Tutorial step
ZG <- tcrossprod(as.matrix(Geno.all)) / ncol(Geno.all)
```

```{r, include = FALSE}
ZG <- readRDS("output/Matrizes/ZG.rds")
```

This section describes the corresponding methodological step in the workflow.

```{r zg-checagem}
# Tutorial step
dim(ZG)
ZG[1:5, 1:5]
```

## 5) Building the phenomic relationship matrix (ZP)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
3.  **Calcular o Kernel (`tcrossprod`)**: Calculamos `ZP`.

This section describes the corresponding methodological step in the workflow.

```{r zp}
kernel_pheno <- build_phenomic_kernel(env_split, first_predictor_col = 6)
NIR.all <- kernel_pheno$NIR.all
dim(NIR.all)
```

```{r zp-2, eval = FALSE}
# Tutorial step
ZP <- tcrossprod(as.matrix(NIR.all)) / ncol(NIR.all)
```

```{r, include = FALSE}
ZP <- readRDS("output/Matrizes/ZP.rds")
```

This section describes the corresponding methodological step in the workflow.

```{r zp-checagem}
# Tutorial step
dim(ZP)
ZP[1:5, 1:5]
```

## 6) Building the environmental relationship matrix (ZE)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r ze-zeze}
# Tutorial step
ZE <- model.matrix(~ as.factor(Pheno$Env) - 1)

# Tutorial step
ZEZE <- tcrossprod(ZE)

# Checagens
cat("Dimensions of matrix ZE:", dim(ZE), "\n")
cat("Dimensions of matrix ZEZE:", dim(ZEZE), "\n")

# Visualiza primeiras linhas
dim(ZE)
ZEZE[1:5, 1:5]
```

## 6.1) Building the environmental relationship matrix with climate covariates (W)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r load-ecs}
# Tutorial step

# Tutorial step
ECs_data <- read.csv("output/climate_results/environmental_covariates.csv") %>%
  dplyr::select(PRECTOTCORR:date, T_eff_min:VPD_STRESS_DAYS)

# Tutorial step
# Tutorial step
ECs_data_scaled <- ECs_data %>%
  dplyr::select(TMAX_AVG:VPD_STRESS_DAYS) %>%
  scale(., center = TRUE, scale = TRUE) %>% unique()

# Tutorial step
ECs.2011 <- matrix(ECs_data_scaled[1, ], nrow = 1)
colnames(ECs.2011) <- colnames(ECs_data_scaled)

# Tutorial step
ECs.2012 <- matrix(ECs_data_scaled[2, ], nrow = 1)
colnames(ECs.2012) <- colnames(ECs_data_scaled)

# Tutorial step
ECs.2011_filtered <- ECs_data %>% filter(year == 2011)
ECs.2011_scaled <- ECs.2011_filtered %>%
  dplyr::select(-year, -date, -(TMAX_AVG:VPD_STRESS_DAYS)) %>%
  scale(. , center = TRUE, scale = TRUE)
ECs.2011_dates <- ECs.2011_filtered$date

# Tutorial step
ECs.2012_filtered <- ECs_data %>% filter(year == 2012)
ECs.2012_scaled <-    ECs.2012_filtered %>%
  dplyr::select(-year, -date, -(TMAX_AVG:VPD_STRESS_DAYS)) %>%
  scale(., center = TRUE, scale = TRUE)
ECs.2012_dates <- ECs.2012_filtered$date

# Tutorial step

# Tutorial step
env_wide_2011 <- as.data.frame(ECs.2011_scaled) %>%
  mutate(date = ECs.2011_dates) %>%
  pivot_wider(names_from = date, values_from = -date)

# Tutorial step
env_wide_2012 <- as.data.frame(ECs.2012_scaled) %>%
  mutate(date = ECs.2012_dates) %>%
  pivot_wider(names_from = date, values_from = -date)

# Tutorial step

# Tutorial step

# Tutorial step
pedigrees_a <- a$Pedigree
ECs.a <- env_wide_2011[rep(1, length(pedigrees_a)), ] # Replicate the daily signature from 2011
ECs.a <- cbind(ECs.a, matrix(
  NA,
  nrow = length(pedigrees_a),
  ncol = length(env_wide_2012),
  dimnames = list(pedigrees_a, colnames(env_wide_2012))
))  # Adiciona colunas NA para 2012
ECs.a <- cbind(ECs.a, ECs.2011) %>% as.matrix() # Add aggregated data from 2011
rownames(ECs.a) <- pedigrees_a
dim(ECs.a)

# Tutorial step
pedigrees_b <- b$Pedigree
ECs.b <- env_wide_2011[rep(1, length(pedigrees_b)), ]
ECs.b <- cbind(ECs.b, matrix(
  NA,
  nrow = length(pedigrees_b),
  ncol = length(env_wide_2012),
  dimnames = list(pedigrees_b, colnames(env_wide_2012))
))  # Add 2012 data
ECs.b <- cbind(ECs.b, ECs.2011) %>% as.matrix()  
rownames(ECs.b) <- pedigrees_b
dim(ECs.b)

# Tutorial step
pedigrees_c <- c$Pedigree
ECs.c <- env_wide_2012[rep(1, length(pedigrees_c)), ] # Replicate the daily signature from 2012
ECs.c <- cbind(matrix(
  NA,
  nrow = length(pedigrees_c),
  ncol = length(env_wide_2011),
  dimnames = list(pedigrees_c, colnames(env_wide_2011))
),  ECs.c) # Adiciona colunas NA para 2011
ECs.c <- cbind(ECs.c, ECs.2012) %>% as.matrix() # Add aggregated data from 2012
rownames(ECs.c) <- pedigrees_c
dim(ECs.c)

# Tutorial step
pedigrees_d <- d$Pedigree
ECs.d <- env_wide_2012[rep(1, length(pedigrees_d)), ]
ECs.d <- cbind(matrix(
  NA,
  nrow = length(pedigrees_d),
  ncol = length(env_wide_2011),
  dimnames = list(pedigrees_d, colnames(env_wide_2011))
), ECs.d)  # Add 2011 data
ECs.d <- cbind(ECs.d, ECs.2012) %>% as.matrix()
rownames(ECs.d) <- pedigrees_d
dim(ECs.d)

# Tutorial step
W <- rbind(ECs.a, ECs.b, ECs.c, ECs.d)

# Tutorial step
dim(W)
W[1:5, 1:5]

# Tutorial step
W[is.na(W)] <- 0

# Tutorial step
ZW <- tcrossprod(as.matrix(W)) / ncol(W)
table(ZW[1,])

# Tutorial step
dim(ZW)
ZW[1:5, 1:5]
```

## 7) Final alignment checks

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r}
Geno.all[str_detect(rownames(Geno.all), "A188"), 1:5]
W[str_detect(rownames(W), "A188"), 1:5]
NIR.all[str_detect(rownames(W), "A188"), 1:5]
```

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
3.  Perfeitamente alinhados entre si (nomes de linhas de ZW == nomes de linhas de ZG, etc.).

This section describes the corresponding methodological step in the workflow.

```{r}
all(rownames(ZW) == colnames(ZW))
all(rownames(ZG) == colnames(ZG))
all(rownames(ZP) == colnames(ZP))

all(rownames(ZW) == rownames(ZG))
all(rownames(ZW) == rownames(ZP))

all(colnames(ZW) == colnames(ZG))
all(colnames(ZW) == colnames(ZP))

all(unique(rownames(ZG)) == unique(colnames(ZG)))
all(unique(rownames(ZP)) == unique(colnames(ZP)))
all(unique(rownames(ZW)) == unique(colnames(ZW)))
```

This section describes the corresponding methodological step in the workflow.

```{r}
ZW[1:5, 1:5]
ZG[1:5, 1:5]
ZP[1:5, 1:5]
```

## 8) Interactions

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r interactions}
# Tutorial step
ZGZE <- ZG * ZEZE
ZGZE[1:5, 1:5]

# Tutorial step
ZPZE <- ZP * ZEZE
ZPZE[1:5, 1:5]

# Tutorial step
ZGZW <- ZG * ZW
ZGZW[1:5, 1:5]

# Tutorial step
ZPZW <- ZP * ZW
ZPZW[1:5, 1:5]
```

## 9) GK and AK kernels

This section describes the corresponding methodological step in the workflow.

**Tipos de kernels:**

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

### 9.1) Gaussian kernel (GK)

This section describes the corresponding methodological step in the workflow.

#### 9.1.1) Gaussian kernel optimization function

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r gk-functions}
# Tutorial step
margh.fun <- function(theta, y, D, q, nu = 0.0001, Sc = 0.0001,
                      nuh = NULL, Sch = NULL, prior = NULL) {
  h <- theta[1]    # Kernel smoothing parameter
  phi <- theta[2]  # Scale parameter
  
  Kh <- exp(-h * D / q)  # Matriz kernel gaussiana
  eigenKh <- eigen(Kh)   # Spectral decomposition
  
  nr <- length(which(eigenKh$values > 1e-10))  # Number of relevant components
  Uh <- eigenKh$vectors[, 1:nr]                # Autovetores
  Sh <- eigenKh$values[1:nr]                   # Autovalores
  
  d <- t(Uh) %*% scale(y, scale = FALSE)    # Data projection
  
  # Tutorial step
  lden <- -0.5 * sum(log(1 + phi * Sh)) -
    (nu + nr - 1) / 2 * log(Sc + sum(d^2 / (1 + phi * Sh)))
  
  # Prior opcional sobre h
  if (!is.null(prior)) {
    lprior <- dgamma(h, nuh, Sch, log = TRUE)
  } else {
    lprior <- 0
  }
  
  return(-(lden + lprior))  # Return the negative log-likelihood
}
```

#### 9.1.2) Building the genomic Gaussian kernel (GGK)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
3.  `solG <- optim(...)`: Executa o otimizador (`optim`) usando `margh.fun` para encontrar o melhor `h` (baseado no trait `Pheno$GY`).
4.  `hG <- solG$par[1]`: Extrai o valor `h` otimizado.
This section describes the corresponding methodological step in the workflow.

```{r ggk}
# Tutorial step
DG <- (as.matrix(dist(Geno)))^2
q05G <- quantile(DG, 0.05)

# Tutorial step
IDs <- as.character(Pheno$Pedigree)
DG <- DG[rownames(DG) %in% IDs, rownames(DG) %in% IDs]

IDs <- factor(IDs, levels = rownames(DG))
Z <- as.matrix(model.matrix(~ IDs - 1))
ZD <- tcrossprod(Z, DG)
DG_proj <- tcrossprod(ZD, Z)
```

```{r ggk-1, eval = FALSE}
# Tutorial step
solG <- optim(c(1, 1), margh.fun, y = Pheno$GY, D = DG_proj, q = q05G,
              method = "L-BFGS-B", lower = c(0.05, 0.05), upper = c(6, 30))
```

```{r, include=FALSE}
solG <- readRDS("output/Matrizes/solG.RDS")
```

```{r ggk-2}
hG <- solG$par[1]

# Tutorial step
GGK <- exp(-hG * DG_proj / median(DG_proj))
```

```{r ggk-checagem}
# Tutorial step
dim(GGK)
GGK[1:5,1:5]
```

#### 9.1.3) Building the phenomic Gaussian kernel (PGK)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r pgk}
# Tutorial step
DP <- (as.matrix(dist(NIR.all)))^2
q05P <- quantile(DP, 0.05)

IDs <- rownames(NIR.all)
DP <- DP[rownames(DP) %in% IDs, rownames(DP) %in% IDs]

IDs <- IDs
Z <- as.matrix(model.matrix(~ IDs - 1))
Z0 <- tcrossprod(Z, Z)
ZD <- tcrossprod(Z0, DP)
DP_proj <- tcrossprod(ZD, Z0)
```

```{r pgk-1, eval = FALSE}
# Tutorial step
solP <- optim(c(1, 1), margh.fun, y = Pheno$GY, D = DP_proj, q = q05P,
              method = "L-BFGS-B", lower = c(0.05, 0.05), upper = c(6, 30))
```

```{r, include=FALSE}
solP <- readRDS("output/Matrizes/solP.RDS")
```

```{r pgk-2}
hP <- solP$par[1]

# Tutorial step
PGK <- exp(-hP * DP_proj / median(DP_proj))
```

```{r pgk-checagem}
# Tutorial step
dim(PGK)
PGK[1:5,1:5]
```

#### 9.1.4) Interactions with Gaussian Kernels

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r interactions-gk}
GGKE    <- GGK * ZEZE
PGKE    <- PGK * ZEZE
GGKW    <- GGK * ZW
PGKW    <- PGK * ZW
```

### 9.2) Arc-cossin Kernels (AK)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

#### 9.2.1) Arc-cosine kernel helper functions

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r ak-functions}
###############################################################
# Tutorial step
# Tutorial step
# Tutorial step
# Tutorial step
###############################################################

# Tutorial step
AK1.fun <- function(X) {
  n <- nrow(X)
  cosalfa <- cor(t(X))  # Correlation between vectors
  cosalfa[cosalfa > 1] <- 1
  cosalfa[cosalfa < -1] <- -1
  angulo <- acos(cosalfa)  # Angle between vectors
  mag <- sqrt(apply(X, 1, function(x) crossprod(x)))  # Magnitude dos vetores
  sxy <- tcrossprod(mag)  # Produto cruzado das magnitudes
  AK1 <- (1 / pi) * sxy * (sin(angulo) + (pi - angulo) * cosalfa)
  AK1 <- AK1 / median(AK1, na.rm = TRUE)  # Normalization
  colnames(AK1) <- rownames(X)
  rownames(AK1) <- rownames(X)
  return(AK1)
}

# Tutorial step
marg.AK <- function(y, AK1, ml) {
  lden.fun <- function(phi, nr, Uh, Sh, d) {
    lden <- -0.5 * sum(log((1 + phi * Sh))) - (nr - 1) / 2 * log(sum(d^2 / ((1 + phi * Sh))))
    return(-lden)
  }
  
  vero <- function(y, GC) {
    Kh <- GC
    eigenKh <- eigen(Kh)
    nr <- length(which(eigenKh$values > 1e-10))
    Uh <- eigenKh$vectors[, 1:nr]
    Sh <- eigenKh$values[1:nr]
    d <- t(Uh) %*% scale(y, scale = FALSE)
    sol <- optimize(lden.fun, nr = nr, Uh = Uh, Sh = Sh, d = d,
                    lower = 0.0005, upper = 200)
    phi <- sol[[1]]
    log.vero <- 0.5 * sum(log((1 + phi * Sh))) -
      (nr - 1) / 2 * log(sum(d^2 / ((1 + phi * Sh))))
    return(log.vero)
  }
  
  GC <- AK1
  l <- 1
  GC2 <- GC
  vero1 <- vero(y = y, GC = GC2)
  m <- 0
  
  while (m == 0 && (l < ml)) {
    l <- l + 1
    GC <- AK.fun(AK1 = GC2, nl = 1)
    GC2 <- GC
    vero2 <- vero(y = y, GC = GC2)
    if (vero2 < vero1) m = 1
    vero1 <- vero2
  }
  
  return(l - 1)  # Return the optimal recursion level
}

# Tutorial step
AK.fun <- function(AK1, nl) {
  n <- nrow(AK1)
  AK <- AK1
  for (l in 1:nl) {
    Aux <- tcrossprod(diag(AK))
    cosalfa <- AK * (Aux^(-1 / 2))
    cosa <- as.vector(cosalfa)
    cosa[cosa > 1] <- 1
    cosa[cosa < -1] <- -1
    angulo <- acos(cosa)
    angulo <- matrix(angulo, n, n)
    AK <- (1 / pi) * (Aux^(1 / 2)) * (sin(angulo) + (pi - angulo) * cos(angulo))
  }
  AK <- AK / median(AK, na.rm = TRUE)
  rownames(AK) <- rownames(AK1)
  colnames(AK) <- colnames(AK1)
  return(AK)
}
```

#### 9.2.2) Building the genomic arc-cosine kernel (GAK)

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r gak}
# Tutorial step
GAK1 <- AK1.fun(Geno)
dim(GAK1)
GAK1[1:5, 1:5]

# Tutorial step
IDs <- as.character(Pheno[, "Pedigree"])
GAK1 <- GAK1[rownames(GAK1) %in% IDs, rownames(GAK1) %in% IDs]

# Tutorial step
IDs <- factor(IDs, levels = rownames(GAK1))
Z <- as.matrix(model.matrix(~ IDs - 1))
G <- tcrossprod(tcrossprod(Z, GAK1), Z)

# Tutorial step
Gl <- marg.AK(y = Pheno$GY, AK1 = G, ml = 50)

# Tutorial step
IDs.L <- Pheno[, "Pedigree"]
IDs.L <- factor(IDs.L, levels = rownames(GAK1))
Z.L <- as.matrix(model.matrix(~ IDs.L - 1))
G.L <- tcrossprod(tcrossprod(Z.L, GAK1), Z.L)

# Tutorial step
GAK <- AK.fun(AK1 = G.L, nl = Gl)
```

```{r gak-checagem}
# Tutorial step
dim(GAK)
GAK[1:5, 1:5]
```

#### 9.2.3) Building the phenomic arc-cosine kernel (PAK)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
2.  `P <- tcrossprod(...)`: Projeta/alinha o *kernel* `PAK1`.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r pak}
# Tutorial step
PAK1 <- AK1.fun(NIR.all)
dim(PAK1)
PAK1[1:5, 1:5]

# Tutorial step
IDs <- as.character(rownames(NIR.all))
PAK1 <- PAK1[rownames(PAK1) %in% IDs, rownames(PAK1) %in% IDs]

# Tutorial step
IDs <- IDs
Z <- as.matrix(model.matrix(~ IDs - 1))
Z0 <- tcrossprod(Z, Z)
ZD <- tcrossprod(Z0, PAK1)
P <- tcrossprod(ZD, Z0)

# Tutorial step
Pl <- marg.AK(y = Pheno$GY, AK1 = P, ml = 50)

# Tutorial step
IDs.L <- rownames(NIR.all)
Z.L <- as.matrix(model.matrix(~ IDs.L - 1))
Z0.L <- tcrossprod(Z.L, Z.L)
P.L <- tcrossprod(tcrossprod(Z0.L, PAK1), Z0.L)

# Tutorial step
PAK <- AK.fun(AK1 = P.L, nl = Pl)
```

```{r pak-checagem}
dim(PAK)
PAK[1:5, 1:5]
```

#### 9.2.4) Interactions with Arc-cosine Kernels

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r interaction-AK}
# Tutorial step
GAKE <- GAK * ZEZE
GAKE[1:5,1:5]

# Tutorial step
PAKE <- PAK * ZEZE
PAKE[1:5,1:5]

# Tutorial step
GAKW <- GAK * ZW
GAKW[1:5,1:5]

# Tutorial step
PAKW <- PAK * ZW
PAKW[1:5,1:5]
```

## 10) Clean environment

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r keep_names, message = FALSE}
# Tutorial step
keep_names <- c(
  # Tutorial step
  "ZE", "ZG", "ZP", "ZGZE", "ZPZE", "ZW", "ZGZW", "ZPZW",
  # Kernels
  "GGK", "PGK", "GGKE", "PGKE", "GGKW", "PGKW", 
  "GAK", "PAK", "GAKE", "PAKE", "GAKW", "PAKW",
  # other important objects
  "Pheno", "Pedigree"
)

# Tutorial step
current_objs <- ls(envir = .GlobalEnv)
to_keep <- intersect(current_objs, keep_names)
to_remove <- setdiff(current_objs, to_keep)

message("Objects to keep (", length(to_keep), "): ", paste(to_keep, collapse = ", "))
message("Objects to remove (", length(to_remove), "): ", paste(head(to_remove, 40), collapse = ", "),
        if (length(to_remove) > 40) paste0(" ... (+", length(to_remove) - 40, " mais)"))

# Tutorial step
for (obj_name in to_keep) {
  saveRDS(get(obj_name), file = file.path("output/Matrizes", paste0(obj_name, ".rds")))
}

# Tutorial step
rm(list = to_remove, envir = .GlobalEnv)
gc()

# Tutorial step
rds_files <- list.files("output/Matrizes", pattern = "\\.rds$", full.names = TRUE)

# Tutorial step
for (file in rds_files) {
  obj_name <- tools::file_path_sans_ext(basename(file))
  assign(obj_name, readRDS(file), envir = .GlobalEnv)
}

# Tutorial step
message("Objetos carregados: ", paste(tools::file_path_sans_ext(basename(rds_files)), collapse = ", "))
```

## 11) Modelos preditivos

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
* O que acontece quando combinamos *kernels* (ex: `G+P`)?

### 11.1) Effect notation

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

#### Interactions with Categorical Environment (E)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

#### Interactions with Climate Environment (W)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

---

### Estrutura dos Modelos

This section describes the corresponding methodological step in the workflow.

#### Group 1: Baseline models (categorical environment E)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
| :--- | :--- |
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
| :--- | :--- |
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
| :--- | :--- |
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

---

#### Group 2: Models with climate covariates (W)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
| :--- | :--- |
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
| **E+W+GGK+PGK** | E+W + G (Gaussiano) + P (Gaussiano) |
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
| **E+W+GAK+PAK** | E+W + G (Arc-Kernel) + P (Arc-Kernel) |

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
| :--- | :--- |
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
| :--- | :--- |
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
| :--- | :--- |
| **W** | Clima (Kernel) |
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
| **W+GGK+PGK** | Clima (Kernel) + G (Gaussiano) + P (Gaussiano) |
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
| **W+GAK+PAK** | Clima (Kernel) + G (Arc-Kernel) + P (Arc-Kernel) |

### 11.2) Model classification table

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r model_info}
# Tutorial step
model_info <- tibble(
  ModelID = 1:62,
  Model   = paste0("M", ModelID),
  Type = case_when(
# Tutorial step
    ModelID %in% c(1:6, 13:15) ~ "Basico",      # E + G/P (efeitos principais)
    ModelID %in% c(7:12, 16:18) ~ "Interacao_E", # E + G/P + GxE/PxE
    
    # Tutorial step
    ModelID %in% 19:27 ~ "Clima_Base_E",   # E + W + G/P (efeitos principais)
    ModelID %in% 28:36 ~ "Clima_IntE",     # E + W + G/P + GxE/PxE
    ModelID %in% 37:45 ~ "Clima_IntW",     # E + W + G/P + GxW/PxW
    
    # Tutorial step
    ModelID %in% 46:54 ~ "Clima_Base_noE", # W + G/P (sem E fixo),
    
    # Grrupo 4: Modelos individuais
    ModelID %in% c(55:62) ~ "Individual", 
    
    TRUE               ~ "Outros"
  ),
  Description = case_when(
    # Tutorial step
    ModelID == 1  ~ "E + G",
    ModelID == 2  ~ "E + P",
    ModelID == 3  ~ "E + GGK",
    ModelID == 4  ~ "E + PGK",     # (Baseado em Eta4)
    ModelID == 5  ~ "E + GAK",     # (Baseado em Eta5)
    ModelID == 6  ~ "E + PAK",

    # Tutorial step
    ModelID == 7  ~ "E + G + GE",
    ModelID == 8  ~ "E + P + PE",
    ModelID == 9  ~ "E + GGK + GGKE",
    ModelID == 10 ~ "E + PGK + PGKE", # (Baseado em Eta10)
    ModelID == 11 ~ "E + GAK + GAKE", # (Baseado em Eta11)
    ModelID == 12 ~ "E + PAK + PAKE",

    # Tutorial step
    ModelID == 13 ~ "E + G + P",
    ModelID == 14 ~ "E + GGK + PGK",
    ModelID == 15 ~ "E + GAK + PAK",

    # Tutorial step
    ModelID == 16 ~ "E + G + P + GE + PE",
    ModelID == 17 ~ "E + GGK + PGK + GGKE + PGKE",
    ModelID == 18 ~ "E + GAK + PAK + GAKE + PAKE",

    # Bloco 5: (Item 10.1) Modelos E + W + G/P (Clima_Base_E)
    ModelID == 19 ~ "E + G + W",
    ModelID == 20 ~ "E + P + W",
    ModelID == 21 ~ "E + G + P + W",
    ModelID == 22 ~ "E + GGK + W",
    ModelID == 23 ~ "E + PGK + W",
    ModelID == 24 ~ "E + GGK + PGK + W",
    ModelID == 25 ~ "E + GAK + W",
    ModelID == 26 ~ "E + PAK + W",
    ModelID == 27 ~ "E + GAK + PAK + W",

    # Bloco 6: (Item 10.1) Modelos E + W + G/P + GxE/PxE (Clima_IntE)
    ModelID == 28 ~ "E + G + W + GE",
    ModelID == 29 ~ "E + P + W + PE",
    ModelID == 30 ~ "E + G + P + W + GE + PE",
    ModelID == 31 ~ "E + GGK + W + GGKE",
    ModelID == 32 ~ "E + PGK + W + PGKE",
    ModelID == 33 ~ "E + GGK + PGK + W + GGKE + PGKE",
    ModelID == 34 ~ "E + GAK + W + GAKE",
    ModelID == 35 ~ "E + PAK + W + PAKE",
    ModelID == 36 ~ "E + GAK + PAK + W + GAKE + PAKE",

    # Bloco 7: (Item 10.1) Modelos E + W + G/P + GxW/PxW (Clima_IntW)
    ModelID == 37 ~ "E + G + W + GW",
    ModelID == 38 ~ "E + P + W + PW",
    ModelID == 39 ~ "E + G + P + W + GW + PW",
    ModelID == 40 ~ "E + GGK + W + GGKW",
    ModelID == 41 ~ "E + PGK + W + PGKW",
    ModelID == 42 ~ "E + GGK + PGK + W + GGKW + PGKW",
    ModelID == 43 ~ "E + GAK + W + GAKW",
    ModelID == 44 ~ "E + PAK + W + PAKW",
    ModelID == 45 ~ "E + GAK + PAK + W + GAKW + PAKW",
    
    # Bloco 8: (Item 10.1) Modelos W + G/P (Sem E) (Clima_Base_noE)
    ModelID == 46 ~ "G + W",
    ModelID == 47 ~ "P + W",
    ModelID == 48 ~ "G + P + W",
    ModelID == 49 ~ "GGK + W",
    ModelID == 50 ~ "PGK + W",
    ModelID == 51 ~ "GGK + PGK + W",
    ModelID == 52 ~ "GAK + W",
    ModelID == 53 ~ "PAK + W",
    ModelID == 54 ~ "GAK + PAK + W",
    
    # Bloco 9: (Item 11.5) Modelos individuais
    ModelID == 55 ~ "E",
    ModelID == 56 ~ "G",
    ModelID == 57 ~ "P",
    ModelID == 58 ~ "GGK",
    ModelID == 59 ~ "PGK",
    ModelID == 60 ~ "GAK",
    ModelID == 61 ~ "PAK",
    ModelID == 62 ~ "W",
    
    TRUE ~ as.character(ModelID) # Fallback
  )
)
```

## 12) Definition of predictive models

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r define-models}
# -------------------------------------------------------------------
# Tutorial step
# (Modelos Eta1 a Eta18)
# -------------------------------------------------------------------

# --- 1.1: Efeitos Principais (E + G/P) ---

# Modelo 1: E + G (Linear)
Eta1 <- list(
  E = list(X = ZE, model = "BRR"),
  G = list(K = ZG, model = "RKHS")
)

# Modelo 2: E + P (Linear)
Eta2 <- list(
  E = list(X = ZE, model = "BRR"),
  P = list(K = ZP, model = "RKHS")
)

# Modelo 3: E + GGK (Gaussiano)
Eta3 <- list(
  E = list(X = ZE, model = "BRR"),
  GGK = list(K = GGK, model = "RKHS")
)

# Modelo 4: E + PGK (Gaussiano)
Eta4 <- list(
  E = list(X = ZE, model = "BRR"),
  PGK = list(K = PGK, model = "RKHS")
)

# Modelo 5: E + GAK (Arc-cosseno)
Eta5 <- list(
  E = list(X = ZE, model = "BRR"),
  GAK = list(K = GAK, model = "RKHS")
)

# Modelo 6: E + PAK (Arc-cosseno)
Eta6 <- list(
  E = list(X = ZE, model = "BRR"),
  PAK = list(K = PAK, model = "RKHS")
)

# Modelo 13: E + G + P (Linear)
Eta13 <- list(
  E = list(X = ZE, model = "BRR"),
  G = list(K = ZG, model = "RKHS"),
  P = list(K = ZP, model = "RKHS")
)

# Modelo 14: E + GGK + PGK (Gaussiano)
Eta14 <- list(
  E = list(X = ZE, model = "BRR"),
  GGK = list(K = GGK, model = "RKHS"),
  PGK = list(K = PGK, model = "RKHS")
)

# Modelo 15: E + GAK + PAK (Arc-cosseno)
Eta15 <- list(
  E = list(X = ZE, model = "BRR"),
  GAK = list(K = GAK, model = "RKHS"),
  PAK = list(K = PAK, model = "RKHS")
)

# Tutorial step

# Modelo 7: E + G + GE (Linear)
Eta7 <- list(
  E = list(X = ZE, model = "BRR"),
  G = list(K = ZG, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS")
)

# Modelo 8: E + P + PE (Linear)
Eta8 <- list(
  E = list(X = ZE, model = "BRR"),
  P = list(K = ZP, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS")
)

# Modelo 9: E + GGK + GGKE (Gaussiano)
Eta9 <- list(
  E = list(X = ZE, model = "BRR"),
  GGK = list(K = GGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS")
)

# Modelo 10: E + PGK + PGKE (Gaussiano)
Eta10 <- list(
  E = list(X = ZE, model = "BRR"),
  PGK = list(K = PGK, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS")
)

# Modelo 11: E + GAK + GAKE (Arc-cosseno)
Eta11 <- list(
  E = list(X = ZE, model = "BRR"),
  GAK = list(K = GAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS")
)

# Modelo 12: E + PAK + PAKE (Arc-cosseno)
Eta12 <- list(
  E = list(X = ZE, model = "BRR"),
  PAK = list(K = PAK, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS")
)

# Modelo 16: E + G + P + GE + PE (Linear)
Eta16 <- list(
  E = list(X = ZE, model = "BRR"),
  G = list(K = ZG, model = "RKHS"),
  P = list(K = ZP, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS")
)

# Modelo 17: E + GGK + PGK + GGKE + PGKE (Gaussiano)
Eta17 <- list(
  E = list(X = ZE, model = "BRR"),
  GGK = list(K = GGK, model = "RKHS"),
  PGK = list(K = PGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS")
)

# Modelo 18: E + GAK + PAK + GAKE + PAKE (Arc-cosseno)
Eta18 <- list(
  E = list(X = ZE, model = "BRR"),
  GAK = list(K = GAK, model = "RKHS"),
  PAK = list(K = PAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS")
)


# -------------------------------------------------------------------
# Tutorial step
# (Modelos Eta19 a Eta46)
# -------------------------------------------------------------------

# --- 2.1: Efeitos Principais (E + W + G/P) ---

# Tutorial step
Eta19 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, model = "RKHS")
)

# Modelo 20: E + W + G (Linear)
Eta20 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  G = list(K = ZG, model = "RKHS")
)

# Modelo 21: E + W + P (Linear)
Eta21 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  P = list(K = ZP, model = "RKHS")
)

# Modelo 22: E + W + G + P (Linear)
Eta22 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  G = list(K = ZG, model = "RKHS"),
  P = list(K = ZP, model = "RKHS")
)

# Modelo 23: E + W + GGK (Gaussiano)
Eta23 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  GGK = list(K = GGK, model = "RKHS")
)

# Modelo 24: E + W + PGK (Gaussiano)
Eta24 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  PGK = list(K = PGK, model = "RKHS")
)

# Modelo 25: E + W + GGK + PGK (Gaussiano)
Eta25 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  GGK = list(K = GGK, model = "RKHS"),
  PGK = list(K = PGK, model = "RKHS")
)

# Modelo 26: E + W + GAK (Arc-cosseno)
Eta26 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  GAK = list(K = GAK, model = "RKHS")
)

# Modelo 27: E + W + PAK (Arc-cosseno)
Eta27 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  PAK = list(K = PAK, model = "RKHS")
)

# Modelo 28: E + W + GAK + PAK (Arc-cosseno)
Eta28 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  GAK = list(K = GAK, model = "RKHS"),
  PAK = list(K = PAK, model = "RKHS")
)

# Tutorial step

# Modelo 29: E + W + G + GE (Linear)
Eta29 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  G = list(K = ZG, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS")
)

# Modelo 30: E + W + P + PE (Linear)
Eta30 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  P = list(K = ZP, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS")
)

# Modelo 31: E + W + G + P + GE + PE (Linear)
Eta31 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  G = list(K = ZG, model = "RKHS"),
  P = list(K = ZP, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS")
)

# Modelo 32: E + W + GGK + GGKE (Gaussiano)
Eta32 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  GGK = list(K = GGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS")
)

# Modelo 33: E + W + PGK + PGKE (Gaussiano)
Eta33 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  PGK = list(K = PGK, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS")
)

# Modelo 34: E + W + GGK + PGK + GGKE + PGKE (Gaussiano)
Eta34 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  GGK = list(K = GGK, model = "RKHS"),
  PGK = list(K = PGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS")
)

# Modelo 35: E + W + GAK + GAKE (Arc-cosseno)
Eta35 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  GAK = list(K = GAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS")
)

# Modelo 36: E + W + PAK + PAKE (Arc-cosseno)
Eta36 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  PAK = list(K = PAK, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS")
)

# Modelo 37: E + W + GAK + PAK + GAKE + PAKE (Arc-cosseno)
Eta37 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  GAK = list(K = GAK, model = "RKHS"),
  PAK = list(K = PAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS")
)

# Tutorial step

# Modelo 38: E + W + G + GW (Linear)
Eta38 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  G = list(K = ZG, model = "RKHS"),
  GW = list(K = ZGZW, model = "RKHS")
)

# Modelo 39: E + W + P + PW (Linear)
Eta39 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  P = list(K = ZP, model = "RKHS"),
  PW = list(K = ZPZW, model = "RKHS")
)

# Modelo 40: E + W + G + P + GW + PW (Linear)
Eta40 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  G = list(K = ZG, model = "RKHS"),
  P = list(K = ZP, model = "RKHS"),
  GW = list(K = ZGZW, model = "RKHS"),
  PW = list(K = ZPZW, model = "RKHS")
)

# Modelo 41: E + W + GGK + GGKW (Gaussiano)
Eta41 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  GGK = list(K = GGK, model = "RKHS"),
  GGKW = list(K = GGKW, model = "RKHS")
)

# Modelo 42: E + W + PGK + PGKW (Gaussiano)
Eta42 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  PGK = list(K = PGK, model = "RKHS"),
  PGKW = list(K = PGKW, model = "RKHS")
)

# Modelo 43: E + W + GGK + PGK + GGKW + PGKW (Gaussiano)
Eta43 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  GGK = list(K = GGK, model = "RKHS"),
  PGK = list(K = PGK, model = "RKHS"),
  GGKW = list(K = GGKW, model = "RKHS"),
  PGKW = list(K = PGKW, model = "RKHS")
)

# Modelo 44: E + W + GAK + GAKW (Arc-cosseno)
Eta44 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  GAK = list(K = GAK, model = "RKHS"),
  GAKW = list(K = GAKW, model = "RKHS")
)

# Modelo 45: E + W + PAK + PAKW (Arc-cosseno)
Eta45 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  PAK = list(K = PAK, model = "RKHS"),
  PAKW = list(K = PAKW, model = "RKHS")
)

# Modelo 46: E + W + GAK + PAK + GAKW + PAKW (Arc-cosseno)
Eta46 <- list(
  E = list(X = ZE, model = "BRR"),
  W = list(K = ZW, "RKHS"),
  GAK = list(K = GAK, model = "RKHS"),
  PAK = list(K = PAK, model = "RKHS"),
  GAKW = list(K = GAKW, model = "RKHS"),
  PAKW = list(K = PAKW, model = "RKHS")
)


# -------------------------------------------------------------------
# Tutorial step
# (Modelos Eta47 a Eta56)
# -------------------------------------------------------------------

# Tutorial step
Eta47 <- list(
  W = list(K = ZW, model = "RKHS")
)

# Modelo 48: W + G (Linear)
Eta48 <- list(
  W = list(K = ZW, model = "RKHS"),
  G = list(K = ZG, model = "RKHS")
)

# Modelo 49: W + P (Linear)
Eta49 <- list(
  W = list(K = ZW, model = "RKHS"),
  P = list(K = ZP, model = "RKHS")
)

# Modelo 50: W + G + P (Linear)
Eta50 <- list(
  W = list(K = ZW, model = "RKHS"),
  G = list(K = ZG, model = "RKHS"),
  P = list(K = ZP, model = "RKHS")
)

# Modelo 51: W + GGK (Gaussiano)
Eta51 <- list(
  W = list(K = ZW, model = "RKHS"),
  GGK = list(K = GGK, model = "RKHS")
)

# Modelo 52: W + PGK (Gaussiano)
Eta52 <- list(
  W = list(K = ZW, model = "RKHS"),
  PGK = list(K = PGK, model = "RKHS")
)

# Modelo 53: W + GGK + PGK (Gaussiano)
Eta53 <- list(
  W = list(K = ZW, model = "RKHS"),
  GGK = list(K = GGK, model = "RKHS"),
  PGK = list(K = PGK, model = "RKHS")
)

# Modelo 54: W + GAK (Arc-cosseno)
Eta54 <- list(
  W = list(K = ZW, model = "RKHS"),
  GAK = list(K = GAK, model = "RKHS")
)

# Modelo 55: W + PAK (Arc-cosseno)
Eta55 <- list(
  W = list(K = ZW, model = "RKHS"),
  PAK = list(K = PAK, model = "RKHS")
)

# Modelo 56: W + GAK + PAK (Arc-cosseno)
Eta56 <- list(
  W = list(K = ZW, model = "RKHS"),
  GAK = list(K = GAK, model = "RKHS"),
  PAK = list(K = PAK, model = "RKHS")
)

# -------------------------------------------------------------------
# Tutorial step
# (Modelos Eta57 a Eta64)
# -------------------------------------------------------------------

# Tutorial step
Eta57 <- list(
  E = list(X = ZE, model = "BRR")
)

# Tutorial step
Eta58 <- list(
  G = list(K = ZG, model = "RKHS")
)

# Tutorial step
Eta59 <- list(
  P = list(K = ZP, model = "RKHS")
)

# Tutorial step
Eta60 <- list(
  GGK = list(K = GGK, model = "RKHS")
)

# Tutorial step
Eta61 <- list(
  PGK = list(K = PGK, model = "RKHS")
)

# Tutorial step
Eta62 <- list(
  GAK = list(K = GAK, model = "RKHS")
)

# Tutorial step
Eta63 <- list(
  PAK = list(K = PAK, model = "RKHS")
)

# Tutorial step
# Tutorial step
Eta64 <- list(
  W = list(K = ZW, model = "RKHS")
)
```

This section describes the corresponding methodological step in the workflow.

```{r consolidate-models}
# Tutorial step
Models <- list(Eta1, Eta2, Eta3, Eta4, Eta5, Eta6, Eta7, Eta8, Eta9, Eta10,
               Eta11, Eta12, Eta13, Eta14, Eta15, Eta16, Eta17, Eta18)
length(Models)  # Deve retornar 18
```

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r consolidate-climate-models-comprehensive}
# 1. Primeiro, agrupamos todos os modelos do Grupo 2 e Grupo 3
# Tutorial step
Climate_Models <- list(Eta20, Eta21, Eta22, Eta23, Eta24, Eta25, Eta26, Eta27, Eta28,
                       Eta29, Eta30, Eta31, Eta32, Eta33, Eta34, Eta35, Eta36, Eta37,
                       Eta38, Eta39, Eta40, Eta41, Eta42, Eta43, Eta44, Eta45, Eta46,
                       Eta48, Eta49, Eta50, Eta51, Eta52, Eta53, Eta54, Eta55, Eta56)

# 2. Adicionamos os modelos de efeitos individuais (Grupo 4)
Climate_Models <- c(Climate_Models, list(Eta57, Eta58, Eta59, Eta60, Eta61, Eta62, Eta63, Eta64))

# Tutorial step
# para criar o "menu" completo de modelos.
Models_full_set <- c(Models, Climate_Models)

# Tutorial step
# Tutorial step
cat("Total de modelos a serem avaliados:", length(Models_full_set))
```

## 13) Cross-validation

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

### 13.0) Visualization of cross-validation schemes


```{r example-cv}
# ------------------------------------------------------------
# Tutorial step
# ------------------------------------------------------------
n_geno <- 50  # number of genotypes
n_env  <- 4
envs <- paste0("Env ", 1:n_env)

# ------------------------------------------------------------
# Tutorial step
# ------------------------------------------------------------
generate_cv_data <- function(type) {
  df <- expand.grid(Genotype = 1:n_geno, Env = envs)
  df$Test <- FALSE
  
  if (type == "CV2") {
    # Tutorial step
    test_genos <- sample(1:nrow(df), nrow(df) * 0.2)
    df$Test[test_genos] <- TRUE
    
    
  } else if (type == "CV1") {
    # Tutorial step
    test_genos <- sample(1:n_geno, n_geno * 0.2)
    df$Test[df$Genotype %in% test_genos] <- TRUE
    
  } else if (type == "CV0") {
    # Tutorial step
    df$Test[df$Env == "Env 2"] <- TRUE

    
  } else if (type == "CV00") {
    # Tutorial step
    test_genos <- sample(1:n_geno, n_geno * 0.2)
    df$Test <- FALSE
    df <- df %>% 
      mutate(Test = ifelse(Env == "Env 2" & Genotype %in% test_genos, TRUE, FALSE)) %>%
      filter(!(Genotype %in% test_genos & Env != "Env 2")) %>% # remove test genotypes from other environments
      filter(!(Env == "Env 2" & !(Genotype %in% test_genos))) # keep only test genotypes in Env 2
  }
  
  # Tutorial step
  df$CV <- type
  df
}

# ------------------------------------------------------------
# Tutorial step
# ------------------------------------------------------------
cv_types <- c("CV2", "CV0", "CV1", "CV00")  # ordem original
data <- do.call(rbind, lapply(cv_types, generate_cv_data))

# ------------------------------------------------------------
# Tutorial step
# ------------------------------------------------------------
plot_cv <- function(df, title) {
  ggplot(df, aes(x = as.character(Env), y = Genotype)) +
    # Linhas azuis finas de fundo apenas para pontos de treino
    geom_tile(data = subset(df, !Test), fill = "#5dade2", height = 0.4, width = 0.9) +
    # Tutorial step
    geom_point(data = subset(df, Test), color = "black", size = 2) +
    theme_minimal(base_size = 12) +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      axis.text.x = element_text(size = 10),
      axis.text.y = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      legend.position = "none"
    ) +
    labs(title = title)
}

# ------------------------------------------------------------
# Tutorial step
# ------------------------------------------------------------
p1 <- plot_cv(filter(data, CV == "CV2"), "CV2")
p2 <- plot_cv(filter(data, CV == "CV0"), "CV0")
p3 <- plot_cv(filter(data, CV == "CV1"), "CV1")
p4 <- plot_cv(filter(data, CV == "CV00"), "CV00")

# ------------------------------------------------------------
# Tutorial step
# ------------------------------------------------------------
grid_arrange_shared_legend <- function(..., ncol = 2, nrow = 2, 
                                       x_label = "Environments", 
                                       y_label = "Genotypes") {
  plots <- list(...)
  grid <- arrangeGrob(grobs = plots, ncol = ncol, nrow = nrow)
  grid <- arrangeGrob(
    grid,
    bottom = textGrob(x_label, gp = gpar(fontsize = 14)),
    left   = textGrob(y_label, rot = 90, gp = gpar(fontsize = 14))
  )
  grid
}

final_plot <- grid_arrange_shared_legend(p1, p2, p3, p4)

grid.newpage()
grid.draw(final_plot)

ggsave("output/cv-schemes.tiff", final_plot, width = 10, height = 8)
```


### 13.1) Parameters for model execution

This section describes the corresponding methodological step in the workflow.

```{r cv-params}
# Tutorial step
nIter  <- 5000   # total number of MCMC chain iterations
burnIn <- 1000   # number of burn-in iterations discarded
thin   <- 10     # sampling interval to reduce autocorrelation
num_rep  <- 20   # number of cross-validation repetitions
num_fold <- 5    # number of folds in each repetition
```

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

### 13.2) Quick checks

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r data-checks}
# Basic consistency checks before long CV runs
if (!exists("Pheno")) stop("`Pheno` object not found in the current environment.")
if (!"Pedigree" %in% names(Pheno)) stop("Column `Pedigree` not found in Pheno.")
if (!"GY" %in% names(Pheno)) stop("Column `GY` (response trait) not found in Pheno.")
if (!"KW" %in% names(Pheno)) stop("Column `KW` (response trait) not found in Pheno.")

if (!exists("Pedigree")) Pedigree <- unique(Pheno$Pedigree)
Pedigree <- unique(Pedigree)
n_hybrids <- length(Pedigree)
stopifnot(n_hybrids > 0)

sanity_table <- run_data_sanity_checks(NIR, Geno, Pheno)
print(sanity_table)
stopifnot(all(sanity_table$passed))
```

### 13.3) Output folders

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r create-dirs}
# Tutorial step
dir.create("output/results", recursive = TRUE, showWarnings = FALSE)
```

### 13.5) Helper functions and CV0/CV00 environments

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
    1.  Roda o `BGLR`.
    2.  Salva os arquivos de log do MCMC (via `saveAt`).
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r functions-setup}
# Tutorial step
envs_leaveout <- c("CS11_WS", "CS11_WW", "CS12_WS", "CS12_WW")

# Tutorial step
# Tutorial step
run_bglr <- function(y2_vector, ETA, nIter, burnIn, thin, save_prefix) {
  # Tutorial step
  y_t <- as.numeric(y2_vector)
  dir.create(dirname(save_prefix), recursive = TRUE, showWarnings = FALSE)
  
  # mede tempo
  start_time <- Sys.time()
  
  fit <- BGLR(y = y_t, ETA = ETA, nIter = nIter, burnIn = burnIn, thin = thin, saveAt = save_prefix)
  
  end_time <- Sys.time()
  elapsed <- as.numeric(difftime(end_time, start_time, units = "secs"))
  
  # Tutorial step
  #    This works for ANY model (with 1 or N effects)
  variancias_U <- list()
  if (!is.null(fit$ETA)) {
      nomes_efeitos <- names(fit$ETA)
      if (is.null(nomes_efeitos)) {
          nomes_efeitos <- paste0("Efeito_", seq_along(fit$ETA))
      }
      
      for (i in seq_along(fit$ETA)) {
          efeito_nome <- nomes_efeitos[i]
          # Tutorial step
          if (!is.null(fit$ETA[[i]]$varU)) {
              variancias_U[[efeito_nome]] <- fit$ETA[[i]]$varU
          }
      }
  }
  
  # Tutorial step
  saida_essencial <- list(
      yHat_mean = fit$yHat,        # Predictions (mean)
      varE_mean = fit$varE,        # Residual variance (mean)
      # --- LINHA ADICIONADA ---
      # Tutorial step
      varB_mean = ifelse(is.null(fit$varB), NA, fit$varB), 
      # --- FIM DA LINHA ADICIONADA ---
      variances_U_mean = variancias_U, # List of kernel variances (mean)
      DIC = ifelse(is.null(fit$fit$DIC), NA, fit$fit$DIC),
      runtime_sec = elapsed
  )
  
  # Tutorial step
  #    Isso substitui o saveRDS(fit, ...)
  saveRDS(saida_essencial, paste0(save_prefix, "_metricas.rds"))
  
  # Tutorial step
  
  # 4. Preparar o retorno para o 'foreach' (isto permanece igual)
  # Tutorial step
  retorno_foreach <- list(yHat = fit$yHat, runtime_sec = elapsed)
  
  # Tutorial step
  rm(fit, saida_essencial, variancias_U)
  gc() # Force garbage collection
  
  # Tutorial step
  return(retorno_foreach)
}
# Tutorial step


# Tutorial step
# Tutorial step
calc_cor_by_env <- function(df_with_yhat, trait_col) {
  df_with_yhat %>%
    dplyr::group_by(Env) %>%
    dplyr::summarize(
      cor = cor(.data[[trait_col]], yhat, use = "complete.obs"),
      .groups = "drop"
    ) %>%
    as.data.frame()
}

# Tutorial step
# Tutorial step
combos <- expand.grid(MODEL = 1:length(Models_full_set), REP = seq_len(num_rep), TRAIT = c("GY", "KW"))
```

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r}
# Tutorial step
rm(list = setdiff(ls(), c("Pheno", "Pedigree", "Models_full_set",
                         "nIter", "burnIn", "thin",
                         "num_rep", "num_fold",
                         "envs_leaveout",
                         "run_bglr", "calc_cor_by_env",
                         "combos", "n_hybrids", "model_info")))
```

## 14) Ajuste de modelos

This section describes the corresponding methodological step in the workflow.

### 14.1) Parallel backend configuration

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r parallel-setup, warning=FALSE, message=FALSE}
# Tutorial step
numCores <- parallel::detectCores()
useCores <- max(1, numCores - 1)    # use all cores except one
cl <- parallel::makeCluster(useCores)
doParallel::registerDoParallel(cl)

cat("Running on ", useCores, " cores.")
```

This section describes the corresponding methodological step in the workflow.

```{r}
gc()  # Clear memory before parallel execution
```

### 14.2) Parallel model execution

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r parallel-model, cache=FALSE, eval=FALSE}
# Tutorial step
foreach(i = seq_len(nrow(combos)),
        .packages = c("BGLR", "dplyr", "plyr")) %dopar% {
          MODEL <- combos$MODEL[i]
          rep_num <- combos$REP[i]
          TRAIT <- as.character(combos$TRAIT[i])
          
          pheno <- data.frame(Pheno)  # copy of Pheno for local manipulation
          
          # Tutorial step
          rep_dir <- file.path("output/ajuste_modelo", paste0("rep_", rep_num))
          dir.create(rep_dir, recursive = TRUE, showWarnings = FALSE)
          eta_dir <- file.path(rep_dir, paste0("Eta", MODEL))
          dir.create(eta_dir, recursive = TRUE, showWarnings = FALSE)
          
          # Tutorial step
          CV1_list <- list()
          CV2_list <- list()
          CV0_lists  <- lapply(envs_leaveout, function(x)
            list())
          CV00_lists <- lapply(envs_leaveout, function(x)
            list())
          names(CV0_lists) <- envs_leaveout
          names(CV00_lists) <- envs_leaveout
          
          # Tutorial step
          #set.seed(1000 + MODEL * 100 + rep_num)
          obs_per_fold <- ceiling(n_hybrids / num_fold)
          folds <- sample(rep(seq_len(num_fold), each = obs_per_fold, length.out = n_hybrids))
          folds_df <- data.frame(Pedigree = Pedigree,
                                 fold = folds,
                                 stringsAsFactors = FALSE)
          
          for (fold_num in seq_len(num_fold)) {
            train_geno <- folds_df$Pedigree[folds_df$fold != fold_num]
            test_geno  <- folds_df$Pedigree[folds_df$fold == fold_num]
            
            # --- CV1 / CV2 ---
            yield <- pheno
            yield$Y2 <- NA
            yield$Y2[yield$Pedigree %in% train_geno] <- yield[[TRAIT]][yield$Pedigree %in% train_geno]
            
            save_prefix_full <- file.path(eta_dir,
                                          paste0("full_Eta", MODEL, "_", TRAIT, "_fold", fold_num, "_"))
            #set.seed(2000 + MODEL * 100 + rep_num * 10 + fold_num)
            res_full <- run_bglr(
              yield$Y2,
              ETA = Models_full_set[[MODEL]],
              nIter = nIter,
              burnIn = burnIn,
              thin = thin,
              save_prefix_full
            )
            yield$yhat <- res_full$yHat
            runtime_full <- res_full$runtime_sec
            
            df_test <- yield[!yield$Pedigree %in% train_geno, ]
            df_train <- yield[yield$Pedigree %in% train_geno, ]
            
            tmp1 <- calc_cor_by_env(df_test, TRAIT)
            tmp2 <- calc_cor_by_env(df_train, TRAIT)
            tmp1$Runtime_sec <- runtime_full
            tmp2$Runtime_sec <- runtime_full
            CV1_list[[length(CV1_list) + 1]] <- tmp1
            CV2_list[[length(CV2_list) + 1]] <- tmp2
            
            # --- CV0 / CV00 ---
            for (env_idx in seq_along(envs_leaveout)) {
              env_name <- envs_leaveout[env_idx]
              
              yield2 <- pheno
              yield2$Y2 <- NA
              yield2$Y2[yield2$Pedigree %in% train_geno] <- yield2[[TRAIT]][yield2$Pedigree %in% train_geno]
              yield2$Y2[yield2$Env == env_name] <- NA
              
              save_prefix_env <- file.path(eta_dir,
                                           paste0(env_name, "_", TRAIT, "_Eta", MODEL, "_fold", fold_num, "_"))
              #set.seed(3000 + MODEL * 100 + rep_num * 10 + fold_num + env_idx)
              res_env <- run_bglr(
                yield2$Y2,
                ETA = Models_full_set[[MODEL]],
                nIter = nIter,
                burnIn = burnIn,
                thin = thin,
                save_prefix_env
              )
              yield2$yhat <- res_env$yHat
              runtime_env <- res_env$runtime_sec
              
              df_test2 <- yield2[!yield2$Pedigree %in% train_geno, ]
              df_train2 <- yield2[yield2$Pedigree %in% train_geno, ]
              
              tmp00 <- calc_cor_by_env(df_test2, TRAIT)
              tmp0  <- calc_cor_by_env(df_train2, TRAIT)
              tmp00$Runtime_sec <- runtime_env
              tmp0$Runtime_sec  <- runtime_env
              CV00_lists[[env_idx]][[length(CV00_lists[[env_idx]]) + 1]] <- tmp00
              CV0_lists[[env_idx]][[length(CV0_lists[[env_idx]]) + 1]] <- tmp0
            }
          }
          
          # Tutorial step
          CV1out <- plyr::ldply(CV1_list, data.frame)
          CV2out <- plyr::ldply(CV2_list, data.frame)
          write.csv(CV1out,
                    file = file.path(
                      "output/results",
                      paste0("CV1_", TRAIT, "_Eta", MODEL, "_rep", rep_num, ".csv")
                    ),
                    row.names = FALSE)
          write.csv(CV2out,
                    file = file.path(
                      "output/results",
                      paste0("CV2_", TRAIT, "_Eta", MODEL, "_rep", rep_num, ".csv")
                    ),
                    row.names = FALSE)
          
          for (env_idx in seq_along(envs_leaveout)) {
            env_name <- envs_leaveout[env_idx]
            CV00out <- plyr::ldply(CV00_lists[[env_idx]], data.frame)
            CV0out  <- plyr::ldply(CV0_lists[[env_idx]], data.frame)
            
            write.csv(CV00out,
                      file = file.path(
                        "output/results",
                        paste0(
                          "CV00_",
                          TRAIT,
                          "_",
                          env_name,
                          "_Eta",
                          MODEL,
                          "_rep",
                          rep_num,
                          ".csv"
                        )
                      ),
                      row.names = FALSE)
            write.csv(CV0out,
                      file = file.path(
                        "output/results",
                        paste0(
                          "CV0_",
                          TRAIT,
                          "_",
                          env_name,
                          "_Eta",
                          MODEL,
                          "_rep",
                          rep_num,
                          ".csv"
                        )
                      ),
                      row.names = FALSE)
          }
          
          NULL
        } # end foreach
```

This section describes the corresponding methodological step in the workflow.

```{r stop_cluster, include=FALSE}
# finalize cluster
stopCluster(cl)
```

## 15) Resultados

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

### 15.1) Segregation of CV00 and CV0 results by environment

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r df_consolidation_CV00_CV0, eval = FALSE}
# Tutorial step
list_csv_files_CV0 <- list.files(
  path = "output/results",
  pattern = "CV0_",
  # Tutorial step
  full.names = TRUE
)

# Tutorial step
list_csv_files_CV00 <- list.files(
  path = "output/results",
  pattern = "CV00_",
  # Tutorial step
  full.names = TRUE
)

# Tutorial step
list_csv_files_CV0_CS12_WS <- list_csv_files_CV0[grep("CS12_WS", list_csv_files_CV0)]
df_CV0_CS12_WS <- read_csv(list_csv_files_CV0_CS12_WS, id = "file_name") %>% as.data.frame()
head(df_CV0_CS12_WS)

write.csv(df_CV0_CS12_WS, "output/Pred.ability/Pred.ability.CV0.CS12_WS.csv")  # Save consolidated output

list_csv_files_CV00_CS12_WS <- list_csv_files_CV00[grep("CS12_WS", list_csv_files_CV00)]
df_CV00_CS12_WS <- read_csv(list_csv_files_CV00_CS12_WS, id = "file_name") %>% as.data.frame()
head(df_CV00_CS12_WS)

write.csv(df_CV00_CS12_WS, "output/Pred.ability/Pred.ability.CV00.CS12_WS.csv")  # Save consolidated output

# Tutorial step
list_csv_files_CV0_CS11_WW <- list_csv_files_CV0[grep("CS11_WW", list_csv_files_CV0)]
df_CV0_CS11_WW <- read_csv(list_csv_files_CV0_CS11_WW, id = "file_name") %>% as.data.frame()
head(df_CV0_CS11_WW)

write.csv(df_CV0_CS11_WW, "output/Pred.ability/Pred.ability.CV0.CS11_WW.csv")  # Save consolidated output

list_csv_files_CV00_CS11_WW <- list_csv_files_CV00[grep("CS11_WW", list_csv_files_CV00)]
df_CV00_CS11_WW <- read_csv(list_csv_files_CV00_CS11_WW, id = "file_name") %>% as.data.frame()
head(df_CV00_CS11_WW)

write.csv(df_CV00_CS11_WW, "output/Pred.ability/Pred.ability.CV00.CS11_WW.csv")  # Save consolidated output

# Tutorial step

list_csv_files_CV0_CS12_WW <- list_csv_files_CV0[grep("CS12_WW", list_csv_files_CV0)]
df_CV0_CS12_WW <- read_csv(list_csv_files_CV0_CS12_WW, id = "file_name") %>% as.data.frame()
head(df_CV0_CS12_WW)

write.csv(df_CV0_CS12_WW, "output/Pred.ability/Pred.ability.CV0.CS12_WW.csv")  # Save consolidated output

list_csv_files_CV00_CS12_WW <- list_csv_files_CV00[grep("CS12_WW", list_csv_files_CV00)]
df_CV00_CS12_WW <- read_csv(list_csv_files_CV00_CS12_WW, id = "file_name") %>% as.data.frame()
head(df_CV00_CS12_WW)

write.csv(df_CV00_CS12_WW, "output/Pred.ability/Pred.ability.CV00.CS12_WW.csv")  # Save consolidated output

# Tutorial step
list_csv_files_CV0_CS11_WS <- list_csv_files_CV0[grep("CS11_WS", list_csv_files_CV0)]
df_CV0_CS11_WS <- read_csv(list_csv_files_CV0_CS11_WS, id = "file_name") %>% as.data.frame()
head(df_CV0_CS11_WS)

write.csv(df_CV0_CS11_WS, "output/Pred.ability/Pred.ability.CV0.CS11_WS.csv")  # Save consolidated output

list_csv_files_CV00_CS11_WS <- list_csv_files_CV00[grep("CS11_WS", list_csv_files_CV00)]
df_CV00_CS11_WS <- read_csv(list_csv_files_CV00_CS11_WS, id = "file_name") %>% as.data.frame()
head(df_CV00_CS11_WS)

write.csv(df_CV00_CS11_WS, "output/Pred.ability/Pred.ability.CV00.CS11_WS.csv")  # Save consolidated output

df_CV0_CV00 <- bind_rows(
  df_CV0_CS12_WS,
  df_CV00_CS12_WS,
  df_CV0_CS11_WW,
  df_CV00_CS11_WW,
  df_CV0_CS12_WW,
  df_CV00_CS12_WW,
  df_CV0_CS11_WS,
  df_CV00_CS11_WS
)
write.csv(df_CV0_CV00, "output/Pred.ability/Pred.ability.CV0.CV00.csv")  # Save consolidated output
```

### 15.2) Resultados CV0 e CV00

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

  * Usa `separate` para quebrar o nome do arquivo (ex: `CV0_GY_CS11_WS_Eta1_rep1`) em colunas limpas: `CV`, `Trait`, `Env_Val` (combinando `Env1` e `Env2`), `Model`, `Rep`.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r consolidation}
# Tutorial step
# Tutorial step

df <- read.csv("output/Pred.ability/Pred.ability.CV0.CV00.csv") %>%
  mutate(
    file_name = basename(file_name),
    # Tutorial step
    file_name = sub("\\.csv$", "", file_name)
  ) %>%    # remove extension
  # Tutorial step
  separate(
    file_name,
    into = c("CV", "Trait", "Env1", "Env2", "ModelRaw", "Rep"),
    sep = "_",
    fill = "right",
    # fill with NA if missing
    remove = FALSE
  ) %>%
  # Cria coluna Env apenas se houver Env1 e Env2
  mutate(
    Env_Val = paste0(Env1, "_", Env2),
    Model = sub("^Eta", "M", sub("\\.csv$", "", ModelRaw))) %>%
  dplyr::select(Model, CV, Trait, Env_Val, Env, Rep, cor)  # remove colunas auxiliares

write.csv(df, "output/Pred.ability/Pred.ability.CV0.CV00.classified.csv")  # Save consolidated output
```

### 15.3) Consolidation of CV1 and CV2 results

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

  * `list.files` encontra todos os arquivos `CV1` ou `CV2`.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r df_consolidation_CV1_CV2, eval = FALSE}
# Tutorial step
list_csv_files <- list.files(
  path = "output/results",
  pattern = "CV1|CV2",
  # Tutorial step
  full.names = TRUE
)

df_CV1_CV2 <- read_csv(list_csv_files, id = "file_name") %>% as.data.frame()
head(df_CV1_CV2)

write.csv(df_CV1_CV2, "output/Pred.ability/Pred.ability.CV1.CV2.csv")  # Save consolidated output
```

### 15.4) Resultados CV1 e CV2

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r consolidation-cv1-cv2-parsing}
# Tutorial step
# Tutorial step

df <- read.csv("output/Pred.ability/Pred.ability.CV1.CV2.csv") %>%
  mutate(
    file_name = basename(file_name),
    # Tutorial step
    file_name = sub("\\.csv$", "", file_name)
  ) %>%    # remove extension
  # Tutorial step
  separate(
    file_name,
    into = c("CV", "Trait", "ModelRaw", "Rep"),
    sep = "_",
    fill = "right",
    # fill with NA if missing
    remove = FALSE
  ) %>%
  # Cria coluna Env apenas se houver Env1 e Env2
  mutate(
    Model = sub("^Eta", "M", sub("\\.csv$", "", ModelRaw))) %>%
  dplyr::select(Model, CV, Trait, Env, Rep, cor)  # remove colunas auxiliares

write.csv(df, "output/Pred.ability/Pred.ability.CV1.CV2.classified.csv")  # Save consolidated output
```

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r}
# Tutorial step

rm(list = setdiff(ls(), c("Pheno", "Pedigree", "model_info")))
gc()  # Clear memory
```

### 15.5) Bar plot of predictive ability for CV0 and CV00

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
  * Agrupamos por `Model`, `Env_Val`, `CV` e `Trait`.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r pred-ability}
# Tutorial step
df <- read.csv("output/Pred.ability/Pred.ability.CV0.CV00.classified.csv")


df_cor <- as.data.frame(df %>%
                          group_by(Model, Env_Val, CV, Trait) %>%
                          dplyr::summarise(M = mean(cor, na.rm = TRUE), SD = sd(cor, na.rm = TRUE)))

head(df_cor)

# Tutorial step
df_cor <- df_cor %>%
  left_join(model_info, by = "Model")

head(df_cor)

write.csv(df_cor, "output/Pred.ability/Pred.ability.CV0.CV00.classified.csv")  # Save consolidated output
```

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
  * `geom_bar()`: Cria as barras.
  * `geom_errorbar()`: Adiciona as barras de erro (±SD).
This section describes the corresponding methodological step in the workflow.

```{r pred-ability-plot}
df_cor <- read.csv("output/Pred.ability/Pred.ability.CV0.CV00.classified.csv")
# Tutorial step

p <- df_cor %>% 
  filter(CV == "CV0" & Trait == "GY") %>% 
  mutate(Description = reorder(Description, ModelID)) %>%
  ggplot(aes(x = Description, y = M, fill = factor(Type))) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.8, show.legend = FALSE) +
  geom_errorbar(aes(ymin = M, ymax = M + SD),
                width = .2,
                position = position_dodge(.9)) +
  theme_bw() +
  facet_wrap(. ~  Env_Val, nrow = 4) +  # One panel for each validation scheme
  scale_y_continuous("Prediction ability", limits = c(-0.1,1.25)) +
  scale_x_discrete("Models") +
  theme(axis.text.x = element_text(
    angle = 90,
    vjust = 0.5,
    hjust = 1,
    size = 6
  ))

# Tutorial step
p

ggsave("output/Pred.ability/grafico_exemplo_GY_CV0.tiff",
       plot = p,
       width = 8,
       height = 8, dpi = 300)
```

```{r}
df_cor <- read.csv("output/Pred.ability/Pred.ability.CV0.CV00.classified.csv")
# Tutorial step

p <- df_cor %>% 
  group_by(Model, CV, Trait) %>% 
  summarise(M = mean(M, na.rm = TRUE), SD = sd(M, na.rm = TRUE)) %>% 
  full_join(model_info) %>% 
  ggplot(aes(x = reorder(Description, ModelID), y = M, fill = factor(Type))) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.8, show.legend = FALSE) +
  geom_errorbar(aes(ymin = M, ymax = M + SD),
                width = .2,
                position = position_dodge(.9)) +
  theme_bw() +
  facet_grid( CV ~  Trait) +  # One panel for each validation scheme
  scale_y_continuous(limits = c(-0.1,1.25)) +
  #scale_x_discrete("Models") +
  theme(axis.text.x = element_text(
    angle = 90,
    vjust = 0.5,
    hjust = 1,
    size = 6
  )) +
  labs(x = "Models",
       y = "Prediction ability")

# Tutorial step
p

ggsave("output/Pred.ability/grafico_exemplo_CV0_CV00.tiff",
       plot = p,
       width = 16,
       height = 10, dpi = 300)
```

### 15.6) Automated plot generation (CV0 and CV00)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
4.  **Conditional Highlighting**:
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
5.  **Plot**:
This section describes the corresponding methodological step in the workflow.
      * `scale_fill_manual` defines colors (blue for "Above", gray for "Below").
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r, eval=FALSE}
#-----------------------------------------------------------------------
# Tutorial step
#-----------------------------------------------------------------------

# Tutorial step
# Tutorial step
df_cor <- read.csv("output/Pred.ability/Pred.ability.CV0.CV00.classified.csv") %>% as.data.frame()

# Tutorial step
# Tutorial step
output_dir <- "output/Pred.ability/graficos_gerados"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}


#-----------------------------------------------------------------------
# Tutorial step
#-----------------------------------------------------------------------

# Tutorial step
combinacoes <- df_cor %>%
  distinct(Trait, CV, Env_Val, Type)

# Tutorial step
print("Combinations found for plot generation:")
head(combinacoes)

# Tutorial step
for (i in 1:nrow(combinacoes)) {
  
  # Tutorial step
  trait_atual <- combinacoes$Trait[i]
  cv_atual <- combinacoes$CV[i]
  type_atual <- combinacoes$Type[i]
  
  # Progress message in console
  message(paste("Generating plot for: Trait =", trait_atual, "| CV =", cv_atual, "| Type =", type_atual))
  
  # Tutorial step
  df_filtrado <- df_cor %>%
    filter(Trait == trait_atual & CV == cv_atual & Type == type_atual)
  
  # Tutorial step
  if(nrow(df_filtrado) == 0) {
    message(" -> No data found. Skipping.")
    next 
  }
  
  # Tutorial step
  
  # Tutorial step
  media_M_grupo <- mean(df_filtrado$M, na.rm = TRUE)
  
  # Tutorial step
  # Tutorial step
  df_plotagem <- df_filtrado %>%
    mutate(
      destaque = ifelse(M > media_M_grupo, 
                        "Above Mean", 
                        "Below/Equal to Mean")
    )
  
  # Tutorial step
  
  
  # Tutorial step
  p <- df_plotagem %>%
    # Tutorial step
    mutate(Description = reorder(Description, ModelID)) %>%
    # Tutorial step
    # 'fill' now uses the new 'highlight' column
    ggplot(aes(x = Description, y = M, fill = destaque)) +
    
    # 'show.legend = TRUE' (or omit) to show the new color legend
    geom_bar(stat = "identity", position = position_dodge(), show.legend = TRUE) +
    
    geom_errorbar(aes(ymin = M, ymax = M + SD),
                  width = .2,
                  position = position_dodge(.9)) +
    
    # Tutorial step
    geom_text(
      aes(label = round(M, 2), y = M + SD), 
      vjust = -0.5,    
      color = "black",  
      position = position_dodge(0.9),
      angle = 0,        
      size = 3.5        
    ) +
    
    facet_grid(Env_Val ~ .) + 
    
    theme_bw() +
    
    # Tutorial step
    scale_y_continuous(
      "Mean Value (M)",
      expand = expansion(mult = c(0, 0.15)) 
    ) +

    # Tutorial step
    scale_fill_manual(
      name = "Performance:", # Legend title
      values = c("Above Mean" = "#0072B2",  # Color for values above mean (e.g., blue)
                 "Below/Equal to Mean" = "grey80") # Color for other values (e.g., gray)
    ) +
    
    labs(
      title = paste("Predictive Ability"),
      subtitle = paste("Trait:", trait_atual, "| CV:", cv_atual, "| Type:", type_atual,
                       # Tutorial step
                       sprintf("\n(Group mean: %.2f)", media_M_grupo)), 
      x = "Model description"
    ) +
    theme(
      axis.text.x = element_text(
        angle = 90,
        vjust = 0.5,
        hjust = 1, 
        size = 8
      ),
      legend.position = "top" # Move legend to top of plot
    )
  
  # Tutorial step
  nome_arquivo <- sprintf("%s/grafico_%s_%s_%s.tiff", 
                          output_dir, 
                          trait_atual, 
                          cv_atual, 
                          type_atual)
  
  # Tutorial step
  ggsave(
    filename = nome_arquivo, 
    plot = p, 
    width = 10,       # Width in inches
    height = 7,       # Height in inches
    dpi = 300         # Resolution (image quality)
  )
}

message("\nProcess finished! Plots saved in '", output_dir, "'")
```

### 15.7) Bar plot of predictive ability for CV1 and CV2

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
  * Agrupamos por `Model`, `CV` e `Trait`.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r pred-ability-cv1-cv2-summary}
# Tutorial step
df <- read.csv("output/Pred.ability/Pred.ability.CV1.CV2.classified.csv")

df_cor <- as.data.frame(df %>%
                          group_by(Model, CV, Trait) %>%
                          dplyr::summarise(M = mean(cor, na.rm = TRUE), SD = sd(cor, na.rm = TRUE)))

head(df_cor)

# Tutorial step
df_cor <- df_cor %>%
  left_join(model_info, by = "Model")

head(df_cor)

write.csv(df_cor, "output/Pred.ability/Pred.ability.CV1.CV2.classified.csv")  # Save consolidated output
```

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

```{r pred-ability-plot-cv1-cv2}
df_cor <- read.csv("output/Pred.ability/Pred.ability.CV1.CV2.classified.csv")
# Tutorial step

p <- df_cor %>% 
  filter(Trait == "GY") %>%
  #mutate(Description = forcats::fct_reorder(Description, ModelID)) %>%
  ggplot(aes(x = reorder(Description, ModelID), y = M, fill = factor(Type))) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.8, show.legend = FALSE) +
  geom_errorbar(aes(ymin = M, ymax = M + SD),
                width = .2,
                position = position_dodge(.9)) +
  theme_bw() +
  facet_grid(CV ~ .) +  # One panel for each validation scheme
  scale_y_continuous("Prediction ability", limits = c(-0.1,1.25)) +
  scale_x_discrete("Models") +
  theme(axis.text.x = element_text(
    angle = 90,
    vjust = 0.5,
    hjust = 1,
    size = 6
  ))

# Tutorial step
p

ggsave("output/Pred.ability/grafico_CV1_CV2.tiff", plot = p, width = 8, height = 8, dpi = 300)
```

### 15.8) Automated plot generation (CV1 and CV2)

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
4.  **Conditional Highlighting**:
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.
This section describes the corresponding methodological step in the workflow.

This section describes the corresponding methodological step in the workflow.

```{r loop-plot-cv1-cv2, eval=FALSE}
#-----------------------------------------------------------------------
# Tutorial step
#-----------------------------------------------------------------------

# Tutorial step
# Tutorial step
df_cor <- read.csv("output/Pred.ability/Pred.ability.CV1.CV2.classified.csv")

# Tutorial step
# Tutorial step
output_dir <- "output/Pred.ability/graficos_gerados"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}


#-----------------------------------------------------------------------
# Tutorial step
#-----------------------------------------------------------------------

# Tutorial step
combinacoes <- df_cor %>%
  distinct(Trait, CV, Type)

# Tutorial step
print("Combinations found for plot generation:")
print(combinacoes)

# Tutorial step
for (i in 1:nrow(combinacoes)) {
  
  # Tutorial step
  trait_atual <- combinacoes$Trait[i]
  cv_atual <- combinacoes$CV[i]
  type_atual <- combinacoes$Type[i]
  
  # Progress message in console
  message(paste("Generating plot for: Trait =", trait_atual, "| CV =", cv_atual, "| Type =", type_atual))
  
  # Tutorial step
  df_filtrado <- df_cor %>%
    filter(Trait == trait_atual & CV == cv_atual & Type == type_atual)
  
  # Tutorial step
  if(nrow(df_filtrado) == 0) {
    message(" -> No data found. Skipping.")
    next 
  }
  
  # Tutorial step
  
  # Tutorial step
  media_M_grupo <- mean(df_filtrado$M, na.rm = TRUE)
  
  # Tutorial step
  # Tutorial step
  df_plotagem <- df_filtrado %>%
    mutate(
      destaque = ifelse(M > media_M_grupo, 
                        "Above Mean", 
                        "Below/Equal to Mean")
    )
  
  # Tutorial step
  
  
  # Tutorial step
  p <- df_plotagem %>%
    # Tutorial step
    mutate(Description = reorder(Description, ModelID)) %>%
    
    # Tutorial step
    # 'fill' now uses the new 'highlight' column
    ggplot(aes(x = Description, y = M, fill = destaque)) +
    
    # 'show.legend = TRUE' (or omit) to show the new color legend
    geom_bar(stat = "identity", position = position_dodge(), show.legend = TRUE) +
    
    geom_errorbar(aes(ymin = M, ymax = M + SD),
                  width = .2,
                  position = position_dodge(.9)) +
    
    # Tutorial step
    geom_text(
      aes(label = round(M, 2), y = M + SD), 
      vjust = -0.5,    
      color = "black",  
      position = position_dodge(0.9),
      angle = 0,        
      size = 3.5        
    ) +
    
    theme_bw() +
    
    # Tutorial step
    scale_y_continuous(
      "Mean Value (M)",
      expand = expansion(mult = c(0, 0.15)) 
    ) +

    # Tutorial step
    scale_fill_manual(
      name = "Performance:", # Legend title
      values = c("Above Mean" = "#0072B2",  # Color for values above mean (e.g., blue)
                 "Below/Equal to Mean" = "grey80") # Color for other values (e.g., gray)
    ) +
    
    labs(
      title = paste("Predictive Ability"),
      subtitle = paste("Trait:", trait_atual, "| CV:", cv_atual, "| Type:", type_atual,
                       # Tutorial step
                       sprintf("\n(Group mean: %.2f)", media_M_grupo)), 
      x = "Model description"
    ) +
    theme(
      axis.text.x = element_text(
        angle = 90,
        vjust = 0.5,
        hjust = 1, 
        size = 8
      ),
      legend.position = "top" # Move legend to top of plot
    )
  
  # Tutorial step
  nome_arquivo <- sprintf("%s/grafico_%s_%s_%s.tiff", 
                          output_dir, 
                          trait_atual, 
                          cv_atual, 
                          type_atual)
  
  # Tutorial step
  ggsave(
    filename = nome_arquivo, 
    plot = p, 
    width = 10,       # Width in inches
    height = 7,       # Height in inches
    dpi = 300         # Resolution (image quality)
  )
}

message("\nProcess finished! Plots saved in '", output_dir, "'")
```
