---
title: "Prediction analyses - didactic tutorial"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

**Language / Idioma:** [English](analysis_prediction.html) | [Português](analysis_prediction_pt.html)

## 1. Introduction

This tutorial performs the prediction analyses for the reduced set of 18 Bayesian multi-kernel models used in the study.

The script combines:

- genomic kernels,
- phenomic kernels,
- the categorical environment effect (`E`),
- and the weather kernel (`W`).

In this workflow, the weather kernel is used mainly in models **M10-M18**. It should be interpreted as a weather-derived source of similarity that complements, but does not replace, the categorical environment effect.

The phenomic component (`P`) should also be interpreted with caution. NIR-based similarity may capture useful predictive structure, but it can also reflect strong physiological response to the environment rather than only stable biological similarity.

A model that only fits the observed data well is not enough; it must be able to **predict data it has never seen**. To simulate that, we use Cross-Validation (CV).

This script implements multiple CV schemes to simulate different real-world challenges in breeding:

- **CV1: genotype-based validation (tests unobserved genotypes).**
    - *Question:* Can the model predict the performance of a genotype (hybrid) that is *completely new*, and that was not present in *any* training environment?
- **CV2: cell-based validation (tests partially observed genotype × environment combinations).**
    - *Question:* Can the model predict the performance of a *known* genotype in an environment that is *new for that genotype* (assuming the genotype was observed in other environments and the environment was observed with other genotypes)?
- **CV0: leave-one-environment-out (tests entirely unobserved environments).**
    - *Question:* The hardest test. Can the model predict the performance of *all* genotypes in a *completely new environment* (a year or location the model has never seen)?
- **CV00: leave-one-environment-out (tests genotypes in unobserved environments).**
    - *Similar to CV0*, but focused on predicting genotypes *within* a held-out environment.

The remaining sections show how these validation schemes are operationalized in the prediction workflow, how the jobs are distributed in parallel, and how the outputs are written for the downstream visualization step in `visualization.Rmd` and `visualization.html`.

## Inputs, outputs and next step

**Inputs of this step**
- matrices and kernels generated by `matrizes.Rmd`
- the reduced set of 18 prediction models adopted in the project

**Outputs of this step**
- general prediction settings
- the structured model catalog
- the split `Eta_list` specification
- the conceptual design of `CV1`, `CV2`, `CV0`, and `CV00`

**Next step**
- `analysis_prediction_run_outputs.Rmd` documents the operational execution of prediction jobs and the organization of the output files used later by `visualization.Rmd`.

## 2. Packages and options

```{r setup}
rm(list = ls())
options(scipen = 999)

library(BGLR)
library(dplyr)
library(readr)
library(stringr)
library(foreach)
library(doParallel)
library(parallel)
```

## 3. Inputs generated by previous scripts

```{r load-inputs}
Pheno <- readRDS("output/Matrizes/Pheno.rds")

ZG   <- readRDS("output/Matrizes/ZG.rds")
ZP   <- readRDS("output/Matrizes/ZP.rds")
ZE   <- readRDS("output/Matrizes/ZE.rds")
ZW   <- readRDS("output/Matrizes/ZW.rds")

ZGZE <- readRDS("output/Matrizes/ZGZE.rds")
ZPZE <- readRDS("output/Matrizes/ZPZE.rds")
ZGZW <- readRDS("output/Matrizes/ZGZW.rds")
ZPZW <- readRDS("output/Matrizes/ZPZW.rds")

GGK  <- readRDS("output/Matrizes/GGK.rds")
PGK  <- readRDS("output/Matrizes/PGK.rds")
GAK  <- readRDS("output/Matrizes/GAK.rds")
PAK  <- readRDS("output/Matrizes/PAK.rds")

GGKE <- readRDS("output/Matrizes/GGKE.rds")
PGKE <- readRDS("output/Matrizes/PGKE.rds")
GAKE <- readRDS("output/Matrizes/GAKE.rds")
PAKE <- readRDS("output/Matrizes/PAKE.rds")

GGKW <- readRDS("output/Matrizes/GGKW.rds")
PGKW <- readRDS("output/Matrizes/PGKW.rds")
GAKW <- readRDS("output/Matrizes/GAKW.rds")
PAKW <- readRDS("output/Matrizes/PAKW.rds")
```

The current `matrizes.Rmd` and `matrizes.html` save the observation-level matrices using the following names:

- `ZE` for the categorical environment design matrix
- `ZGZE` and `ZPZE` for genomic-by-environment and phenomic-by-environment interactions
- `ZGZW` and `ZPZW` for genomic-by-weather and phenomic-by-weather interactions

For that reason, this prediction script reads those matrices directly from the files produced by `matrizes.Rmd`.

## 4. MCMC and parallel settings

```{r mcmc-settings}
nIter <- 10000
burnIn <- 1000
thin <- 10
num_rep <- 20

n_cores <- max(1, parallel::detectCores() - 1)

nIter
burnIn
thin
num_rep
n_cores
```

## 5. Model catalog

```{r model-catalog}
model_catalog <- data.frame(
  ModelNumber = 1:18,
  Model = sprintf("M%02d", 1:18),
  Description = c(
    "E + G + GE",
    "E + P + PE",
    "E + GGK + GGKE",
    "E + PGK + PGKE",
    "E + GAK + GAKE",
    "E + PAK + PAKE",
    "E + G + P + GE + PE",
    "E + GGK + PGK + GGKE + PGKE",
    "E + GAK + PAK + GAKE + PAKE",
    "E + W + G + GE + GW",
    "E + W + P + PE + PW",
    "E + W + G + P + GE + PE + GW + PW",
    "E + W + GGK + GGKE + GGKW",
    "E + W + PGK + PGKE + PGKW",
    "E + W + GGK + PGK + GGKE + PGKE + GGKW + PGKW",
    "E + W + GAK + GAKE + GAKW",
    "E + W + PAK + PAKE + PAKW",
    "E + W + GAK + PAK + GAKE + PAKE + GAKW + PAKW"
  ),
  stringsAsFactors = FALSE
)

model_catalog
```

## 6. Model specification

```{r eta-list}
Eta_list <- list()

Eta_list[[1]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"))

Eta_list[[2]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"))

Eta_list[[3]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"))

Eta_list[[4]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = PGKE, model = "RKHS"))

Eta_list[[5]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"))

Eta_list[[6]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = PAKE, model = "RKHS"))

Eta_list[[7]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"))

Eta_list[[8]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"),
                       list(K = PGKE, model = "RKHS"))

Eta_list[[9]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"),
                       list(K = PAKE, model = "RKHS"))

Eta_list[[10]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"),
                       list(K = ZGZW, model = "RKHS"))

Eta_list[[11]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"),
                       list(K = ZPZW, model = "RKHS"))

Eta_list[[12]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"),
                       list(K = ZGZW, model = "RKHS"),
                       list(K = ZPZW, model = "RKHS"))

Eta_list[[13]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"),
                       list(K = GGKW, model = "RKHS"))

Eta_list[[14]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = PGKE, model = "RKHS"),
                       list(K = PGKW, model = "RKHS"))

Eta_list[[15]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"),
                       list(K = PGKE, model = "RKHS"),
                       list(K = GGKW, model = "RKHS"),
                       list(K = PGKW, model = "RKHS"))

Eta_list[[16]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"),
                       list(K = GAKW, model = "RKHS"))

Eta_list[[17]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = PAKE, model = "RKHS"),
                       list(K = PAKW, model = "RKHS"))

Eta_list[[18]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"),
                       list(K = PAKE, model = "RKHS"),
                       list(K = GAKW, model = "RKHS"),
                       list(K = PAKW, model = "RKHS"))
```

## Stage navigation

**Previous stage:** [Variance components](variance_components.html)  
**Next stage:** [Prediction — Part II](analysis_prediction_run_outputs.html)
