---
title: "Prediction analyses I - setup, models, and CV design"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

**Language / Idioma:** [English](analysis_prediction_01_setup_models_cv.html) | [Português](analysis_prediction_01_setup_models_cv_pt.html)

**Module navigation / Navegação do módulo:** [Prediction overview](analysis_prediction.html) | [Part II: run prediction and outputs](analysis_prediction_02_runprediction_outputs.html)

## 1. Introduction

This file is the first didactic submodule of the prediction stage. Its purpose is to prepare the reader before the computationally expensive execution stage.

Here we focus on three questions:

1. what objects enter the prediction step;
2. which reduced models are compared;
3. how the cross-validation schemes are conceptually defined.

By separating setup, model specification, and CV design from the execution stage, the tutorial becomes easier to follow and maintain.

## 2. Packages and options

This chunk resets the R session for a clean run, disables scientific notation in printed outputs, and loads the packages used later for Bayesian modeling, data handling, and parallel execution.

```{r setup}
rm(list = ls())
options(scipen = 999)

library(BGLR)
library(dplyr)
library(readr)
library(stringr)
library(foreach)
library(doParallel)
library(parallel)
```

## 3. Inputs generated by previous scripts

This chunk reads all matrices and kernels produced in the previous data-harmonization stage so that the predictive analyses start from the saved artifacts rather than rebuilding them.

```{r load-inputs}
Pheno <- readRDS("output/Matrizes/Pheno.rds")

ZG   <- readRDS("output/Matrizes/ZG.rds")
ZP   <- readRDS("output/Matrizes/ZP.rds")
ZE   <- readRDS("output/Matrizes/ZE.rds")
ZW   <- readRDS("output/Matrizes/ZW.rds")

ZGZE <- readRDS("output/Matrizes/ZGZE.rds")
ZPZE <- readRDS("output/Matrizes/ZPZE.rds")
ZGZW <- readRDS("output/Matrizes/ZGZW.rds")
ZPZW <- readRDS("output/Matrizes/ZPZW.rds")

GGK  <- readRDS("output/Matrizes/GGK.rds")
PGK  <- readRDS("output/Matrizes/PGK.rds")
GAK  <- readRDS("output/Matrizes/GAK.rds")
PAK  <- readRDS("output/Matrizes/PAK.rds")

GGKE <- readRDS("output/Matrizes/GGKE.rds")
PGKE <- readRDS("output/Matrizes/PGKE.rds")
GAKE <- readRDS("output/Matrizes/GAKE.rds")
PAKE <- readRDS("output/Matrizes/PAKE.rds")

GGKW <- readRDS("output/Matrizes/GGKW.rds")
PGKW <- readRDS("output/Matrizes/PGKW.rds")
GAKW <- readRDS("output/Matrizes/GAKW.rds")
PAKW <- readRDS("output/Matrizes/PAKW.rds")
```

The current `matrizes.Rmd` and `matrizes.html` save the observation-level matrices using the following names:

- `ZE` for the categorical environment design matrix;
- `ZGZE` and `ZPZE` for genomic-by-environment and phenomic-by-environment interactions;
- `ZGZW` and `ZPZW` for genomic-by-weather and phenomic-by-weather interactions.

For that reason, this prediction submodule reads those matrices directly from the files produced by `matrizes.Rmd`.

## 4. MCMC and general settings

This chunk defines the MCMC configuration, the number of repetitions, and the number of processor cores available, because these values control both computational cost and posterior stability.

```{r mcmc-settings}
nIter <- 10000
burnIn <- 1000
thin <- 10
num_rep <- 20

n_cores <- max(1, parallel::detectCores() - 1)

nIter
burnIn
thin
num_rep
n_cores
```

## 5. Model catalog

This chunk creates the catalog of the 18 reduced models used throughout the prediction stage, linking each model identifier to the corresponding biological and kernel specification.

```{r model-catalog}
model_catalog <- data.frame(
  ModelNumber = 1:18,
  Model = sprintf("M%02d", 1:18),
  Description = c(
    "E + G + GE",
    "E + P + PE",
    "E + GGK + GGKE",
    "E + PGK + PGKE",
    "E + GAK + GAKE",
    "E + PAK + PAKE",
    "E + G + P + GE + PE",
    "E + GGK + PGK + GGKE + PGKE",
    "E + GAK + PAK + GAKE + PAKE",
    "E + W + G + GE + GW",
    "E + W + P + PE + PW",
    "E + W + G + P + GE + PE + GW + PW",
    "E + W + GGK + GGKE + GGKW",
    "E + W + PGK + PGKE + PGKW",
    "E + W + GGK + PGK + GGKE + PGKE + GGKW + PGKW",
    "E + W + GAK + GAKE + GAKW",
    "E + W + PAK + PAKE + PAKW",
    "E + W + GAK + PAK + GAKE + PAKE + GAKW + PAKW"
  ),
  stringsAsFactors = FALSE
)

model_catalog
```

## 6. Model specification

The next blocks translate each conceptual model into a concrete BGLR `ETA` specification. To keep the tutorial more readable, the model list is split into smaller groups.

### 6.1 Initialize the ETA container

This chunk creates the empty list that will store the 18 model specifications.

```{r eta-list-initialize}
Eta_list <- list()
```

### 6.2 Models 1 to 6: single-source pathways with `E`

These are the simplest models in the reduced catalog. Each one combines the categorical environment effect with one main kernel and its interaction counterpart.

```{r eta-list-models-1-6}
Eta_list[[1]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"))

Eta_list[[2]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"))

Eta_list[[3]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"))

Eta_list[[4]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = PGKE, model = "RKHS"))

Eta_list[[5]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"))

Eta_list[[6]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = PAKE, model = "RKHS"))
```

### 6.3 Models 7 to 9: integrated genomic and phenomic pathways without `W`

This group combines two main information sources at the same time, still without the weather kernel.

```{r eta-list-models-7-9}
Eta_list[[7]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"))

Eta_list[[8]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"),
                       list(K = PGKE, model = "RKHS"))

Eta_list[[9]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"),
                       list(K = PAKE, model = "RKHS"))
```

### 6.4 Models 10 to 12: weather-augmented linear pathways

This group introduces the weather kernel and its interactions with the linear genomic or phenomic structures.

```{r eta-list-models-10-12}
Eta_list[[10]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"),
                       list(K = ZGZW, model = "RKHS"))

Eta_list[[11]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"),
                       list(K = ZPZW, model = "RKHS"))

Eta_list[[12]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"),
                       list(K = ZGZW, model = "RKHS"),
                       list(K = ZPZW, model = "RKHS"))
```

### 6.5 Models 13 to 18: weather-augmented nonlinear pathways

The final group contains Gaussian and arc-cosine kernels enriched with weather-derived information.

```{r eta-list-models-13-18}
Eta_list[[13]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"),
                       list(K = GGKW, model = "RKHS"))

Eta_list[[14]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = PGKE, model = "RKHS"),
                       list(K = PGKW, model = "RKHS"))

Eta_list[[15]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"),
                       list(K = PGKE, model = "RKHS"),
                       list(K = GGKW, model = "RKHS"),
                       list(K = PGKW, model = "RKHS"))

Eta_list[[16]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"),
                       list(K = GAKW, model = "RKHS"))

Eta_list[[17]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = PAKE, model = "RKHS"),
                       list(K = PAKW, model = "RKHS"))

Eta_list[[18]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"),
                       list(K = PAKE, model = "RKHS"),
                       list(K = GAKW, model = "RKHS"),
                       list(K = PAKW, model = "RKHS"))
```

### 6.6 Final integrity check of the ETA list

This final chunk confirms that the number of ETA objects matches the number of rows in the model catalog.

```{r eta-list-check}
stopifnot(length(Eta_list) == nrow(model_catalog))
length(Eta_list)
```

## 7. Cross-validation design

The next blocks do not run the expensive prediction stage yet. Instead, they document the design logic of the four validation schemes used later in the execution submodule.

### 7.1 Summary table of the validation scenarios

This chunk creates a compact table showing what is masked and what scientific question each scenario is intended to answer.

```{r cv-scenarios-table}
cv_scenarios <- data.frame(
  CV = c("CV1", "CV2", "CV0", "CV00"),
  WhatIsMasked = c(
    "Entire genotypes",
    "Selected genotype × environment cells",
    "One environment at a time",
    "Genotypes within a held-out environment"
  ),
  MainQuestion = c(
    "Can the model predict completely new genotypes?",
    "Can the model predict a known genotype in a new environment for that genotype?",
    "Can the model predict all genotypes in a completely new environment?",
    "Can the model predict new genotypes within a completely new environment?"
  ),
  stringsAsFactors = FALSE
)

cv_scenarios
```

### 7.2 Operational logic of CV1 and CV2

This chunk summarizes the logic behind CV1 and CV2, emphasizing the difference between fully unobserved genotypes and partially unobserved genotype-by-environment cells.

```{r cv1-cv2-design}
cv12_design <- data.frame(
  Scenario = c("CV1", "CV2"),
  TrainingSet = c(
    "All observations except those from held-out genotypes",
    "All observations except selected masked cells"
  ),
  EvaluationTarget = c(
    "Predictions for unobserved genotypes",
    "Predictions for masked genotype × environment cells"
  ),
  stringsAsFactors = FALSE
)

cv12_design
```

### 7.3 Operational logic of CV0 and CV00

This chunk summarizes the logic behind CV0 and CV00, emphasizing the role of the held-out environment in the most difficult prediction scenarios.

```{r cv0-cv00-design}
cvloo_design <- data.frame(
  Scenario = c("CV0", "CV00"),
  TrainingSet = c(
    "All observations except those from the held-out environment",
    "All observations except target genotypes within the held-out environment"
  ),
  EvaluationTarget = c(
    "Predictions for all genotypes in the held-out environment",
    "Predictions for new genotypes in the held-out environment"
  ),
  stringsAsFactors = FALSE
)

cvloo_design
```

### 7.4 File-naming patterns expected downstream

This chunk records the naming convention used later by the execution step and by `visualization.Rmd`, so that the reader can connect the computational outputs to the downstream summaries.

```{r filename-patterns}
prediction_file_patterns <- data.frame(
  Scenario = c("CV1", "CV2", "CV0", "CV00"),
  ExampleFile = c(
    "CV1_GY_M05_rep2_fold5.csv",
    "CV2_KW_M12_rep8_fold3.csv",
    "CV0_GY_CS11_WS_M10_rep1_fold1.csv",
    "CV00_KW_CS12_WW_M18_rep1_fold1.csv"
  ),
  stringsAsFactors = FALSE
)

prediction_file_patterns
```

## 8. Transition to the execution submodule

At this point, the tutorial has defined the inputs, the reduced model set, and the conceptual logic of the validation schemes. The next submodule, `analysis_prediction_02_runprediction_outputs.Rmd`, concentrates the computationally expensive execution stage and the organization of the output files.
