---
title: "Prediction analyses - run prediction and outputs"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

**Language / Idioma:** [English](analysis_prediction_run_outputs.html) | [Português](analysis_prediction_run_outputs_pt.html)

**Module navigation / Navegação do módulo:** [Part I: setup, models and CV design](analysis_prediction.html)

## 1. Introduction

This submodule is the execution counterpart of `analysis_prediction.Rmd`. Its role is to run the prediction jobs for the reduced catalog of 18 models, save the CSV files under `output/results/`, and inventory the outputs that will later be consumed by the visualization stage.

The execution contract used here has four operational rules:

- `CV1` and `CV2` are derived from the same fit within each fold;
- `CV0` and `CV00` are derived from the same fit for each held-out environment within each fold;
- the fold partition is deterministic for each `MODEL × REP × TRAIT` task and is reused in both CV blocks;
- output files are aggregated by repetition, with the fold recorded in the internal `.id` column rather than in the file name.

The heavy fitting chunks remain disabled by default in the tutorial. Even so, the code shown below is the executable version that should be used when the prediction stage is rerun.

## 2. Packages and project paths

The execution module needs both the matrix objects created upstream and the packages used by the parallel fitting layer. The path logic below keeps the file runnable from either the project root or the `analysis/` directory.


```{r setup}
rm(list = ls())
options(scipen = 999)

library(readr)
library(dplyr)
library(stringr)
library(knitr)
library(BGLR)
library(foreach)
library(doParallel)
library(parallel)
library(tidyr)

project_root <- getwd()

if (!dir.exists(file.path(project_root, "output")) &&
    dir.exists(file.path(project_root, "..", "output"))) {
  project_root <- normalizePath(file.path(project_root, ".."))
}

if (!dir.exists(file.path(project_root, "output"))) {
  stop("Could not find the folder 'output/'. Render this file from the project root or from analysis/.")
}

results_dir <- file.path(project_root, "output", "results")
analysis_tables_dir <- file.path(project_root, "output", "tables", "analysis_prediction")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(analysis_tables_dir, recursive = TRUE, showWarnings = FALSE)
```


## 3. Reload the execution objects from Part I

This chunk rebuilds the execution environment when the file is rendered on its own. It respects the contract of `analysis_prediction.Rmd` and recreates the reduced set of 18 ETA objects exactly as defined in Part I.

```{r load-required-objects, eval=FALSE}
matrices_dir <- file.path(project_root, "output", "Matrizes")

required_matrix_files <- c(
  "Pheno.rds",
  "ZG.rds", "ZP.rds", "ZE.rds", "ZW.rds",
  "ZGZE.rds", "ZPZE.rds", "ZGZW.rds", "ZPZW.rds",
  "GGK.rds", "PGK.rds", "GAK.rds", "PAK.rds",
  "GGKE.rds", "PGKE.rds", "GAKE.rds", "PAKE.rds",
  "GGKW.rds", "PGKW.rds", "GAKW.rds", "PAKW.rds"
)

required_matrix_paths <- file.path(matrices_dir, required_matrix_files)

if (!all(file.exists(required_matrix_paths))) {
  missing_matrix_files <- required_matrix_files[!file.exists(required_matrix_paths)]
  stop(
    paste0(
      "The execution objects required by analysis_prediction_run_outputs were not found in output/Matrizes/. Missing files: ",
      paste(missing_matrix_files, collapse = ", ")
    )
  )
}

Pheno <- readRDS(file.path(matrices_dir, "Pheno.rds"))
ZG   <- readRDS(file.path(matrices_dir, "ZG.rds"))
ZP   <- readRDS(file.path(matrices_dir, "ZP.rds"))
ZE   <- readRDS(file.path(matrices_dir, "ZE.rds"))
ZW   <- readRDS(file.path(matrices_dir, "ZW.rds"))
ZGZE <- readRDS(file.path(matrices_dir, "ZGZE.rds"))
ZPZE <- readRDS(file.path(matrices_dir, "ZPZE.rds"))
ZGZW <- readRDS(file.path(matrices_dir, "ZGZW.rds"))
ZPZW <- readRDS(file.path(matrices_dir, "ZPZW.rds"))
GGK  <- readRDS(file.path(matrices_dir, "GGK.rds"))
PGK  <- readRDS(file.path(matrices_dir, "PGK.rds"))
GAK  <- readRDS(file.path(matrices_dir, "GAK.rds"))
PAK  <- readRDS(file.path(matrices_dir, "PAK.rds"))
GGKE <- readRDS(file.path(matrices_dir, "GGKE.rds"))
PGKE <- readRDS(file.path(matrices_dir, "PGKE.rds"))
GAKE <- readRDS(file.path(matrices_dir, "GAKE.rds"))
PAKE <- readRDS(file.path(matrices_dir, "PAKE.rds"))
GGKW <- readRDS(file.path(matrices_dir, "GGKW.rds"))
PGKW <- readRDS(file.path(matrices_dir, "PGKW.rds"))
GAKW <- readRDS(file.path(matrices_dir, "GAKW.rds"))
PAKW <- readRDS(file.path(matrices_dir, "PAKW.rds"))

if (!exists("Pedigree")) Pedigree <- unique(Pheno$Pedigree)

candidate_traits <- c("GY", "KW")
traits_to_run <- candidate_traits[candidate_traits %in% names(Pheno)]
if (length(traits_to_run) == 0) {
  stop("None of the expected traits (GY, KW) were found in Pheno.")
}

if (!exists("nIter"))   nIter   <- 10000
if (!exists("burnIn"))  burnIn  <- 1000
if (!exists("thin"))    thin    <- 10
if (!exists("num_rep")) num_rep <- 20
if (!exists("n_cores")) n_cores <- max(1L, parallel::detectCores() - 1L)

model_catalog <- data.frame(
  ModelNumber = 1:18,
  Model = paste0("Eta", 1:18),
  Description = c(
    "E + G + GE",
    "E + P + PE",
    "E + GGK + GGKE",
    "E + PGK + PGKE",
    "E + GAK + GAKE",
    "E + PAK + PAKE",
    "E + G + P + GE + PE",
    "E + GGK + PGK + GGKE + PGKE",
    "E + GAK + PAK + GAKE + PAKE",
    "E + W + G + GE + GW",
    "E + W + P + PE + PW",
    "E + W + G + P + GE + PE + GW + PW",
    "E + W + GGK + GGKE + GGKW",
    "E + W + PGK + PGKE + PGKW",
    "E + W + GGK + PGK + GGKE + PGKE + GGKW + PGKW",
    "E + W + GAK + GAKE + GAKW",
    "E + W + PAK + PAKE + PAKW",
    "E + W + GAK + PAK + GAKE + PAKE + GAKW + PAKW"
  ),
  stringsAsFactors = FALSE
)

env_design <- model.matrix(~ Pheno$Env - 1)

Eta_list <- vector("list", 18)
Eta_list[[1]]  <- list(list(X = env_design, model = "BRR"), list(K = ZG, model = "RKHS"), list(K = ZGZE, model = "RKHS"))
Eta_list[[2]]  <- list(list(X = env_design, model = "BRR"), list(K = ZP, model = "RKHS"), list(K = ZPZE, model = "RKHS"))
Eta_list[[3]]  <- list(list(X = env_design, model = "BRR"), list(K = GGK, model = "RKHS"), list(K = GGKE, model = "RKHS"))
Eta_list[[4]]  <- list(list(X = env_design, model = "BRR"), list(K = PGK, model = "RKHS"), list(K = PGKE, model = "RKHS"))
Eta_list[[5]]  <- list(list(X = env_design, model = "BRR"), list(K = GAK, model = "RKHS"), list(K = GAKE, model = "RKHS"))
Eta_list[[6]]  <- list(list(X = env_design, model = "BRR"), list(K = PAK, model = "RKHS"), list(K = PAKE, model = "RKHS"))
Eta_list[[7]]  <- list(list(X = env_design, model = "BRR"), list(K = ZG, model = "RKHS"), list(K = ZP, model = "RKHS"), list(K = ZGZE, model = "RKHS"), list(K = ZPZE, model = "RKHS"))
Eta_list[[8]]  <- list(list(X = env_design, model = "BRR"), list(K = GGK, model = "RKHS"), list(K = PGK, model = "RKHS"), list(K = GGKE, model = "RKHS"), list(K = PGKE, model = "RKHS"))
Eta_list[[9]]  <- list(list(X = env_design, model = "BRR"), list(K = GAK, model = "RKHS"), list(K = PAK, model = "RKHS"), list(K = GAKE, model = "RKHS"), list(K = PAKE, model = "RKHS"))
Eta_list[[10]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = ZG, model = "RKHS"), list(K = ZGZE, model = "RKHS"), list(K = ZGZW, model = "RKHS"))
Eta_list[[11]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = ZP, model = "RKHS"), list(K = ZPZE, model = "RKHS"), list(K = ZPZW, model = "RKHS"))
Eta_list[[12]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = ZG, model = "RKHS"), list(K = ZP, model = "RKHS"), list(K = ZGZE, model = "RKHS"), list(K = ZPZE, model = "RKHS"), list(K = ZGZW, model = "RKHS"), list(K = ZPZW, model = "RKHS"))
Eta_list[[13]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = GGK, model = "RKHS"), list(K = GGKE, model = "RKHS"), list(K = GGKW, model = "RKHS"))
Eta_list[[14]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = PGK, model = "RKHS"), list(K = PGKE, model = "RKHS"), list(K = PGKW, model = "RKHS"))
Eta_list[[15]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = GGK, model = "RKHS"), list(K = PGK, model = "RKHS"), list(K = GGKE, model = "RKHS"), list(K = PGKE, model = "RKHS"), list(K = GGKW, model = "RKHS"), list(K = PGKW, model = "RKHS"))
Eta_list[[16]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = GAK, model = "RKHS"), list(K = GAKE, model = "RKHS"), list(K = GAKW, model = "RKHS"))
Eta_list[[17]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = PAK, model = "RKHS"), list(K = PAKE, model = "RKHS"), list(K = PAKW, model = "RKHS"))
Eta_list[[18]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = GAK, model = "RKHS"), list(K = PAK, model = "RKHS"), list(K = GAKE, model = "RKHS"), list(K = PAKE, model = "RKHS"), list(K = GAKW, model = "RKHS"), list(K = PAKW, model = "RKHS"))

```


## 4. Define the execution contract

This block restores the operational objects used by the current workflow: the fixed leave-one-environment-out set, a lightweight `run_bglr()` wrapper, an environment-wise accuracy helper, a deterministic fold builder shared across CV blocks, and the `MODEL × REP × TRAIT` task grid for the 18-model catalog.

```{r define-run-objects, eval=FALSE}
envs_leaveout <- c("CS11_WS", "CS11_WW", "CS12_WS", "CS12_WW")
envs_leaveout <- envs_leaveout[envs_leaveout %in% unique(Pheno$Env)]

run_bglr <- function(y2_vector, ETA, nIter, burnIn, thin, save_prefix = NULL) {
  y_t <- as.numeric(y2_vector)

  if (is.null(save_prefix)) {
    save_prefix <- tempfile(pattern = "bglr_", tmpdir = tempdir())
  }

  start_time <- Sys.time()
  fit <- BGLR(y = y_t, ETA = ETA, nIter = nIter, burnIn = burnIn, thin = thin, saveAt = save_prefix)
  end_time <- Sys.time()
  elapsed <- as.numeric(difftime(end_time, start_time, units = "secs"))

  generated_files <- Sys.glob(paste0(save_prefix, "*"))
  if (length(generated_files) > 0) {
    unlink(generated_files, recursive = FALSE, force = TRUE)
  }

  retorno_foreach <- list(yHat = fit$yHat, runtime_sec = elapsed)

  rm(fit)
  gc()

  return(retorno_foreach)
}

calc_cor_by_env <- function(df_with_yhat, trait_col) {
  df_with_yhat %>%
    dplyr::group_by(Env) %>%
    dplyr::summarize(
      cor = cor(.data[[trait_col]], yhat, use = "complete.obs"),
      .groups = "drop"
    ) %>%
    as.data.frame()
}

build_fold_map <- function(Pedigree, num_fold, MODEL, rep_num, TRAIT) {
  trait_id <- match(TRAIT, sort(unique(traits_to_run)))
  if (is.na(trait_id)) {
    trait_id <- 1L
  }

  seed_value <- as.integer(100000L * MODEL + 100L * rep_num + trait_id)

  old_seed_exists <- exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
  if (old_seed_exists) {
    old_seed <- get(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
  }

  set.seed(seed_value)
  obs_per_fold <- ceiling(length(Pedigree) / num_fold)
  folds <- sample(rep(seq_len(num_fold), each = obs_per_fold, length.out = length(Pedigree)))

  if (old_seed_exists) {
    assign(".Random.seed", old_seed, envir = .GlobalEnv)
  } else if (exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)) {
    rm(".Random.seed", envir = .GlobalEnv)
  }

  data.frame(Pedigree = Pedigree, fold = folds, stringsAsFactors = FALSE)
}

num_fold <- 5L
n_hybrids <- length(Pedigree)
combos <- expand.grid(
  MODEL = seq_along(Eta_list),
  REP = seq_len(num_rep),
  TRAIT = traits_to_run,
  stringsAsFactors = FALSE
)
```

## 5. Start the parallel backend

The backend below mirrors the original workflow: one worker receives one combination of model, repetition, and trait.

```{r start-parallel-backend, eval=FALSE}
useCores <- max(1L, n_cores)
cl <- parallel::makeCluster(useCores)
doParallel::registerDoParallel(cl)

parallel::clusterEvalQ(cl, {
  library(BGLR)
  library(dplyr)
  library(plyr)
  NULL
})

cat("Running on ", useCores, " cores.")
```


## 6. Run prediction analyses for CV1 and CV2

This block implements genotype-based validation for the 18-model workflow. The same deterministic fold map is rebuilt for each `MODEL × REP × TRAIT` task, one fit is run per fold, and the resulting predictions are split into `CV1` (held-out genotypes) and `CV2` (training genotypes under the same fit).

```{r run-predictions-cv12, eval=FALSE}
foreach(i = seq_len(nrow(combos)),
        .packages = c("BGLR", "dplyr", "plyr")) %dopar% {
  MODEL <- combos$MODEL[i]
  rep_num <- combos$REP[i]
  TRAIT <- as.character(combos$TRAIT[i])

  pheno <- data.frame(Pheno)
  CV1_list <- list()
  CV2_list <- list()

  folds_df <- build_fold_map(Pedigree, num_fold, MODEL, rep_num, TRAIT)

  for (fold_num in seq_len(num_fold)) {
    train_geno <- folds_df$Pedigree[folds_df$fold != fold_num]

    yield <- pheno
    yield$Y2 <- NA
    yield$Y2[yield$Pedigree %in% train_geno] <- yield[[TRAIT]][yield$Pedigree %in% train_geno]

    save_prefix_full <- tempfile(
      pattern = paste0("full_Eta", MODEL, "_", TRAIT, "_rep", rep_num, "_fold", fold_num, "_"),
      tmpdir = tempdir()
    )

    res_full <- run_bglr(
      yield$Y2,
      ETA = Eta_list[[MODEL]],
      nIter = nIter,
      burnIn = burnIn,
      thin = thin,
      save_prefix = save_prefix_full
    )

    yield$yhat <- res_full$yHat
    runtime_full <- res_full$runtime_sec

    df_test <- yield[!yield$Pedigree %in% train_geno, ]
    df_train <- yield[yield$Pedigree %in% train_geno, ]

    tmp1 <- calc_cor_by_env(df_test, TRAIT)
    tmp2 <- calc_cor_by_env(df_train, TRAIT)
    tmp1$Runtime_sec <- runtime_full
    tmp2$Runtime_sec <- runtime_full

    CV1_list[[length(CV1_list) + 1L]] <- tmp1
    CV2_list[[length(CV2_list) + 1L]] <- tmp2
  }

  CV1out <- plyr::ldply(CV1_list, data.frame)
  CV2out <- plyr::ldply(CV2_list, data.frame)

  write.csv(
    CV1out,
    file = file.path(results_dir, paste0("CV1_", TRAIT, "_Eta", MODEL, "_rep", rep_num, ".csv")),
    row.names = FALSE
  )
  write.csv(
    CV2out,
    file = file.path(results_dir, paste0("CV2_", TRAIT, "_Eta", MODEL, "_rep", rep_num, ".csv")),
    row.names = FALSE
  )

  NULL
}
```

## 7. Run prediction analyses for CV0 and CV00

This block implements leave-one-environment-out validation for the same 18-model workflow. For each task, the deterministic fold map is rebuilt with the same rule used in `CV1/CV2`, one fit is run for each `env_leave × fold` combination, and the predictions are split into `CV0` (training genotypes under the environment mask) and `CV00` (held-out genotypes under the same mask).

```{r run-predictions-cv0-cv00, eval=FALSE}
foreach(i = seq_len(nrow(combos)),
        .packages = c("BGLR", "dplyr", "plyr")) %dopar% {
  MODEL <- combos$MODEL[i]
  rep_num <- combos$REP[i]
  TRAIT <- as.character(combos$TRAIT[i])

  pheno <- data.frame(Pheno)
  CV0_lists <- lapply(envs_leaveout, function(x) list())
  CV00_lists <- lapply(envs_leaveout, function(x) list())
  names(CV0_lists) <- envs_leaveout
  names(CV00_lists) <- envs_leaveout

  folds_df <- build_fold_map(Pedigree, num_fold, MODEL, rep_num, TRAIT)

  for (fold_num in seq_len(num_fold)) {
    train_geno <- folds_df$Pedigree[folds_df$fold != fold_num]

    for (env_idx in seq_along(envs_leaveout)) {
      env_name <- envs_leaveout[env_idx]

      yield2 <- pheno
      yield2$Y2 <- NA
      yield2$Y2[yield2$Pedigree %in% train_geno] <- yield2[[TRAIT]][yield2$Pedigree %in% train_geno]
      yield2$Y2[yield2$Env == env_name] <- NA

      save_prefix_env <- tempfile(
        pattern = paste0(env_name, "_", TRAIT, "_Eta", MODEL, "_rep", rep_num, "_fold", fold_num, "_"),
        tmpdir = tempdir()
      )

      res_env <- run_bglr(
        yield2$Y2,
        ETA = Eta_list[[MODEL]],
        nIter = nIter,
        burnIn = burnIn,
        thin = thin,
        save_prefix = save_prefix_env
      )

      yield2$yhat <- res_env$yHat
      runtime_env <- res_env$runtime_sec

      df_test2 <- yield2[!yield2$Pedigree %in% train_geno & yield2$Env == env_name, ]
      df_train2 <- yield2[yield2$Pedigree %in% train_geno & yield2$Env == env_name, ]

      tmp00 <- calc_cor_by_env(df_test2, TRAIT)
      tmp0  <- calc_cor_by_env(df_train2, TRAIT)
      tmp00$Runtime_sec <- runtime_env
      tmp0$Runtime_sec <- runtime_env

      CV00_lists[[env_idx]][[length(CV00_lists[[env_idx]]) + 1L]] <- tmp00
      CV0_lists[[env_idx]][[length(CV0_lists[[env_idx]]) + 1L]] <- tmp0
    }
  }

  for (env_idx in seq_along(envs_leaveout)) {
    env_name <- envs_leaveout[env_idx]
    CV00out <- plyr::ldply(CV00_lists[[env_idx]], data.frame)
    CV0out <- plyr::ldply(CV0_lists[[env_idx]], data.frame)

    write.csv(
      CV00out,
      file = file.path(results_dir, paste0("CV00_", TRAIT, "_", env_name, "_Eta", MODEL, "_rep", rep_num, ".csv")),
      row.names = FALSE
    )
    write.csv(
      CV0out,
      file = file.path(results_dir, paste0("CV0_", TRAIT, "_", env_name, "_Eta", MODEL, "_rep", rep_num, ".csv")),
      row.names = FALSE
    )
  }

  NULL
}
```

## 8. Stop the parallel backend

```{r stop-parallel-backend, eval=FALSE}
stopCluster(cl)
foreach::registerDoSEQ()
```


## 9. Inventory of the prediction output files

The inventories below summarize the CSV files produced by the execution stage using the current 18-model naming contract.

```{r prediction-output-inventory}
prediction_files <- character(0)

if (dir.exists(results_dir)) {
  prediction_files <- list.files(results_dir, pattern = "\\.csv$", full.names = TRUE)
}

inventory_cv12 <- data.frame()
inventory_cvloo <- data.frame()

if (length(prediction_files) > 0) {
  for (current_file in prediction_files) {
    current_name <- basename(current_file)
    current_name_no_ext <- sub("\\.csv$", "", current_name)
    current_parts <- strsplit(current_name_no_ext, "_")[[1]]

    if (length(current_parts) == 4 && current_parts[1] %in% c("CV1", "CV2")) {
      inventory_cv12 <- bind_rows(
        inventory_cv12,
        data.frame(
          file_name = current_name,
          CV = current_parts[1],
          Trait = current_parts[2],
          Model = current_parts[3],
          Rep = current_parts[4],
          Fold = NA_character_,
          stringsAsFactors = FALSE
        )
      )
    }

    if (length(current_parts) >= 6 && current_parts[1] %in% c("CV0", "CV00")) {
      inventory_cvloo <- bind_rows(
        inventory_cvloo,
        data.frame(
          file_name = current_name,
          CV = current_parts[1],
          Trait = current_parts[2],
          Env_leave = paste(current_parts[3:(length(current_parts) - 2)], collapse = "_"),
          Model = current_parts[length(current_parts) - 1],
          Rep = current_parts[length(current_parts)],
          Fold = NA_character_,
          stringsAsFactors = FALSE
        )
      )
    }
  }
}

write_csv(inventory_cv12, file.path(analysis_tables_dir, "analysis_prediction_inventory_cv12.csv"))
write_csv(inventory_cvloo, file.path(analysis_tables_dir, "analysis_prediction_inventory_cvloo.csv"))
```


## 10. Counts of files by validation scheme and trait

```{r prediction-output-counts}
counts_by_cv <- data.frame()
counts_by_trait <- data.frame()

if (nrow(inventory_cv12) > 0 || nrow(inventory_cvloo) > 0) {
  combined_inventory <- bind_rows(
    inventory_cv12 %>% mutate(Env_leave = NA_character_),
    inventory_cvloo
  )

  counts_by_cv <- combined_inventory %>%
    count(CV, name = "n_files") %>%
    arrange(CV)

  counts_by_trait <- combined_inventory %>%
    count(CV, Trait, name = "n_files") %>%
    arrange(CV, Trait)
}

write_csv(counts_by_cv, file.path(analysis_tables_dir, "analysis_prediction_counts_by_cv.csv"))
write_csv(counts_by_trait, file.path(analysis_tables_dir, "analysis_prediction_counts_by_trait.csv"))

if (nrow(counts_by_cv) > 0) {
  knitr::kable(counts_by_cv, caption = "Number of prediction files by validation scheme.", align = "cc")
} else {
  cat("Prediction counts by validation scheme are not available yet.
")
}
```


## 11. Examples of the output naming pattern

```{r prediction-output-examples}
output_examples <- bind_rows(
  tibble::tibble(
    ExampleType = "CV1/CV2",
    file_name = c("CV1_GY_Eta1_rep1.csv", "CV2_KW_Eta18_rep20.csv"),
    Meaning = c(
      "Prediction accuracy by environment for CV1, trait GY, model Eta1, repetition 1.",
      "Prediction accuracy by environment for CV2, trait KW, model Eta18, repetition 20."
    )
  ),
  tibble::tibble(
    ExampleType = "CV0/CV00",
    file_name = c("CV0_GY_CS11_WS_Eta1_rep1.csv", "CV00_KW_CS12_WW_Eta18_rep20.csv"),
    Meaning = c(
      "Prediction accuracy by environment for CV0, trait GY, held-out environment CS11_WS, model Eta1, repetition 1.",
      "Prediction accuracy by environment for CV00, trait KW, held-out environment CS12_WW, model Eta18, repetition 20."
    )
  )
)

write_csv(output_examples, file.path(analysis_tables_dir, "analysis_prediction_output_examples.csv"))
knitr::kable(output_examples, caption = "Examples of file names produced by the prediction workflow.", align = "c")
```


## 12. Note on temporary `BGLR` artifacts

This execution stage uses two different kinds of files, and it is useful to distinguish them clearly in tutorial form.

**Auxiliary fitting artifacts.**  
During each call to `BGLR(...)`, the argument `saveAt` points to a prefix created with `tempdir()`. These files are temporary working files used internally by `BGLR` while the model is being fitted. In other words, they help the fitting process run, but they are not the analytical products that will be interpreted later.

**Final analytical outputs.**  
After the fit finishes, the script collects the predictions, runtime, and CV summaries in memory and then writes the final CSV files to `output/results/`. These CSV files are the outputs that should be preserved, shared, and consumed by the downstream visualization and summary modules.

**Why the temporary files can be removed.**  
The temporary `saveAt` files are deleted only after `BGLR(...)` has returned and the script has already extracted the objects needed for the workflow. Therefore, the temporary files support the fit while it is running, but they are not needed anymore once the final summaries have been created.

**Practical interpretation for this module.**  
When this file inventories outputs, it inventories only the final CSV files stored under `output/results/`. The temporary `BGLR` artifacts in `tempdir()` are treated only as short-lived fitting aids, not as reproducible analytical deliverables.

## 13. Final note

This file closes the execution layer of the prediction stage while preserving consistency with the reduced 18-model preprocessing module documented in `analysis_prediction.Rmd` and with the current CSV contract used by the downstream summaries.

## Stage navigation

**Previous stage:** [Prediction — Part I](analysis_prediction.html)  
**Next stage:** [Visualization](visualization.html)

