---
title: "Prediction analyses - run prediction and outputs"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

**Language / Idioma:** [English](analysis_prediction_run_outputs.html) | [Português](analysis_prediction_run_outputs_pt.html)

**Module navigation / Navegação do módulo:** [Part I: setup, models and CV design](analysis_prediction.html)

## 1. Introduction

This submodule is the execution counterpart of `analysis_prediction.Rmd`. Its role is to run the prediction jobs for the reduced catalog of 18 models, save the CSV files under `output/results/`, and inventory the outputs that will later be consumed by the visualization stage.

The execution contract used here has five operational rules:

- `CV1` and `CV2` are derived from the same fit within each fold;
- `CV0` and `CV00` are derived from the same fit for each held-out environment within each fold;
- the fold partition is deterministic for each `MODEL × REP × TRAIT` task and is reused in both CV blocks;
- output files are organized explicitly by validation scheme, environment when applicable, trait, repetition, and fold, with the fold represented both in the directory tree and in the file name;
- persistent `BGLR` artifacts are saved under the primary fit folders (`CV1` and `CV0`), renamed with explicit component labels, and then mirrored to the paired folders (`CV2` and `CV00`).

The heavy fitting chunks remain disabled by default in the tutorial. Even so, the code shown below is the executable version that should be used when the prediction stage is rerun.

## 2. Packages and project paths

The execution module needs both the matrix objects created upstream and the packages used by the parallel fitting layer. The path logic below keeps the file runnable from either the project root or the `analysis/` directory.


```{r setup}
rm(list = ls())
options(scipen = 999)

library(readr)
library(dplyr)
library(stringr)
library(knitr)
library(BGLR)
library(foreach)
library(doParallel)
library(parallel)
library(tidyr)

project_root <- getwd()

if (!dir.exists(file.path(project_root, "output")) &&
    dir.exists(file.path(project_root, "..", "output"))) {
  project_root <- normalizePath(file.path(project_root, ".."))
}

if (!dir.exists(file.path(project_root, "output"))) {
  stop("Could not find the folder 'output/'. Render this file from the project root or from analysis/.")
}

results_dir <- file.path(project_root, "output", "results")
analysis_tables_dir <- file.path(project_root, "output", "tables", "analysis_prediction")
bglr_runs_dir <- file.path(results_dir, "bglr_runs")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(analysis_tables_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(bglr_runs_dir, recursive = TRUE, showWarnings = FALSE)
```


## 3. Reload the execution objects from Part I

This section rebuilds the execution environment when the file is rendered on its own. It follows the contract of `analysis_prediction.Rmd` and recreates the reduced set of 18 ETA objects exactly as defined in Part I.

```{r load-required-objects, eval=FALSE}
matrices_dir <- file.path(project_root, "output", "Matrizes")

required_matrix_files <- c(
  "Pheno.rds",
  "ZG.rds", "ZP.rds", "ZE.rds", "ZW.rds",
  "ZGZE.rds", "ZPZE.rds", "ZGZW.rds", "ZPZW.rds",
  "GGK.rds", "PGK.rds", "GAK.rds", "PAK.rds",
  "GGKE.rds", "PGKE.rds", "GAKE.rds", "PAKE.rds",
  "GGKW.rds", "PGKW.rds", "GAKW.rds", "PAKW.rds"
)

required_matrix_paths <- file.path(matrices_dir, required_matrix_files)

if (!all(file.exists(required_matrix_paths))) {
  missing_matrix_files <- required_matrix_files[!file.exists(required_matrix_paths)]
  stop(
    paste0(
      "The execution objects required by analysis_prediction_run_outputs were not found in output/Matrizes/. Missing files: ",
      paste(missing_matrix_files, collapse = ", ")
    )
  )
}

Pheno <- readRDS(file.path(matrices_dir, "Pheno.rds"))
ZG   <- readRDS(file.path(matrices_dir, "ZG.rds"))
ZP   <- readRDS(file.path(matrices_dir, "ZP.rds"))
ZE   <- readRDS(file.path(matrices_dir, "ZE.rds"))
ZW   <- readRDS(file.path(matrices_dir, "ZW.rds"))
ZGZE <- readRDS(file.path(matrices_dir, "ZGZE.rds"))
ZPZE <- readRDS(file.path(matrices_dir, "ZPZE.rds"))
ZGZW <- readRDS(file.path(matrices_dir, "ZGZW.rds"))
ZPZW <- readRDS(file.path(matrices_dir, "ZPZW.rds"))
GGK  <- readRDS(file.path(matrices_dir, "GGK.rds"))
PGK  <- readRDS(file.path(matrices_dir, "PGK.rds"))
GAK  <- readRDS(file.path(matrices_dir, "GAK.rds"))
PAK  <- readRDS(file.path(matrices_dir, "PAK.rds"))
GGKE <- readRDS(file.path(matrices_dir, "GGKE.rds"))
PGKE <- readRDS(file.path(matrices_dir, "PGKE.rds"))
GAKE <- readRDS(file.path(matrices_dir, "GAKE.rds"))
PAKE <- readRDS(file.path(matrices_dir, "PAKE.rds"))
GGKW <- readRDS(file.path(matrices_dir, "GGKW.rds"))
PGKW <- readRDS(file.path(matrices_dir, "PGKW.rds"))
GAKW <- readRDS(file.path(matrices_dir, "GAKW.rds"))
PAKW <- readRDS(file.path(matrices_dir, "PAKW.rds"))

if (!exists("Pedigree")) Pedigree <- unique(Pheno$Pedigree)

candidate_traits <- c("GY", "KW")
traits_to_run <- candidate_traits[candidate_traits %in% names(Pheno)]
if (length(traits_to_run) == 0) {
  stop("None of the expected traits (GY, KW) were found in Pheno.")
}

if (!exists("nIter"))   nIter   <- 5000
if (!exists("burnIn"))  burnIn  <- 1000
if (!exists("thin"))    thin    <- 10
if (!exists("num_rep")) num_rep <- 10
if (!exists("n_cores")) n_cores <- max(1L, parallel::detectCores() - 1L)

model_catalog <- data.frame(
  ModelNumber = 1:18,
  Model = paste0("Eta", 1:18),
  Description = c(
    "E + G + GE",
    "E + P + PE",
    "E + GGK + GGKE",
    "E + PGK + PGKE",
    "E + GAK + GAKE",
    "E + PAK + PAKE",
    "E + G + P + GE + PE",
    "E + GGK + PGK + GGKE + PGKE",
    "E + GAK + PAK + GAKE + PAKE",
    "E + W + G + GE + GW",
    "E + W + P + PE + PW",
    "E + W + G + P + GE + PE + GW + PW",
    "E + W + GGK + GGKE + GGKW",
    "E + W + PGK + PGKE + PGKW",
    "E + W + GGK + PGK + GGKE + PGKE + GGKW + PGKW",
    "E + W + GAK + GAKE + GAKW",
    "E + W + PAK + PAKE + PAKW",
    "E + W + GAK + PAK + GAKE + PAKE + GAKW + PAKW"
  ),
  stringsAsFactors = FALSE
)

env_design <- ZE

Eta_list <- vector("list", 18)

Eta_list[[1]] <- list(
  E = list(X = env_design, model = "BRR"),
  G = list(K = ZG, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS")
)

Eta_list[[2]] <- list(
  E = list(X = env_design, model = "BRR"),
  P = list(K = ZP, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS")
)

Eta_list[[3]] <- list(
  E = list(X = env_design, model = "BRR"),
  GGK = list(K = GGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS")
)

Eta_list[[4]] <- list(
  E = list(X = env_design, model = "BRR"),
  PGK = list(K = PGK, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS")
)

Eta_list[[5]] <- list(
  E = list(X = env_design, model = "BRR"),
  GAK = list(K = GAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS")
)

Eta_list[[6]] <- list(
  E = list(X = env_design, model = "BRR"),
  PAK = list(K = PAK, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS")
)

Eta_list[[7]] <- list(
  E = list(X = env_design, model = "BRR"),
  G = list(K = ZG, model = "RKHS"),
  P = list(K = ZP, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS")
)

Eta_list[[8]] <- list(
  E = list(X = env_design, model = "BRR"),
  GGK = list(K = GGK, model = "RKHS"),
  PGK = list(K = PGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS")
)

Eta_list[[9]] <- list(
  E = list(X = env_design, model = "BRR"),
  GAK = list(K = GAK, model = "RKHS"),
  PAK = list(K = PAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS")
)

Eta_list[[10]] <- list(
  E = list(X = env_design, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  G = list(K = ZG, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS"),
  GW = list(K = ZGZW, model = "RKHS")
)

Eta_list[[11]] <- list(
  E = list(X = env_design, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  P = list(K = ZP, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS"),
  PW = list(K = ZPZW, model = "RKHS")
)

Eta_list[[12]] <- list(
  E = list(X = env_design, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  G = list(K = ZG, model = "RKHS"),
  P = list(K = ZP, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS"),
  GW = list(K = ZGZW, model = "RKHS"),
  PW = list(K = ZPZW, model = "RKHS")
)

Eta_list[[13]] <- list(
  E = list(X = env_design, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  GGK = list(K = GGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS"),
  GGKW = list(K = GGKW, model = "RKHS")
)

Eta_list[[14]] <- list(
  E = list(X = env_design, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  PGK = list(K = PGK, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS"),
  PGKW = list(K = PGKW, model = "RKHS")
)

Eta_list[[15]] <- list(
  E = list(X = env_design, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  GGK = list(K = GGK, model = "RKHS"),
  PGK = list(K = PGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS"),
  GGKW = list(K = GGKW, model = "RKHS"),
  PGKW = list(K = PGKW, model = "RKHS")
)

Eta_list[[16]] <- list(
  E = list(X = env_design, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  GAK = list(K = GAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS"),
  GAKW = list(K = GAKW, model = "RKHS")
)

Eta_list[[17]] <- list(
  E = list(X = env_design, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  PAK = list(K = PAK, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS"),
  PAKW = list(K = PAKW, model = "RKHS")
)

Eta_list[[18]] <- list(
  E = list(X = env_design, model = "BRR"),
  W = list(K = ZW, model = "RKHS"),
  GAK = list(K = GAK, model = "RKHS"),
  PAK = list(K = PAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS"),
  GAKW = list(K = GAKW, model = "RKHS"),
  PAKW = list(K = PAKW, model = "RKHS")
)

Eta_labels <- lapply(Eta_list, function(current_model) {
  paste0(
    names(current_model),
    "_",
    vapply(current_model, function(component) component$model, character(1))
  )
})

```


## 4. Define the execution contract

This block restores the operational objects used by the current workflow: the fixed leave-one-environment-out set, the number of folds, and the `MODEL × REP × TRAIT` task grid for the 18-model catalog. The fitting and summarization logic are written directly inside the CV blocks so that the execution path stays explicit.

```{r define-run-objects, eval=FALSE}
envs_leaveout <- c("CS11_WS", "CS11_WW", "CS12_WS", "CS12_WW")
envs_leaveout <- envs_leaveout[envs_leaveout %in% unique(Pheno$Env)]

num_fold <- 5L
n_hybrids <- length(Pedigree)
combos <- expand.grid(
  MODEL = seq_along(Eta_list),
  REP = seq_len(num_rep),
  TRAIT = traits_to_run,
  stringsAsFactors = FALSE
)
```

## 5. Start the parallel backend

The backend below mirrors the original workflow: one worker receives one combination of model, repetition, and trait.

```{r start-parallel-backend, eval=FALSE}
useCores <- max(1L, n_cores)
cl <- parallel::makeCluster(useCores)
doParallel::registerDoParallel(cl)

parallel::clusterEvalQ(cl, {
  library(BGLR)
  library(dplyr)
  library(plyr)
  NULL
})

cat("Running on ", useCores, " cores.")
```


## 6. Run prediction analyses for CV1 and CV2

This block implements genotype-based validation for the 18-model workflow. The deterministic fold map, the BGLR fit, and the environment-wise summary are all written directly inside the loop so that each step stays explicit. One fit is run per fold, the resulting predictions are split into `CV1` (held-out genotypes) and `CV2` (training genotypes under the same fit), and the persistent artifacts are first saved under `bglr_runs/CV1/...` before being mirrored to `bglr_runs/CV2/...` with explicit component labels.

```{r run-predictions-cv12, eval=FALSE}
foreach(i = seq_len(nrow(combos)),
        .packages = c("BGLR", "dplyr", "plyr")) %dopar% {
  MODEL <- combos$MODEL[i]
  rep_num <- combos$REP[i]
  TRAIT <- as.character(combos$TRAIT[i])

  pheno <- data.frame(Pheno)

  trait_id <- match(TRAIT, trait_reference_order)
  if (is.na(trait_id)) {
    trait_id <- 1L
  }

  seed_value <- as.integer(100000L * MODEL + 100L * rep_num + trait_id)
  old_seed_exists <- exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
  if (old_seed_exists) {
    old_seed <- get(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
  }

  set.seed(seed_value)
  obs_per_fold <- ceiling(length(Pedigree) / num_fold)
  folds <- sample(rep(seq_len(num_fold), each = obs_per_fold, length.out = length(Pedigree)))

  if (old_seed_exists) {
    assign(".Random.seed", old_seed, envir = .GlobalEnv)
  } else if (exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)) {
    rm(".Random.seed", envir = .GlobalEnv)
  }

  folds_df <- data.frame(Pedigree = Pedigree, fold = folds, stringsAsFactors = FALSE)

  for (fold_num in seq_len(num_fold)) {
    rep_label <- sprintf("rep%02d", rep_num)
    fold_label <- sprintf("fold%02d", fold_num)

    cv1_dir <- file.path(results_dir, "CV1", TRAIT, rep_label, fold_label)
    cv2_dir <- file.path(results_dir, "CV2", TRAIT, rep_label, fold_label)
    dir.create(cv1_dir, recursive = TRUE, showWarnings = FALSE)
    dir.create(cv2_dir, recursive = TRUE, showWarnings = FALSE)

    cv1_file <- file.path(cv1_dir, paste0("CV1_", TRAIT, "_Eta", MODEL, "_", rep_label, "_", fold_label, ".csv"))
    cv2_file <- file.path(cv2_dir, paste0("CV2_", TRAIT, "_Eta", MODEL, "_", rep_label, "_", fold_label, ".csv"))

    if (file.exists(cv1_file) && file.exists(cv2_file)) {
      next
    }

    train_geno <- folds_df$Pedigree[folds_df$fold != fold_num]

    yield <- pheno
    yield$Y2 <- NA
    yield$Y2[yield$Pedigree %in% train_geno] <- yield[[TRAIT]][yield$Pedigree %in% train_geno]

    run_dir_cv1 <- file.path(bglr_runs_dir, "CV1", TRAIT, rep_label, fold_label, paste0("Eta", MODEL))
    run_dir_cv2 <- file.path(bglr_runs_dir, "CV2", TRAIT, rep_label, fold_label, paste0("Eta", MODEL))
    dir.create(run_dir_cv1, recursive = TRUE, showWarnings = FALSE)
    dir.create(run_dir_cv2, recursive = TRUE, showWarnings = FALSE)

    save_prefix_full <- file.path(run_dir_cv1, paste0("CV1_", TRAIT, "_Eta", MODEL, "_", rep_label, "_", fold_label, "_"))

    y_t <- as.numeric(yield$Y2)
    start_time <- Sys.time()
    fit <- BGLR(y = y_t, ETA = Eta_list[[MODEL]], nIter = nIter, burnIn = burnIn, thin = thin, saveAt = save_prefix_full)
    end_time <- Sys.time()
    runtime_full <- as.numeric(difftime(end_time, start_time, units = "secs"))

    yield$yhat <- fit$yHat
    rm(fit)
    gc()

    current_files <- list.files(run_dir_cv1, full.names = TRUE, recursive = FALSE)
    for (current_file in current_files) {
      current_name <- basename(current_file)
      renamed_name <- current_name
      for (eta_idx in seq_along(Eta_labels[[MODEL]])) {
        renamed_name <- gsub(paste0("ETA_", eta_idx), Eta_labels[[MODEL]][eta_idx], renamed_name, fixed = TRUE)
      }
      if (!identical(current_name, renamed_name)) {
        file.rename(current_file, file.path(run_dir_cv1, renamed_name))
      }
    }

    current_files <- list.files(run_dir_cv1, full.names = TRUE, recursive = FALSE)
    for (current_file in current_files) {
      mirrored_name <- sub("^CV1_", "CV2_", basename(current_file))
      file.copy(current_file, file.path(run_dir_cv2, mirrored_name), overwrite = TRUE)
    }

    df_test <- yield[!yield$Pedigree %in% train_geno, ]
    df_train <- yield[yield$Pedigree %in% train_geno, ]

    tmp1 <- df_test %>%
      dplyr::group_by(Env) %>%
      dplyr::summarize(cor = cor(.data[[TRAIT]], yhat, use = "complete.obs"), .groups = "drop") %>%
      as.data.frame()

    tmp2 <- df_train %>%
      dplyr::group_by(Env) %>%
      dplyr::summarize(cor = cor(.data[[TRAIT]], yhat, use = "complete.obs"), .groups = "drop") %>%
      as.data.frame()

    tmp1$Runtime_sec <- runtime_full
    tmp2$Runtime_sec <- runtime_full
    tmp1$Fold <- fold_label
    tmp2$Fold <- fold_label

    write.csv(tmp1, file = cv1_file, row.names = FALSE)
    write.csv(tmp2, file = cv2_file, row.names = FALSE)
  }

  NULL
}
```
## 7. Run prediction analyses for CV0 and CV00

This block implements leave-one-environment-out validation for the same 18-model workflow. For each task, the deterministic fold map is rebuilt with the same rule used in `CV1/CV2`, one fit is run for each `env_leave × fold` combination, and the predictions are split into `CV0` (training genotypes under the environment mask) and `CV00` (held-out genotypes under the same mask). As in the genotype-based block, the fold construction, the BGLR fit, and the environment-wise summary are all written directly inside the loop so that the execution path remains explicit. The artifacts are first saved under `bglr_runs/CV0/...`, renamed with explicit component labels, and then mirrored to `bglr_runs/CV00/...`.

```{r run-predictions-cv0-cv00, eval=FALSE}
foreach(i = seq_len(nrow(combos)),
        .packages = c("BGLR", "dplyr", "plyr")) %dopar% {
  MODEL <- combos$MODEL[i]
  rep_num <- combos$REP[i]
  TRAIT <- as.character(combos$TRAIT[i])

  pheno <- data.frame(Pheno)

  trait_id <- match(TRAIT, trait_reference_order)
  if (is.na(trait_id)) {
    trait_id <- 1L
  }

  seed_value <- as.integer(100000L * MODEL + 100L * rep_num + trait_id)
  old_seed_exists <- exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
  if (old_seed_exists) {
    old_seed <- get(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
  }

  set.seed(seed_value)
  obs_per_fold <- ceiling(length(Pedigree) / num_fold)
  folds <- sample(rep(seq_len(num_fold), each = obs_per_fold, length.out = length(Pedigree)))

  if (old_seed_exists) {
    assign(".Random.seed", old_seed, envir = .GlobalEnv)
  } else if (exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)) {
    rm(".Random.seed", envir = .GlobalEnv)
  }

  folds_df <- data.frame(Pedigree = Pedigree, fold = folds, stringsAsFactors = FALSE)

  for (fold_num in seq_len(num_fold)) {
    train_geno <- folds_df$Pedigree[folds_df$fold != fold_num]
    rep_label <- sprintf("rep%02d", rep_num)
    fold_label <- sprintf("fold%02d", fold_num)

    for (env_idx in seq_along(envs_leaveout)) {
      env_name <- envs_leaveout[env_idx]

      cv00_dir <- file.path(results_dir, "CV00", env_name, TRAIT, rep_label, fold_label)
      cv0_dir <- file.path(results_dir, "CV0", env_name, TRAIT, rep_label, fold_label)
      dir.create(cv00_dir, recursive = TRUE, showWarnings = FALSE)
      dir.create(cv0_dir, recursive = TRUE, showWarnings = FALSE)

      cv00_file <- file.path(cv00_dir, paste0("CV00_", TRAIT, "_", env_name, "_Eta", MODEL, "_", rep_label, "_", fold_label, ".csv"))
      cv0_file <- file.path(cv0_dir, paste0("CV0_", TRAIT, "_", env_name, "_Eta", MODEL, "_", rep_label, "_", fold_label, ".csv"))

      if (file.exists(cv00_file) && file.exists(cv0_file)) {
        next
      }

      yield2 <- pheno
      yield2$Y2 <- NA
      yield2$Y2[yield2$Pedigree %in% train_geno] <- yield2[[TRAIT]][yield2$Pedigree %in% train_geno]
      yield2$Y2[yield2$Env == env_name] <- NA

      run_dir_cv0 <- file.path(bglr_runs_dir, "CV0", env_name, TRAIT, rep_label, fold_label, paste0("Eta", MODEL))
      run_dir_cv00 <- file.path(bglr_runs_dir, "CV00", env_name, TRAIT, rep_label, fold_label, paste0("Eta", MODEL))
      dir.create(run_dir_cv0, recursive = TRUE, showWarnings = FALSE)
      dir.create(run_dir_cv00, recursive = TRUE, showWarnings = FALSE)

      save_prefix_env <- file.path(run_dir_cv0, paste0("CV0_", env_name, "_", TRAIT, "_Eta", MODEL, "_", rep_label, "_", fold_label, "_"))

      y_t <- as.numeric(yield2$Y2)
      start_time <- Sys.time()
      fit <- BGLR(y = y_t, ETA = Eta_list[[MODEL]], nIter = nIter, burnIn = burnIn, thin = thin, saveAt = save_prefix_env)
      end_time <- Sys.time()
      runtime_env <- as.numeric(difftime(end_time, start_time, units = "secs"))

      yield2$yhat <- fit$yHat
      rm(fit)
      gc()

      current_files <- list.files(run_dir_cv0, full.names = TRUE, recursive = FALSE)
      for (current_file in current_files) {
        current_name <- basename(current_file)
        renamed_name <- current_name
        for (eta_idx in seq_along(Eta_labels[[MODEL]])) {
          renamed_name <- gsub(paste0("ETA_", eta_idx), Eta_labels[[MODEL]][eta_idx], renamed_name, fixed = TRUE)
        }
        if (!identical(current_name, renamed_name)) {
          file.rename(current_file, file.path(run_dir_cv0, renamed_name))
        }
      }

      current_files <- list.files(run_dir_cv0, full.names = TRUE, recursive = FALSE)
      for (current_file in current_files) {
        mirrored_name <- sub("^CV0_", "CV00_", basename(current_file))
        file.copy(current_file, file.path(run_dir_cv00, mirrored_name), overwrite = TRUE)
      }

      df_test2 <- yield2[!yield2$Pedigree %in% train_geno & yield2$Env == env_name, , drop = FALSE]
      df_train2 <- yield2[yield2$Pedigree %in% train_geno & yield2$Env == env_name, , drop = FALSE]

      tmp00 <- df_test2 %>%
        dplyr::group_by(Env) %>%
        dplyr::summarize(cor = cor(.data[[TRAIT]], yhat, use = "complete.obs"), .groups = "drop") %>%
        as.data.frame()

      tmp0 <- df_train2 %>%
        dplyr::group_by(Env) %>%
        dplyr::summarize(cor = cor(.data[[TRAIT]], yhat, use = "complete.obs"), .groups = "drop") %>%
        as.data.frame()

      tmp00$Runtime_sec <- runtime_env
      tmp0$Runtime_sec <- runtime_env
      tmp00$Fold <- fold_label
      tmp0$Fold <- fold_label

      write.csv(tmp00, file = cv00_file, row.names = FALSE)
      write.csv(tmp0, file = cv0_file, row.names = FALSE)
    }
  }

  NULL
}
```
## 8. Stop the parallel backend

```{r stop-parallel-backend, eval=FALSE}
stopCluster(cl)
foreach::registerDoSEQ()
```


## 9. Inventory of the prediction output files

The inventories below summarize the CSV files produced by the execution stage using the current rep/fold folder contract for the 18-model workflow.

```{r prediction-output-inventory}
prediction_files <- character(0)

if (dir.exists(results_dir)) {
  prediction_files <- list.files(results_dir, pattern = "\\.csv$", full.names = TRUE, recursive = TRUE)
}

inventory_cv12 <- data.frame()
inventory_cvloo <- data.frame()

if (length(prediction_files) > 0) {
  results_root <- normalizePath(results_dir, winslash = "/")

  for (current_file in prediction_files) {
    current_norm <- normalizePath(current_file, winslash = "/")
    rel_path <- substring(current_norm, nchar(results_root) + 2)
    path_parts <- strsplit(rel_path, "/", fixed = TRUE)[[1]]
    current_name <- basename(current_file)
    current_name_no_ext <- sub("\\.csv$", "", current_name)
    current_parts <- strsplit(current_name_no_ext, "_")[[1]]

    if (length(path_parts) >= 5 && current_parts[1] %in% c("CV1", "CV2")) {
      inventory_cv12 <- dplyr::bind_rows(
        inventory_cv12,
        data.frame(
          relative_path = rel_path,
          file_name = current_name,
          CV = current_parts[1],
          Trait = path_parts[2],
          Rep = path_parts[3],
          Fold = path_parts[4],
          Model = current_parts[3],
          stringsAsFactors = FALSE
        )
      )
    }

    if (length(path_parts) >= 6 && current_parts[1] %in% c("CV0", "CV00")) {
      inventory_cvloo <- dplyr::bind_rows(
        inventory_cvloo,
        data.frame(
          relative_path = rel_path,
          file_name = current_name,
          CV = current_parts[1],
          Env_leave = path_parts[2],
          Trait = path_parts[3],
          Rep = path_parts[4],
          Fold = path_parts[5],
          Model = current_parts[length(current_parts) - 2],
          stringsAsFactors = FALSE
        )
      )
    }
  }
}

write_csv(inventory_cv12, file.path(analysis_tables_dir, "analysis_prediction_inventory_cv12.csv"))
write_csv(inventory_cvloo, file.path(analysis_tables_dir, "analysis_prediction_inventory_cvloo.csv"))
```
## 10. Counts of files by validation scheme and trait

```{r prediction-output-counts}
counts_by_cv <- data.frame()
counts_by_trait <- data.frame()

if (nrow(inventory_cv12) > 0 || nrow(inventory_cvloo) > 0) {
  combined_inventory <- bind_rows(
    inventory_cv12 %>% mutate(Env_leave = NA_character_),
    inventory_cvloo
  )

  counts_by_cv <- combined_inventory %>%
    count(CV, name = "n_files") %>%
    arrange(CV)

  counts_by_trait <- combined_inventory %>%
    count(CV, Trait, name = "n_files") %>%
    arrange(CV, Trait)
}

write_csv(counts_by_cv, file.path(analysis_tables_dir, "analysis_prediction_counts_by_cv.csv"))
write_csv(counts_by_trait, file.path(analysis_tables_dir, "analysis_prediction_counts_by_trait.csv"))

if (nrow(counts_by_cv) > 0) {
  knitr::kable(counts_by_cv, caption = "Number of prediction files by validation scheme.", align = "cc")
} else {
  cat("Prediction counts by validation scheme are not available yet.
")
}
```


## 11. Examples of the output path pattern

```{r prediction-output-examples}
output_examples <- bind_rows(
  tibble::tibble(
    ExampleType = "CV1/CV2",
    relative_path = c(
      "CV1/GY/rep01/fold01/CV1_GY_Eta1_rep01_fold01.csv",
      "CV2/KW/rep10/fold05/CV2_KW_Eta18_rep10_fold05.csv"
    ),
    bglr_example = c(
      "bglr_runs/CV1/GY/rep01/fold01/Eta1/CV1_GY_Eta1_rep01_fold01_GE_RKHS_varU.dat",
      "bglr_runs/CV2/KW/rep10/fold05/Eta18/CV2_KW_Eta18_rep10_fold05_PAKW_RKHS_varU.dat"
    ),
    Meaning = c(
      "Prediction accuracy for CV1, trait GY, model Eta1, repetition 1, fold 1.",
      "Prediction accuracy for CV2, trait KW, model Eta18, repetition 10, fold 5."
    )
  ),
  tibble::tibble(
    ExampleType = "CV0/CV00",
    relative_path = c(
      "CV0/CS11_WS/GY/rep01/fold01/CV0_GY_CS11_WS_Eta1_rep01_fold01.csv",
      "CV00/CS12_WW/KW/rep10/fold02/CV00_KW_CS12_WW_Eta18_rep10_fold02.csv"
    ),
    bglr_example = c(
      "bglr_runs/CV0/CS11_WS/GY/rep01/fold01/Eta1/CV0_CS11_WS_GY_Eta1_rep01_fold01_GW_RKHS_varU.dat",
      "bglr_runs/CV00/CS12_WW/KW/rep10/fold02/Eta18/CV00_CS12_WW_KW_Eta18_rep10_fold02_PAKW_RKHS_varU.dat"
    ),
    Meaning = c(
      "Prediction accuracy for CV0, trait GY, held-out environment CS11_WS, model Eta1, repetition 1, fold 1.",
      "Prediction accuracy for CV00, trait KW, held-out environment CS12_WW, model Eta18, repetition 10, fold 2."
    )
  )
)

write_csv(output_examples, file.path(analysis_tables_dir, "analysis_prediction_output_examples.csv"))
knitr::kable(output_examples, caption = "Examples of output files stored under the current rep/fold layout, together with the scenario-specific prefixes used for the persistent BGLR artifacts.", align = "c")
```
## 12. Note on temporary `BGLR` artifacts

In the current 18-model workflow, persistent `saveAt` artifacts are written under `output/results/bglr_runs/` during each fit and preserved for later convergence diagnostics. The scenario-specific paths now separate `CV1`, `CV2`, `CV0`, and `CV00`, and the persisted filenames use the scenario prefix together with explicit effect labels instead of generic `ETA_1`, `ETA_2`, and similar placeholders. Therefore, this execution module inventories both the final CSV outputs stored under `output/results/` and the hierarchical paths that locate the corresponding BGLR run folders.

## 13. Final note

This file closes the execution layer of the prediction stage while preserving consistency with the reduced 18-model preprocessing module documented in `analysis_prediction.Rmd` and with the current rep/fold output contract used by the downstream summaries.

## Stage navigation

**Previous stage:** [Prediction — Part I](analysis_prediction.html)  
**Next stage:** [Visualization](visualization.html)

