---
title: "Prediction Analyses"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

**Language / Idioma:** [English](analysis_prediction.html) | [Português](analysis_prediction_pt.html)

## 1. Introduction

This tutorial performs the prediction analyses for the reduced set of 18 Bayesian multi-kernel models used in the study.

The script combines:

- genomic kernels,
- phenomic kernels,
- the categorical environment effect (`E`),
- and the weather kernel (`W`).

In this workflow, the weather kernel is used mainly in models **M10-M18**. It should be interpreted as a weather-derived source of similarity that complements, but does not replace, the categorical environment effect.

The phenomic component (`P`) should also be interpreted with caution. NIR-based similarity may capture useful predictive structure, but it can also reflect strong physiological response to the environment rather than only stable biological similarity.

The script evaluates the 18 models under four prediction scenarios:

- `CV1`: untested genotypes
- `CV2`: partially observed genotypes
- `CV0`: untested environment
- `CV00`: untested genotypes in an untested environment

In the current version of the workflow:

- `CV1` and `CV2` are run with **5 folds and 20 repetitions**
- `CV0` and `CV00` are run as **leave-one-environment-out (LOEO)** analyses
- the model-fitting step is parallelized with **`foreach`**
- raw prediction CSV files are saved in `output/results/`
- auxiliary files generated by `BGLR` through `saveAt` are saved in structured subfolders inside `output/results/bglr_runs/`

## 2. Packages and options

```{r setup}
rm(list = ls())
options(scipen = 999)

library(BGLR)
library(dplyr)
library(readr)
library(stringr)
library(foreach)
library(doParallel)
library(parallel)
```

## 3. Inputs generated by previous scripts

```{r load-inputs}
Pheno <- readRDS("output/Matrizes/Pheno.rds")

ZG   <- readRDS("output/Matrizes/ZG.rds")
ZP   <- readRDS("output/Matrizes/ZP.rds")
ZE   <- readRDS("output/Matrizes/ZE.rds")
ZW   <- readRDS("output/Matrizes/ZW.rds")

ZGZE <- readRDS("output/Matrizes/ZGZE.rds")
ZPZE <- readRDS("output/Matrizes/ZPZE.rds")
ZGZW <- readRDS("output/Matrizes/ZGZW.rds")
ZPZW <- readRDS("output/Matrizes/ZPZW.rds")

GGK  <- readRDS("output/Matrizes/GGK.rds")
PGK  <- readRDS("output/Matrizes/PGK.rds")
GAK  <- readRDS("output/Matrizes/GAK.rds")
PAK  <- readRDS("output/Matrizes/PAK.rds")

GGKE <- readRDS("output/Matrizes/GGKE.rds")
PGKE <- readRDS("output/Matrizes/PGKE.rds")
GAKE <- readRDS("output/Matrizes/GAKE.rds")
PAKE <- readRDS("output/Matrizes/PAKE.rds")

GGKW <- readRDS("output/Matrizes/GGKW.rds")
PGKW <- readRDS("output/Matrizes/PGKW.rds")
GAKW <- readRDS("output/Matrizes/GAKW.rds")
PAKW <- readRDS("output/Matrizes/PAKW.rds")
```

The current `matrizes.Rmd` saves the observation-level matrices using the following names:

- `ZE` for the categorical environment design matrix
- `ZGZE` and `ZPZE` for genomic-by-environment and phenomic-by-environment interactions
- `ZGZW` and `ZPZW` for genomic-by-weather and phenomic-by-weather interactions

For that reason, this prediction script reads those matrices directly instead of the older names used in previous versions of the project.

## 4. MCMC and parallel settings

```{r mcmc-settings}
nIter <- 10000
burnIn <- 1000
thin <- 10
num_rep <- 20

n_cores <- max(1, parallel::detectCores() - 1)

nIter
burnIn
thin
num_rep
n_cores
```

## 5. Model catalog

```{r model-catalog}
model_catalog <- data.frame(
  ModelNumber = 1:18,
  Model = sprintf("M%02d", 1:18),
  Description = c(
    "E + G + GE",
    "E + P + PE",
    "E + GGK + GGKE",
    "E + PGK + PGKE",
    "E + GAK + GAKE",
    "E + PAK + PAKE",
    "E + G + P + GE + PE",
    "E + GGK + PGK + GGKE + PGKE",
    "E + GAK + PAK + GAKE + PAKE",
    "E + W + G + GE + GW",
    "E + W + P + PE + PW",
    "E + W + G + P + GE + PE + GW + PW",
    "E + W + GGK + GGKE + GGKW",
    "E + W + PGK + PGKE + PGKW",
    "E + W + GGK + PGK + GGKE + PGKE + GGKW + PGKW",
    "E + W + GAK + GAKE + GAKW",
    "E + W + PAK + PAKE + PAKW",
    "E + W + GAK + PAK + GAKE + PAKE + GAKW + PAKW"
  ),
  stringsAsFactors = FALSE
)

model_catalog
```

## 6. Model specification

```{r eta-list}
Eta_list <- list()

Eta_list[[1]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"))

Eta_list[[2]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"))

Eta_list[[3]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"))

Eta_list[[4]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = PGKE, model = "RKHS"))

Eta_list[[5]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"))

Eta_list[[6]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = PAKE, model = "RKHS"))

Eta_list[[7]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"))

Eta_list[[8]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"),
                       list(K = PGKE, model = "RKHS"))

Eta_list[[9]]  <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"),
                       list(K = PAKE, model = "RKHS"))

Eta_list[[10]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"),
                       list(K = ZGZW, model = "RKHS"))

Eta_list[[11]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"),
                       list(K = ZPZW, model = "RKHS"))

Eta_list[[12]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = ZG,   model = "RKHS"),
                       list(K = ZP,   model = "RKHS"),
                       list(K = ZGZE, model = "RKHS"),
                       list(K = ZPZE, model = "RKHS"),
                       list(K = ZGZW, model = "RKHS"),
                       list(K = ZPZW, model = "RKHS"))

Eta_list[[13]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"),
                       list(K = GGKW, model = "RKHS"))

Eta_list[[14]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = PGKE, model = "RKHS"),
                       list(K = PGKW, model = "RKHS"))

Eta_list[[15]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GGK,  model = "RKHS"),
                       list(K = PGK,  model = "RKHS"),
                       list(K = GGKE, model = "RKHS"),
                       list(K = PGKE, model = "RKHS"),
                       list(K = GGKW, model = "RKHS"),
                       list(K = PGKW, model = "RKHS"))

Eta_list[[16]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"),
                       list(K = GAKW, model = "RKHS"))

Eta_list[[17]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = PAKE, model = "RKHS"),
                       list(K = PAKW, model = "RKHS"))

Eta_list[[18]] <- list(list(X = model.matrix(~ Pheno$Env - 1), model = "BRR"),
                       list(K = ZW,   model = "RKHS"),
                       list(K = GAK,  model = "RKHS"),
                       list(K = PAK,  model = "RKHS"),
                       list(K = GAKE, model = "RKHS"),
                       list(K = PAKE, model = "RKHS"),
                       list(K = GAKW, model = "RKHS"),
                       list(K = PAKW, model = "RKHS"))
```

## 7. Running prediction analyses in parallel with foreach

This chunk is kept with `eval = FALSE` because it is the computationally expensive step. It generates the CSV files later used by `visualization.Rmd`.

```{r run-predictions, eval=FALSE}
dir.create("output/results", recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("output", "results", "bglr_runs"), recursive = TRUE, showWarnings = FALSE)

traits_to_run <- c("GY", "KW")
env_levels <- unique(Pheno$Env)
pedigrees <- unique(Pheno$Pedigree)

cl <- parallel::makeCluster(n_cores)
doParallel::registerDoParallel(cl)
on.exit({
  try(parallel::stopCluster(cl), silent = TRUE)
  foreach::registerDoSEQ()
}, add = TRUE)

parallel::clusterEvalQ(cl, {
  library(BGLR)
  NULL
})

for (trait_name in traits_to_run) {

  y_full <- Pheno[[trait_name]]

  # ---------------------------------------------
  # CV1 and CV2: 5 folds x 20 repetitions
  # ---------------------------------------------
  set.seed(123)

  for (rep_id in 1:num_rep) {

    fold_id <- sample(rep(1:5, length.out = length(pedigrees)))
    fold_map <- data.frame(Pedigree = pedigrees, Fold = fold_id, stringsAsFactors = FALSE)

    for (fold in 1:5) {

      pedigrees_in_fold <- fold_map$Pedigree[fold_map$Fold == fold]

      # CV1: untested genotypes
      y_cv1 <- y_full
      y_cv1[Pheno$Pedigree %in% pedigrees_in_fold] <- NA

      cv1_results <- foreach(
        m = 1:18,
        .packages = c("BGLR"),
        .export = c("Pheno", "Eta_list", "model_catalog", "y_cv1", "y_full", "env_levels",
                    "pedigrees_in_fold", "nIter", "burnIn", "thin", "trait_name", "rep_id", "fold")
      ) %dopar% {
        model_label <- model_catalog$Model[m]

        run_dir <- file.path(
          "output", "results", "bglr_runs",
          "CV1",
          trait_name,
          model_label,
          paste0("rep", rep_id),
          paste0("fold", fold)
        )

        dir.create(run_dir, recursive = TRUE, showWarnings = FALSE)

        existing_files <- list.files(run_dir, full.names = TRUE, all.files = TRUE, no.. = TRUE)
        if (length(existing_files) > 0) {
          unlink(existing_files, recursive = TRUE, force = TRUE)
        }

        save_prefix <- file.path(
          run_dir,
          paste0("CV1_", trait_name, "_", model_label, "_rep", rep_id, "_fold", fold, "_")
        )

        fit_cv1 <- BGLR(
          y = y_cv1,
          ETA = Eta_list[[m]],
          nIter = nIter,
          burnIn = burnIn,
          thin = thin,
          saveAt = save_prefix,
          verbose = FALSE
        )

        pred_cv1 <- fit_cv1$yHat

        out_cv1 <- data.frame(
          ObsID = Pheno$ObsID,
          Pedigree = Pheno$Pedigree,
          Env = Pheno$Env,
          observed = y_full,
          predicted = pred_cv1,
          cor = NA_real_,
          stringsAsFactors = FALSE
        )

        for (env_now in env_levels) {
          idx_env <- which(out_cv1$Env == env_now & out_cv1$Pedigree %in% pedigrees_in_fold)

          if (length(idx_env) > 1) {
            out_cv1$cor[idx_env] <- cor(
              out_cv1$observed[idx_env],
              out_cv1$predicted[idx_env],
              use = "complete.obs"
            )
          }
        }

        list(model_number = m, model_label = model_label, result_table = out_cv1)
      }

      for (k in seq_along(cv1_results)) {
        m <- cv1_results[[k]]$model_number
        model_label <- cv1_results[[k]]$model_label
        out_cv1 <- cv1_results[[k]]$result_table

        write.csv(
          out_cv1,
          file = paste0(
            "output/results/CV1_",
            trait_name,
            "_", model_label,
            "_rep", rep_id,
            "_fold", fold,
            ".csv"
          ),
          row.names = FALSE
        )
      }

      # CV2: partially observed genotypes
      y_cv2 <- y_full

      for (ped_now in pedigrees_in_fold) {
        idx_ped <- which(Pheno$Pedigree == ped_now)
        envs_ped <- unique(Pheno$Env[idx_ped])

        if (length(envs_ped) > 1) {
          env_to_mask <- sample(envs_ped, size = floor(length(envs_ped) / 2))
          idx_mask <- which(Pheno$Pedigree == ped_now & Pheno$Env %in% env_to_mask)
          y_cv2[idx_mask] <- NA
        }
      }

      cv2_results <- foreach(
        m = 1:18,
        .packages = c("BGLR"),
        .export = c("Pheno", "Eta_list", "model_catalog", "y_cv2", "y_full", "env_levels",
                    "nIter", "burnIn", "thin", "trait_name", "rep_id", "fold")
      ) %dopar% {
        model_label <- model_catalog$Model[m]

        run_dir <- file.path(
          "output", "results", "bglr_runs",
          "CV2",
          trait_name,
          model_label,
          paste0("rep", rep_id),
          paste0("fold", fold)
        )

        dir.create(run_dir, recursive = TRUE, showWarnings = FALSE)

        existing_files <- list.files(run_dir, full.names = TRUE, all.files = TRUE, no.. = TRUE)
        if (length(existing_files) > 0) {
          unlink(existing_files, recursive = TRUE, force = TRUE)
        }

        save_prefix <- file.path(
          run_dir,
          paste0("CV2_", trait_name, "_", model_label, "_rep", rep_id, "_fold", fold, "_")
        )

        fit_cv2 <- BGLR(
          y = y_cv2,
          ETA = Eta_list[[m]],
          nIter = nIter,
          burnIn = burnIn,
          thin = thin,
          saveAt = save_prefix,
          verbose = FALSE
        )

        pred_cv2 <- fit_cv2$yHat

        out_cv2 <- data.frame(
          ObsID = Pheno$ObsID,
          Pedigree = Pheno$Pedigree,
          Env = Pheno$Env,
          observed = y_full,
          predicted = pred_cv2,
          cor = NA_real_,
          stringsAsFactors = FALSE
        )

        idx_na <- which(is.na(y_cv2))

        for (env_now in env_levels) {
          idx_env <- which(out_cv2$Env == env_now)
          idx_eval <- intersect(idx_env, idx_na)

          if (length(idx_eval) > 1) {
            out_cv2$cor[idx_eval] <- cor(
              out_cv2$observed[idx_eval],
              out_cv2$predicted[idx_eval],
              use = "complete.obs"
            )
          }
        }

        list(model_number = m, model_label = model_label, result_table = out_cv2)
      }

      for (k in seq_along(cv2_results)) {
        m <- cv2_results[[k]]$model_number
        model_label <- cv2_results[[k]]$model_label
        out_cv2 <- cv2_results[[k]]$result_table

        write.csv(
          out_cv2,
          file = paste0(
            "output/results/CV2_",
            trait_name,
            "_", model_label,
            "_rep", rep_id,
            "_fold", fold,
            ".csv"
          ),
          row.names = FALSE
        )
      }
    }
  }

  # ---------------------------------------------
  # CV0 and CV00: LOEO
  # ---------------------------------------------
  for (env_leave in env_levels) {

    env_parts <- strsplit(env_leave, "_")[[1]]

    # CV0: untested environment
    y_cv0 <- y_full
    idx_env_leave <- which(Pheno$Env == env_leave)
    y_cv0[idx_env_leave] <- NA

    cv0_results <- foreach(
      m = 1:18,
      .packages = c("BGLR"),
      .export = c("Pheno", "Eta_list", "model_catalog", "y_cv0", "y_full", "idx_env_leave",
                  "nIter", "burnIn", "thin", "trait_name", "env_leave")
    ) %dopar% {
      model_label <- model_catalog$Model[m]

      run_dir <- file.path(
        "output", "results", "bglr_runs",
        "CV0",
        trait_name,
        model_label,
        "rep1",
        "fold1",
        env_leave
      )

      dir.create(run_dir, recursive = TRUE, showWarnings = FALSE)

      existing_files <- list.files(run_dir, full.names = TRUE, all.files = TRUE, no.. = TRUE)
      if (length(existing_files) > 0) {
        unlink(existing_files, recursive = TRUE, force = TRUE)
      }

      save_prefix <- file.path(
        run_dir,
        paste0("CV0_", trait_name, "_", model_label, "_rep1_fold1_", env_leave, "_")
      )

      fit_cv0 <- BGLR(
        y = y_cv0,
        ETA = Eta_list[[m]],
        nIter = nIter,
        burnIn = burnIn,
        thin = thin,
        saveAt = save_prefix,
        verbose = FALSE
      )

      pred_cv0 <- fit_cv0$yHat
      cor_cv0 <- cor(y_full[idx_env_leave], pred_cv0[idx_env_leave], use = "complete.obs")

      out_cv0 <- data.frame(
        ObsID = Pheno$ObsID[idx_env_leave],
        Pedigree = Pheno$Pedigree[idx_env_leave],
        Env = Pheno$Env[idx_env_leave],
        observed = y_full[idx_env_leave],
        predicted = pred_cv0[idx_env_leave],
        cor = cor_cv0,
        stringsAsFactors = FALSE
      )

      list(model_number = m, model_label = model_label, result_table = out_cv0)
    }

    for (k in seq_along(cv0_results)) {
      m <- cv0_results[[k]]$model_number
      model_label <- cv0_results[[k]]$model_label
      out_cv0 <- cv0_results[[k]]$result_table

      write.csv(
        out_cv0,
        file = paste0(
          "output/results/CV0_",
          trait_name,
          "_", env_parts[1], "_", env_parts[2],
          "_", model_label,
          "_rep1_fold1.csv"
        ),
        row.names = FALSE
      )
    }

    # CV00: untested genotypes in an untested environment
    pedigrees_test <- unique(Pheno$Pedigree[Pheno$Env == env_leave])
    pedigrees_train <- unique(Pheno$Pedigree[Pheno$Env != env_leave])
    pedigrees_new <- setdiff(pedigrees_test, pedigrees_train)

    idx_cv00 <- which(Pheno$Env == env_leave & Pheno$Pedigree %in% pedigrees_new)

    if (length(idx_cv00) > 1) {

      y_cv00 <- y_full
      y_cv00[idx_cv00] <- NA

      cv00_results <- foreach(
        m = 1:18,
        .packages = c("BGLR"),
        .export = c("Pheno", "Eta_list", "model_catalog", "y_cv00", "y_full", "idx_cv00",
                    "nIter", "burnIn", "thin", "trait_name", "env_leave")
      ) %dopar% {
        model_label <- model_catalog$Model[m]

        run_dir <- file.path(
          "output", "results", "bglr_runs",
          "CV00",
          trait_name,
          model_label,
          "rep1",
          "fold1",
          env_leave
        )

        dir.create(run_dir, recursive = TRUE, showWarnings = FALSE)

        existing_files <- list.files(run_dir, full.names = TRUE, all.files = TRUE, no.. = TRUE)
        if (length(existing_files) > 0) {
          unlink(existing_files, recursive = TRUE, force = TRUE)
        }

        save_prefix <- file.path(
          run_dir,
          paste0("CV00_", trait_name, "_", model_label, "_rep1_fold1_", env_leave, "_")
        )

        fit_cv00 <- BGLR(
          y = y_cv00,
          ETA = Eta_list[[m]],
          nIter = nIter,
          burnIn = burnIn,
          thin = thin,
          saveAt = save_prefix,
          verbose = FALSE
        )

        pred_cv00 <- fit_cv00$yHat
        cor_cv00 <- cor(y_full[idx_cv00], pred_cv00[idx_cv00], use = "complete.obs")

        out_cv00 <- data.frame(
          ObsID = Pheno$ObsID[idx_cv00],
          Pedigree = Pheno$Pedigree[idx_cv00],
          Env = Pheno$Env[idx_cv00],
          observed = y_full[idx_cv00],
          predicted = pred_cv00[idx_cv00],
          cor = cor_cv00,
          stringsAsFactors = FALSE
        )

        list(model_number = m, model_label = model_label, result_table = out_cv00)
      }

      for (k in seq_along(cv00_results)) {
        m <- cv00_results[[k]]$model_number
        model_label <- cv00_results[[k]]$model_label
        out_cv00 <- cv00_results[[k]]$result_table

        write.csv(
          out_cv00,
          file = paste0(
            "output/results/CV00_",
            trait_name,
            "_", env_parts[1], "_", env_parts[2],
            "_", model_label,
            "_rep1_fold1.csv"
          ),
          row.names = FALSE
        )
      }
    }
  }
}
```

## 8. Final note

This script produces the raw prediction outputs. The summaries, weighted averages, final tables, and final figures for the manuscript are generated in `visualization.Rmd`.

The current output structure is fully standardized to the new pipeline.

The CSV filenames in `output/results/` now include only:

- CV type
- trait
- model label (`M01` to `M18`)
- repetition
- fold

Examples:

- `CV1_GY_M05_rep2_fold5.csv`
- `CV2_KW_M12_rep8_fold3.csv`
- `CV0_GY_CS11_WS_M10_rep1_fold1.csv`
- `CV00_KW_CS12_WW_M18_rep1_fold1.csv`

The auxiliary files generated by `BGLR` through `saveAt` are stored in subfolders under:

`output/results/bglr_runs/`

organized by:

- CV type
- trait
- model
- repetition
- fold
- held-out environment when applicable
