---
title: "Prediction Analyses - Didactic Tutorial"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

## 1. Introduction

This file is the **didactic version** of the final prediction script. The objective here is not only to run the analyses, but also to make clear **what each block does**, **why each decision was taken**, and **which files are produced for the next stages of the pipeline**.

This script runs the prediction analyses for the reduced set of **18 Bayesian multi-kernel models** used in the study. The models combine:

- genomic kernels,
- phenomic kernels,
- the categorical environment effect (`E`),
- and, when appropriate, the weather kernel (`W`).

The models are evaluated under four cross-validation scenarios:

- `CV1`: untested genotypes;
- `CV2`: training-set performance, following the **reference script**;
- `CV0`: leave-one-environment-out **inside the fold structure**;
- `CV00`: untested genotypes in a leave-one-environment-out setting, also **inside the fold structure**.

This is important because the final corrected version restores the logic of the original `analysis.Rmd` while keeping the reduced catalog of 18 final models used in the updated repository.

### What this script writes

After execution, the script produces:

- prediction summary CSV files in `output/results/`;
- auxiliary `saveAt` files and compact BGLR metrics in `output/results/bglr_runs/`;
- a job-level diagnostics table in `output/tables/`.

### Main implementation decisions

- `CV0` and `CV00` are reconstructed inside the same `rep × fold` structure of the reference script;
- `CV2` follows the original reference logic, even if later the article text needs to be harmonized with that definition;
- parallelization is done at the **beginning of the workflow**, with one job for each `MODEL × REP × TRAIT` combination;
- outputs are saved in a way that the next scripts of the pipeline can read them directly.

### How to read this tutorial file

The file is organized in the same order in which the analysis is actually executed:

1. load matrices and metadata;
2. define MCMC and cross-validation settings;
3. define the 18 models;
4. run the full cross-validation workflow in parallel;
5. save aggregated results for `CV1`, `CV2`, `CV0`, and `CV00`.

So, even if you are reading this file for learning purposes, the order of the sections mirrors the real computational logic of the analysis.

## 2. Packages and options

Here we clear the workspace, avoid scientific notation in printed output, and load the packages used for model fitting, data manipulation, and parallel execution.

```{r setup}
rm(list = ls())
options(scipen = 999)

library(BGLR)
library(dplyr)
library(readr)
library(stringr)
library(foreach)
library(doParallel)
library(parallel)
library(plyr)
```

## 3. Inputs generated by previous scripts

This script assumes that the matrix-building stage of the repository has already been executed. Therefore, all kernels and design matrices are read directly from `output/Matrizes/`.

```{r load-inputs}
Pheno <- readRDS("output/Matrizes/Pheno.rds")

ZG   <- readRDS("output/Matrizes/ZG.rds")
ZP   <- readRDS("output/Matrizes/ZP.rds")
ZE   <- readRDS("output/Matrizes/ZE.rds")
ZW   <- readRDS("output/Matrizes/ZW.rds")

ZGZE <- readRDS("output/Matrizes/ZGZE.rds")
ZPZE <- readRDS("output/Matrizes/ZPZE.rds")
ZGZW <- readRDS("output/Matrizes/ZGZW.rds")
ZPZW <- readRDS("output/Matrizes/ZPZW.rds")

GGK  <- readRDS("output/Matrizes/GGK.rds")
PGK  <- readRDS("output/Matrizes/PGK.rds")
GAK  <- readRDS("output/Matrizes/GAK.rds")
PAK  <- readRDS("output/Matrizes/PAK.rds")

GGKE <- readRDS("output/Matrizes/GGKE.rds")
PGKE <- readRDS("output/Matrizes/PGKE.rds")
GAKE <- readRDS("output/Matrizes/GAKE.rds")
PAKE <- readRDS("output/Matrizes/PAKE.rds")

GGKW <- readRDS("output/Matrizes/GGKW.rds")
PGKW <- readRDS("output/Matrizes/PGKW.rds")
GAKW <- readRDS("output/Matrizes/GAKW.rds")
PAKW <- readRDS("output/Matrizes/PAKW.rds")
```

## 4. MCMC and parallel settings

In this section we define the MCMC configuration, the number of repetitions and folds, the environments used in leave-one-environment-out validation, and the number of cores used in parallel execution.

```{r mcmc-settings}
nIter <- 5000
burnIn <- 1000
thin <- 10
num_rep <- 20
num_fold <- 5

n_cores <- max(1, parallel::detectCores() - 1)

envs_leaveout <- c("CS11_WS", "CS11_WW", "CS12_WS", "CS12_WW")
traits_to_run <- c("GY", "KW")
Pedigree <- unique(Pheno$Pedigree)
n_hybrids <- length(Pedigree)

nIter
burnIn
thin
num_rep
num_fold
n_cores
n_hybrids
```

## 5. Model catalog

The catalog below is the reduced final set of 18 models. Each model receives an identifier (`M01` to `M18`) and a short description that will later be propagated to the result tables.

```{r model-catalog}
model_catalog <- data.frame(
  ModelNumber = 1:18,
  Model = sprintf("M%02d", 1:18),
  Description = c(
    "E + G + GE",
    "E + P + PE",
    "E + GGK + GGKE",
    "E + PGK + PGKE",
    "E + GAK + GAKE",
    "E + PAK + PAKE",
    "E + G + P + GE + PE",
    "E + GGK + PGK + GGKE + PGKE",
    "E + GAK + PAK + GAKE + PAKE",
    "E + W + G + GE + GW",
    "E + W + P + PE + PW",
    "E + W + G + P + GE + PE + GW + PW",
    "E + W + GGK + GGKE + GGKW",
    "E + W + PGK + PGKE + PGKW",
    "E + W + GGK + PGK + GGKE + PGKE + GGKW + PGKW",
    "E + W + GAK + GAKE + GAKW",
    "E + W + PAK + PAKE + PAKW",
    "E + W + GAK + PAK + GAKE + PAKE + GAKW + PAKW"
  ),
  stringsAsFactors = FALSE
)

model_catalog
```

## 6. Model specification

This section defines the `ETA` objects used by the 18 final models. To make the material easier to study, the model list is now split into smaller conceptual groups rather than written as one long block.

### 6.1 Initialize the `Eta_list` object

We start by creating the empty container that will store the 18 model specifications.

```{r eta-list-initialize}
Eta_list <- list()
```

### 6.2 Models 1 to 6: environment plus one main kernel block

These are the simplest models in the catalog. Each one combines the categorical environment effect with one main kernel and its interaction with environment.

```{r eta-list-models-1-6}
Eta_list[[1]]  <- list(
  E  = list(X = ZE,   model = "BRR"),
  G  = list(K = ZG,   model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS")
)

Eta_list[[2]]  <- list(
  E  = list(X = ZE,   model = "BRR"),
  P  = list(K = ZP,   model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS")
)

Eta_list[[3]]  <- list(
  E    = list(X = ZE,   model = "BRR"),
  GGK  = list(K = GGK,  model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS")
)

Eta_list[[4]]  <- list(
  E    = list(X = ZE,   model = "BRR"),
  PGK  = list(K = PGK,  model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS")
)

Eta_list[[5]]  <- list(
  E    = list(X = ZE,   model = "BRR"),
  GAK  = list(K = GAK,  model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS")
)

Eta_list[[6]]  <- list(
  E    = list(X = ZE,   model = "BRR"),
  PAK  = list(K = PAK,  model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS")
)
```

### 6.3 Models 7 to 9: combined genomic and phenomic information without weather

This group combines two main kernels at once, still without the weather kernel.

```{r eta-list-models-7-9}
Eta_list[[7]]  <- list(
  E  = list(X = ZE,   model = "BRR"),
  G  = list(K = ZG,   model = "RKHS"),
  P  = list(K = ZP,   model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS")
)

Eta_list[[8]]  <- list(
  E    = list(X = ZE,   model = "BRR"),
  GGK  = list(K = GGK,  model = "RKHS"),
  PGK  = list(K = PGK,  model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS")
)

Eta_list[[9]]  <- list(
  E    = list(X = ZE,   model = "BRR"),
  GAK  = list(K = GAK,  model = "RKHS"),
  PAK  = list(K = PAK,  model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS")
)
```

### 6.4 Models 10 to 12: weather added to the baseline combinations

The next group introduces the weather kernel (`W`) and its interactions with genomic or phenomic information.

```{r eta-list-models-10-12}
Eta_list[[10]] <- list(
  E  = list(X = ZE,   model = "BRR"),
  W  = list(K = ZW,   model = "RKHS"),
  G  = list(K = ZG,   model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS"),
  GW = list(K = ZGZW, model = "RKHS")
)

Eta_list[[11]] <- list(
  E  = list(X = ZE,   model = "BRR"),
  W  = list(K = ZW,   model = "RKHS"),
  P  = list(K = ZP,   model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS"),
  PW = list(K = ZPZW, model = "RKHS")
)

Eta_list[[12]] <- list(
  E  = list(X = ZE,   model = "BRR"),
  W  = list(K = ZW,   model = "RKHS"),
  G  = list(K = ZG,   model = "RKHS"),
  P  = list(K = ZP,   model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS"),
  GW = list(K = ZGZW, model = "RKHS"),
  PW = list(K = ZPZW, model = "RKHS")
)
```

### 6.5 Models 13 to 18: Gaussian and additive kernels combined with weather

The final group contains the models with Gaussian or additive kernels plus the weather effect.

```{r eta-list-models-13-18}
Eta_list[[13]] <- list(
  E    = list(X = ZE,   model = "BRR"),
  W    = list(K = ZW,   model = "RKHS"),
  GGK  = list(K = GGK,  model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS"),
  GGKW = list(K = GGKW, model = "RKHS")
)

Eta_list[[14]] <- list(
  E    = list(X = ZE,   model = "BRR"),
  W    = list(K = ZW,   model = "RKHS"),
  PGK  = list(K = PGK,  model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS"),
  PGKW = list(K = PGKW, model = "RKHS")
)

Eta_list[[15]] <- list(
  E    = list(X = ZE,   model = "BRR"),
  W    = list(K = ZW,   model = "RKHS"),
  GGK  = list(K = GGK,  model = "RKHS"),
  PGK  = list(K = PGK,  model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS"),
  GGKW = list(K = GGKW, model = "RKHS"),
  PGKW = list(K = PGKW, model = "RKHS")
)

Eta_list[[16]] <- list(
  E    = list(X = ZE,   model = "BRR"),
  W    = list(K = ZW,   model = "RKHS"),
  GAK  = list(K = GAK,  model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS"),
  GAKW = list(K = GAKW, model = "RKHS")
)

Eta_list[[17]] <- list(
  E    = list(X = ZE,   model = "BRR"),
  W    = list(K = ZW,   model = "RKHS"),
  PAK  = list(K = PAK,  model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS"),
  PAKW = list(K = PAKW, model = "RKHS")
)

Eta_list[[18]] <- list(
  E    = list(X = ZE,   model = "BRR"),
  W    = list(K = ZW,   model = "RKHS"),
  GAK  = list(K = GAK,  model = "RKHS"),
  PAK  = list(K = PAK,  model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS"),
  GAKW = list(K = GAKW, model = "RKHS"),
  PAKW = list(K = PAKW, model = "RKHS")
)
```

### 6.6 Check the final list before model fitting

This last chunk confirms that the number of ETA objects matches the number of rows in the model catalog.

```{r eta-list-check}
stopifnot(length(Eta_list) == nrow(model_catalog))
length(Eta_list)
```

## 7. Running prediction analyses in parallel with foreach

This is the computational core of the script. In the didactic version, this section is split into smaller chunks so the reader can understand the workflow step by step.

All chunks below are kept with `eval = FALSE` because this stage is computationally expensive. The idea is that the reader can first understand the structure and then decide whether to run the full prediction step.

The execution logic remains exactly the same as in the final operational script:

1. prepare output folders and remove old prediction files;
2. define the compact BGLR runner and the correlation summarization step;
3. create the full grid of jobs (`MODEL × REP × TRAIT`);
4. initialize the parallel backend;
5. run all folds and all validation scenarios inside each job;
6. aggregate the fold-level results and save one CSV per repetition.

### 7.1 Prepare output folders and remove old prediction files

This first chunk creates the directories used by the prediction stage and removes previous prediction outputs. This cleanup is important because the script rewrites all CV tables from scratch.

```{r prepare-output-folders, eval=FALSE}
dir.create("output/results", recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("output", "results", "bglr_runs"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("output", "tables"), recursive = TRUE, showWarnings = FALSE)

old_prediction_csv <- list.files(
  "output/results",
  pattern = "^(CV1|CV2|CV0|CV00)_.*\\.csv$",
  full.names = TRUE
)
if (length(old_prediction_csv) > 0) {
  unlink(old_prediction_csv, force = TRUE)
}

for (cv_dir in c("CV1_CV2", "CV0_CV00")) {
  old_run_dir <- file.path("output", "results", "bglr_runs", cv_dir)
  if (dir.exists(old_run_dir)) {
    unlink(old_run_dir, recursive = TRUE, force = TRUE)
  }
}
```

### 7.2 Define the helper functions used during model execution

To keep the workflow readable, the execution helper and the summarization helper are now shown in separate chunks.

#### 7.2.1 Function to fit one BGLR model and save compact metrics

This function reproduces the fitting logic of the final operational script. It stores the essential metrics in an `.rds` file and returns only the predicted values and runtime needed by the rest of the pipeline.

```{r helper-run-bglr, eval=FALSE}
run_bglr <- function(y2_vector, ETA, nIter, burnIn, thin, save_prefix) {
  y_t <- as.numeric(y2_vector)
  dir.create(dirname(save_prefix), recursive = TRUE, showWarnings = FALSE)

  start_time <- Sys.time()

  fit <- BGLR(
    y = y_t,
    ETA = ETA,
    nIter = nIter,
    burnIn = burnIn,
    thin = thin,
    saveAt = save_prefix,
    verbose = FALSE
  )

  end_time <- Sys.time()
  elapsed <- as.numeric(difftime(end_time, start_time, units = "secs"))

  variancias_U <- list()
  if (!is.null(fit$ETA)) {
    nomes_efeitos <- names(fit$ETA)
    if (is.null(nomes_efeitos)) {
      nomes_efeitos <- paste0("Efeito_", seq_along(fit$ETA))
    }

    for (j in seq_along(fit$ETA)) {
      efeito_nome <- nomes_efeitos[j]
      if (!is.null(fit$ETA[[j]]$varU)) {
        variancias_U[[efeito_nome]] <- fit$ETA[[j]]$varU
      }
    }
  }

  saida_essencial <- list(
    yHat_mean = fit$yHat,
    varE_mean = fit$varE,
    varB_mean = ifelse(is.null(fit$varB), NA, fit$varB),
    variances_U_mean = variancias_U,
    DIC = ifelse(is.null(fit$fit$DIC), NA, fit$fit$DIC),
    runtime_sec = elapsed
  )

  saveRDS(saida_essencial, paste0(save_prefix, "_metricas.rds"))

  retorno_foreach <- list(
    yHat = fit$yHat,
    runtime_sec = elapsed
  )

  rm(fit, saida_essencial, variancias_U)
  gc()

  return(retorno_foreach)
}
```

#### 7.2.2 Function to summarize correlations by environment

This second helper receives a data frame with observed and predicted values and returns the correlation for each environment.

```{r helper-calc-cor-by-env, eval=FALSE}
calc_cor_by_env <- function(df_with_yhat, trait_col) {
  df_with_yhat %>%
    dplyr::group_by(Env) %>%
    dplyr::summarize(
      cor = cor(.data[[trait_col]], yhat, use = "complete.obs"),
      .groups = "drop"
    ) %>%
    as.data.frame()
}
```

### 7.3 Create the full grid of parallel jobs

The prediction stage is parallelized at the level of `MODEL × REP × TRAIT`. This is the main efficiency decision of the final version because it distributes larger independent jobs across workers instead of parallelizing only inside a single model loop.

```{r define-job-grid, eval=FALSE}
combos <- expand.grid(
  MODEL = 1:18,
  REP = seq_len(num_rep),
  TRAIT = traits_to_run,
  stringsAsFactors = FALSE
)
```

### 7.4 Initialize the parallel backend

This chunk starts the cluster, registers the workers for `foreach`, and loads the packages that each worker needs during execution.

```{r start-parallel-backend, eval=FALSE}
cl <- parallel::makeCluster(n_cores)
doParallel::registerDoParallel(cl)
on.exit({
  try(parallel::stopCluster(cl), silent = TRUE)
  foreach::registerDoSEQ()
}, add = TRUE)

parallel::clusterEvalQ(cl, {
  library(BGLR)
  library(dplyr)
  library(plyr)
  NULL
})

gc()
```

### 7.5 Run all jobs in parallel

The original long execution block is now decomposed into smaller stages. The code still follows exactly the same computational logic, but the reader can inspect each stage separately.

#### 7.5.1 Prepare one job before entering the fold loop

This helper creates the job-specific objects: a local copy of the phenotype table, empty result lists, the model label, and the pedigree-based fold assignment for the current `MODEL × REP × TRAIT` combination.

```{r define-job-initialization, eval=FALSE}
initialize_prediction_job <- function(MODEL, rep_num, TRAIT) {
  model_label <- model_catalog$Model[MODEL]
  pheno <- data.frame(Pheno)

  CV1_list <- list()
  CV2_list <- list()
  CV0_lists <- lapply(envs_leaveout, function(x) list())
  CV00_lists <- lapply(envs_leaveout, function(x) list())
  names(CV0_lists) <- envs_leaveout
  names(CV00_lists) <- envs_leaveout

  set.seed(100000 + MODEL * 100 + rep_num + ifelse(TRAIT == "GY", 0, 50000))
  obs_per_fold <- ceiling(n_hybrids / num_fold)
  folds <- sample(rep(seq_len(num_fold), each = obs_per_fold, length.out = n_hybrids))

  folds_df <- data.frame(
    Pedigree = Pedigree,
    fold = folds,
    stringsAsFactors = FALSE
  )

  list(
    model_label = model_label,
    pheno = pheno,
    CV1_list = CV1_list,
    CV2_list = CV2_list,
    CV0_lists = CV0_lists,
    CV00_lists = CV00_lists,
    folds_df = folds_df
  )
}
```

#### 7.5.2 Run `CV1` and `CV2` for one fold

This helper reproduces the reference-script logic for the base prediction fit. It prepares the training mask, fits the model, and returns the fold-level tables for `CV1` and `CV2`.

```{r define-cv1-cv2-step, eval=FALSE}
run_cv1_cv2_for_fold <- function(fold_num, folds_df, pheno, TRAIT, MODEL, model_label, rep_num) {
  train_geno <- folds_df$Pedigree[folds_df$fold != fold_num]
  test_geno <- folds_df$Pedigree[folds_df$fold == fold_num]

  yield <- pheno
  yield$Y2 <- NA_real_
  yield$Y2[yield$Pedigree %in% train_geno] <- yield[[TRAIT]][yield$Pedigree %in% train_geno]

  cv12_run_dir <- file.path(
    "output", "results", "bglr_runs",
    "CV1_CV2",
    TRAIT,
    model_label,
    paste0("rep", rep_num),
    paste0("fold", fold_num)
  )
  dir.create(cv12_run_dir, recursive = TRUE, showWarnings = FALSE)

  existing_cv12_files <- list.files(cv12_run_dir, full.names = TRUE, all.files = TRUE, no.. = TRUE)
  if (length(existing_cv12_files) > 0) {
    unlink(existing_cv12_files, recursive = TRUE, force = TRUE)
  }

  save_prefix_full <- file.path(
    cv12_run_dir,
    paste0("full_", TRAIT, "_", model_label, "_rep", rep_num, "_fold", fold_num, "_")
  )

  res_full <- run_bglr(
    yield$Y2,
    ETA = Eta_list[[MODEL]],
    nIter = nIter,
    burnIn = burnIn,
    thin = thin,
    save_prefix = save_prefix_full
  )

  yield$yhat <- res_full$yHat
  runtime_full <- res_full$runtime_sec

  df_test <- yield[!yield$Pedigree %in% train_geno, ]
  df_train <- yield[yield$Pedigree %in% train_geno, ]

  tmp1 <- calc_cor_by_env(df_test, TRAIT)
  tmp2 <- calc_cor_by_env(df_train, TRAIT)

  tmp1$Runtime_sec <- runtime_full
  tmp2$Runtime_sec <- runtime_full

  list(
    train_geno = train_geno,
    test_geno = test_geno,
    cv1_result = tmp1,
    cv2_result = tmp2
  )
}
```

#### 7.5.3 Run `CV0` and `CV00` for one fold and all leave-one-environment-out cases

This helper keeps the same logic used in the final corrected script: for each fold, it reuses `train_geno`, removes one environment at a time, fits the model again, and stores fold-level results for `CV0` and `CV00`.

```{r define-cv0-cv00-step, eval=FALSE}
run_cv0_cv00_for_fold <- function(fold_num, train_geno, pheno, TRAIT, MODEL, model_label, rep_num) {
  CV0_fold_results <- lapply(envs_leaveout, function(x) NULL)
  CV00_fold_results <- lapply(envs_leaveout, function(x) NULL)

  for (env_idx in seq_along(envs_leaveout)) {
    env_name <- envs_leaveout[env_idx]

    yield2 <- pheno
    yield2$Y2 <- NA_real_
    yield2$Y2[yield2$Pedigree %in% train_geno] <- yield2[[TRAIT]][yield2$Pedigree %in% train_geno]
    yield2$Y2[yield2$Env == env_name] <- NA_real_

    cv0_cv00_run_dir <- file.path(
      "output", "results", "bglr_runs",
      "CV0_CV00",
      TRAIT,
      model_label,
      paste0("rep", rep_num),
      paste0("fold", fold_num),
      env_name
    )
    dir.create(cv0_cv00_run_dir, recursive = TRUE, showWarnings = FALSE)

    existing_cv0_cv00_files <- list.files(cv0_cv00_run_dir, full.names = TRUE, all.files = TRUE, no.. = TRUE)
    if (length(existing_cv0_cv00_files) > 0) {
      unlink(existing_cv0_cv00_files, recursive = TRUE, force = TRUE)
    }

    save_prefix_env <- file.path(
      cv0_cv00_run_dir,
      paste0(
        env_name, "_", TRAIT, "_", model_label,
        "_rep", rep_num, "_fold", fold_num, "_"
      )
    )

    res_env <- run_bglr(
      yield2$Y2,
      ETA = Eta_list[[MODEL]],
      nIter = nIter,
      burnIn = burnIn,
      thin = thin,
      save_prefix = save_prefix_env
    )

    yield2$yhat <- res_env$yHat
    runtime_env <- res_env$runtime_sec

    df_test2 <- yield2[!yield2$Pedigree %in% train_geno, ]
    df_train2 <- yield2[yield2$Pedigree %in% train_geno, ]

    tmp00 <- calc_cor_by_env(df_test2, TRAIT)
    tmp0 <- calc_cor_by_env(df_train2, TRAIT)

    tmp00$Runtime_sec <- runtime_env
    tmp0$Runtime_sec <- runtime_env

    CV00_fold_results[[env_idx]] <- tmp00
    CV0_fold_results[[env_idx]] <- tmp0
  }

  list(
    cv0_fold_results = CV0_fold_results,
    cv00_fold_results = CV00_fold_results
  )
}
```

#### 7.5.4 Write the aggregated outputs for one job

After all folds are processed, this helper collapses the fold-level tables with `plyr::ldply(...)` and writes the final CSV files expected by the next scripts in the pipeline.

```{r define-output-writing-step, eval=FALSE}
write_prediction_outputs <- function(CV1_list, CV2_list, CV0_lists, CV00_lists, TRAIT, model_label, rep_num) {
  CV1out <- plyr::ldply(CV1_list, data.frame)
  CV2out <- plyr::ldply(CV2_list, data.frame)

  write.csv(
    CV1out,
    file = file.path(
      "output/results",
      paste0("CV1_", TRAIT, "_", model_label, "_rep", rep_num, ".csv")
    ),
    row.names = FALSE
  )

  write.csv(
    CV2out,
    file = file.path(
      "output/results",
      paste0("CV2_", TRAIT, "_", model_label, "_rep", rep_num, ".csv")
    ),
    row.names = FALSE
  )

  for (env_idx in seq_along(envs_leaveout)) {
    env_name <- envs_leaveout[env_idx]
    env_parts <- strsplit(env_name, "_")[[1]]

    CV00out <- plyr::ldply(CV00_lists[[env_idx]], data.frame)
    CV0out <- plyr::ldply(CV0_lists[[env_idx]], data.frame)

    write.csv(
      CV00out,
      file = file.path(
        "output/results",
        paste0(
          "CV00_", TRAIT, "_",
          env_parts[1], "_", env_parts[2], "_",
          model_label, "_rep", rep_num, ".csv"
        )
      ),
      row.names = FALSE
    )

    write.csv(
      CV0out,
      file = file.path(
        "output/results",
        paste0(
          "CV0_", TRAIT, "_",
          env_parts[1], "_", env_parts[2], "_",
          model_label, "_rep", rep_num, ".csv"
        )
      ),
      row.names = FALSE
    )
  }
}
```

#### 7.5.5 Execute the full job grid with `foreach`

Now the parallel call itself becomes shorter and easier to follow: each worker initializes one job, iterates over the five folds, updates the result lists, writes the aggregated outputs, and finally returns a compact diagnostics row.

```{r run-jobs, eval=FALSE}
job_diagnostics <- foreach(
  i = seq_len(nrow(combos)),
  .combine = rbind,
  .packages = c("BGLR", "dplyr", "plyr"),
  .export = c(
    "Pheno", "Pedigree", "n_hybrids", "num_fold", "num_rep",
    "envs_leaveout", "traits_to_run", "Eta_list", "model_catalog",
    "run_bglr", "calc_cor_by_env", "initialize_prediction_job",
    "run_cv1_cv2_for_fold", "run_cv0_cv00_for_fold",
    "write_prediction_outputs", "nIter", "burnIn", "thin"
  )
) %dopar% {
  MODEL <- combos$MODEL[i]
  rep_num <- combos$REP[i]
  TRAIT <- as.character(combos$TRAIT[i])

  job_state <- initialize_prediction_job(MODEL, rep_num, TRAIT)
  model_label <- job_state$model_label
  pheno <- job_state$pheno
  CV1_list <- job_state$CV1_list
  CV2_list <- job_state$CV2_list
  CV0_lists <- job_state$CV0_lists
  CV00_lists <- job_state$CV00_lists
  folds_df <- job_state$folds_df

  for (fold_num in seq_len(num_fold)) {
    cv12_step <- run_cv1_cv2_for_fold(
      fold_num = fold_num,
      folds_df = folds_df,
      pheno = pheno,
      TRAIT = TRAIT,
      MODEL = MODEL,
      model_label = model_label,
      rep_num = rep_num
    )

    CV1_list[[length(CV1_list) + 1]] <- cv12_step$cv1_result
    CV2_list[[length(CV2_list) + 1]] <- cv12_step$cv2_result

    cv0_cv00_step <- run_cv0_cv00_for_fold(
      fold_num = fold_num,
      train_geno = cv12_step$train_geno,
      pheno = pheno,
      TRAIT = TRAIT,
      MODEL = MODEL,
      model_label = model_label,
      rep_num = rep_num
    )

    for (env_idx in seq_along(envs_leaveout)) {
      CV00_lists[[env_idx]][[length(CV00_lists[[env_idx]]) + 1]] <- cv0_cv00_step$cv00_fold_results[[env_idx]]
      CV0_lists[[env_idx]][[length(CV0_lists[[env_idx]]) + 1]] <- cv0_cv00_step$cv0_fold_results[[env_idx]]
    }
  }

  write_prediction_outputs(
    CV1_list = CV1_list,
    CV2_list = CV2_list,
    CV0_lists = CV0_lists,
    CV00_lists = CV00_lists,
    TRAIT = TRAIT,
    model_label = model_label,
    rep_num = rep_num
  )

  data.frame(
    Trait = TRAIT,
    Model = model_label,
    Rep = rep_num,
    Status = "ok",
    stringsAsFactors = FALSE
  )
}
```

### 7.6 Save the job diagnostics table

The final chunk of this section stores a compact table showing which jobs were completed. This is useful for checking whether all combinations of trait, model, and repetition finished correctly.

```{r save-job-diagnostics, eval=FALSE}
write.csv(
  job_diagnostics,
  file.path("output", "tables", "analysis_prediction_reference_aligned_jobs.csv"),
  row.names = FALSE
)
```

## 8. Final note

This corrected version produces summary prediction outputs aligned with the reference-script CV definitions.

The CSV filenames in `output/results/` include:

- CV type
- trait
- model label (`M01` to `M18`)
- repetition
- held-out environment when applicable

Examples:

- `CV1_GY_M05_rep2.csv`
- `CV2_KW_M12_rep8.csv`
- `CV0_GY_CS11_WS_M10_rep1.csv`
- `CV00_KW_CS12_WW_M18_rep1.csv`

The files generated by `BGLR` through `saveAt` are stored in subfolders under:

`output/results/bglr_runs/`

organized by:

- `CV1_CV2` or `CV0_CV00`
- trait
- model
- repetition
- fold
- held-out environment when applicable

The outputs of `CV1`, `CV2`, `CV0`, and `CV00` are aggregated across the 5 folds inside each repetition, matching the reference-script workflow.
