---
title: "Prediction analyses - run prediction and outputs"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

**Language / Idioma:** [English](analysis_prediction_run_outputs.html) | [Português](analysis_prediction_run_outputs_pt.html)

**Module navigation / Navegação do módulo:** [Part I: setup, models and CV design](analysis_prediction.html)

## 1. Introduction

This submodule is the execution counterpart of `analysis_prediction.Rmd`. Its role is to run the prediction jobs for the reduced catalog of 18 models, save the CSV files under `output/results/`, and inventory the outputs that will later be consumed by the visualization stage.

The execution contract used here has four operational rules:

- `CV1` and `CV2` are derived from the same fit within each fold;
- `CV0` and `CV00` are derived from the same fit for each held-out environment within each fold;
- the fold partition is deterministic for each `MODEL × REP × TRAIT` task and is reused in both CV blocks;
- output files are organized explicitly by validation scheme, environment when applicable, trait, repetition, and fold, with the fold represented both in the directory tree and in the file name.

The heavy fitting chunks remain disabled by default in the tutorial. Even so, the code shown below is the executable version that should be used when the prediction stage is rerun.

## 2. Packages and project paths

The execution module needs both the matrix objects created upstream and the packages used by the parallel fitting layer. The path logic below uses the current working directory as the project root.


```{r setup}
rm(list = ls())
options(scipen = 999)

library(readr)
library(dplyr)
library(stringr)
library(knitr)
library(kableExtra)
library(BGLR)
library(foreach)
library(doParallel)
library(parallel)
library(tidyr)

project_root <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)

results_dir <- file.path(project_root, "output", "results")
analysis_tables_dir <- file.path(project_root, "output", "tables", "analysis_prediction")
bglr_runs_dir <- file.path(results_dir, "bglr_runs")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(analysis_tables_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(bglr_runs_dir, recursive = TRUE, showWarnings = FALSE)
```


## 3. Reload the execution objects from Part I

This section rebuilds the execution environment when the file is rendered on its own. It follows the contract of `analysis_prediction.Rmd` and recreates the reduced set of 18 ETA objects exactly as defined in Part I.

```{r load-required-objects, eval=FALSE}
matrices_dir <- file.path(project_root, "output", "Matrizes")

required_matrix_files <- c(
  "Pheno.rds",
  "ZG.rds", "ZP.rds", "ZE.rds", "ZW.rds",
  "ZGZE.rds", "ZPZE.rds", "ZGZW.rds", "ZPZW.rds",
  "GGK.rds", "PGK.rds", "GAK.rds", "PAK.rds",
  "GGKE.rds", "PGKE.rds", "GAKE.rds", "PAKE.rds",
  "GGKW.rds", "PGKW.rds", "GAKW.rds", "PAKW.rds"
)

Pheno <- readRDS(file.path(matrices_dir, "Pheno.rds"))
ZG   <- readRDS(file.path(matrices_dir, "ZG.rds"))
ZP   <- readRDS(file.path(matrices_dir, "ZP.rds"))
ZE   <- readRDS(file.path(matrices_dir, "ZE.rds"))
ZW   <- readRDS(file.path(matrices_dir, "ZW.rds"))
ZGZE <- readRDS(file.path(matrices_dir, "ZGZE.rds"))
ZPZE <- readRDS(file.path(matrices_dir, "ZPZE.rds"))
ZGZW <- readRDS(file.path(matrices_dir, "ZGZW.rds"))
ZPZW <- readRDS(file.path(matrices_dir, "ZPZW.rds"))
GGK  <- readRDS(file.path(matrices_dir, "GGK.rds"))
PGK  <- readRDS(file.path(matrices_dir, "PGK.rds"))
GAK  <- readRDS(file.path(matrices_dir, "GAK.rds"))
PAK  <- readRDS(file.path(matrices_dir, "PAK.rds"))
GGKE <- readRDS(file.path(matrices_dir, "GGKE.rds"))
PGKE <- readRDS(file.path(matrices_dir, "PGKE.rds"))
GAKE <- readRDS(file.path(matrices_dir, "GAKE.rds"))
PAKE <- readRDS(file.path(matrices_dir, "PAKE.rds"))
GGKW <- readRDS(file.path(matrices_dir, "GGKW.rds"))
PGKW <- readRDS(file.path(matrices_dir, "PGKW.rds"))
GAKW <- readRDS(file.path(matrices_dir, "GAKW.rds"))
PAKW <- readRDS(file.path(matrices_dir, "PAKW.rds"))

Pedigree <- unique(Pheno$Pedigree)

traits_to_run <- c("GY", "KW")

nIter <- 5000
burnIn <- 1000
thin <- 10
num_rep <- 10
n_cores <- max(1L, parallel::detectCores() - 1L)

model_catalog <- data.frame(
  ModelNumber = 1:18,
  Model = paste0("M", 1:18),
  EtaName = paste0("Eta", 1:18),
  Description = c(
    "E + G + GE",
    "E + P + PE",
    "E + GGK + GGKE",
    "E + PGK + PGKE",
    "E + GAK + GAKE",
    "E + PAK + PAKE",
    "E + G + P + GE + PE",
    "E + GGK + PGK + GGKE + PGKE",
    "E + GAK + PAK + GAKE + PAKE",
    "E + W + G + GE + GW",
    "E + W + P + PE + PW",
    "E + W + G + P + GE + PE + GW + PW",
    "E + W + GGK + GGKE + GGKW",
    "E + W + PGK + PGKE + PGKW",
    "E + W + GGK + PGK + GGKE + PGKE + GGKW + PGKW",
    "E + W + GAK + GAKE + GAKW",
    "E + W + PAK + PAKE + PAKW",
    "E + W + GAK + PAK + GAKE + PAKE + GAKW + PAKW"
  ),
  FullEffects = c(
    "E + G + GE",
    "E + P + PE",
    "E + GGK + GGKE",
    "E + PGK + PGKE",
    "E + GAK + GAKE",
    "E + PAK + PAKE",
    "E + G + P + GE + PE",
    "E + GGK + PGK + GGKE + PGKE",
    "E + GAK + PAK + GAKE + PAKE",
    "E + W + G + GE + GW",
    "E + W + P + PE + PW",
    "E + W + G + P + GE + PE + GW + PW",
    "E + W + GGK + GGKE + GGKW",
    "E + W + PGK + PGKE + PGKW",
    "E + W + GGK + PGK + GGKE + PGKE + GGKW + PGKW",
    "E + W + GAK + GAKE + GAKW",
    "E + W + PAK + PAKE + PAKW",
    "E + W + GAK + PAK + GAKE + PAKE + GAKW + PAKW"
  ),
  InterpretiveDescription = c(
    "Linear genomic model with environment and G × E interaction",
    "Linear phenomic model with environment and P × E interaction",
    "Gaussian genomic model with environment and GGK × E interaction",
    "Gaussian phenomic model with environment and PGK × E interaction",
    "Arccosine genomic model with environment and GAK × E interaction",
    "Arccosine phenomic model with environment and PAK × E interaction",
    "Linear combined genomic + phenomic model with environment, GE, and PE",
    "Gaussian combined genomic + phenomic model with environment, GGKE, and PGKE",
    "Arccosine combined genomic + phenomic model with environment, GAKE, and PAKE",
    "Linear genomic-weather model with environment, weather, GE, and GW",
    "Linear phenomic-weather model with environment, weather, PE, and PW",
    "Linear combined genomic + phenomic + weather model with GE, PE, GW, and PW",
    "Gaussian genomic-weather model with environment, weather, GGKE, and GGKW",
    "Gaussian phenomic-weather model with environment, weather, PGKE, and PGKW",
    "Gaussian combined genomic + phenomic + weather model with GGKE, PGKE, GGKW, and PGKW",
    "Arccosine genomic-weather model with environment, weather, GAKE, and GAKW",
    "Arccosine phenomic-weather model with environment, weather, PAKE, and PAKW",
    "Arccosine combined genomic + phenomic + weather model with GAKE, PAKE, GAKW, and PAKW"
  ),
  stringsAsFactors = FALSE
)

# The column Model gives the manuscript label used in tables and figures
# (M1-M18), whereas EtaName keeps the computational identifier used
# internally in file names and BGLR run folders (Eta1-Eta18).

env_design <- model.matrix(~ Pheno$Env - 1)

Eta_list <- vector("list", 18)
Eta_list[[1]]  <- list(list(X = env_design, model = "BRR"), list(K = ZG, model = "RKHS"), list(K = ZGZE, model = "RKHS"))
Eta_list[[2]]  <- list(list(X = env_design, model = "BRR"), list(K = ZP, model = "RKHS"), list(K = ZPZE, model = "RKHS"))
Eta_list[[3]]  <- list(list(X = env_design, model = "BRR"), list(K = GGK, model = "RKHS"), list(K = GGKE, model = "RKHS"))
Eta_list[[4]]  <- list(list(X = env_design, model = "BRR"), list(K = PGK, model = "RKHS"), list(K = PGKE, model = "RKHS"))
Eta_list[[5]]  <- list(list(X = env_design, model = "BRR"), list(K = GAK, model = "RKHS"), list(K = GAKE, model = "RKHS"))
Eta_list[[6]]  <- list(list(X = env_design, model = "BRR"), list(K = PAK, model = "RKHS"), list(K = PAKE, model = "RKHS"))
Eta_list[[7]]  <- list(list(X = env_design, model = "BRR"), list(K = ZG, model = "RKHS"), list(K = ZP, model = "RKHS"), list(K = ZGZE, model = "RKHS"), list(K = ZPZE, model = "RKHS"))
Eta_list[[8]]  <- list(list(X = env_design, model = "BRR"), list(K = GGK, model = "RKHS"), list(K = PGK, model = "RKHS"), list(K = GGKE, model = "RKHS"), list(K = PGKE, model = "RKHS"))
Eta_list[[9]]  <- list(list(X = env_design, model = "BRR"), list(K = GAK, model = "RKHS"), list(K = PAK, model = "RKHS"), list(K = GAKE, model = "RKHS"), list(K = PAKE, model = "RKHS"))
Eta_list[[10]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = ZG, model = "RKHS"), list(K = ZGZE, model = "RKHS"), list(K = ZGZW, model = "RKHS"))
Eta_list[[11]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = ZP, model = "RKHS"), list(K = ZPZE, model = "RKHS"), list(K = ZPZW, model = "RKHS"))
Eta_list[[12]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = ZG, model = "RKHS"), list(K = ZP, model = "RKHS"), list(K = ZGZE, model = "RKHS"), list(K = ZPZE, model = "RKHS"), list(K = ZGZW, model = "RKHS"), list(K = ZPZW, model = "RKHS"))
Eta_list[[13]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = GGK, model = "RKHS"), list(K = GGKE, model = "RKHS"), list(K = GGKW, model = "RKHS"))
Eta_list[[14]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = PGK, model = "RKHS"), list(K = PGKE, model = "RKHS"), list(K = PGKW, model = "RKHS"))
Eta_list[[15]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = GGK, model = "RKHS"), list(K = PGK, model = "RKHS"), list(K = GGKE, model = "RKHS"), list(K = PGKE, model = "RKHS"), list(K = GGKW, model = "RKHS"), list(K = PGKW, model = "RKHS"))
Eta_list[[16]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = GAK, model = "RKHS"), list(K = GAKE, model = "RKHS"), list(K = GAKW, model = "RKHS"))
Eta_list[[17]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = PAK, model = "RKHS"), list(K = PAKE, model = "RKHS"), list(K = PAKW, model = "RKHS"))
Eta_list[[18]] <- list(list(X = env_design, model = "BRR"), list(K = ZW, model = "RKHS"), list(K = GAK, model = "RKHS"), list(K = PAK, model = "RKHS"), list(K = GAKE, model = "RKHS"), list(K = PAKE, model = "RKHS"), list(K = GAKW, model = "RKHS"), list(K = PAKW, model = "RKHS"))

```


## 4. Define the execution contract

This block restores the operational objects used by the current workflow: the fixed leave-one-environment-out set, the number of folds, and the `MODEL × REP × TRAIT` task grid for the 18-model catalog. The fitting and summarization logic are written directly inside the CV blocks so that the execution path stays explicit.

```{r define-run-objects, eval=FALSE}
envs_leaveout <- c("CS11_WS", "CS11_WW", "CS12_WS", "CS12_WW")
envs_leaveout <- envs_leaveout[envs_leaveout %in% unique(Pheno$Env)]

num_fold <- 5L
n_hybrids <- length(Pedigree)
combos <- expand.grid(
  MODEL = seq_along(Eta_list),
  REP = seq_len(num_rep),
  TRAIT = traits_to_run,
  stringsAsFactors = FALSE
)
```

## 5. Start the parallel backend

The backend below mirrors the original workflow: one worker receives one combination of model, repetition, and trait.

```{r start-parallel-backend, eval=FALSE}
useCores <- max(1L, n_cores)
cl <- parallel::makeCluster(useCores)
doParallel::registerDoParallel(cl)

parallel::clusterEvalQ(cl, {
  library(BGLR)
  library(dplyr)
  library(plyr)
  NULL
})

cat("Running on ", useCores, " cores.")
```


## 6. Run prediction analyses for CV1 and CV2

This block implements genotype-based validation for the 18-model workflow. The deterministic fold map, the BGLR fit, and the environment-wise summary are all written directly inside the loop so that each step stays explicit. One fit is run per fold, and the resulting predictions are split into `CV1` (held-out genotypes) and `CV2` (training genotypes under the same fit). In the current standard, both the final CSV summaries and the BGLR fitting artifacts are stored in a rep/fold hierarchy.

```{r run-predictions-cv12, eval=FALSE}
foreach(i = seq_len(nrow(combos)),
        .packages = c("BGLR", "dplyr", "plyr")) %dopar% {
  MODEL <- combos$MODEL[i]
  rep_num <- combos$REP[i]
  trait_id <- as.character(combos$TRAIT[i])

  pheno <- data.frame(Pheno)

  seed_value <- as.integer(100000L * MODEL + 100L * rep_num + as.numeric(as.factor(combos$TRAIT[i])))
  set.seed(seed_value)
  obs_per_fold <- ceiling(length(Pedigree) / num_fold)
  folds <- sample(rep(seq_len(num_fold), each = obs_per_fold, length.out = length(Pedigree)))

  folds_df <- data.frame(Pedigree = Pedigree, fold = folds, stringsAsFactors = FALSE)

  for (fold_num in seq_len(num_fold)) {
    rep_label <- sprintf("rep%02d", rep_num)
    fold_label <- sprintf("fold%02d", fold_num)

    cv1_dir <- file.path(results_dir, "CV1", trait_id, rep_label, fold_label)
    cv2_dir <- file.path(results_dir, "CV2", trait_id, rep_label, fold_label)
    dir.create(cv1_dir, recursive = TRUE, showWarnings = FALSE)
    dir.create(cv2_dir, recursive = TRUE, showWarnings = FALSE)

    cv1_file <- file.path(cv1_dir, paste0("CV1_", trait_id, "_Eta", MODEL, "_", rep_label, "_", fold_label, ".csv"))
    cv2_file <- file.path(cv2_dir, paste0("CV2_", trait_id, "_Eta", MODEL, "_", rep_label, "_", fold_label, ".csv"))

    train_geno <- folds_df$Pedigree[folds_df$fold != fold_num]

    yield <- pheno
    yield$Y2 <- NA
    yield$Y2[yield$Pedigree %in% train_geno] <- yield[[trait_id]][yield$Pedigree %in% train_geno]

    run_dir <- file.path(bglr_runs_dir, "CV1", trait_id, rep_label, fold_label, paste0("Eta", MODEL))
    dir.create(run_dir, recursive = TRUE, showWarnings = FALSE)

    save_prefix_full <- file.path(run_dir, paste0("CV1_", trait_id, "_Eta", MODEL, "_", rep_label, "_", fold_label, "_"))

    y_t <- as.numeric(yield$Y2)
    start_time <- Sys.time()
    fit <- BGLR(y = y_t, ETA = Eta_list[[MODEL]], nIter = nIter, burnIn = burnIn, thin = thin, saveAt = save_prefix_full)
    end_time <- Sys.time()
    runtime_full <- as.numeric(difftime(end_time, start_time, units = "secs"))

    yield$yhat <- fit$yHat
    rm(fit)
    gc()

    df_test <- yield[!yield$Pedigree %in% train_geno, ]
    df_train <- yield[yield$Pedigree %in% train_geno, ]

    tmp1 <- df_test %>%
      dplyr::group_by(Env) %>%
      dplyr::summarize(cor = cor(.data[[trait_id]], yhat, use = "complete.obs"), .groups = "drop") %>%
      as.data.frame()

    tmp2 <- df_train %>%
      dplyr::group_by(Env) %>%
      dplyr::summarize(cor = cor(.data[[trait_id]], yhat, use = "complete.obs"), .groups = "drop") %>%
      as.data.frame()

    tmp1$Runtime_sec <- runtime_full
    tmp2$Runtime_sec <- runtime_full
    tmp1$Fold <- fold_label
    tmp2$Fold <- fold_label

    write.csv(tmp1, file = cv1_file, row.names = FALSE)
    write.csv(tmp2, file = cv2_file, row.names = FALSE)
  }

  NULL
}
```
## 7. Run prediction analyses for CV0 and CV00

This block implements leave-one-environment-out validation for the same 18-model workflow. For each task, the deterministic fold map is rebuilt with the same rule used in `CV1/CV2`, one fit is run for each `env_leave × fold` combination, and the predictions are split into `CV0` (training genotypes under the environment mask) and `CV00` (held-out genotypes under the same mask). As in the genotype-based block, the fold construction, the BGLR fit, and the environment-wise summary are all written directly inside the loop so that the execution path remains explicit. The outputs are written in a rep/fold hierarchy and the BGLR artifacts are preserved in matching persistent folders.

```{r run-predictions-cv0-cv00, eval=FALSE}
foreach(i = seq_len(nrow(combos)),
        .packages = c("BGLR", "dplyr", "plyr")) %dopar% {
  MODEL <- combos$MODEL[i]
  rep_num <- combos$REP[i]
  trait_id <- as.character(combos$TRAIT[i])

  pheno <- data.frame(Pheno)

  seed_value <- as.integer(100000L * MODEL + 100L * rep_num + as.numeric(as.factor(combos$TRAIT[i])))
  set.seed(seed_value)
  obs_per_fold <- ceiling(length(Pedigree) / num_fold)
  folds <- sample(rep(seq_len(num_fold), each = obs_per_fold, length.out = length(Pedigree)))

  folds_df <- data.frame(Pedigree = Pedigree, fold = folds, stringsAsFactors = FALSE)

  for (fold_num in seq_len(num_fold)) {
    train_geno <- folds_df$Pedigree[folds_df$fold != fold_num]
    rep_label <- sprintf("rep%02d", rep_num)
    fold_label <- sprintf("fold%02d", fold_num)

    for (env_idx in seq_along(envs_leaveout)) {
      env_name <- envs_leaveout[env_idx]

      cv00_dir <- file.path(results_dir, "CV00", env_name, trait_id, rep_label, fold_label)
      cv0_dir <- file.path(results_dir, "CV0", env_name, trait_id, rep_label, fold_label)
      dir.create(cv00_dir, recursive = TRUE, showWarnings = FALSE)
      dir.create(cv0_dir, recursive = TRUE, showWarnings = FALSE)

      cv00_file <- file.path(cv00_dir, paste0("CV00_", trait_id, "_", env_name, "_Eta", MODEL, "_", rep_label, "_", fold_label, ".csv"))
      cv0_file <- file.path(cv0_dir, paste0("CV0_", trait_id, "_", env_name, "_Eta", MODEL, "_", rep_label, "_", fold_label, ".csv"))

      yield2 <- pheno
      yield2$Y2 <- NA
      yield2$Y2[yield2$Pedigree %in% train_geno] <- yield2[[trait_id]][yield2$Pedigree %in% train_geno]
      yield2$Y2[yield2$Env == env_name] <- NA

      run_dir <- file.path(bglr_runs_dir, "CV0", env_name, trait_id, rep_label, fold_label, paste0("Eta", MODEL))
      dir.create(run_dir, recursive = TRUE, showWarnings = FALSE)

      save_prefix_env <- file.path(run_dir, paste0("CV0_", env_name, "_", trait_id, "_Eta", MODEL, "_", rep_label, "_", fold_label, "_"))

      y_t <- as.numeric(yield2$Y2)
      start_time <- Sys.time()
      fit <- BGLR(y = y_t, ETA = Eta_list[[MODEL]], nIter = nIter, burnIn = burnIn, thin = thin, saveAt = save_prefix_env)
      end_time <- Sys.time()
      runtime_env <- as.numeric(difftime(end_time, start_time, units = "secs"))

      yield2$yhat <- fit$yHat
      rm(fit)
      gc()

      df_test2 <- yield2[!yield2$Pedigree %in% train_geno & yield2$Env == env_name, , drop = FALSE]
      df_train2 <- yield2[yield2$Pedigree %in% train_geno & yield2$Env == env_name, , drop = FALSE]

      tmp00 <- df_test2 %>%
        dplyr::group_by(Env) %>%
        dplyr::summarize(cor = cor(.data[[trait_id]], yhat, use = "complete.obs"), .groups = "drop") %>%
        as.data.frame()

      tmp0 <- df_train2 %>%
        dplyr::group_by(Env) %>%
        dplyr::summarize(cor = cor(.data[[trait_id]], yhat, use = "complete.obs"), .groups = "drop") %>%
        as.data.frame()

      tmp00$Runtime_sec <- runtime_env
      tmp0$Runtime_sec <- runtime_env
      tmp00$Fold <- fold_label
      tmp0$Fold <- fold_label

      write.csv(tmp00, file = cv00_file, row.names = FALSE)
      write.csv(tmp0, file = cv0_file, row.names = FALSE)
    }
  }

  NULL
}
```
## 8. Stop the parallel backend

```{r stop-parallel-backend, eval=FALSE}
stopCluster(cl)
foreach::registerDoSEQ()
```


## 9. Inventory of the prediction output files

The inventories below summarize the CSV files produced by the execution stage using the current rep/fold folder contract for the 18-model workflow.

```{r prediction-output-inventory}
prediction_files <- list.files(results_dir, pattern = "\\.csv$", full.names = TRUE, recursive = TRUE)
results_root <- normalizePath(results_dir, winslash = "/", mustWork = FALSE)

cv12_files <- prediction_files[
  stringr::str_detect(basename(prediction_files), "^CV[12]_")
]

cvloo_files <- prediction_files[
  stringr::str_detect(basename(prediction_files), "^CV0{1,2}_")
]

inventory_cv12 <- data.frame()

for (current_file in cv12_files) {
  current_norm <- normalizePath(current_file, winslash = "/", mustWork = FALSE)
  rel_path <- substring(current_norm, nchar(results_root) + 2)
  path_parts <- strsplit(rel_path, "/", fixed = TRUE)[[1]]
  current_name <- basename(current_file)
  current_name_no_ext <- sub("\\.csv$", "", current_name)
  current_parts <- strsplit(current_name_no_ext, "_")[[1]]

  inventory_cv12 <- dplyr::bind_rows(
    inventory_cv12,
    data.frame(
      relative_path = rel_path,
      file_name = current_name,
      CV = current_parts[1],
      Trait = path_parts[2],
      Rep = path_parts[3],
      Fold = path_parts[4],
      EtaName = current_parts[3],
      Model = stringr::str_replace(current_parts[3], "Eta", "M"),
      stringsAsFactors = FALSE
    )
  )
}

inventory_cvloo <- data.frame()

for (current_file in cvloo_files) {
  current_norm <- normalizePath(current_file, winslash = "/", mustWork = FALSE)
  rel_path <- substring(current_norm, nchar(results_root) + 2)
  path_parts <- strsplit(rel_path, "/", fixed = TRUE)[[1]]
  current_name <- basename(current_file)
  current_name_no_ext <- sub("\\.csv$", "", current_name)
  current_parts <- strsplit(current_name_no_ext, "_")[[1]]

  inventory_cvloo <- dplyr::bind_rows(
    inventory_cvloo,
    data.frame(
      relative_path = rel_path,
      file_name = current_name,
      CV = current_parts[1],
      Env_leave = path_parts[2],
      Trait = path_parts[3],
      Rep = path_parts[4],
      Fold = path_parts[5],
      EtaName = current_parts[length(current_parts) - 2],
      Model = stringr::str_replace(current_parts[length(current_parts) - 2], "Eta", "M"),
      stringsAsFactors = FALSE
    )
  )
}

write_csv(inventory_cv12, file.path(analysis_tables_dir, "analysis_prediction_inventory_cv12.csv"))
write_csv(inventory_cvloo, file.path(analysis_tables_dir, "analysis_prediction_inventory_cvloo.csv"))
```
## 10. Counts of files by validation scheme and trait

```{r prediction-output-counts}
combined_inventory <- bind_rows(
  inventory_cv12 %>% mutate(Env_leave = NA_character_),
  inventory_cvloo
)

counts_by_cv <- combined_inventory %>%
  count(CV, name = "n_files") %>%
  arrange(CV)

counts_by_trait <- combined_inventory %>%
  count(CV, Trait, name = "n_files") %>%
  arrange(CV, Trait)

write_csv(counts_by_cv, file.path(analysis_tables_dir, "analysis_prediction_counts_by_cv.csv"))
write_csv(counts_by_trait, file.path(analysis_tables_dir, "analysis_prediction_counts_by_trait.csv"))

counts_by_cv %>%
  knitr::kable(
    caption = "Number of prediction files by validation scheme.",
    align = "cc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = TRUE,
    position = "center",
    font_size = 13
  ) %>%
  kableExtra::row_spec(0, bold = TRUE)
```

## 11. Note on temporary `BGLR` artifacts

In the current 18-model workflow, persistent `saveAt` artifacts are written under `output/results/bglr_runs/` during each fit and preserved for later convergence diagnostics. The scenario-specific paths now separate `CV1`, `CV2`, `CV0`, and `CV00`, and the persisted filenames use the scenario prefix together with explicit effect labels instead of generic `ETA_1`, `ETA_2`, and similar placeholders. Therefore, this execution module inventories both the final CSV outputs stored under `output/results/` and the hierarchical paths that locate the corresponding BGLR run folders.

## 12. Final note

This file closes the execution layer of the prediction stage while preserving consistency with the reduced 18-model preprocessing module documented in `analysis_prediction.Rmd` and with the current rep/fold output contract used by the downstream summaries.

## Stage navigation

**Previous stage:** [Prediction — Part I](analysis_prediction.html)  
**Next stage:** [Visualization](visualization.html)

