---
title: "Construction of Genomic, Phenomic, Environmental, and Interaction Kernels"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

# 1. Introduction

This tutorial documents how the project constructs the core kernels used in the genomic prediction analyses. The goal is not simply to read files and calculate matrices. The goal is to transform heterogeneous information sources into aligned scientific objects that can be used in multi-environment prediction models.

The script integrates four data layers:

1. **Phenotypic records** (`Pheno`): grain yield (`GY`) and kernel weight (`KW`) by genotype and environment.
2. **Genomic markers** (`Geno`): SNP markers processed in numerical format.
3. **NIR spectra** (`NIR`): phenomic measurements used to build a spectral similarity kernel.
4. **Environmental covariates** (`W`): weather-derived descriptors generated in the climate tutorial.

The final outputs of this script are the matrices used later in the prediction models:

- **Linear kernels**: `ZG`, `ZP`, `ZEZE`, `ZW`
- **Interaction kernels**: `ZGZE`, `ZPZE`, `ZGZW`, `ZPZW`
- **Nonlinear Gaussian kernels**: `GGK`, `PGK`, `GGKE`, `PGKE`, `GGKW`, `PGKW`
- **Nonlinear arc-cosine kernels**: `GAK`, `PAK`, `GAKE`, `PAKE`, `GAKW`, `PAKW`

A key practical objective of this tutorial is to guarantee that all matrices are constructed in the **same observation order**. If the kernels are not perfectly aligned, the downstream prediction analyses become invalid.

## 2. R environment and dependencies

We first load the packages required for data handling, kernel construction, visualization, and memory management.

- `tidyverse`: data wrangling and plotting.
- `data.table`: fast reading of large files.
- `BGLR`: downstream Bayesian genomic prediction framework.
- `parallel` and `doParallel`: support for computationally intensive tasks.
- `MASS`: additional statistical utilities.
- `patchwork`: multi-panel figure composition for diagnostic plots.

```{r load-libraries}
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(BGLR)
  library(parallel)
  library(doParallel)
  library(MASS)
  library(gridExtra)
  library(grid)
  library(patchwork)
})

dir.create("output/Matrizes", recursive = TRUE, showWarnings = FALSE)
dir.create("output/Matrizes/figures", recursive = TRUE, showWarnings = FALSE)
```

## 2.1 Graphical helper functions

This script is primarily matrix-oriented, but a small set of diagnostic figures is useful to verify whether the main kernels have the expected block structure and alignment. The helpers below standardize the visual style for article-oriented diagnostics.

```{r helper-functions}
theme_article_matrix <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
      plot.subtitle = element_text(size = 10, hjust = 0.5),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(size = 8),
      legend.title = element_text(face = "bold"),
      legend.text = element_text(size = 8),
      panel.grid = element_blank(),
      strip.text = element_text(face = "bold")
    )
}

matrix_to_long <- function(K, n_show = 60) {
  n_show <- min(n_show, nrow(K))
  K_sub <- K[seq_len(n_show), seq_len(n_show), drop = FALSE]
  df <- as.data.frame(as.table(K_sub))
  colnames(df) <- c("Row", "Column", "Value")
  df$Row <- factor(df$Row, levels = rev(rownames(K_sub)))
  df$Column <- factor(df$Column, levels = colnames(K_sub))
  df
}

plot_kernel_excerpt <- function(K, title, subtitle = NULL, n_show = 60) {
  df <- matrix_to_long(K, n_show = n_show)

  ggplot(df, aes(x = Column, y = Row, fill = Value)) +
    geom_raster() +
    scale_fill_gradient2(
      low = "#2166AC",
      mid = "white",
      high = "#B2182B",
      midpoint = median(df$Value, na.rm = TRUE),
      name = "Kernel value"
    ) +
    labs(
      title = title,
      subtitle = subtitle,
      x = "Observation index",
      y = "Observation index"
    ) +
    theme_article_matrix() +
    theme(
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank()
    )
}

check_kernel_alignment <- function(K, reference_ids) {
  tibble(
    n_rows = nrow(K),
    n_cols = ncol(K),
    is_square = nrow(K) == ncol(K),
    symmetric_names = identical(rownames(K), colnames(K)),
    aligned_with_reference = identical(rownames(K), reference_ids)
  )
}
```

## 3. Load the raw data

This section reads the raw inputs used in the kernel construction pipeline. At this stage, the goal is only to import the data and inspect the basic structure of each source.

### 3.1 Load NIR records and phenotypes

The `NIR.csv` file contains identifiers, environments, phenotypic traits, and the spectral bands used later to build the phenomic kernel.

```{r load-nir-data}
NIR <- fread("data/NIR.csv") %>% as.data.frame()

head(NIR[, 1:7])
unique(NIR$Env)
```

### 3.2 Load genomic markers

The genotype file contains one row per genotype and one column per marker. The first column (`taxa`) is the genotype identifier used to align genomic and phenotypic information.

```{r load-genotypic-data}
Geno <- fread("data/GAPIT.Genotype.Numerical.txt") %>% as.data.frame()

head(Geno[, 1:5])
```

### 3.3 Build the phenotype table used downstream

The modeling scripts later require a clean phenotype table containing only the identifiers and the target traits. We extract those columns directly from the NIR file to guarantee that the phenotype table remains synchronized with the observation-level spectral records.

We also create an **observation identifier** (`ObsID`) combining environment and genotype. This is useful because the same genotype can appear in more than one environment, and unique observation labels greatly improve traceability and alignment checks.

```{r build-pheno-table}
Pheno <- NIR %>%
  dplyr::select(Pedigree, Env, GY, KW) %>%
  arrange(Env, Pedigree) %>%
  mutate(ObsID = paste(Env, Pedigree, sep = "__"))

head(Pheno)
```

## 4. Harmonize genotype IDs across data sources

This is a critical validation step. The genomic file and the NIR/phenotypic file must refer to the same genotypes. If the IDs are not harmonized before matrix construction, every downstream kernel will be misaligned.

```{r harmonize-genotypes}
Pedigree <- intersect(Geno$taxa, NIR$Pedigree)

NIR <- NIR[NIR$Pedigree %in% Pedigree, ] %>%
  arrange(Env, Pedigree) %>%
  mutate(ObsID = paste(Env, Pedigree, sep = "__"))

Geno <- Geno[Geno$taxa %in% Pedigree, ] %>%
  arrange(taxa)

Pheno <- Pheno[Pheno$Pedigree %in% Pedigree, ] %>%
  arrange(Env, Pedigree) %>%
  mutate(ObsID = paste(Env, Pedigree, sep = "__"))

cat("Shared genotypes in NIR :", length(unique(NIR$Pedigree)), "\n")
cat("Shared genotypes in Geno:", nrow(Geno), "\n")
cat("Shared genotypes in Pheno:", length(unique(Pheno$Pedigree)), "\n")
```

## 5. Split NIR observations by environment

The study has four environments:

- `CS11_WS`
- `CS11_WW`
- `CS12_WS`
- `CS12_WW`

The kernel construction later uses environment-specific blocks. For that reason, we separate the NIR observations into four environment-specific data frames while preserving the observation identifiers.

```{r split-by-environment}
a <- NIR %>% filter(Env == "CS11_WS") %>% arrange(Pedigree)
b <- NIR %>% filter(Env == "CS11_WW") %>% arrange(Pedigree)
c <- NIR %>% filter(Env == "CS12_WS") %>% arrange(Pedigree)
d <- NIR %>% filter(Env == "CS12_WW") %>% arrange(Pedigree)

rownames(a) <- a$ObsID
rownames(b) <- b$ObsID
rownames(c) <- c$ObsID
rownames(d) <- d$ObsID
```

### 5.1 Identify genotypes observed in all environments

This object is informative for balanced-data checks and for interpreting the overlap across trials.

```{r shared-genotypes-all-envs}
Pedigree_communs <- Reduce(
  intersect,
  list(Geno$taxa, a$Pedigree, b$Pedigree, c$Pedigree, d$Pedigree)
)

length(Pedigree_communs)
```

## 6. Build the linear genomic kernel (K_G stored as `ZG`)

The genomic kernel measures similarity among observations based on SNP markers.

The legacy pipeline used in this project standardizes marker values **within each environment-specific observation block** before stacking the data. Although this is not the only possible strategy, we preserve it here to maintain compatibility with the downstream analyses already developed for the article.

The practical steps are:

1. Sort the genotype matrix and remove the ID column.
2. Subset the genotype matrix according to the genotypes observed in each environment.
3. Standardize markers within each block.
4. Stack the environment-specific matrices in the same order used by the observation table.
5. Compute the linear kernel by the cross-product of the standardized marker matrix.

```{r build-zg}
Geno <- Geno[order(Geno$taxa), ]
rownames(Geno) <- Geno$taxa
Geno <- Geno[, -1, drop = FALSE]

Geno.a <- scale(Geno[a$Pedigree, , drop = FALSE], center = TRUE, scale = TRUE)
Geno.b <- scale(Geno[b$Pedigree, , drop = FALSE], center = TRUE, scale = TRUE)
Geno.c <- scale(Geno[c$Pedigree, , drop = FALSE], center = TRUE, scale = TRUE)
Geno.d <- scale(Geno[d$Pedigree, , drop = FALSE], center = TRUE, scale = TRUE)

rownames(Geno.a) <- a$ObsID
rownames(Geno.b) <- b$ObsID
rownames(Geno.c) <- c$ObsID
rownames(Geno.d) <- d$ObsID

Geno.all <- rbind(Geno.a, Geno.b, Geno.c, Geno.d)

ZG <- tcrossprod(as.matrix(Geno.all)) / ncol(Geno.all)
rownames(ZG) <- rownames(Geno.all)
colnames(ZG) <- rownames(Geno.all)

dim(ZG)
ZG[1:5, 1:5]
```

## 7. Build the linear phenomic kernel (K_P stored as `ZP`)

The phenomic kernel uses the NIR spectral bands. The same blockwise logic is used here to keep the observation order synchronized with the genomic and environmental kernels.

```{r build-zp}
# Identify spectral variables defensively: keep only numeric columns and remove
# metadata/trait columns that should not enter the phenomic kernel.
metadata_cols <- c("Pedigree", "Env", "GY", "KW", "ObsID")

nir_feature_cols <- setdiff(colnames(NIR), metadata_cols)
nir_feature_cols <- nir_feature_cols[
  vapply(NIR[, nir_feature_cols, drop = FALSE], is.numeric, logical(1))
]

if (length(nir_feature_cols) == 0) {
  stop("No numeric NIR feature columns were found after filtering metadata columns.")
}

excluded_non_numeric <- setdiff(setdiff(colnames(NIR), metadata_cols), nir_feature_cols)

cat("Number of NIR feature columns used in ZP:", length(nir_feature_cols), "\n")
if (length(excluded_non_numeric) > 0) {
  cat("Excluded non-numeric columns:", paste(excluded_non_numeric, collapse = ", "), "\n")
}

NIR.a <- scale(as.matrix(a[, nir_feature_cols, drop = FALSE]), center = TRUE, scale = TRUE)
NIR.b <- scale(as.matrix(b[, nir_feature_cols, drop = FALSE]), center = TRUE, scale = TRUE)
NIR.c <- scale(as.matrix(c[, nir_feature_cols, drop = FALSE]), center = TRUE, scale = TRUE)
NIR.d <- scale(as.matrix(d[, nir_feature_cols, drop = FALSE]), center = TRUE, scale = TRUE)

rownames(NIR.a) <- a$ObsID
rownames(NIR.b) <- b$ObsID
rownames(NIR.c) <- c$ObsID
rownames(NIR.d) <- d$ObsID

NIR.all <- rbind(NIR.a, NIR.b, NIR.c, NIR.d)

ZP <- tcrossprod(NIR.all) / ncol(NIR.all)
rownames(ZP) <- rownames(NIR.all)
colnames(ZP) <- rownames(NIR.all)

dim(ZP)
ZP[1:5, 1:5]
```

## 8. Build the categorical environmental kernel (K_E)

The simplest environmental representation treats each environment as a categorical factor. We first build the incidence matrix `ZE`, then convert it into a kernel (`ZEZE`) through the cross-product.

This kernel assigns full environmental similarity to observations from the same environment and zero similarity to observations from different environments.

```{r build-ze}
Pheno <- Pheno %>% arrange(match(ObsID, rownames(ZG)))

ZE <- model.matrix(~ as.factor(Pheno$Env) - 1)
rownames(ZE) <- Pheno$ObsID
colnames(ZE) <- gsub("^as.factor\\(Pheno\\$Env\\)", "", colnames(ZE))

ZEZE <- tcrossprod(ZE)
rownames(ZEZE) <- rownames(ZE)
colnames(ZEZE) <- rownames(ZE)

cat("Dimensions of ZE   :", dim(ZE), "\n")
cat("Dimensions of ZEZE :", dim(ZEZE), "\n")

ZEZE[1:5, 1:5]
```

## 9. Build the weather-derived environmental kernel (K_W stored as `ZW`)

The weather kernel uses the climate outputs created in the `climate_data` tutorial. The updated climate workflow now exports two files:

1. `environmental_covariates.csv`: one row per year with aggregated environmental summaries.
2. `environmental_covariates_expanded.csv`: daily records linked to those annual summaries.

This script uses **both** files:

- The annual file provides the summary covariates.
- The expanded file provides the daily weather trajectories used to describe the within-season climatic signature of each year.

### 9.1 Load climate files

```{r load-climate-files}
annual_ec_file <- "output/climate_results/environmental_covariates.csv"
expanded_ec_file <- "output/climate_results/environmental_covariates_expanded.csv"

if (!file.exists(annual_ec_file)) {
  stop("The file 'output/climate_results/environmental_covariates.csv' was not found. Run climate_data.Rmd first.")
}

if (!file.exists(expanded_ec_file)) {
  stop("The file 'output/climate_results/environmental_covariates_expanded.csv' was not found. Run climate_data.Rmd first.")
}

ECs_annual <- read.csv(annual_ec_file)
ECs_expanded <- read.csv(expanded_ec_file)

annual_cols <- c(
  "TMAX_AVG", "TMIN_AVG", "GDD_CUM", "HEAT_STRESS_DAYS",
  "PRECTOT", "DRY_DAYS", "RH_AVG", "WS2M_AVG", "RAD_CUM",
  "T_AVG", "DTR_AVG", "COLD_STRESS_DAYS", "RAINY_DAYS",
  "PREC_INTENSITY", "LOW_RH_DAYS", "VPD_AVG", "VPD_STRESS_DAYS"
)

# Keep only true daily weather variables. Exclude identifiers, date encodings,
# coordinate fields, and any annual summary columns that may also be present in
# the expanded file.
daily_id_cols <- c(
  "year", "Year", "YEAR",
  "date", "Date",
  "YYYYMMDD", "MMDD", "DOY",
  "month", "Month", "MM",
  "day", "DD",
  "LON", "LAT"
)
candidate_daily_cols <- setdiff(
  colnames(ECs_expanded),
  c(daily_id_cols, annual_cols)
)

daily_cols <- candidate_daily_cols[
  vapply(ECs_expanded[, candidate_daily_cols, drop = FALSE], is.numeric, logical(1))
]

excluded_daily_cols <- setdiff(candidate_daily_cols, daily_cols)

if (length(daily_cols) == 0) {
  stop("No numeric daily climate variables were found in environmental_covariates_expanded.csv.")
}

cat("Number of daily climate variables used in ZW:", length(daily_cols), "\n")
if (length(excluded_daily_cols) > 0) {
  cat("Excluded non-climatic or non-numeric columns:", paste(excluded_daily_cols, collapse = ", "), "\n")
}

ECs_annual <- ECs_annual %>%
  mutate(year = as.integer(year)) %>%
  dplyr::select(year, all_of(annual_cols))

ECs_expanded <- ECs_expanded %>%
  mutate(
    year = as.integer(year),
    date = as.character(date)
  )
```

### 9.2 Standardize annual summaries and daily trajectories

The purpose of this step is to place heterogeneous environmental variables on a comparable scale before kernel construction.

```{r standardize-climate-data}
annual_scaled <- ECs_annual %>%
  dplyr::select(all_of(annual_cols)) %>%
  scale(center = TRUE, scale = TRUE) %>%
  as.data.frame()

annual_scaled <- bind_cols(
  year = ECs_annual$year,
  annual_scaled
)

daily_scaled <- ECs_expanded %>%
  dplyr::select(year, date, all_of(daily_cols)) %>%
  group_by(year) %>%
  mutate(across(all_of(daily_cols), ~ as.numeric(scale(.x, center = TRUE, scale = TRUE)))) %>%
  ungroup()
```

### 9.3 Create year-specific climatic signatures

For each year, we transform the daily climatic trajectories into a single wide-row signature. This gives each year a reproducible vector representing its within-season weather profile.

```{r create-year-signatures}
daily_signature_2011 <- daily_scaled %>%
  filter(year == 2011) %>%
  dplyr::select(date, all_of(daily_cols)) %>%
  pivot_longer(cols = -date, names_to = "variable", values_to = "value") %>%
  mutate(signature_col = paste(variable, date, sep = "__")) %>%
  dplyr::select(signature_col, value) %>%
  pivot_wider(names_from = signature_col, values_from = value)

daily_signature_2012 <- daily_scaled %>%
  filter(year == 2012) %>%
  dplyr::select(date, all_of(daily_cols)) %>%
  pivot_longer(cols = -date, names_to = "variable", values_to = "value") %>%
  mutate(signature_col = paste(variable, date, sep = "__")) %>%
  dplyr::select(signature_col, value) %>%
  pivot_wider(names_from = signature_col, values_from = value)

annual_signature_2011 <- annual_scaled %>%
  filter(year == 2011) %>%
  dplyr::select(-year)

annual_signature_2012 <- annual_scaled %>%
  filter(year == 2012) %>%
  dplyr::select(-year)
```

### 9.4 Build the observation-level W matrix

Each observation inherits the climatic signature of the year associated with its environment. The 2011 environments receive the 2011 daily and annual signatures; the 2012 environments receive the 2012 signatures. Missing year-specific blocks are filled with `NA` and later replaced with zero so that the final matrix can be used in a cross-product.

```{r build-w}
build_weather_block <- function(obs_ids, daily_sig, annual_sig, other_daily_sig = NULL) {
  main_daily <- as.data.frame(daily_sig[rep(1, length(obs_ids)), , drop = FALSE])
  rownames(main_daily) <- obs_ids

  if (is.null(other_daily_sig)) {
    other_daily <- NULL
  } else {
    other_daily <- as.data.frame(
      matrix(
        NA_real_,
        nrow = length(obs_ids),
        ncol = ncol(other_daily_sig),
        dimnames = list(obs_ids, colnames(other_daily_sig))
      )
    )
  }

  annual_block <- as.data.frame(annual_sig[rep(1, length(obs_ids)), , drop = FALSE])
  rownames(annual_block) <- obs_ids

  if (is.null(other_daily)) {
    out <- cbind(main_daily, annual_block)
  } else {
    out <- cbind(main_daily, other_daily, annual_block)
  }

  out <- as.data.frame(out)
  rownames(out) <- obs_ids
  out
}

ids_a <- a$ObsID
ids_b <- b$ObsID
ids_c <- c$ObsID
ids_d <- d$ObsID

ECs.a <- build_weather_block(ids_a, daily_signature_2011, annual_signature_2011, daily_signature_2012)
ECs.b <- build_weather_block(ids_b, daily_signature_2011, annual_signature_2011, daily_signature_2012)
ECs.c <- build_weather_block(ids_c, daily_signature_2012, annual_signature_2012, daily_signature_2011)
ECs.d <- build_weather_block(ids_d, daily_signature_2012, annual_signature_2012, daily_signature_2011)

W <- rbind(ECs.a, ECs.b, ECs.c, ECs.d)
W <- as.data.frame(W)
rownames(W) <- c(ids_a, ids_b, ids_c, ids_d)

W <- W[Pheno$ObsID, , drop = FALSE]
W[is.na(W)] <- 0

ZW <- tcrossprod(as.matrix(W)) / ncol(W)
rownames(ZW) <- rownames(W)
colnames(ZW) <- rownames(W)

print(paste("Rows in W :", nrow(W)))
print(paste("Rows in ZG:", nrow(ZG)))
print(paste("ZW and ZG aligned:", identical(rownames(ZW), rownames(ZG))))
print(paste("ZW and ZP aligned:", identical(rownames(ZW), rownames(ZP))))
```

## 10. Check kernel dimensions and alignment

Before building interactions and nonlinear kernels, we summarize the main linear kernels. This table is useful both for quality control and for tutorial purposes.

```{r kernel-summary-linear}
reference_ids <- Pheno$ObsID

kernel_inventory_linear <- bind_rows(
  check_kernel_alignment(ZG, reference_ids) %>% mutate(kernel = "ZG (genomic)"),
  check_kernel_alignment(ZP, reference_ids) %>% mutate(kernel = "ZP (phenomic/NIR)"),
  check_kernel_alignment(ZEZE, reference_ids) %>% mutate(kernel = "ZEZE (categorical environment)"),
  check_kernel_alignment(ZW, reference_ids) %>% mutate(kernel = "ZW (weather-derived)")
) %>%
  dplyr::select(kernel, everything())

knitr::kable(
  kernel_inventory_linear,
  caption = "Diagnostic summary of the main linear kernels."
)
```

### 10.1 Spot-check selected observations

We inspect the same pedigree across matrices to verify that the stacked observation ordering remains coherent.

```{r spot-check}
Geno.all[str_detect(rownames(Geno.all), "A188"), 1:5]
W[str_detect(rownames(W), "A188"), 1:5]
NIR.all[str_detect(rownames(NIR.all), "A188"), 1:5]
```

### 10.2 Global alignment tests

All checks below should return `TRUE`.

```{r alignment-tests}
all(rownames(ZW) == colnames(ZW))
all(rownames(ZG) == colnames(ZG))
all(rownames(ZP) == colnames(ZP))

all(rownames(ZW) == rownames(ZG))
all(rownames(ZW) == rownames(ZP))

all(colnames(ZW) == colnames(ZG))
all(colnames(ZW) == colnames(ZP))
```

### 10.3 Diagnostic figures will be generated after all kernel families are built

We postpone the final diagnostic figures until the Gaussian and arc-cosine kernels are available. This allows us to compare the linear and nonlinear representations within the genomic and phenomic spaces separately, which is more informative for the article than mixing weather and phenomic matrices in a single panel.


## 11. Build interaction kernels

The interaction kernels are obtained through the Hadamard product (element-wise multiplication). This operation restricts genomic or phenomic similarity to the environmental context represented by `ZEZE` or `ZW`.

```{r build-interactions}
ZGZE <- ZG * ZEZE
ZPZE <- ZP * ZEZE
ZGZW <- ZG * ZW
ZPZW <- ZP * ZW

ZGZE[1:5, 1:5]
ZPZE[1:5, 1:5]
ZGZW[1:5, 1:5]
ZPZW[1:5, 1:5]

saveRDS(ZGZE, "output/Matrizes/ZGZE.rds")
saveRDS(ZPZE, "output/Matrizes/ZPZE.rds")
```

## 12. Build nonlinear Gaussian kernels

The linear kernels above capture additive similarity. To explore nonlinear relationships, we also build Gaussian kernels for the genomic and phenomic spaces.

### 12.1 Optimization function for the Gaussian kernel

The function below optimizes the Gaussian bandwidth parameter by maximizing the marginal likelihood.

```{r gk-functions}
margh.fun <- function(theta, y, D, q, nu = 0.0001, Sc = 0.0001,
                      nuh = NULL, Sch = NULL, prior = NULL) {
  h <- theta[1]
  phi <- theta[2]

  Kh <- exp(-h * D / q)
  eigenKh <- eigen(Kh)

  nr <- length(which(eigenKh$values > 1e-10))
  Uh <- eigenKh$vectors[, 1:nr, drop = FALSE]
  Sh <- eigenKh$values[1:nr]

  d <- t(Uh) %*% scale(y, scale = FALSE)

  lden <- -0.5 * sum(log(1 + phi * Sh)) -
    (nu + nr - 1) / 2 * log(Sc + sum(d^2 / (1 + phi * Sh)))

  if (!is.null(prior)) {
    lprior <- dgamma(h, nuh, Sch, log = TRUE)
  } else {
    lprior <- 0
  }

  return(-(lden + lprior))
}
```

### 12.2 Genomic Gaussian kernel (`GGK`)

We first compute genomic distances among unique genotypes, then project them into the observation space.

```{r ggk}
DG <- (as.matrix(dist(Geno)))^2
q05G <- quantile(DG, 0.05)

IDs <- as.character(Pheno$Pedigree)
DG <- DG[rownames(DG) %in% IDs, rownames(DG) %in% IDs, drop = FALSE]

IDs <- factor(IDs, levels = rownames(DG))
Z <- as.matrix(model.matrix(~ IDs - 1))
rownames(Z) <- Pheno$ObsID

ZD <- tcrossprod(Z, DG)
DG_proj <- tcrossprod(ZD, Z)

rownames(DG_proj) <- Pheno$ObsID
colnames(DG_proj) <- Pheno$ObsID
```

```{r ggk-optimize, eval=FALSE}
solG <- optim(
  c(1, 1),
  margh.fun,
  y = Pheno$GY,
  D = DG_proj,
  q = q05G,
  method = "L-BFGS-B",
  lower = c(0.05, 0.05),
  upper = c(6, 30)
)
```

```{r ggk-read-solution, include=FALSE}
solG <- readRDS("output/Matrizes/solG.RDS")
```

```{r ggk-final}
hG <- solG$par[1]

GGK <- exp(-hG * DG_proj / q05G)
rownames(GGK) <- Pheno$ObsID
colnames(GGK) <- Pheno$ObsID

dim(GGK)
GGK[1:5, 1:5]
```

### 12.3 Phenomic Gaussian kernel (`PGK`)

The phenomic Gaussian kernel uses Euclidean distances derived from the standardized NIR matrix.

```{r pgk}
DP <- (as.matrix(dist(NIR.all)))^2
q05P <- quantile(DP, 0.05)

IDs <- rownames(NIR.all)
DP <- DP[rownames(DP) %in% IDs, rownames(DP) %in% IDs, drop = FALSE]

Z <- as.matrix(model.matrix(~ IDs - 1))
rownames(Z) <- IDs

Z0 <- tcrossprod(Z, Z)
ZD <- tcrossprod(Z0, DP)
DP_proj <- tcrossprod(ZD, Z0)

rownames(DP_proj) <- IDs
colnames(DP_proj) <- IDs
```

```{r pgk-optimize, eval=FALSE}
solP <- optim(
  c(1, 1),
  margh.fun,
  y = Pheno$GY,
  D = DP_proj,
  q = q05P,
  method = "L-BFGS-B",
  lower = c(0.05, 0.05),
  upper = c(6, 30)
)
```

```{r pgk-read-solution, include=FALSE}
solP <- readRDS("output/Matrizes/solP.RDS")
```

```{r pgk-final}
hP <- solP$par[1]

PGK <- exp(-hP * DP_proj / q05P)
rownames(PGK) <- rownames(NIR.all)
colnames(PGK) <- rownames(NIR.all)

dim(PGK)
PGK[1:5, 1:5]
```

### 12.4 Gaussian interaction kernels

```{r interactions-gk}
GGKE <- GGK * ZEZE
PGKE <- PGK * ZEZE
GGKW <- GGK * ZW
PGKW <- PGK * ZW
```

## 13. Build nonlinear arc-cosine kernels

Arc-cosine kernels offer a second nonlinear representation. They are inspired by neural-network geometry and can capture hierarchical nonlinear patterns through recursive transformations.

### 13.1 Helper functions for arc-cosine kernels

```{r ak-functions}
AK1.fun <- function(X) {
  cosalfa <- cor(t(X))
  cosalfa[cosalfa > 1] <- 1
  cosalfa[cosalfa < -1] <- -1
  angulo <- acos(cosalfa)
  mag <- sqrt(apply(X, 1, function(x) crossprod(x)))
  sxy <- tcrossprod(mag)
  AK1 <- (1 / pi) * sxy * (sin(angulo) + (pi - angulo) * cosalfa)
  AK1 <- AK1 / median(AK1, na.rm = TRUE)
  colnames(AK1) <- rownames(X)
  rownames(AK1) <- rownames(X)
  return(AK1)
}

marg.AK <- function(y, AK1, ml) {
  lden.fun <- function(phi, nr, Uh, Sh, d) {
    lden <- -0.5 * sum(log(1 + phi * Sh)) -
      (nr - 1) / 2 * log(sum(d^2 / (1 + phi * Sh)))
    return(-lden)
  }

  vero <- function(y, GC) {
    Kh <- GC
    eigenKh <- eigen(Kh)
    nr <- length(which(eigenKh$values > 1e-10))
    Uh <- eigenKh$vectors[, 1:nr, drop = FALSE]
    Sh <- eigenKh$values[1:nr]
    d <- t(Uh) %*% scale(y, scale = FALSE)
    sol <- optimize(lden.fun, nr = nr, Uh = Uh, Sh = Sh, d = d,
                    lower = 0.0005, upper = 200)
    phi <- sol[[1]]
    log.vero <- 0.5 * sum(log(1 + phi * Sh)) -
      (nr - 1) / 2 * log(sum(d^2 / (1 + phi * Sh)))
    return(log.vero)
  }

  GC <- AK1
  l <- 1
  GC2 <- GC
  vero1 <- vero(y = y, GC = GC2)
  m <- 0

  while (m == 0 && (l < ml)) {
    l <- l + 1
    GC <- AK.fun(AK1 = GC2, nl = 1)
    GC2 <- GC
    vero2 <- vero(y = y, GC = GC2)
    if (vero2 < vero1) m <- 1
    vero1 <- vero2
  }

  return(l - 1)
}

AK.fun <- function(AK1, nl) {
  n <- nrow(AK1)
  AK <- AK1
  for (l in 1:nl) {
    Aux <- tcrossprod(diag(AK))
    cosalfa <- AK * (Aux^(-1 / 2))
    cosa <- as.vector(cosalfa)
    cosa[cosa > 1] <- 1
    cosa[cosa < -1] <- -1
    angulo <- acos(cosa)
    angulo <- matrix(angulo, n, n)
    AK <- (1 / pi) * (Aux^(1 / 2)) * (sin(angulo) + (pi - angulo) * cos(angulo))
  }
  AK <- AK / median(AK, na.rm = TRUE)
  rownames(AK) <- rownames(AK1)
  colnames(AK) <- colnames(AK1)
  return(AK)
}
```

### 13.2 Genomic arc-cosine kernel (`GAK`)

```{r gak}
GAK1 <- AK1.fun(Geno)
dim(GAK1)
GAK1[1:5, 1:5]

IDs <- as.character(Pheno$Pedigree)
GAK1 <- GAK1[rownames(GAK1) %in% IDs, rownames(GAK1) %in% IDs, drop = FALSE]

IDs <- factor(IDs, levels = rownames(GAK1))
Z <- as.matrix(model.matrix(~ IDs - 1))
rownames(Z) <- Pheno$ObsID
G <- tcrossprod(tcrossprod(Z, GAK1), Z)

Gl <- marg.AK(y = Pheno$GY, AK1 = G, ml = 50)

IDs.L <- factor(Pheno$Pedigree, levels = rownames(GAK1))
Z.L <- as.matrix(model.matrix(~ IDs.L - 1))
rownames(Z.L) <- Pheno$ObsID
G.L <- tcrossprod(tcrossprod(Z.L, GAK1), Z.L)

GAK <- AK.fun(AK1 = G.L, nl = Gl)
rownames(GAK) <- Pheno$ObsID
colnames(GAK) <- Pheno$ObsID
```

```{r gak-check}
dim(GAK)
GAK[1:5, 1:5]
```

### 13.3 Phenomic arc-cosine kernel (`PAK`)

```{r pak}
PAK1 <- AK1.fun(NIR.all)
dim(PAK1)
PAK1[1:5, 1:5]

IDs <- rownames(NIR.all)
PAK1 <- PAK1[rownames(PAK1) %in% IDs, rownames(PAK1) %in% IDs, drop = FALSE]

Z <- as.matrix(model.matrix(~ IDs - 1))
rownames(Z) <- IDs
Z0 <- tcrossprod(Z, Z)
ZD <- tcrossprod(Z0, PAK1)
P <- tcrossprod(ZD, Z0)

Pl <- marg.AK(y = Pheno$GY, AK1 = P, ml = 50)

IDs.L <- rownames(NIR.all)
Z.L <- as.matrix(model.matrix(~ IDs.L - 1))
rownames(Z.L) <- IDs.L
Z0.L <- tcrossprod(Z.L, Z.L)
P.L <- tcrossprod(tcrossprod(Z0.L, PAK1), Z0.L)

PAK <- AK.fun(AK1 = P.L, nl = Pl)
rownames(PAK) <- rownames(NIR.all)
colnames(PAK) <- rownames(NIR.all)
```

```{r pak-check}
dim(PAK)
PAK[1:5, 1:5]
```

### 13.4 Arc-cosine interaction kernels

```{r interaction-ak}
GAKE <- GAK * ZEZE
PAKE <- PAK * ZEZE
GAKW <- GAK * ZW
PAKW <- PAK * ZW

GAKE[1:5, 1:5]
PAKE[1:5, 1:5]
GAKW[1:5, 1:5]
PAKW[1:5, 1:5]
```

## 14. Summarize nonlinear kernels

The table below provides a final integrity check for the nonlinear kernels before export.

```{r kernel-summary-nonlinear}
kernel_inventory_nonlinear <- bind_rows(
  check_kernel_alignment(GGK, reference_ids) %>% mutate(kernel = "GGK"),
  check_kernel_alignment(PGK, reference_ids) %>% mutate(kernel = "PGK"),
  check_kernel_alignment(GAK, reference_ids) %>% mutate(kernel = "GAK"),
  check_kernel_alignment(PAK, reference_ids) %>% mutate(kernel = "PAK"),
  check_kernel_alignment(GGKE, reference_ids) %>% mutate(kernel = "GGKE"),
  check_kernel_alignment(PGKE, reference_ids) %>% mutate(kernel = "PGKE"),
  check_kernel_alignment(GGKW, reference_ids) %>% mutate(kernel = "GGKW"),
  check_kernel_alignment(PGKW, reference_ids) %>% mutate(kernel = "PGKW"),
  check_kernel_alignment(GAKE, reference_ids) %>% mutate(kernel = "GAKE"),
  check_kernel_alignment(PAKE, reference_ids) %>% mutate(kernel = "PAKE"),
  check_kernel_alignment(GAKW, reference_ids) %>% mutate(kernel = "GAKW"),
  check_kernel_alignment(PAKW, reference_ids) %>% mutate(kernel = "PAKW")
) %>%
  dplyr::select(kernel, everything())

knitr::kable(
  kernel_inventory_nonlinear,
  caption = "Diagnostic summary of the nonlinear and interaction kernels."
)
```

## 15. Diagnostic figures for article reporting

To keep the visual narrative aligned with the manuscript, we generate a single integrated diagnostic figure with two rows and three columns. The first row contains the genomic kernels and the second row contains the phenomic kernels. This layout improves side-by-side comparison between linear, Gaussian, and arc-cosine structures while preserving a compact figure format for the article.

### 15.1 Integrated genomic and phenomic kernel figure

```{r kernel-diagnostic-figure-integrated, fig.width=12, fig.height=8, fig.cap="Figure 1. Integrated diagnostic view of the genomic and phenomic kernel families. The first row shows the genomic kernels: linear genomic kernel (ZG), Gaussian genomic kernel (GGK), and arc-cosine genomic kernel (GAK). The second row shows the phenomic kernels: linear phenomic kernel (ZP), Gaussian phenomic kernel (PGK), and arc-cosine phenomic kernel (PAK)."}
p_zg <- plot_kernel_excerpt(
  ZG,
  title = "ZG",
  subtitle = "Linear genomic kernel",
  n_show = 60
)

p_ggk <- plot_kernel_excerpt(
  GGK,
  title = "GGK",
  subtitle = "Gaussian genomic kernel",
  n_show = 60
)

p_gak <- plot_kernel_excerpt(
  GAK,
  title = "GAK",
  subtitle = "Arc-cosine genomic kernel",
  n_show = 60
)

p_zp <- plot_kernel_excerpt(
  ZP,
  title = "ZP",
  subtitle = "Linear phenomic kernel",
  n_show = 60
)

p_pgk <- plot_kernel_excerpt(
  PGK,
  title = "PGK",
  subtitle = "Gaussian phenomic kernel",
  n_show = 60
)

p_pak <- plot_kernel_excerpt(
  PAK,
  title = "PAK",
  subtitle = "Arc-cosine phenomic kernel",
  n_show = 60
)

kernel_integrated_plot <- patchwork::wrap_plots(
  p_zg, p_ggk, p_gak,
  p_zp, p_pgk, p_pak,
  ncol = 3
) +
  patchwork::plot_annotation(
    title = "Integrated diagnostic view of genomic and phenomic kernels",
    subtitle = "Top row: genomic kernels | Bottom row: phenomic kernels",
    theme = theme(
      plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
      plot.subtitle = element_text(size = 10.5, hjust = 0.5)
    )
  )

print(kernel_integrated_plot)

ggsave(
  filename = "output/Matrizes/figures/kernel_integrated_diagnostics.tiff",
  plot = kernel_integrated_plot,
  width = 14,
  height = 9,
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
```

## 16. Save outputs and reset the environment

The kernel objects created in this tutorial are computationally expensive to reproduce. For that reason, the final step saves them as `.rds` files and cleans the workspace so the next modeling script can start from a lighter memory state.

```{r save-and-clean, message=FALSE}
keep_names <- c(
  "ZE", "ZG", "ZP", "ZGZE", "ZPZE", "ZW", "ZGZW", "ZPZW",
  "GGK", "PGK", "GGKE", "PGKE", "GGKW", "PGKW",
  "GAK", "PAK", "GAKE", "PAKE", "GAKW", "PAKW",
  "Pheno", "Pedigree", "Pedigree_communs", "Geno.all", "NIR.all", "W"
)

current_objs <- ls(envir = .GlobalEnv)
to_keep <- intersect(current_objs, keep_names)
to_remove <- setdiff(current_objs, to_keep)

message("Objects kept (", length(to_keep), "): ", paste(to_keep, collapse = ", "))
message(
  "Objects removed (", length(to_remove), "): ",
  paste(head(to_remove, 40), collapse = ", "),
  if (length(to_remove) > 40) paste0(" ... (+", length(to_remove) - 40, " more)")
)

for (obj_name in to_keep) {
  saveRDS(get(obj_name), file = file.path("output/Matrizes", paste0(obj_name, ".rds")))
}

write.csv(
  kernel_inventory_linear,
  file = "output/Matrizes/kernel_inventory_linear.csv",
  row.names = FALSE
)

write.csv(
  kernel_inventory_nonlinear,
  file = "output/Matrizes/kernel_inventory_nonlinear.csv",
  row.names = FALSE
)

rm(list = to_remove, envir = .GlobalEnv)
gc()

rds_files <- list.files("output/Matrizes", pattern = "\\.rds$", full.names = TRUE)

for (file in rds_files) {
  obj_name <- tools::file_path_sans_ext(basename(file))
  assign(obj_name, readRDS(file), envir = .GlobalEnv)
}

message("Objects reloaded: ", paste(tools::file_path_sans_ext(basename(rds_files)), collapse = ", "))
```

## 16. Final remarks

This tutorial converts raw genomic, phenomic, and climatic information into a reproducible collection of aligned kernels ready for downstream prediction models.

From a scientific workflow perspective, this script plays a structural role in the project:

- it defines the observation order used by the modeling stage,
- it makes the environmental representation explicit,
- it stores the kernels as reusable computational artifacts,
- and it documents how the linear and nonlinear relationships were encoded.

In the next stage of the project, these kernels become the inputs for the genomic prediction models compared in the article.
