---
title: "Variance Components"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

This chunk sets the knitting options for the document so that tables, code, and figures are displayed consistently throughout the tutorial.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 12,
  fig.height = 7,
  dpi = 300
)
```

**Language / Idioma:** [English](variance_components.html) | [Português](variance_components_pt.html)

## 1. Introduction

This tutorial summarizes **variance components** for the predictive models used in this study. In the current workflow, this file is used mainly to **consolidate and interpret results that were already estimated previously**, although it still keeps the code structure needed to fit the models when that step must be rerun.

The goal is to quantify how much of the total phenotypic variance is attributed to the main effects and interaction terms included in each model. In practical terms, this script helps answer questions such as:

- How much variance is associated with the **categorical environment effect (E)**?
- How much is captured by the **genomic kernels**?
- How much is captured by the **phenomic/NIRS kernels (P)**?
- How much is associated with the **weather kernel (W)**, which is used mainly in models **M10-M18**?
- How relevant are the interaction terms such as **G × E**, **G × W**, **P × E**, and **P × W**?

This script is organized into four stages:

1. **Load kernels and phenotypic data** produced in `matrizes.Rmd`.
2. **Define the 18 variance-component models**.
3. **Read or rerun the BGLR outputs**, depending on whether the saved posterior summaries already exist.
4. **Process and visualize the variance decomposition** for Grain Yield (GY) and Kernel Weight (KW).

A high variance share for a given source should not be interpreted automatically as evidence of biological complementarity or better predictive performance. This point is especially important for **P**, because NIRS-based similarity may reflect strong physiological response to the environment in addition to any stable biological signal, and for **W**, because the weather kernel complements but does not replace the categorical environment effect.

## Inputs, outputs and next step

**Inputs of this step**
- matrices and kernels generated by `matrizes.Rmd`
- the reduced catalog of 18 models used in the project

**Outputs of this step**
- raw variance-component results
- processed variance-component summaries
- percentage tables and diagnostic figures

**Next step**
- this stage complements the structural interpretation of the models and should be read together with the predictive results generated later in `analysis_prediction.Rmd`.

## 2. R Environment and Required Inputs

First, we load the packages used to manipulate the data, fit the Bayesian models, and generate the final article-quality figures.

This chunk loads the packages needed for reading the saved artifacts, reshaping the results, plotting summaries, and running the heavy estimation step when required.

```{r load-packages}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(tibble)
library(ggplot2)
library(BGLR)
library(parallel)
library(doParallel)
library(foreach)
library(ggthemes)
library(scales)
library(fs)
library(grid)
library(knitr)
library(kableExtra)
```

### 2.1. File paths

All kernels and phenotypic objects are expected to be available in `output/Matrizes/`, which is the output directory created in `matrizes.Rmd`. The variance-component results generated here are stored in `output/variance_components/`.

This chunk defines the project directories used to find matrices, store outputs, and organize the saved variance-component files.

```{r define-paths}
kernel_dir <- "output/Matrizes"
results_dir <- "output/variance_components"
figure_dir <- file.path(results_dir, "figures")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
```

### 2.2. Input checks

Before reading the files, we check whether the expected `.rds` objects are present in the matrices output folder. This makes the tutorial easier to follow because the script stops immediately if a required object is missing.

This chunk checks that the required input files and directories exist before the script moves into model definition or result processing.

```{r input-checks}
required_files <- c(
  "ZG.rds", "ZP.rds", "ZE.rds", "ZEZE.rds", "ZW.rds",
  "ZGZE.rds", "ZPZE.rds", "ZGZW.rds", "ZPZW.rds",
  "GGK.rds", "PGK.rds", "GAK.rds", "PAK.rds",
  "GGKE.rds", "PGKE.rds", "GGKW.rds", "PGKW.rds",
  "GAKE.rds", "PAKE.rds", "GAKW.rds", "PAKW.rds",
  "Pheno.rds", "Pedigree.rds"
)

missing_files <- required_files[!file.exists(file.path(kernel_dir, required_files))]

if (length(missing_files) > 0) {
  stop(
    "The following required files were not found in output/Matrizes/: ",
    paste(missing_files, collapse = ", "),
    ". Run matrizes.Rmd first."
  )
}
```

### 2.3. Labels used in the final tables and figures

These vectors define how the variance components will appear in the final outputs. Keeping these labels visible in the main script helps the reader understand how internal model names are translated into article-ready labels.

This chunk defines the labels and naming conventions used later in tables, output files, and figures.

```{r labels-for-output}
component_display <- c(
  E = "Environment (E)",
  G = "Genomic linear (G)",
  P = "Phenomic linear (P)",
  W = "Weather kernel (W)",
  GE = "G × E",
  PE = "P × E",
  GW = "G × W",
  PW = "P × W",
  GGK = "Genomic Gaussian (GGK)",
  PGK = "Phenomic Gaussian (PGK)",
  GAK = "Genomic arc-cosine (GAK)",
  PAK = "Phenomic arc-cosine (PAK)",
  GGKE = "GGK × E",
  PGKE = "PGK × E",
  GAKE = "GAK × E",
  PAKE = "PAK × E",
  GGKW = "GGK × W",
  PGKW = "PGK × W",
  GAKW = "GAK × W",
  PAKW = "PAK × W",
  Residual = "Residual"
)

component_order <- c(
  "Environment (E)",
  "Genomic linear (G)",
  "Phenomic linear (P)",
  "Weather kernel (W)",
  "G × E",
  "P × E",
  "G × W",
  "P × W",
  "Genomic Gaussian (GGK)",
  "Phenomic Gaussian (PGK)",
  "Genomic arc-cosine (GAK)",
  "Phenomic arc-cosine (PAK)",
  "GGK × E",
  "PGK × E",
  "GAK × E",
  "PAK × E",
  "GGK × W",
  "PGK × W",
  "GAK × W",
  "PAKW × W",
  "Residual"
)

component_order[component_order == "PAKW × W"] <- "PAK × W"
```

## 3. Loading the Kernels and Phenotypic Objects

This section loads the matrices generated in `matrizes.Rmd`. These objects are the core inputs for the variance-component analysis.

This chunk reads the matrices and kernels generated in the previous module so the variance-component models use the same aligned inputs as the prediction stage.

```{r load-matrices}
ZG    <- readRDS(file.path(kernel_dir, "ZG.rds"))
ZP    <- readRDS(file.path(kernel_dir, "ZP.rds"))
ZE    <- readRDS(file.path(kernel_dir, "ZE.rds"))
ZEZE  <- readRDS(file.path(kernel_dir, "ZEZE.rds"))
ZW    <- readRDS(file.path(kernel_dir, "ZW.rds"))

ZGZE  <- readRDS(file.path(kernel_dir, "ZGZE.rds"))
ZPZE  <- readRDS(file.path(kernel_dir, "ZPZE.rds"))
ZGZW  <- readRDS(file.path(kernel_dir, "ZGZW.rds"))
ZPZW  <- readRDS(file.path(kernel_dir, "ZPZW.rds"))

GGK   <- readRDS(file.path(kernel_dir, "GGK.rds"))
PGK   <- readRDS(file.path(kernel_dir, "PGK.rds"))
GAK   <- readRDS(file.path(kernel_dir, "GAK.rds"))
PAK   <- readRDS(file.path(kernel_dir, "PAK.rds"))

GGKE  <- readRDS(file.path(kernel_dir, "GGKE.rds"))
PGKE  <- readRDS(file.path(kernel_dir, "PGKE.rds"))
GGKW  <- readRDS(file.path(kernel_dir, "GGKW.rds"))
PGKW  <- readRDS(file.path(kernel_dir, "PGKW.rds"))

GAKE  <- readRDS(file.path(kernel_dir, "GAKE.rds"))
PAKE  <- readRDS(file.path(kernel_dir, "PAKE.rds"))
GAKW  <- readRDS(file.path(kernel_dir, "GAKW.rds"))
PAKW  <- readRDS(file.path(kernel_dir, "PAKW.rds"))

Pheno    <- readRDS(file.path(kernel_dir, "Pheno.rds"))
Pedigree <- readRDS(file.path(kernel_dir, "Pedigree.rds"))

n_obs <- nrow(Pheno)
cat("Number of phenotypic observations:", n_obs, "\n")
cat("Number of pedigrees:", length(Pedigree), "\n")
```

### 3.1. Alignment diagnostics

Before fitting the models, it is important to verify that the kernels are square and aligned to the same observation order used in `Pheno`. In a tutorial context, this is an important check because many downstream errors are caused by mismatched row and column orders.

This chunk verifies that the loaded kernels are compatible in dimension and ordering before they are combined in the Bayesian models.

```{r kernel-alignment-check}
reference_ids <- Pheno$ObsID

ZE <- ZE[match(reference_ids, rownames(ZE)), , drop = FALSE]

ZG   <- ZG[reference_ids, reference_ids, drop = FALSE]
ZP   <- ZP[reference_ids, reference_ids, drop = FALSE]
ZEZE <- ZEZE[reference_ids, reference_ids, drop = FALSE]
ZW   <- ZW[reference_ids, reference_ids, drop = FALSE]
ZGZE <- ZGZE[reference_ids, reference_ids, drop = FALSE]
ZPZE <- ZPZE[reference_ids, reference_ids, drop = FALSE]
ZGZW <- ZGZW[reference_ids, reference_ids, drop = FALSE]
ZPZW <- ZPZW[reference_ids, reference_ids, drop = FALSE]
GGK  <- GGK[reference_ids, reference_ids, drop = FALSE]
PGK  <- PGK[reference_ids, reference_ids, drop = FALSE]
GAK  <- GAK[reference_ids, reference_ids, drop = FALSE]
PAK  <- PAK[reference_ids, reference_ids, drop = FALSE]
GGKE <- GGKE[reference_ids, reference_ids, drop = FALSE]
PGKE <- PGKE[reference_ids, reference_ids, drop = FALSE]
GGKW <- GGKW[reference_ids, reference_ids, drop = FALSE]
PGKW <- PGKW[reference_ids, reference_ids, drop = FALSE]
GAKE <- GAKE[reference_ids, reference_ids, drop = FALSE]
PAKE <- PAKE[reference_ids, reference_ids, drop = FALSE]
GAKW <- GAKW[reference_ids, reference_ids, drop = FALSE]
PAKW <- PAKW[reference_ids, reference_ids, drop = FALSE]

kernel_inventory <- tibble(
  kernel = c("ZG", "ZP", "ZEZE", "ZW", "GGK", "PGK", "GAK", "PAK"),
  n_rows = c(nrow(ZG), nrow(ZP), nrow(ZEZE), nrow(ZW), nrow(GGK), nrow(PGK), nrow(GAK), nrow(PAK)),
  n_cols = c(ncol(ZG), ncol(ZP), ncol(ZEZE), ncol(ZW), ncol(GGK), ncol(PGK), ncol(GAK), ncol(PAK)),
  is_square = c(
    nrow(ZG) == ncol(ZG),
    nrow(ZP) == ncol(ZP),
    nrow(ZEZE) == ncol(ZEZE),
    nrow(ZW) == ncol(ZW),
    nrow(GGK) == ncol(GGK),
    nrow(PGK) == ncol(PGK),
    nrow(GAK) == ncol(GAK),
    nrow(PAK) == ncol(PAK)
  ),
  symmetric_names = c(
    identical(rownames(ZG), colnames(ZG)),
    identical(rownames(ZP), colnames(ZP)),
    identical(rownames(ZEZE), colnames(ZEZE)),
    identical(rownames(ZW), colnames(ZW)),
    identical(rownames(GGK), colnames(GGK)),
    identical(rownames(PGK), colnames(PGK)),
    identical(rownames(GAK), colnames(GAK)),
    identical(rownames(PAK), colnames(PAK))
  ),
  aligned_with_reference = c(
    identical(rownames(ZG), reference_ids),
    identical(rownames(ZP), reference_ids),
    identical(rownames(ZEZE), reference_ids),
    identical(rownames(ZW), reference_ids),
    identical(rownames(GGK), reference_ids),
    identical(rownames(PGK), reference_ids),
    identical(rownames(GAK), reference_ids),
    identical(rownames(PAK), reference_ids)
  )
)

kernel_inventory %>%
  dplyr::mutate(
    n_rows = format(n_rows, big.mark = ","),
    n_cols = format(n_cols, big.mark = ",")
  ) %>%
  knitr::kable(
    caption = "Diagnostic summary of the main kernels used in the variance-component analysis after explicit alignment to the phenotypic observation order.",
    align = "lcccccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "260px")
```

## 4. Definition of the 18 Predictive Models

Each model is defined as a list in the format required by the `ETA` argument of `BGLR`. The environment term (`E`) is included as a fixed-effect design matrix with `model = "BRR"`, while the kernels are fitted as random effects with `model = "RKHS"`.

This chunk defines the 18 reduced Bayesian multi-kernel models in the form required for the variance-component workflow.

```{r define-models}
models <- list()

models[[1]] <- list(
  E  = list(X = ZE, model = "BRR"),
  G  = list(K = ZG, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS")
)

models[[2]] <- list(
  E  = list(X = ZE, model = "BRR"),
  P  = list(K = ZP, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS")
)

models[[3]] <- list(
  E    = list(X = ZE, model = "BRR"),
  GGK  = list(K = GGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS")
)

models[[4]] <- list(
  E    = list(X = ZE, model = "BRR"),
  PGK  = list(K = PGK, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS")
)

models[[5]] <- list(
  E    = list(X = ZE, model = "BRR"),
  GAK  = list(K = GAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS")
)

models[[6]] <- list(
  E    = list(X = ZE, model = "BRR"),
  PAK  = list(K = PAK, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS")
)

models[[7]] <- list(
  E  = list(X = ZE, model = "BRR"),
  G  = list(K = ZG, model = "RKHS"),
  P  = list(K = ZP, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS")
)

models[[8]] <- list(
  E    = list(X = ZE, model = "BRR"),
  GGK  = list(K = GGK, model = "RKHS"),
  PGK  = list(K = PGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS")
)

models[[9]] <- list(
  E    = list(X = ZE, model = "BRR"),
  GAK  = list(K = GAK, model = "RKHS"),
  PAK  = list(K = PAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS")
)

models[[10]] <- list(
  E  = list(X = ZE, model = "BRR"),
  W  = list(K = ZW, model = "RKHS"),
  G  = list(K = ZG, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS"),
  GW = list(K = ZGZW, model = "RKHS")
)

models[[11]] <- list(
  E  = list(X = ZE, model = "BRR"),
  W  = list(K = ZW, model = "RKHS"),
  P  = list(K = ZP, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS"),
  PW = list(K = ZPZW, model = "RKHS")
)

models[[12]] <- list(
  E  = list(X = ZE, model = "BRR"),
  W  = list(K = ZW, model = "RKHS"),
  G  = list(K = ZG, model = "RKHS"),
  P  = list(K = ZP, model = "RKHS"),
  GE = list(K = ZGZE, model = "RKHS"),
  PE = list(K = ZPZE, model = "RKHS"),
  GW = list(K = ZGZW, model = "RKHS"),
  PW = list(K = ZPZW, model = "RKHS")
)

models[[13]] <- list(
  E    = list(X = ZE, model = "BRR"),
  W    = list(K = ZW, model = "RKHS"),
  GGK  = list(K = GGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS"),
  GGKW = list(K = GGKW, model = "RKHS")
)

models[[14]] <- list(
  E    = list(X = ZE, model = "BRR"),
  W    = list(K = ZW, model = "RKHS"),
  PGK  = list(K = PGK, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS"),
  PGKW = list(K = PGKW, model = "RKHS")
)

models[[15]] <- list(
  E    = list(X = ZE, model = "BRR"),
  W    = list(K = ZW, model = "RKHS"),
  GGK  = list(K = GGK, model = "RKHS"),
  PGK  = list(K = PGK, model = "RKHS"),
  GGKE = list(K = GGKE, model = "RKHS"),
  PGKE = list(K = PGKE, model = "RKHS"),
  GGKW = list(K = GGKW, model = "RKHS"),
  PGKW = list(K = PGKW, model = "RKHS")
)

models[[16]] <- list(
  E    = list(X = ZE, model = "BRR"),
  W    = list(K = ZW, model = "RKHS"),
  GAK  = list(K = GAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS"),
  GAKW = list(K = GAKW, model = "RKHS")
)

models[[17]] <- list(
  E    = list(X = ZE, model = "BRR"),
  W    = list(K = ZW, model = "RKHS"),
  PAK  = list(K = PAK, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS"),
  PAKW = list(K = PAKW, model = "RKHS")
)

models[[18]] <- list(
  E    = list(X = ZE, model = "BRR"),
  W    = list(K = ZW, model = "RKHS"),
  GAK  = list(K = GAK, model = "RKHS"),
  PAK  = list(K = PAK, model = "RKHS"),
  GAKE = list(K = GAKE, model = "RKHS"),
  PAKE = list(K = PAKE, model = "RKHS"),
  GAKW = list(K = GAKW, model = "RKHS"),
  PAKW = list(K = PAKW, model = "RKHS")
)

names(models) <- sprintf("Eta%02d", seq_along(models))
```

### 4.1. Model catalog

This table is useful both for internal checks and for linking figure labels to the formal description of each model.

This chunk builds a compact model catalog that links each model identifier to its category and descriptive label for later summaries.

```{r model-catalog}
model_info <- tibble(
  ModelID = 1:18,
  Model = sprintf("M%02d", 1:18),
  ETA = names(models),
  Category = c(
    rep("Single-source + E", 6),
    rep("Integrated G/P + E", 3),
    rep("Weather-augmented", 9)
  ),
  Description = c(
    "E + G + GE",
    "E + P + PE",
    "E + GGK + GGKE",
    "E + PGK + PGKE",
    "E + GAK + GAKE",
    "E + PAK + PAKE",
    "E + G + P + GE + PE",
    "E + GGK + PGK + GGKE + PGKE",
    "E + GAK + PAK + GAKE + PAKE",
    "E + W + G + GE + GW",
    "E + W + P + PE + PW",
    "E + W + G + P + GE + PE + GW + PW",
    "E + W + GGK + GGKE + GGKW",
    "E + W + PGK + PGKE + PGKW",
    "E + W + GGK + PGK + GGKE + PGKE + GGKW + PGKW",
    "E + W + GAK + GAKE + GAKW",
    "E + W + PAK + PAKE + PAKW",
    "E + W + GAK + PAK + GAKE + PAKE + GAKW + PAKW"
  )
)

readr::write_csv(model_info, file.path(results_dir, "model_catalog.csv"))

model_info %>%
  knitr::kable(
    caption = "Catalog of the 18 models evaluated in the variance-component analysis.",
    align = "cll"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "340px")
```

## 5. MCMC Settings

These values define the Gibbs sampler used by `BGLR`. Adjust them here if a longer chain is required for convergence diagnostics or sensitivity analyses.

This chunk defines the MCMC settings and execution controls that regulate the estimation stage when the heavy computation is turned on.

```{r mcmc-parameters}
nIter   <- 10000
burnIn  <- 1000
thin    <- 10
num_rep <- 20
traits  <- c("GY", "KW")
```

## 6. Running BGLR for Each Model

This is the computationally expensive step. The chunk is kept with `eval=FALSE` so that the tutorial can be rendered without rerunning all Bayesian models every time. When you need updated variance components, change `eval` to `TRUE` and run the chunk.

This chunk contains the computationally expensive parallel estimation stage that fits the variance-component models and saves the raw posterior outputs.

```{r parallel-run, eval=FALSE}
numCores <- parallel::detectCores()
useCores <- max(1, numCores - 1)
cl <- parallel::makeCluster(useCores)
doParallel::registerDoParallel(cl)

combos <- expand.grid(
  Model = 1:18,
  REP = 1:num_rep,
  Trait = traits,
  stringsAsFactors = FALSE
)

results_list <- foreach(
  i = 1:nrow(combos),
  .packages = c("BGLR", "dplyr", "tibble")
) %dopar% {
  Model <- combos$Model[i]
  rep_num <- combos$REP[i]
  Trait <- combos$Trait[i]

  pheno <- data.frame(Pheno)
  y_data <- as.numeric(pheno[[Trait]])

  out_dir <- file.path(results_dir, paste0("rep_", rep_num), sprintf("Eta%02d", Model))
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  save_prefix <- file.path(out_dir, Trait)

  start_time <- Sys.time()
  fit <- BGLR(
    y = y_data,
    ETA = models[[Model]],
    nIter = nIter,
    burnIn = burnIn,
    thin = thin,
    saveAt = save_prefix
  )
  end_time <- Sys.time()

  variance_samples <- list()

  if (!is.null(fit$ETA)) {
    effect_names <- names(fit$ETA)

    if (is.null(effect_names)) {
      effect_names <- paste0("Effect_", seq_along(fit$ETA))
    }

    for (j in seq_along(fit$ETA)) {
      parameter_names <- names(fit$ETA[[j]])
      var_parameter <- parameter_names[grepl("^var", parameter_names)]

      if (length(var_parameter) > 0) {
        variance_samples[[effect_names[j]]] <- fit$ETA[[j]][[var_parameter[1]]]
      }
    }
  }

  metrics_object <- list(
    varE_samples = fit$varE,
    variances_samples = variance_samples,
    DIC = fit$fit$DIC,
    runtime_sec = as.numeric(difftime(end_time, start_time, units = "secs"))
  )

  saveRDS(metrics_object, paste0(save_prefix, "_metrics.rds"))

  result_table <- tibble(
    Component = "Residual",
    MeanVar = mean(metrics_object$varE_samples, na.rm = TRUE)
  )

  if (length(metrics_object$variances_samples) > 0) {
    for (nm in names(metrics_object$variances_samples)) {
      result_table <- bind_rows(
        result_table,
        tibble(
          Component = nm,
          MeanVar = mean(metrics_object$variances_samples[[nm]], na.rm = TRUE)
        )
      )
    }
  }

  rm(fit)
  gc()

  result_table %>%
    mutate(Model = Model, Rep = rep_num, Trait = Trait)
}

stopCluster(cl)

all_variance <- bind_rows(results_list)
write_csv(all_variance, file.path(results_dir, "variance_components_raw.csv"))
saveRDS(all_variance, file.path(results_dir, "variance_components_raw.rds"))
```

## 7. Consolidating Saved Posterior Summaries

In the current version of the workflow, this is the section used most often. Instead of refitting all models every time, the script reads the previously saved `_metrics.rds` files and consolidates them into one object. This keeps the tutorial practical and reproducible, while making it clear that this page is mainly a **result-consolidation step** unless the user intentionally reruns the full BGLR estimation stage.

This chunk consolidates the saved outputs produced by the estimation stage into a single structure that can be processed more easily later.

```{r consolidate-saved-results}
rds_files <- list.files(
  path = results_dir,
  pattern = "_(metrics|metricas)\\.rds$",
  recursive = TRUE,
  full.names = TRUE
)

if (length(rds_files) > 0) {
  all_tables <- vector("list", length(rds_files))

  for (i in seq_along(rds_files)) {
    current_file <- normalizePath(rds_files[i], winslash = "/", mustWork = FALSE)
    path_parts <- str_split(current_file, "/")[[1]]

    rep_part <- path_parts[str_detect(path_parts, "^rep_")][1]
    eta_part <- path_parts[str_detect(path_parts, "^Eta")][1]
    file_part <- fs::path_file(current_file)

    rep_value <- as.integer(str_remove(rep_part, "^rep_"))
    model_value <- as.integer(str_remove(eta_part, "^Eta0*"))
    trait_value <- str_remove(file_part, "_(metrics|metricas)\\.rds$")

    current_object <- readRDS(current_file)

    current_table <- tibble(
      Rep = rep_value,
      Model = model_value,
      Trait = trait_value,
      Component = "Residual",
      MeanVar = mean(current_object$varE_samples, na.rm = TRUE),
      DIC = current_object$DIC,
      runtime_sec = current_object$runtime_sec
    )

    if (length(current_object$variances_samples) > 0) {
      for (nm in names(current_object$variances_samples)) {
        current_table <- bind_rows(
          current_table,
          tibble(
            Rep = rep_value,
            Model = model_value,
            Trait = trait_value,
            Component = nm,
            MeanVar = mean(current_object$variances_samples[[nm]], na.rm = TRUE),
            DIC = current_object$DIC,
            runtime_sec = current_object$runtime_sec
          )
        )
      }
    }

    all_tables[[i]] <- current_table
  }

  all_variance <- bind_rows(all_tables) %>%
    arrange(Model, Rep, Trait, Component)

  saveRDS(all_variance, file.path(results_dir, "variance_components_all.rds"))
}
```

## 8. Loading the Consolidated Variance-Component Object

This section loads the consolidated results and prepares them for summarization. If the file does not exist, the script stops and instructs the user to execute the model-fitting stage first.

This chunk loads the already consolidated variance-component results so the document can summarize and visualize them without repeating the expensive estimation.

```{r load-results, error=TRUE}
raw_file <- file.path(results_dir, "variance_components_all.rds")

if (!file.exists(raw_file)) {
  stop(
    "The file '", raw_file, "' was not found. ",
    "Run the Bayesian fitting stage first or keep the previously saved _metrics.rds files in output/variance_components/."
  )
}

all_variance <- readRDS(raw_file)

all_variance <- all_variance %>%
  mutate(
    Trait = str_extract(Trait, "GY|KW"),
    Trait = ifelse(is.na(Trait), Trait, Trait),
    Model = as.integer(Model)
  )

cat("Number of rows in consolidated variance object:", nrow(all_variance), "\n")
```

## 9. Summarizing the Variance Components

Now we average the posterior means across repetitions for each model, trait, and component. Then we convert the results to **percentages of total variance**, which makes the models directly comparable.

This chunk reshapes and annotates the loaded results so that posterior means can be compared across traits, models, and component categories.

```{r process-variance-components}
variance_summary <- all_variance %>%
  group_by(Model, Trait, Component) %>%
  summarise(
    AvgVar = mean(MeanVar, na.rm = TRUE),
    MeanDIC = mean(DIC, na.rm = TRUE),
    MeanRuntimeSec = mean(runtime_sec, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(AvgVar), AvgVar > 0) %>%
  group_by(Model, Trait) %>%
  mutate(
    TotalVar = sum(AvgVar),
    Percentage = AvgVar / TotalVar
  ) %>%
  ungroup() %>%
  left_join(model_info, by = c("Model" = "ModelID")) %>%
  mutate(
    Model = factor(sprintf("M%02d", Model), levels = sprintf("M%02d", 1:18)),
    ComponentLabel = dplyr::recode(Component, !!!component_display),
    ComponentLabel = factor(ComponentLabel, levels = component_order),
    Trait = factor(Trait, levels = c("GY", "KW"))
  )

readr::write_csv(variance_summary, file.path(results_dir, "variance_components_processed.csv"))
```

### 9.1. Wide-format summary table

This table is helpful when you need a compact numerical summary for the manuscript or supplementary material.

This chunk builds the main summary table used to inspect the relative contribution of each variance component across models.

```{r summary-table}
variance_table <- variance_summary %>%
  dplyr::select(Model, Trait, ComponentLabel, Percentage) %>%
  dplyr::mutate(Percentage = round(100 * Percentage, 2)) %>%
  tidyr::pivot_wider(names_from = ComponentLabel, values_from = Percentage)

readr::write_csv(variance_table, file.path(results_dir, "variance_components_table_percent.csv"))

head(variance_table, 12) %>%
  knitr::kable(
    caption = "Preview of the processed variance-component table (percent of total variance).",
    align = "lcccccccccccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

## 10. Figure for the Article

The next figure summarizes the percentage contribution of each variance component across the 18 models for both traits. To keep the x-axis readable, the models are displayed as `M01` to `M18`, and the detailed descriptions remain available in the model catalog saved above.

This figure should be read as a structural summary of the fitted models, not as direct proof that combining more data sources automatically leads to stronger biological complementarity or better predictive performance.

This chunk draws the final variance-component figure, turning the processed table into a compact visual summary for interpretation.

```{r variance-figure, fig.cap="Figure 3. Posterior mean percentage of total variance attributed to each model component across the 18 predictive models for grain yield (GY) and kernel weight (KW). Each stacked bar represents one model (M01-M18), averaged across repetitions, and colors identify the variance components included in the model structure. The weather kernel (W) and its associated interaction terms appear mainly in models M10-M18. Variance decomposition should be interpreted as a description of model structure and not automatically as evidence of stronger biological complementarity among information sources.", fig.height=10, fig.width=13}
trait_labels <- c(
  GY = "A. Grain yield (GY)",
  KW = "B. Kernel weight (KW)"
)

variance_plot <- variance_summary %>%
  ggplot(aes(x = Model, y = Percentage, fill = ComponentLabel)) +
  geom_col(color = "white", linewidth = 0.15, width = 0.82) +
  facet_wrap(~ Trait, ncol = 1, labeller = as_labeller(trait_labels)) +
  ggthemes::scale_fill_gdocs(name = "Variance component", drop = FALSE) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    x = "Model ID",
    y = "Proportion of total variance"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    strip.text = element_text(face = "bold", size = 11),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8),
    axis.text.y = element_text(size = 9),
    legend.position = "right",
    legend.box = "vertical",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 8),
    legend.key.height = unit(0.45, "cm"),
    legend.key.width = unit(0.55, "cm"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  guides(fill = guide_legend(ncol = 2, byrow = TRUE))

print(variance_plot)

ggsave(
  filename = file.path(figure_dir, "variance_components_article.tiff"),
  plot = variance_plot,
  width = 14,
  height = 10,
  dpi = 300,
  compression = "lzw",
  bg = "white"
)

ggsave(
  filename = file.path(figure_dir, "variance_components_article.pdf"),
  plot = variance_plot,
  width = 14,
  height = 10,
  bg = "white"
)
```

## 11. Interpretation Guide

This final section clarifies how the figure should be interpreted.

- A large **Environment (E)** segment indicates that the categorical macro-environment effect dominates the variance partition.
- Large genomic terms such as **G**, **GGK**, or **GAK** indicate that genomic similarity captures an important share of the trait variability.
- Large phenomic terms such as **P**, **PGK**, or **PAK** indicate stronger contribution of the NIRS-derived similarity, but these terms should be interpreted with caution because NIRS may also reflect strong physiological response to the environment.
- Large weather-related terms such as **W**, **G × W**, or **P × W** indicate that weather-derived covariates explain additional structure. These terms appear mainly in models **M10-M18** and should be interpreted as complementary to the categorical environment effect, not as a replacement for **E**.
- A large **Residual** segment indicates that important sources of variability remain unexplained by the fitted model.

In the context of this article, these variance decompositions complement the predictive analyses by showing **which information sources contribute to model structure**. However, a large variance share for a component should not be read automatically as evidence of stronger biological complementarity or of better predictive ability.

## Stage navigation

**Previous stage:** [Matrices](matrizes.html)  
**Next stage:** [Prediction — Part I](analysis_prediction.html)