---
title: "Prediction Results Visualization"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 13,
  fig.height = 8,
  out.width = "100%"
)
```

# Prediction Results Visualization

This tutorial consolidates the cross-validation results produced in `analysis_prediction.Rmd` and turns them into summary tables and publication-ready figures.

In this workflow, the **weather kernel (`W`)** is used mainly in models **M10-M18**, whereas the categorical environment effect is represented separately by `E`. The phenomic component (`P`) should be interpreted with caution, because NIRS-based similarity may reflect strong physiological response to the environment in addition to potentially useful predictive structure.

To make comparison easier, the models are ordered to help answer three practical questions:

1. does changing from linear to Gaussian or arc-cosine kernels improve predictive ability?
2. does adding the weather kernel (`W`) improve predictive ability?
3. does combining genomic and phenomic information improve predictive ability?

## 1. Load packages

```{r load-packages}
library(tidyverse)
library(knitr)
library(ggthemes)
library(kableExtra)
```

## 2. Define the model catalog in comparison order

```{r model-catalog}
model_catalog <- tibble(
  ModelNumber = c(1, 10, 3, 13, 5, 16,
                  2, 11, 4, 14, 6, 17,
                  7, 12, 8, 15, 9, 18),
  Model = sprintf("M%02d", c(1, 10, 3, 13, 5, 16,
                             2, 11, 4, 14, 6, 17,
                             7, 12, 8, 15, 9, 18)),
  OriginalEta = c("Eta7", "Eta65", "Eta9", "Eta68", "Eta11", "Eta71",
                  "Eta8", "Eta66", "Eta10", "Eta69", "Eta12", "Eta72",
                  "Eta16", "Eta67", "Eta17", "Eta70", "Eta18", "Eta73"),
  Pathway = c(
    rep("Genomic pathway", 6),
    rep("Phenomic pathway", 6),
    rep("Combined G + P pathway", 6)
  ),
  KernelFamily = c(
    "Linear", "Linear",
    "Gaussian", "Gaussian",
    "Arc-cosine", "Arc-cosine",
    "Linear", "Linear",
    "Gaussian", "Gaussian",
    "Arc-cosine", "Arc-cosine",
    "Linear", "Linear",
    "Gaussian", "Gaussian",
    "Arc-cosine", "Arc-cosine"
  ),
  WeatherStatus = c(
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W"
  ),
  Group = c(
    "Genomic baseline",
    "Weather single-source",
    "Genomic baseline",
    "Weather single-source",
    "Genomic baseline",
    "Weather single-source",
    "Phenomic baseline",
    "Weather single-source",
    "Phenomic baseline",
    "Weather single-source",
    "Phenomic baseline",
    "Weather single-source",
    "Combined baseline",
    "Weather combined",
    "Combined baseline",
    "Weather combined",
    "Combined baseline",
    "Weather combined"
  ),
  Description = c(
    "Genomic linear + E",
    "Genomic linear + W",
    "Genomic Gaussian + E",
    "Genomic Gaussian + W",
    "Genomic arc-cosine + E",
    "Genomic arc-cosine + W",
    "Phenomic linear + E",
    "Phenomic linear + W",
    "Phenomic Gaussian + E",
    "Phenomic Gaussian + W",
    "Phenomic arc-cosine + E",
    "Phenomic arc-cosine + W",
    "Combined linear + E",
    "Combined linear + W",
    "Combined Gaussian + E",
    "Combined Gaussian + W",
    "Combined arc-cosine + E",
    "Combined arc-cosine + W"
  ),
  ShortDescription = c(
    "G + E",
    "G + W",
    "GGK + E",
    "GGK + W",
    "GAK + E",
    "GAK + W",
    "P + E",
    "P + W",
    "PGK + E",
    "PGK + W",
    "PAK + E",
    "PAK + W",
    "G+P + E",
    "G+P + W",
    "GGK+PGK + E",
    "GGK+PGK + W",
    "GAK+PAK + E",
    "GAK+PAK + W"
  )
)

model_catalog
```

## 3. Pairwise table for comparing the effect of W

```{r comparison-pairs}
comparison_pairs <- tibble(
  Comparison = c(
    "Genomic linear",
    "Genomic Gaussian",
    "Genomic arc-cosine",
    "Phenomic linear",
    "Phenomic Gaussian",
    "Phenomic arc-cosine",
    "Combined linear",
    "Combined Gaussian",
    "Combined arc-cosine"
  ),
  WithoutW = c("M01", "M03", "M05", "M02", "M04", "M06", "M07", "M08", "M09"),
  WithW = c("M10", "M13", "M16", "M11", "M14", "M17", "M12", "M15", "M18")
)

comparison_pairs
```

## 4. Locate the result files

```{r locate-files}
results_dir <- "output/results"
figures_dir <- "output/figures"
tables_dir <- "output/tables"

dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tables_dir, recursive = TRUE, showWarnings = FALSE)

all_result_files <- list.files(results_dir, pattern = "\\.csv$", full.names = TRUE)
all_result_names <- basename(all_result_files)

cv12_files <- all_result_files[grepl("^CV1_|^CV2_", all_result_names)]
cvloo_files <- all_result_files[grepl("^CV0_|^CV00_", all_result_names)]

cat("Number of CV1/CV2 files:", length(cv12_files), "\n")
cat("Number of CV0/CV00 files:", length(cvloo_files), "\n")
```

## 5. Read and classify CV1 and CV2 files

```{r read-cv12}
cv12_list <- vector("list", length(cv12_files))

for (i in seq_along(cv12_files)) {
  current_file <- cv12_files[i]
  file_name <- basename(current_file)
  file_name <- sub("\\.csv$", "", file_name)
  parts <- strsplit(file_name, "_")[[1]]

  current_cv <- parts[1]
  current_trait <- parts[2]
  current_model_number <- as.numeric(gsub("Eta", "", parts[3]))
  current_rep <- as.numeric(gsub("rep", "", parts[4]))
  current_fold <- as.numeric(gsub("fold", "", parts[5]))

  current_data <- readr::read_csv(current_file, show_col_types = FALSE)
  current_data$CV <- current_cv
  current_data$Trait <- current_trait
  current_data$ModelNumber <- current_model_number
  current_data$Model <- sprintf("M%02d", current_model_number)
  current_data$Rep <- current_rep
  current_data$Fold <- current_fold

  cv12_list[[i]] <- current_data
}

cv12_raw <- bind_rows(cv12_list) %>%
  left_join(model_catalog, by = c("ModelNumber", "Model"))

head(cv12_raw)
```

## 6. Read and classify CV0 and CV00 files

```{r read-cvloo}
cvloo_list <- vector("list", length(cvloo_files))

for (i in seq_along(cvloo_files)) {
  current_file <- cvloo_files[i]
  file_name <- basename(current_file)
  file_name <- sub("\\.csv$", "", file_name)
  parts <- strsplit(file_name, "_")[[1]]

  current_cv <- parts[1]
  current_trait <- parts[2]
  current_env_leave <- paste(parts[3], parts[4], sep = "_")
  current_model_number <- as.numeric(gsub("Eta", "", parts[5]))
  current_rep <- as.numeric(gsub("rep", "", parts[6]))

  current_data <- readr::read_csv(current_file, show_col_types = FALSE)
  current_data$CV <- current_cv
  current_data$Trait <- current_trait
  current_data$Env_leave <- current_env_leave
  current_data$ModelNumber <- current_model_number
  current_data$Model <- sprintf("M%02d", current_model_number)
  current_data$Rep <- current_rep

  cvloo_list[[i]] <- current_data
}

cvloo_raw <- bind_rows(cvloo_list) %>%
  left_join(model_catalog, by = c("ModelNumber", "Model"))

head(cvloo_raw)
```

## 7. Save the raw consolidated tables

```{r save-raw-tables}
readr::write_csv(cv12_raw, file.path(tables_dir, "prediction_cv1_cv2_raw.csv"))
readr::write_csv(cvloo_raw, file.path(tables_dir, "prediction_cv0_cv00_raw.csv"))
readr::write_csv(model_catalog, file.path(tables_dir, "prediction_model_catalog.csv"))
```

## 8. Summarize CV1 and CV2 results by repetition and fold

```{r summarize-cv12}
cv12_env_fold <- cv12_raw %>%
  group_by(ModelNumber, Model, OriginalEta, Group, Pathway, KernelFamily, WeatherStatus,
           Description, ShortDescription, Trait, CV, Rep, Fold, Env) %>%
  summarise(
    EnvCorrelation = mean(unique(cor[!is.na(cor)]), na.rm = TRUE),
    ValidationCount = sum(!is.na(cor)),
    .groups = "drop"
  ) %>%
  mutate(
    EnvCorrelation = ifelse(is.nan(EnvCorrelation), NA_real_, EnvCorrelation)
  ) %>%
  filter(!is.na(EnvCorrelation), ValidationCount > 0)

cv12_env_rep <- cv12_env_fold %>%
  group_by(ModelNumber, Model, OriginalEta, Group, Pathway, KernelFamily, WeatherStatus,
           Description, ShortDescription, Trait, CV, Rep, Env) %>%
  summarise(
    EnvCorrelation = weighted.mean(EnvCorrelation, w = ValidationCount, na.rm = TRUE),
    ValidationCount = sum(ValidationCount, na.rm = TRUE),
    .groups = "drop"
  )

cv12_rep_summary <- cv12_env_rep %>%
  group_by(ModelNumber, Model, OriginalEta, Group, Pathway, KernelFamily, WeatherStatus,
           Description, ShortDescription, Trait, CV, Rep) %>%
  summarise(
    WeightedCorrelation = weighted.mean(EnvCorrelation, w = ValidationCount, na.rm = TRUE),
    .groups = "drop"
  )

cv12_summary <- cv12_rep_summary %>%
  group_by(ModelNumber, Model, OriginalEta, Group, Pathway, KernelFamily, WeatherStatus,
           Description, ShortDescription, Trait, CV) %>%
  summarise(
    MeanCorrelation = mean(WeightedCorrelation, na.rm = TRUE),
    SDCorrelation = sd(WeightedCorrelation, na.rm = TRUE),
    NumberOfRepetitions = n(),
    .groups = "drop"
  ) %>%
  mutate(
    SDCorrelation = ifelse(is.na(SDCorrelation), 0, SDCorrelation)
  )

cv12_summary
```

## 9. Summarize CV0 and CV00 under LOEO

```{r summarize-cvloo}
cvloo_env_filtered <- cvloo_raw %>%
  mutate(Env = as.character(Env), Env_leave = as.character(Env_leave)) %>%
  filter(Env == Env_leave)

cvloo_file_summary <- cvloo_env_filtered %>%
  group_by(ModelNumber, Model, OriginalEta, Group, Pathway, KernelFamily, WeatherStatus,
           Description, ShortDescription, Trait, CV, Env_leave, Rep) %>%
  summarise(
    LOEOCorrelation = mean(unique(cor[!is.na(cor)]), na.rm = TRUE),
    ValidationCount = sum(!is.na(cor)),
    .groups = "drop"
  ) %>%
  mutate(
    LOEOCorrelation = ifelse(is.nan(LOEOCorrelation), NA_real_, LOEOCorrelation)
  ) %>%
  filter(!is.na(LOEOCorrelation), ValidationCount > 0)

cvloo_summary <- cvloo_file_summary %>%
  group_by(ModelNumber, Model, OriginalEta, Group, Pathway, KernelFamily, WeatherStatus,
           Description, ShortDescription, Trait, CV, Env_leave) %>%
  summarise(
    MeanCorrelation = mean(LOEOCorrelation, na.rm = TRUE),
    SDCorrelation = sd(LOEOCorrelation, na.rm = TRUE),
    NumberOfRuns = n(),
    .groups = "drop"
  ) %>%
  mutate(
    SDCorrelation = ifelse(is.na(SDCorrelation), 0, SDCorrelation)
  )

cvloo_summary
```

## 10. Save the summary tables

```{r save-summary-tables}
readr::write_csv(cv12_summary, file.path(tables_dir, "prediction_cv1_cv2_summary.csv"))
readr::write_csv(cvloo_summary, file.path(tables_dir, "prediction_cv0_cv00_summary.csv"))
readr::write_csv(cv12_rep_summary, file.path(tables_dir, "prediction_cv1_cv2_by_rep.csv"))
readr::write_csv(cv12_env_fold, file.path(tables_dir, "prediction_cv1_cv2_by_rep_fold_env.csv"))
readr::write_csv(cvloo_file_summary, file.path(tables_dir, "prediction_cv0_cv00_by_run.csv"))
```

## 11. Pairwise comparison tables for the effect of W

```{r comparison-tables}
cv12_w_effect <- comparison_pairs %>%
  left_join(
    cv12_summary %>%
      select(Trait, CV, Model, MeanCorrelation, SDCorrelation),
    by = c("WithoutW" = "Model")
  ) %>%
  rename(
    MeanWithoutW = MeanCorrelation,
    SDWithoutW = SDCorrelation
  ) %>%
  left_join(
    cv12_summary %>%
      select(Trait, CV, Model, MeanCorrelation, SDCorrelation),
    by = c("WithW" = "Model", "Trait", "CV")
  ) %>%
  rename(
    MeanWithW = MeanCorrelation,
    SDWithW = SDCorrelation
  ) %>%
  mutate(
    DeltaWithW = MeanWithW - MeanWithoutW
  ) %>%
  arrange(Trait, CV, Comparison)

cvloo_w_effect <- comparison_pairs %>%
  left_join(
    cvloo_summary %>%
      select(Trait, CV, Env_leave, Model, MeanCorrelation, SDCorrelation),
    by = c("WithoutW" = "Model")
  ) %>%
  rename(
    MeanWithoutW = MeanCorrelation,
    SDWithoutW = SDCorrelation
  ) %>%
  left_join(
    cvloo_summary %>%
      select(Trait, CV, Env_leave, Model, MeanCorrelation, SDCorrelation),
    by = c("WithW" = "Model", "Trait", "CV", "Env_leave")
  ) %>%
  rename(
    MeanWithW = MeanCorrelation,
    SDWithW = SDCorrelation
  ) %>%
  mutate(
    DeltaWithW = MeanWithW - MeanWithoutW
  ) %>%
  arrange(Trait, CV, Env_leave, Comparison)

readr::write_csv(cv12_w_effect, file.path(tables_dir, "prediction_cv1_cv2_w_effect_pairs.csv"))
readr::write_csv(cvloo_w_effect, file.path(tables_dir, "prediction_cv0_cv00_w_effect_pairs.csv"))
```

## 12. Comparison tables for kernel families

```{r kernel-family-comparison}
cv12_kernel_family_compare <- cv12_summary %>%
  select(Trait, CV, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  mutate(
    Delta_Gaussian_vs_Linear = Gaussian - Linear,
    Delta_ArcCos_vs_Linear = `Arc-cosine` - Linear,
    Delta_ArcCos_vs_Gaussian = `Arc-cosine` - Gaussian
  ) %>%
  arrange(Trait, CV, Pathway, WeatherStatus)

cvloo_kernel_family_compare <- cvloo_summary %>%
  select(Trait, CV, Env_leave, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  mutate(
    Delta_Gaussian_vs_Linear = Gaussian - Linear,
    Delta_ArcCos_vs_Linear = `Arc-cosine` - Linear,
    Delta_ArcCos_vs_Gaussian = `Arc-cosine` - Gaussian
  ) %>%
  arrange(Trait, CV, Env_leave, Pathway, WeatherStatus)

readr::write_csv(cv12_kernel_family_compare, file.path(tables_dir, "prediction_cv1_cv2_kernel_family_compare.csv"))
readr::write_csv(cvloo_kernel_family_compare, file.path(tables_dir, "prediction_cv0_cv00_kernel_family_compare.csv"))
```

## 13. Compact tables for manuscript use

```{r compact-tables}
cv12_table_display <- cv12_summary %>%
  arrange(Trait, CV, Pathway, KernelFamily, WeatherStatus) %>%
  mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3)
  ) %>%
  select(Trait, CV, Pathway, KernelFamily, WeatherStatus,
         ShortDescription, MeanCorrelation, SDCorrelation, NumberOfRepetitions)

cvloo_table_display <- cvloo_summary %>%
  arrange(Trait, CV, Env_leave, Pathway, KernelFamily, WeatherStatus) %>%
  mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3)
  ) %>%
  select(Trait, CV, Env_leave, Pathway, KernelFamily, WeatherStatus,
         ShortDescription, MeanCorrelation, SDCorrelation, NumberOfRuns)

cv12_top3 <- cv12_summary %>%
  group_by(Trait, CV) %>%
  slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3)
  ) %>%
  select(Trait, CV, ShortDescription, MeanCorrelation, SDCorrelation)

cvloo_top3 <- cvloo_summary %>%
  group_by(Trait, CV, Env_leave) %>%
  slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3)
  ) %>%
  select(Trait, CV, Env_leave, ShortDescription, MeanCorrelation, SDCorrelation)

readr::write_csv(cv12_table_display, file.path(tables_dir, "prediction_cv1_cv2_display_table.csv"))
readr::write_csv(cvloo_table_display, file.path(tables_dir, "prediction_cv0_cv00_display_table.csv"))
readr::write_csv(cv12_top3, file.path(tables_dir, "prediction_cv1_cv2_top3.csv"))
readr::write_csv(cvloo_top3, file.path(tables_dir, "prediction_cv0_cv00_top3.csv"))
```

### 13.1 Top models for CV1 and CV2

```{r table-cv12-top3}
cv12_top3 %>%
  knitr::kable(
    caption = "Top 3 models for each trait and validation scheme in CV1 and CV2.",
    align = "llccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  )
```

### 13.2 Top models for CV0 and CV00

```{r table-cvloo-top3}
cvloo_top3 %>%
  knitr::kable(
    caption = "Top 3 models for each trait, validation scheme, and held-out environment in CV0 and CV00.",
    align = "lllccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  )
```

### 13.3 Pairwise effect of W in CV1 and CV2

```{r table-cv12-w-effect}
cv12_w_effect %>%
  mutate(
    MeanWithoutW = round(MeanWithoutW, 3),
    MeanWithW = round(MeanWithW, 3),
    DeltaWithW = round(DeltaWithW, 3)
  ) %>%
  knitr::kable(
    caption = "Pairwise comparison of models without and with the weather kernel (W) in CV1 and CV2.",
    align = "lllccccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

### 13.4 Kernel family comparison in CV1 and CV2

```{r table-cv12-kernel-family}
cv12_kernel_family_compare %>%
  mutate(
    Linear = round(Linear, 3),
    Gaussian = round(Gaussian, 3),
    `Arc-cosine` = round(`Arc-cosine`, 3),
    Delta_Gaussian_vs_Linear = round(Delta_Gaussian_vs_Linear, 3),
    Delta_ArcCos_vs_Linear = round(Delta_ArcCos_vs_Linear, 3),
    Delta_ArcCos_vs_Gaussian = round(Delta_ArcCos_vs_Gaussian, 3)
  ) %>%
  knitr::kable(
    caption = "Comparison of kernel families within each analytical pathway in CV1 and CV2.",
    align = "llllcccccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

## 14. Plot CV1 and CV2

```{r figure-cv12, fig.cap="Predictive ability of the 18 selected models under CV1 and CV2 for grain yield (GY) and kernel weight (KW). Models are ordered to facilitate comparison of kernel family, addition of the weather kernel (W), and combination of information sources. Bars represent the mean weighted predictive correlation across repetitions, and error bars represent one standard deviation across repetitions."}
cv12_plot_data <- cv12_summary %>%
  mutate(
    ShortDescription = factor(ShortDescription, levels = model_catalog$ShortDescription)
  )

lower_limit_cv12 <- min(cv12_plot_data$MeanCorrelation - cv12_plot_data$SDCorrelation, na.rm = TRUE)
upper_limit_cv12 <- max(cv12_plot_data$MeanCorrelation + cv12_plot_data$SDCorrelation, na.rm = TRUE)

p_cv12 <- ggplot(cv12_plot_data, aes(x = ShortDescription, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30") +
  geom_errorbar(
    aes(
      ymin = MeanCorrelation - SDCorrelation,
      ymax = MeanCorrelation + SDCorrelation
    ),
    width = 0.20,
    linewidth = 0.35
  ) +
  facet_grid(Trait ~ CV) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  scale_y_continuous(
    name = "Predictive correlation",
    limits = c(lower_limit_cv12 - 0.03, upper_limit_cv12 + 0.03)
  ) +
  scale_x_discrete(name = "Model description") +
  ggthemes::scale_fill_gdocs(name = "Comparison group") +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
    strip.text = element_text(face = "bold", size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7),
    axis.text.y = element_text(size = 9),
    axis.title = element_text(size = 10),
    legend.position = "top",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  ) +
  labs(title = "Predictive ability under CV1 and CV2")

p_cv12

ggsave(file.path(figures_dir, "prediction_cv1_cv2.tiff"), plot = p_cv12, width = 14, height = 8, dpi = 300, bg = "white")
```

## 15. Plot CV0 and CV00

```{r figure-cvloo, fig.cap="Predictive ability of the 18 selected models under CV0 and CV00 in each held-out environment. Models are ordered to facilitate comparison of kernel family, addition of the weather kernel (W), and combination of information sources. Bars represent the mean predictive correlation for the leave-one-environment-out target, and error bars represent variation only when more than one run is available for the same target environment."}
cvloo_plot_data <- cvloo_summary %>%
  mutate(
    ShortDescription = factor(ShortDescription, levels = model_catalog$ShortDescription),
    Env_leave = factor(Env_leave, levels = c("CS11_WS", "CS11_WW", "CS12_WS", "CS12_WW"))
  )

lower_limit_cvloo <- min(cvloo_plot_data$MeanCorrelation - cvloo_plot_data$SDCorrelation, na.rm = TRUE)
upper_limit_cvloo <- max(cvloo_plot_data$MeanCorrelation + cvloo_plot_data$SDCorrelation, na.rm = TRUE)

p_cvloo <- ggplot(cvloo_plot_data, aes(x = ShortDescription, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30") +
  geom_errorbar(
    aes(
      ymin = MeanCorrelation - SDCorrelation,
      ymax = MeanCorrelation + SDCorrelation
    ),
    width = 0.20,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  facet_grid(Env_leave ~ Trait + CV) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  scale_y_continuous(
    name = "Predictive correlation",
    limits = c(lower_limit_cvloo - 0.03, upper_limit_cvloo + 0.03)
  ) +
  scale_x_discrete(name = "Model description") +
  ggthemes::scale_fill_gdocs(name = "Comparison group") +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
    axis.text.y = element_text(size = 8),
    axis.title = element_text(size = 10),
    legend.position = "top",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  ) +
  labs(title = "Predictive ability under CV0 and CV00")

p_cvloo

ggsave(file.path(figures_dir, "prediction_cv0_cv00.tiff"), plot = p_cvloo, width = 16, height = 10, dpi = 300, bg = "white")
```

## 16. Return dedicated figures for KW

```{r figure-kw-only}
p_cv12_kw <- p_cv12 %+% filter(cv12_plot_data, Trait == "KW") +
  facet_grid(. ~ CV) +
  labs(title = "Kernel weight (KW) under CV1 and CV2")

p_cvloo_kw <- p_cvloo %+% filter(cvloo_plot_data, Trait == "KW") +
  facet_grid(Env_leave ~ CV) +
  labs(title = "Kernel weight (KW) under CV0 and CV00")

p_cv12_kw
p_cvloo_kw

ggsave(file.path(figures_dir, "prediction_cv1_cv2_KW.tiff"), plot = p_cv12_kw, width = 14, height = 5.5, dpi = 300, bg = "white")
ggsave(file.path(figures_dir, "prediction_cv0_cv00_KW.tiff"), plot = p_cvloo_kw, width = 16, height = 8, dpi = 300, bg = "white")
```

## 17. Final validation notes

Compared with the previous version, this revised visualization script makes the following points explicit:

1. the model order is now designed for direct comparison of kernel family, addition of the weather kernel, and combination of information sources;
2. the model groups were reduced to five broader comparison groups to make interpretation easier;
3. the x-axis uses shorter model descriptions for cleaner visualization;
4. the script exports dedicated KW figures in addition to the full trait figures;
5. the weather effect can be inspected directly in the exported pairwise comparison tables;
6. the kernel family effect can be inspected directly in the exported kernel comparison tables.
