---
title: "Prediction summaries and figures - didactic tutorial"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
---

**Language / Idioma:** [English](visualization.html) | [Português](visualization_pt.html)

This file is the didactic version of the visualization stage. Its purpose is to take the raw prediction outputs produced in `analysis_prediction_run_outputs.Rmd`, reorganize them into interpretable summary tables, and then translate those summaries into compact comparisons and final figures.

At this point in the pipeline, the main analytical question is no longer how each model is specified, but how the fitted models behave under different prediction problems. For that reason, the visualization stage moves from model construction to model interpretation. It brings together performance summaries for `CV1`, `CV2`, `CV0`, and `CV00`, highlights contrasts between models with and without the weather kernel (`W`), and organizes the results into tables and figures that are easier to inspect than the raw prediction files themselves.

The current prediction workflow saves raw outputs in `output/results/` using filenames that encode the validation scenario, trait, fitted model, repetition, and fold. For `CV0` and `CV00`, the held-out environment label is also recorded in the filename and is later matched to `Env` through the filter `Env == Env_leave`. This naming structure is essential because it connects the execution stage to the interpretation stage.

Throughout this tutorial, the **weather kernel (`W`)** should be interpreted as a weather-derived source of similarity that complements, but does not replace, the categorical environment effect (`E`). The phenomic component (`P`) should also be interpreted carefully, because NIRS-based similarity can reflect strong environmental response in addition to potentially useful predictive signal.

### What this script does

The workflow is organized around four practical tasks:

1. locate and read the prediction CSV files;
2. parse the filenames and recover trait, model, repetition, and leave-one-environment-out information;
3. summarize results for `CV1`, `CV2`, `CV0`, and `CV00`;
4. export clean tables and figures for the next manuscript-oriented stage.

To make interpretation easier, the models are ordered to help answer three practical questions:

1. does moving from linear to Gaussian or arc-cosine kernels improve predictive ability?
2. does adding the weather kernel (`W`) improve predictive ability?
3. does combining genomic and phenomic information improve predictive ability?

## Inputs, outputs and next step

**Inputs of this step**
- raw prediction outputs written by `analysis_prediction_run_outputs.Rmd`
- light prediction metadata tables generated in the previous stage

**Outputs of this step**
- final summary tables
- model rankings
- W-effect comparisons
- kernel-family comparisons
- final figures for downstream interpretation

**Next step**
- the outputs generated here feed the manuscript-oriented summaries and the final interpretation of the project results.

## 1. Setup

The document begins by defining a consistent rendering style for code, tables, and figures. This helps the tutorial remain readable when it is inspected in HTML and keeps the visual presentation stable across the different outputs produced in this stage.

```{r setup}
knitr::opts_chunk$set( echo = TRUE,
 warning = FALSE,
 message = FALSE,
 fig.width = 13,
 fig.height = 8,
  out.width = "100%" )
```

## 2. Load packages and locate project folders

The visualization stage depends on a small set of packages for data manipulation, figure construction, and table formatting. It also needs to locate the project folders correctly, because the summaries generated here must be written in stable output paths that can later be reused by the site and by manuscript-oriented materials.

This section standardizes the project root, locates the folders that contain the prediction outputs, and prepares the directories in which the final tables and figures will be written.

```{r packages-and-paths}
rm(list = ls())
options(scipen = 999)
packages_needed <- c( "dplyr",
 "tidyr",
 "readr",
 "tibble",
 "ggplot2",
 "ggthemes",
 "knitr",
 "kableExtra",
  "stringr" )
packages_missing <- packages_needed[!(packages_needed %in% installed.packages()[, "Package"])]
if (length(packages_missing) > 0) {
  install.packages(packages_missing)
}
library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(ggplot2)
library(ggthemes)
library(knitr)
library(kableExtra)
library(stringr)
project_root <- getwd()
if (!dir.exists(file.path(project_root, "output")) && dir.exists(file.path(project_root, "..", "output"))) {
  project_root <- normalizePath(file.path(project_root, ".."))
}
if (!dir.exists(file.path(project_root, "output"))) {
  stop("Could not find the folder 'output/'.\nRun this script from the project root or from analysis/.")
}
results_dir <- file.path(project_root, "output", "results")
figures_dir <- file.path(project_root, "output", "figures")
tables_dir <- file.path(project_root, "output", "tables")
dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tables_dir, recursive = TRUE, showWarnings = FALSE)
```

## 3. Model catalog used in the reduced pipeline

The visualization stage uses the same reduced catalog of 18 models adopted in the prediction workflow. Here that catalog is reorganized so that the summaries are easier to interpret in analytical terms. Instead of showing only model numbers, the table emphasizes the pathway being represented, the kernel family being used, and whether weather information is present.

This organization is important because the comparisons shown later in the document are not arbitrary. They are designed to help the reader understand where predictive gains come from and whether those gains are associated with weather information, kernel choice, or integration between genomic and phenomic sources.

```{r model-catalog}
model_catalog <- tibble( ModelNumber = c(1, 10, 3, 13, 5, 16, 2, 11, 4, 14, 6, 17, 7, 12, 8, 15, 9, 18),
 Model = paste0("Eta", c(1, 10, 3, 13, 5, 16, 2, 11, 4, 14, 6, 17, 7, 12, 8, 15, 9, 18)),
 Pathway = c( rep("Genomic pathway", 6), rep("Phenomic pathway", 6), rep("Combined G + P pathway", 6) ),
 KernelFamily = c( "Linear", "Linear", "Gaussian", "Gaussian", "Arc-cosine", "Arc-cosine", "Linear", "Linear", "Gaussian", "Gaussian", "Arc-cosine", "Arc-cosine", "Linear", "Linear", "Gaussian", "Gaussian", "Arc-cosine", "Arc-cosine" ),
 WeatherStatus = c( "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W" ),
 Group = c( "Genomic baseline", "Weather single-source", "Genomic baseline", "Weather single-source", "Genomic baseline", "Weather single-source", "Phenomic baseline", "Weather single-source", "Phenomic baseline", "Weather single-source", "Phenomic baseline", "Weather single-source", "Combined baseline", "Weather combined", "Combined baseline", "Weather combined", "Combined baseline", "Weather combined" ),
 Description = c( "Genomic linear + E", "Genomic linear + W", "Genomic Gaussian + E", "Genomic Gaussian + W", "Genomic arc-cosine + E", "Genomic arc-cosine + W", "Phenomic linear + E", "Phenomic linear + W", "Phenomic Gaussian + E", "Phenomic Gaussian + W", "Phenomic arc-cosine + E", "Phenomic arc-cosine + W", "Combined linear + E", "Combined linear + W", "Combined Gaussian + E", "Combined Gaussian + W", "Combined arc-cosine + E", "Combined arc-cosine + W" ),
  ShortDescription = c( "G + E", "G + W", "GGK + E", "GGK + W", "GAK + E", "GAK + W", "P + E", "P + W", "PGK + E", "PGK + W", "PAK + E", "PAK + W", "G+P + E", "G+P + W", "GGK+PGK + E", "GGK+PGK + W", "GAK+PAK + E", "GAK+PAK + W" ) )
model_catalog
```

## 4. Comparison pairs used for W versus no-W contrasts

A central question in this project is whether adding the weather kernel (`W`) improves predictive ability. To answer that question fairly, the comparison must always be made between models that are structurally similar and differ only by the inclusion of weather information.

The comparison pairs defined below provide exactly that structure. They are later used to build tables that quantify how much predictive performance changes when `W` is introduced within the same pathway and kernel family.

```{r comparison-pairs}
comparison_pairs <- tibble( Comparison = c( "Genomic linear", "Genomic Gaussian", "Genomic arc-cosine", "Phenomic linear", "Phenomic Gaussian", "Phenomic arc-cosine", "Combined linear", "Combined Gaussian", "Combined arc-cosine" ),
 WithoutW = c("Eta1", "Eta3", "Eta5", "Eta2", "Eta4", "Eta6", "Eta7", "Eta8", "Eta9"),
  WithW = c("Eta10", "Eta13", "Eta16", "Eta11", "Eta14", "Eta17", "Eta12", "Eta15", "Eta18") )
comparison_pairs
```

## 5. Locate prediction files

The first operational step of the visualization stage is to locate the raw prediction files produced by the previous module. Under the current pipeline standard, these files are saved recursively inside `output/results/`, with an explicit directory hierarchy by validation scheme, trait, repetition, and fold.

At this stage, the script reads the results folder recursively, keeps only the prediction CSVs stored inside the structured `CV1`, `CV2`, `CV0`, and `CV00` branches, and ignores other artifacts such as `bglr_runs`. This makes the downstream summaries compatible with the new `Rep/Fold` contract while preserving the distinction between genotype-based and leave-one-environment-out validation problems.

```{r locate-files}
all_result_rel_paths <- list.files(results_dir, pattern = "\\.csv$", full.names = FALSE, recursive = TRUE)
all_result_rel_paths <- gsub("\\", "/", all_result_rel_paths)
structured_idx <- grepl("^CV1/|^CV2/|^CV0/|^CV00/", all_result_rel_paths)
all_result_rel_paths <- all_result_rel_paths[structured_idx]
all_result_files <- file.path(results_dir, all_result_rel_paths)
cv12_idx <- grepl("^CV1/|^CV2/", all_result_rel_paths)
cvloo_idx <- grepl("^CV0/|^CV00/", all_result_rel_paths)
cv12_files <- all_result_files[cv12_idx]
cv12_rel_paths <- all_result_rel_paths[cv12_idx]
cvloo_files <- all_result_files[cvloo_idx]
cvloo_rel_paths <- all_result_rel_paths[cvloo_idx]
cat("Number of CV1/CV2 files:", length(cv12_files), "
")
cat("Number of CV0/CV00 files:", length(cvloo_files), "
")
if (length(cv12_files) == 0 && length(cvloo_files) == 0) {
  stop("No prediction CSV files were found in output/results/.
Run analysis_prediction_run_outputs first.")
}
```

## 6. Read CV1 and CV2 files generated by the final prediction script

The `CV1` and `CV2` files contain the raw outputs needed to summarize prediction performance when the challenge is defined on the genotype axis. In the current workflow, repetition and fold are encoded both in the directory hierarchy and in the filename, so the script reconstructs those metadata explicitly before building the summaries.

This design is useful because it keeps the file system organized for quick inspection while still allowing the visualization stage to rebuild unified tables across repetitions and folds.

```{r read-cv12}
cv12_list <- vector("list", length(cv12_files))
for (i in seq_along(cv12_files)) {
  current_file <- cv12_files[i]
  current_rel_path <- cv12_rel_paths[i]
  path_parts <- strsplit(current_rel_path, "/")[[1]]

  if (length(path_parts) != 5) {
    stop(
      "Unexpected directory structure for CV1/CV2 file: ",
      current_rel_path,
      ".
Expected pattern CV1/<trait>/repXX/foldXX/<file>.csv or CV2/<trait>/repXX/foldXX/<file>.csv"
    )
  }

  current_cv <- path_parts[1]
  trait_from_path <- path_parts[2]
  rep_from_path <- as.numeric(gsub("^rep", "", path_parts[3]))
  fold_from_path <- as.numeric(gsub("^fold", "", path_parts[4]))

  file_name <- sub("\.csv$", "", basename(current_file))
  file_parts <- strsplit(file_name, "_")[[1]]

  if (length(file_parts) != 5) {
    stop(
      "Unexpected filename for CV1/CV2 file: ",
      file_name,
      ".
Expected pattern CV1_<trait>_Eta<model>_repXX_foldXX.csv or CV2_<trait>_Eta<model>_repXX_foldXX.csv"
    )
  }

  current_trait <- file_parts[2]
  current_model_label <- file_parts[3]
  current_model_number <- as.numeric(gsub("^Eta", "", current_model_label))
  current_rep <- as.numeric(gsub("^rep", "", file_parts[4]))
  current_fold <- as.numeric(gsub("^fold", "", file_parts[5]))

  if (!identical(current_cv, file_parts[1]) || !identical(trait_from_path, current_trait) || !identical(rep_from_path, current_rep) || !identical(fold_from_path, current_fold)) {
    stop("Mismatch between path and filename metadata in: ", current_rel_path)
  }

  current_data <- readr::read_csv(current_file, show_col_types = FALSE)

  if (!("Env" %in% names(current_data)) || !("cor" %in% names(current_data))) {
    stop(
      "Unexpected column structure in file: ",
      file_name,
      ".
The final prediction script should export columns including Env and cor."
    )
  }

  current_data <- current_data %>%
    mutate(
      RelativePath = current_rel_path,
      Fold = current_fold,
      CV = current_cv,
      Trait = current_trait,
      ModelNumber = current_model_number,
      Model = current_model_label,
      Rep = current_rep
    )

  cv12_list[[i]] <- current_data
}

cv12_raw <- bind_rows(cv12_list) %>%
  left_join(model_catalog, by = c("ModelNumber", "Model"))

if (!("Runtime_sec" %in% names(cv12_raw))) cv12_raw$Runtime_sec <- NA_real_
readr::write_csv(cv12_raw, file.path(tables_dir, "prediction_cv1_cv2_raw.csv"))
```

## 7. Read CV0 and CV00 files generated by the final prediction script

The `CV0` and `CV00` files represent the most demanding prediction problems in the study because they involve a held-out environment. Under the current pipeline standard, the held-out environment is encoded both in the directory hierarchy and in the filename, so the script reconstructs that information explicitly before the leave-one-environment-out summaries are built.

This section reads those files, checks the consistency of the structured paths, and prepares the metadata needed to keep `Env_leave` explicit throughout the summaries.

```{r read-cvloo}
cvloo_list <- vector("list", length(cvloo_files))
for (i in seq_along(cvloo_files)) {
  current_file <- cvloo_files[i]
  current_rel_path <- cvloo_rel_paths[i]
  path_parts <- strsplit(current_rel_path, "/")[[1]]

  if (length(path_parts) != 6) {
    stop(
      "Unexpected directory structure for CV0/CV00 file: ",
      current_rel_path,
      ".
Expected pattern CV0/<Env_leave>/<trait>/repXX/foldXX/<file>.csv or CV00/<Env_leave>/<trait>/repXX/foldXX/<file>.csv"
    )
  }

  current_cv <- path_parts[1]
  env_from_path <- path_parts[2]
  trait_from_path <- path_parts[3]
  rep_from_path <- as.numeric(gsub("^rep", "", path_parts[4]))
  fold_from_path <- as.numeric(gsub("^fold", "", path_parts[5]))

  file_name <- sub("\.csv$", "", basename(current_file))
  file_parts <- strsplit(file_name, "_")[[1]]

  if (length(file_parts) < 6) {
    stop(
      "Unexpected filename for CV0/CV00 file: ",
      file_name,
      ".
Expected pattern CV0_<trait>_<Env_leave>_Eta<model>_repXX_foldXX.csv or CV00_<trait>_<Env_leave>_Eta<model>_repXX_foldXX.csv"
    )
  }

  current_trait <- file_parts[2]
  current_model_label <- file_parts[length(file_parts) - 2]
  current_model_number <- as.numeric(gsub("^Eta", "", current_model_label))
  current_rep <- as.numeric(gsub("^rep", "", file_parts[length(file_parts) - 1]))
  current_fold <- as.numeric(gsub("^fold", "", file_parts[length(file_parts)]))
  current_env_leave <- paste(file_parts[3:(length(file_parts) - 3)], collapse = "_")

  if (!identical(current_cv, file_parts[1]) || !identical(env_from_path, current_env_leave) || !identical(trait_from_path, current_trait) || !identical(rep_from_path, current_rep) || !identical(fold_from_path, current_fold)) {
    stop("Mismatch between path and filename metadata in: ", current_rel_path)
  }

  current_data <- readr::read_csv(current_file, show_col_types = FALSE)

  if (!("Env" %in% names(current_data)) || !("cor" %in% names(current_data))) {
    stop(
      "Unexpected column structure in file: ",
      file_name,
      ".
The final prediction script should export columns including Env and cor."
    )
  }

  current_data <- current_data %>%
    mutate(
      RelativePath = current_rel_path,
      Fold = current_fold,
      CV = current_cv,
      Trait = current_trait,
      Env_leave = current_env_leave,
      ModelNumber = current_model_number,
      Model = current_model_label,
      Rep = current_rep
    )

  cvloo_list[[i]] <- current_data
}
cvloo_raw <- bind_rows(cvloo_list) %>%
  left_join(model_catalog, by = c("ModelNumber", "Model"))
if (!("Runtime_sec" %in% names(cvloo_raw))) cvloo_raw$Runtime_sec <- NA_real_
readr::write_csv(cvloo_raw, file.path(tables_dir, "prediction_cv0_cv00_raw.csv"))
readr::write_csv(model_catalog, file.path(tables_dir, "prediction_model_catalog.csv"))
```

## 8. Summarize CV1 and CV2

This stage builds the summaries for `CV1` and `CV2` in a stepwise way: first by environment and fold, then by environment within repetition, then by repetition, and finally by model-level means and standard deviations across repetitions.

Once the raw `CV1` and `CV2` files have been assembled, the next goal is to summarize them in a form that supports interpretation rather than file inspection. The summaries move from lower-level information to increasingly aggregated views, so that each model can be evaluated in terms of its typical predictive behavior across repetitions, folds, and environments.

This hierarchy is important because raw prediction files are too granular for interpretation. What matters at this stage is the stable performance pattern of each model under each validation scenario.

```{r summarize-cv12}
cv12_env_fold <- cv12_raw %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Rep,
   Fold,
    Env ) %>%
  summarise( EnvCorrelation = mean(cor, na.rm = TRUE),
   ValidationCount = sum(!is.na(cor)),
   MeanRuntimeSec = mean(Runtime_sec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( EnvCorrelation = ifelse(is.nan(EnvCorrelation), NA_real_, EnvCorrelation),
    MeanRuntimeSec = ifelse(is.nan(MeanRuntimeSec), NA_real_, MeanRuntimeSec) ) %>%
  filter(!is.na(EnvCorrelation), ValidationCount > 0)
cv12_env_rep <- cv12_env_fold %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Rep,
    Env ) %>%
  summarise( EnvCorrelation = weighted.mean(EnvCorrelation, w = ValidationCount, na.rm = TRUE),
   ValidationCount = sum(ValidationCount, na.rm = TRUE),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" )
cv12_rep_summary <- cv12_env_rep %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
    Rep ) %>%
  summarise( WeightedCorrelation = weighted.mean(EnvCorrelation, w = ValidationCount, na.rm = TRUE),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" )
cv12_summary <- cv12_rep_summary %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
    CV ) %>%
  summarise( MeanCorrelation = mean(WeightedCorrelation, na.rm = TRUE),
   SDCorrelation = sd(WeightedCorrelation, na.rm = TRUE),
   NumberOfRepetitions = n(),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( SDCorrelation = ifelse(is.na(SDCorrelation), 0, SDCorrelation) )
readr::write_csv(cv12_env_fold, file.path(tables_dir, "prediction_cv1_cv2_by_rep_fold_env.csv"))
readr::write_csv(cv12_env_rep, file.path(tables_dir, "prediction_cv1_cv2_by_rep_env.csv"))
readr::write_csv(cv12_rep_summary, file.path(tables_dir, "prediction_cv1_cv2_by_rep.csv"))
readr::write_csv(cv12_summary, file.path(tables_dir, "prediction_cv1_cv2_summary.csv"))
```

## 9. Summarize CV0 and CV00

For leave-one-environment-out scenarios, the script first restricts the summaries to rows in which `Env == Env_leave`. This keeps only the held-out-environment component that is conceptually relevant for `CV0` and `CV00`.

The summaries for `CV0` and `CV00` follow the same general logic as the `CV1/CV2` summaries, but they keep the held-out environment explicit. This distinction is essential because predictive performance in leave-one-environment-out settings can vary strongly across environments.

By preserving `Env_leave` in the final tables, the script allows the reader to see not only which models perform well on average, but also whether that behavior is stable across the different environmental contexts.

```{r summarize-cvloo}
cvloo_env_filtered <- cvloo_raw %>%
  filter(Env == Env_leave)
cvloo_env_fold <- cvloo_env_filtered %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Env_leave,
   Rep,
   Fold,
    Env ) %>%
  summarise( LOEOCorrelation = mean(cor, na.rm = TRUE),
   ValidationCount = sum(!is.na(cor)),
   MeanRuntimeSec = mean(Runtime_sec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( LOEOCorrelation = ifelse(is.nan(LOEOCorrelation), NA_real_, LOEOCorrelation),
    MeanRuntimeSec = ifelse(is.nan(MeanRuntimeSec), NA_real_, MeanRuntimeSec) ) %>%
  filter(!is.na(LOEOCorrelation), ValidationCount > 0)
cvloo_rep_summary <- cvloo_env_fold %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Env_leave,
    Rep ) %>%
  summarise( LOEOCorrelation = weighted.mean(LOEOCorrelation, w = ValidationCount, na.rm = TRUE),
   ValidationCount = sum(ValidationCount, na.rm = TRUE),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" )
cvloo_summary <- cvloo_rep_summary %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
    Env_leave ) %>%
  summarise( MeanCorrelation = mean(LOEOCorrelation, na.rm = TRUE),
   SDCorrelation = sd(LOEOCorrelation, na.rm = TRUE),
   NumberOfRuns = n(),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( SDCorrelation = ifelse(is.na(SDCorrelation), 0, SDCorrelation) )
readr::write_csv(cvloo_env_fold, file.path(tables_dir, "prediction_cv0_cv00_by_rep_fold_env.csv"))
readr::write_csv(cvloo_rep_summary, file.path(tables_dir, "prediction_cv0_cv00_by_run.csv"))
readr::write_csv(cvloo_summary, file.path(tables_dir, "prediction_cv0_cv00_summary.csv"))
```

## 10. Comparison tables

The summary tables describe model performance, but they do not yet answer the comparative questions that motivated the reduced pipeline. Those questions require derived contrasts, especially between models without and with `W`, and between different kernel families within the same pathway.

The tables built in this section reorganize the summaries into analytical comparisons that are easier to interpret and easier to carry into manuscript-oriented outputs.

```{r comparison-tables}
cv12_w_effect <- comparison_pairs %>%
  left_join( cv12_summary %>% select(Trait, CV, Model, MeanCorrelation, SDCorrelation),
    by = c("WithoutW" = "Model") ) %>%
  rename( MeanWithoutW = MeanCorrelation, SDWithoutW = SDCorrelation ) %>%
  left_join( cv12_summary %>% select(Trait, CV, Model, MeanCorrelation, SDCorrelation),
    by = c("WithW" = "Model", "Trait", "CV") ) %>%
  rename( MeanWithW = MeanCorrelation, SDWithW = SDCorrelation ) %>%
  mutate( DeltaWithW = MeanWithW - MeanWithoutW ) %>%
  arrange(Trait, CV, Comparison)
cvloo_w_effect <- comparison_pairs %>%
  left_join( cvloo_summary %>% select(Trait, CV, Env_leave, Model, MeanCorrelation, SDCorrelation),
    by = c("WithoutW" = "Model") ) %>%
  rename( MeanWithoutW = MeanCorrelation, SDWithoutW = SDCorrelation ) %>%
  left_join( cvloo_summary %>% select(Trait, CV, Env_leave, Model, MeanCorrelation, SDCorrelation),
    by = c("WithW" = "Model", "Trait", "CV", "Env_leave") ) %>%
  rename( MeanWithW = MeanCorrelation, SDWithW = SDCorrelation ) %>%
  mutate( DeltaWithW = MeanWithW - MeanWithoutW ) %>%
  arrange(Trait, CV, Env_leave, Comparison)
cv12_kernel_family_compare <- cv12_summary %>%
  select(Trait, CV, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  mutate( Delta_Gaussian_vs_Linear = Gaussian - Linear,
   Delta_ArcCos_vs_Linear = `Arc-cosine` - Linear,
    Delta_ArcCos_vs_Gaussian = `Arc-cosine` - Gaussian ) %>%
  arrange(Trait, CV, Pathway, WeatherStatus)
cvloo_kernel_family_compare <- cvloo_summary %>%
  select(Trait, CV, Env_leave, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  mutate( Delta_Gaussian_vs_Linear = Gaussian - Linear,
   Delta_ArcCos_vs_Linear = `Arc-cosine` - Linear,
    Delta_ArcCos_vs_Gaussian = `Arc-cosine` - Gaussian ) %>%
  arrange(Trait, CV, Env_leave, Pathway, WeatherStatus)
readr::write_csv(cv12_w_effect, file.path(tables_dir, "prediction_cv1_cv2_w_effect_pairs.csv"))
readr::write_csv(cvloo_w_effect, file.path(tables_dir, "prediction_cv0_cv00_w_effect_pairs.csv"))
readr::write_csv(cv12_kernel_family_compare,
  file.path(tables_dir, "prediction_cv1_cv2_kernel_family_compare.csv"))
readr::write_csv(cvloo_kernel_family_compare,
  file.path(tables_dir, "prediction_cv0_cv00_kernel_family_compare.csv"))
```

## 11. Compact display tables

The full summaries are useful for traceability, but they are often too dense for quick inspection. For that reason, this stage also produces compact display tables that preserve the main analytical information while making the results easier to read in reports, supplementary material, or classroom discussion.

These reduced tables are especially useful when the reader wants to compare pathways, kernel families, or the relative position of the best-performing models without scanning the full outputs.

```{r display-tables}
cv12_table_display <- cv12_summary %>%
  arrange(Trait, CV, Pathway, KernelFamily, WeatherStatus) %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3),
   SDCorrelation = round(SDCorrelation, 3),
    MeanRuntimeSec = round(MeanRuntimeSec, 2) ) %>%
  select( Trait,
   CV,
   Pathway,
   KernelFamily,
   WeatherStatus,
   ShortDescription,
   MeanCorrelation,
   SDCorrelation,
   NumberOfRepetitions,
    MeanRuntimeSec )
cvloo_table_display <- cvloo_summary %>%
  arrange(Trait, CV, Env_leave, Pathway, KernelFamily, WeatherStatus) %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3),
   SDCorrelation = round(SDCorrelation, 3),
    MeanRuntimeSec = round(MeanRuntimeSec, 2) ) %>%
  select( Trait,
   CV,
   Env_leave,
   Pathway,
   KernelFamily,
   WeatherStatus,
   ShortDescription,
   MeanCorrelation,
   SDCorrelation,
   NumberOfRuns,
    MeanRuntimeSec )
cv12_top3 <- cv12_summary %>%
  group_by(Trait, CV) %>%
  slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3), SDCorrelation = round(SDCorrelation, 3) ) %>%
  select(Trait, CV, ShortDescription, MeanCorrelation, SDCorrelation)
cvloo_top3 <- cvloo_summary %>%
  group_by(Trait, CV, Env_leave) %>%
  slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3), SDCorrelation = round(SDCorrelation, 3) ) %>%
  select(Trait, CV, Env_leave, ShortDescription, MeanCorrelation, SDCorrelation)
readr::write_csv(cv12_table_display, file.path(tables_dir, "prediction_cv1_cv2_display_table.csv"))
readr::write_csv(cvloo_table_display, file.path(tables_dir, "prediction_cv0_cv00_display_table.csv"))
readr::write_csv(cv12_top3, file.path(tables_dir, "prediction_cv1_cv2_top3.csv"))
readr::write_csv(cvloo_top3, file.path(tables_dir, "prediction_cv0_cv00_top3.csv"))
```

## 12. Top-3 tables

A compact ranking is often the easiest way to communicate the strongest results of a prediction study. The Top-3 tables below identify the best-performing models within each validation setting and help the reader quickly see which pathways dominate under each prediction problem.

The first table focuses on `CV1` and `CV2`, where the challenge is defined around new or partially observed genotypes.

```{r table-cv12-top3}
cv12_top3 %>%
  knitr::kable( caption = "Top 3 models for each trait and validation scheme in CV1 and CV2.",
    align = "llccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" )
```

The second table focuses on `CV0` and `CV00`, where performance is also conditioned by the held-out environment.

```{r table-cvloo-top3}
cvloo_top3 %>%
  knitr::kable( caption = "Top 3 models for each trait, validation scheme, and held-out environment in CV0 and CV00.",
    align = "lllccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" )
```

## 13. W-effect tables

These tables isolate one of the main scientific questions of the project: whether adding weather information improves prediction. Because the comparisons are defined through carefully matched model pairs, the resulting contrasts are easier to interpret than raw model-by-model inspection.

The first table presents the effect of adding `W` under `CV1` and `CV2`.

```{r table-cv12-w-effect}
cv12_w_effect %>%
  mutate( MeanWithoutW = round(MeanWithoutW, 3),
   MeanWithW = round(MeanWithW, 3),
    DeltaWithW = round(DeltaWithW, 3) ) %>%
  knitr::kable( caption = "Pairwise comparison of models without and with the weather kernel (W) in CV1 and CV2.",
    align = "lllccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

The second table presents the same type of contrast under `CV0` and `CV00`, where generalization to new environments becomes part of the prediction problem.

```{r table-cvloo-w-effect}
cvloo_w_effect %>%
  mutate( MeanWithoutW = round(MeanWithoutW, 3),
   MeanWithW = round(MeanWithW, 3),
    DeltaWithW = round(DeltaWithW, 3) ) %>%
  knitr::kable( caption = "Pairwise comparison of models without and with the weather kernel (W) in CV0 and CV00.",
    align = "llllccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

## 14. Kernel-family comparison tables

Kernel choice is another central analytical decision in this project. Rather than treating the linear, Gaussian, and arc-cosine formulations as isolated models, this section compares them within each pathway so that the effect of kernel family can be interpreted directly.

The first table reports these comparisons for `CV1` and `CV2`.

```{r table-cv12-kernel-family}
cv12_kernel_family_compare %>%
  mutate( Linear = round(Linear, 3),
   Gaussian = round(Gaussian, 3),
   `Arc-cosine` = round(`Arc-cosine`, 3),
   Delta_Gaussian_vs_Linear = round(Delta_Gaussian_vs_Linear, 3),
   Delta_ArcCos_vs_Linear = round(Delta_ArcCos_vs_Linear, 3),
    Delta_ArcCos_vs_Gaussian = round(Delta_ArcCos_vs_Gaussian, 3) ) %>%
  knitr::kable( caption = "Comparison of kernel families within each analytical pathway in CV1 and CV2.",
    align = "llllcccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

The second table reports the same comparisons for `CV0` and `CV00`, where the results must be read together with the held-out environment structure.

```{r table-cvloo-kernel-family}
cvloo_kernel_family_compare %>%
  mutate( Linear = round(Linear, 3),
   Gaussian = round(Gaussian, 3),
   `Arc-cosine` = round(`Arc-cosine`, 3),
   Delta_Gaussian_vs_Linear = round(Delta_Gaussian_vs_Linear, 3),
   Delta_ArcCos_vs_Linear = round(Delta_ArcCos_vs_Linear, 3),
    Delta_ArcCos_vs_Gaussian = round(Delta_ArcCos_vs_Gaussian, 3) ) %>%
  knitr::kable( caption = "Comparison of kernel families within each analytical pathway in CV0 and CV00.",
    align = "lllllcccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

## 15. Figures

The figures produced in this stage are visual summaries of the final model catalog. Their role is to condense the numerical tables into patterns that can be recognized quickly, especially when the reader wants to compare pathways, kernel families, and the presence or absence of `W`.

The first figure summarizes predictive ability under `CV1` and `CV2`, combining mean predictive correlations with uncertainty bars across models.

```{r figure-cv12}
cv12_plot_data <- cv12_summary %>%
  mutate( ShortDescription = factor(ShortDescription, levels = model_catalog$ShortDescription) )
lower_limit_cv12 <- min(cv12_plot_data$MeanCorrelation - cv12_plot_data$SDCorrelation, na.rm = TRUE)
upper_limit_cv12 <- max(cv12_plot_data$MeanCorrelation + cv12_plot_data$SDCorrelation, na.rm = TRUE)
p_cv12 <- ggplot(cv12_plot_data, aes(x = ShortDescription, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30") +
  geom_errorbar( aes( ymin = MeanCorrelation - SDCorrelation, ymax = MeanCorrelation + SDCorrelation ),
   width = 0.20,
    linewidth = 0.35 ) +
  facet_grid(Trait ~ CV) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  coord_cartesian(ylim = c(lower_limit_cv12 - 0.03, upper_limit_cv12 + 0.03)) +
  scale_x_discrete(name = "Model description") +
  scale_y_continuous(name = "Predictive correlation") +
  ggthemes::scale_fill_gdocs(name = "Comparison group") +
  theme_bw() +
  theme( plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
   strip.text = element_text(face = "bold", size = 10),
   axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7),
   axis.text.y = element_text(size = 9),
   axis.title = element_text(size = 10),
   legend.position = "top",
   legend.title = element_text(size = 9),
    legend.text = element_text(size = 8) ) +
  labs(title = "Predictive ability under CV1 and CV2")
p_cv12
ggsave(file.path(figures_dir, "prediction_cv1_cv2.tiff"),
 plot = p_cv12,
 width = 14,
 height = 8,
 dpi = 300,
  bg = "white")
```

The second figure summarizes predictive ability under `CV0` and `CV00`, keeping the held-out environments explicit in the layout so that environmental differences remain visible.

```{r figure-cvloo}
cvloo_plot_data <- cvloo_summary %>%
  mutate( ShortDescription = factor(ShortDescription, levels = model_catalog$ShortDescription),
    Env_leave = factor(Env_leave, levels = c("CS11_WS", "CS11_WW", "CS12_WS", "CS12_WW")) )
lower_limit_cvloo <- min(cvloo_plot_data$MeanCorrelation - cvloo_plot_data$SDCorrelation, na.rm = TRUE)
upper_limit_cvloo <- max(cvloo_plot_data$MeanCorrelation + cvloo_plot_data$SDCorrelation, na.rm = TRUE)
p_cvloo <- ggplot(cvloo_plot_data, aes(x = ShortDescription, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30") +
  geom_errorbar( aes( ymin = MeanCorrelation - SDCorrelation, ymax = MeanCorrelation + SDCorrelation ),
   width = 0.20,
   linewidth = 0.35,
    na.rm = TRUE ) +
  facet_grid(Env_leave ~ Trait + CV) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  coord_cartesian(ylim = c(lower_limit_cvloo - 0.03, upper_limit_cvloo + 0.03)) +
  scale_x_discrete(name = "Model description") +
  scale_y_continuous(name = "Predictive correlation") +
  ggthemes::scale_fill_gdocs(name = "Comparison group") +
  theme_bw() +
  theme( plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
   strip.text = element_text(face = "bold", size = 9),
   axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
   axis.text.y = element_text(size = 8),
   axis.title = element_text(size = 10),
   legend.position = "top",
   legend.title = element_text(size = 9),
    legend.text = element_text(size = 8) ) +
  labs(title = "Predictive ability under CV0 and CV00")
p_cvloo
ggsave(file.path(figures_dir, "prediction_cv0_cv00.tiff"),
 plot = p_cvloo,
 width = 16,
 height = 12,
 dpi = 300,
  bg = "white")
```

## 16. Final notes

This tutorialized version is aligned with the current prediction workflow based on `analysis_prediction_run_outputs.Rmd`. It expects files in which:

- repetition and fold are encoded in both the directory hierarchy and the filename;
- `CV0` and `CV00` carry the held-out environment label in both the directory hierarchy and the filename;
- prediction CSV files are read recursively from `output/results/`;
- `CV0` and `CV00` still require the `Env == Env_leave` filter for the held-out-environment summaries.

After this stage, the pipeline is ready to move from technical summarization to manuscript-oriented interpretation. The tables written to `output/tables/` are the main bridge between the computational workflow and the final discussion of results.

## Stage navigation

**Previous stage:** [Prediction — Part II](analysis_prediction_run_outputs.html)  
**Next reference in the site:** [Home](index.html)

Outside the site menu, the next analytical step is to use the outputs created here in manuscript-oriented summaries, interpretative tables, and pipeline review materials.
