---
title: "Prediction summaries and figures"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 13,
  fig.height = 8,
  out.width = "100%"
)
```

**Language / Idioma:** [English](visualization.html) | [Português](visualization_pt.html)

This module prepares the prediction summaries used in the manuscript tables, automated result summaries, and final figures. It starts from the prediction-result tables available in `output/tables/`, attaches the model catalog, computes compact comparison tables, and writes the final visualization-oriented outputs.


## 1. Load packages and locate project folders

This step loads the packages used for table manipulation, figure construction, and HTML table formatting. The paths below assume that the document is rendered from the project root.

```{r packages-and-paths}
rm(list = ls())
options(scipen = 999)

library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(ggplot2)
library(ggthemes)
library(knitr)
library(kableExtra)
library(stringr)
library(patchwork)
library(grid)

project_root <- getwd()
results_dir <- file.path(project_root, "output", "results")
figures_dir <- file.path(project_root, "output", "figures")
tables_dir <- file.path(project_root, "output", "tables")

dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tables_dir, recursive = TRUE, showWarnings = FALSE)
```

## 2. Model catalog used in the prediction summaries

The catalog defines the 18 models compared in the prediction study. The column `Model` gives the label used in manuscript tables and figures (`M1`--`M18`), whereas `EtaName` keeps the computational identifier used internally in file names and BGLR run folders (`Eta1`--`Eta18`).

```{r model-catalog}
model_catalog <- tibble(
  ModelNumber = c(1, 10, 3, 13, 5, 16, 2, 11, 4, 14, 6, 17, 7, 12, 8, 15, 9, 18),
  Model = paste0("M", c(1, 10, 3, 13, 5, 16, 2, 11, 4, 14, 6, 17, 7, 12, 8, 15, 9, 18)),
  EtaName = paste0("Eta", c(1, 10, 3, 13, 5, 16, 2, 11, 4, 14, 6, 17, 7, 12, 8, 15, 9, 18)),
  Pathway = c(
    rep("Genomic pathway", 6),
    rep("Phenomic pathway", 6),
    rep("Combined G + P pathway", 6)
  ),
  KernelFamily = c(
    "Linear", "Linear", "Gaussian", "Gaussian", "Arccosine", "Arccosine",
    "Linear", "Linear", "Gaussian", "Gaussian", "Arccosine", "Arccosine",
    "Linear", "Linear", "Gaussian", "Gaussian", "Arccosine", "Arccosine"
  ),
  WeatherStatus = c(
    "Without W", "With W", "Without W", "With W", "Without W", "With W",
    "Without W", "With W", "Without W", "With W", "Without W", "With W",
    "Without W", "With W", "Without W", "With W", "Without W", "With W"
  ),
  Group = c(
    rep("Genomic", 6),
    rep("Phenomic", 6),
    rep("Combined", 6)
  ),
  Description = c(
    "Genomic linear + E", "Genomic linear + W", "Genomic Gaussian + E", "Genomic Gaussian + W", "Genomic arccosine + E", "Genomic arccosine + W",
    "Phenomic linear + E", "Phenomic linear + W", "Phenomic Gaussian + E", "Phenomic Gaussian + W", "Phenomic arccosine + E", "Phenomic arccosine + W",
    "Combined linear + E", "Combined linear + W", "Combined Gaussian + E", "Combined Gaussian + W", "Combined arccosine + E", "Combined arccosine + W"
  ),
  FullEffects = c(
    "E + G + GE",
    "E + W + G + GE + GW",
    "E + GGK + GGKE",
    "E + W + GGK + GGKE + GGKW",
    "E + GAK + GAKE",
    "E + W + GAK + GAKE + GAKW",
    "E + P + PE",
    "E + W + P + PE + PW",
    "E + PGK + PGKE",
    "E + W + PGK + PGKE + PGKW",
    "E + PAK + PAKE",
    "E + W + PAK + PAKE + PAKW",
    "E + G + P + GE + PE",
    "E + W + G + P + GE + PE + GW + PW",
    "E + GGK + PGK + GGKE + PGKE",
    "E + W + GGK + PGK + GGKE + PGKE + GGKW + PGKW",
    "E + GAK + PAK + GAKE + PAKE",
    "E + W + GAK + PAK + GAKE + PAKE + GAKW + PAKW"
  ),
  InterpretiveDescription = c(
    "Linear genomic model with environment and GxE interaction.",
    "Linear genomic-weather model with environment, weather, GxE, and GxW interaction.",
    "Gaussian genomic model with environment and genomic-by-environment interaction.",
    "Gaussian genomic-weather model with environment, weather, genomic-by-environment, and genomic-by-weather interaction.",
    "Arccosine genomic model with environment and genomic-by-environment interaction.",
    "Arccosine genomic-weather model with environment, weather, genomic-by-environment, and genomic-by-weather interaction.",
    "Linear phenomic model with environment and PxE interaction.",
    "Linear phenomic-weather model with environment, weather, PxE, and PxW interaction.",
    "Gaussian phenomic model with environment and phenomic-by-environment interaction.",
    "Gaussian phenomic-weather model with environment, weather, phenomic-by-environment, and phenomic-by-weather interaction.",
    "Arccosine phenomic model with environment and phenomic-by-environment interaction.",
    "Arccosine phenomic-weather model with environment, weather, phenomic-by-environment, and phenomic-by-weather interaction.",
    "Linear combined genomic-phenomic model with environment and both GxE and PxE interaction terms.",
    "Linear combined genomic-phenomic-weather model with environment, weather, GxE, PxE, GxW, and PxW interaction terms.",
    "Gaussian combined genomic-phenomic model with environment and both genomic and phenomic interaction terms.",
    "Gaussian combined genomic-phenomic-weather model with environment, weather, and the full set of genomic and phenomic interaction terms.",
    "Arccosine combined genomic-phenomic model with environment and both genomic and phenomic interaction terms.",
    "Arccosine combined genomic-phenomic-weather model with environment, weather, and the full set of genomic and phenomic interaction terms."
  )
)

model_catalog
```

## 3. Comparison pairs used for contrasts between without W and with W

The pairs below compare the same modeling pathway and kernel family with and without the weather kernel.

```{r comparison-pairs}
comparison_pairs <- tibble(
  Comparison = c(
    "Genomic linear", "Genomic Gaussian", "Genomic arccosine",
    "Phenomic linear", "Phenomic Gaussian", "Phenomic arccosine",
    "Combined linear", "Combined Gaussian", "Combined arccosine"
  ),
  WithoutW = c("M1", "M3", "M5", "M2", "M4", "M6", "M7", "M8", "M9"),
  WithW = c("M10", "M13", "M16", "M11", "M14", "M17", "M12", "M15", "M18")
)

comparison_pairs
```

## 4. Load prediction summary tables

```{r load-visualization-tables, include=FALSE}
cv12_summary <- readr::read_csv(
  file.path(tables_dir, "prediction_cv1_cv2_summary.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(-dplyr::any_of(c(
    "Pathway", "KernelFamily", "WeatherStatus", "Group",
    "FullEffects", "InterpretiveDescription"
  ))) %>%
  dplyr::mutate(
    Model = stringr::str_replace(as.character(Model), "Eta", "M"),
    EtaName = stringr::str_replace(Model, "M", "Eta")
  ) %>%
  dplyr::left_join(
    model_catalog %>%
      dplyr::select(
        ModelNumber, Model, EtaName, Pathway, KernelFamily, WeatherStatus,
        Group, FullEffects, InterpretiveDescription
      ),
    by = c("ModelNumber", "Model", "EtaName")
  ) %>%
  dplyr::mutate(
    KernelFamily = stringr::str_replace_all(
      as.character(KernelFamily),
      paste0("(?i)", "arc", "-", "cosine"),
      "Arccosine"
    ),
    KernelFamily = dplyr::recode(
      KernelFamily,
      "Gaussiano" = "Gaussian",
      .default = KernelFamily
    )
  )

cvloo_summary <- readr::read_csv(
  file.path(tables_dir, "prediction_cv0_cv00_summary.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(-dplyr::any_of(c(
    "Pathway", "KernelFamily", "WeatherStatus", "Group",
    "FullEffects", "InterpretiveDescription"
  ))) %>%
  dplyr::mutate(
    Model = stringr::str_replace(as.character(Model), "Eta", "M"),
    EtaName = stringr::str_replace(Model, "M", "Eta")
  ) %>%
  dplyr::left_join(
    model_catalog %>%
      dplyr::select(
        ModelNumber, Model, EtaName, Pathway, KernelFamily, WeatherStatus,
        Group, FullEffects, InterpretiveDescription
      ),
    by = c("ModelNumber", "Model", "EtaName")
  ) %>%
  dplyr::mutate(
    KernelFamily = stringr::str_replace_all(
      as.character(KernelFamily),
      paste0("(?i)", "arc", "-", "cosine"),
      "Arccosine"
    ),
    KernelFamily = dplyr::recode(
      KernelFamily,
      "Gaussiano" = "Gaussian",
      .default = KernelFamily
    )
  )

cv12_display <- cv12_summary %>%
  dplyr::arrange(Trait, CV, Pathway, KernelFamily, WeatherStatus, ModelNumber) %>%
  dplyr::mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3),
    MeanRuntimeSec = round(MeanRuntimeSec, 2)
  ) %>%
  dplyr::select(
    Trait, CV, Model, EtaName, Pathway, KernelFamily, WeatherStatus,
    FullEffects, MeanCorrelation, SDCorrelation, NumberOfRepetitions, MeanRuntimeSec
  )

cvloo_display <- cvloo_summary %>%
  dplyr::arrange(Trait, CV, Env_leave, Pathway, KernelFamily, WeatherStatus, ModelNumber) %>%
  dplyr::mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3),
    MeanRuntimeSec = round(MeanRuntimeSec, 2)
  ) %>%
  dplyr::select(
    Trait, CV, Env_leave, Model, EtaName, Pathway, KernelFamily, WeatherStatus,
    FullEffects, MeanCorrelation, SDCorrelation, NumberOfRuns, MeanRuntimeSec
  )

cv12_top3 <- cv12_summary %>%
  dplyr::group_by(Trait, CV) %>%
  dplyr::slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3)
  ) %>%
  dplyr::select(Trait, CV, Model, EtaName, FullEffects, MeanCorrelation, SDCorrelation)

cvloo_top3 <- cvloo_summary %>%
  dplyr::group_by(Trait, CV, Env_leave) %>%
  dplyr::slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3)
  ) %>%
  dplyr::select(Trait, CV, Env_leave, Model, EtaName, FullEffects, MeanCorrelation, SDCorrelation)

cv12_w_effect <- cv12_summary %>%
  dplyr::select(Trait, CV, Model, MeanCorrelation) %>%
  dplyr::inner_join(comparison_pairs, by = c("Model" = "WithoutW")) %>%
  dplyr::rename(MeanWithoutW = MeanCorrelation) %>%
  dplyr::inner_join(
    cv12_summary %>%
      dplyr::select(Trait, CV, Model, MeanCorrelation),
    by = c("Trait", "CV", "WithW" = "Model")
  ) %>%
  dplyr::rename(MeanWithW = MeanCorrelation) %>%
  dplyr::mutate(DeltaWithW = MeanWithW - MeanWithoutW) %>%
  dplyr::select(Comparison, Trait, CV, MeanWithoutW, MeanWithW, DeltaWithW)

cvloo_w_effect <- cvloo_summary %>%
  dplyr::select(Trait, CV, Env_leave, Model, MeanCorrelation) %>%
  dplyr::inner_join(comparison_pairs, by = c("Model" = "WithoutW")) %>%
  dplyr::rename(MeanWithoutW = MeanCorrelation) %>%
  dplyr::inner_join(
    cvloo_summary %>%
      dplyr::select(Trait, CV, Env_leave, Model, MeanCorrelation),
    by = c("Trait", "CV", "Env_leave", "WithW" = "Model")
  ) %>%
  dplyr::rename(MeanWithW = MeanCorrelation) %>%
  dplyr::mutate(DeltaWithW = MeanWithW - MeanWithoutW) %>%
  dplyr::select(Comparison, Trait, CV, Env_leave, MeanWithoutW, MeanWithW, DeltaWithW)

cv12_kernel_family_compare <- cv12_summary %>%
  dplyr::select(Trait, CV, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  tidyr::pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  dplyr::mutate(
    Delta_Gaussian_vs_Linear = Gaussian - Linear,
    Delta_Arccosine_vs_Linear = Arccosine - Linear
  ) %>%
  dplyr::arrange(Trait, CV, Pathway, WeatherStatus)

cvloo_kernel_family_compare <- cvloo_summary %>%
  dplyr::select(Trait, CV, Env_leave, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  tidyr::pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  dplyr::mutate(
    Delta_Gaussian_vs_Linear = Gaussian - Linear,
    Delta_Arccosine_vs_Linear = Arccosine - Linear
  ) %>%
  dplyr::arrange(Trait, CV, Env_leave, Pathway, WeatherStatus)
```

```{r save-visualization-tables, include=FALSE}
readr::write_csv(cv12_display, file.path(tables_dir, "prediction_cv1_cv2_display_table.csv"))
readr::write_csv(cvloo_display, file.path(tables_dir, "prediction_cv0_cv00_display_table.csv"))
readr::write_csv(cv12_top3, file.path(tables_dir, "prediction_cv1_cv2_top3.csv"))
readr::write_csv(cvloo_top3, file.path(tables_dir, "prediction_cv0_cv00_top3.csv"))
readr::write_csv(cv12_w_effect, file.path(tables_dir, "prediction_cv1_cv2_w_effect_pairs.csv"))
readr::write_csv(cvloo_w_effect, file.path(tables_dir, "prediction_cv0_cv00_w_effect_pairs.csv"))
readr::write_csv(cv12_kernel_family_compare, file.path(tables_dir, "prediction_cv1_cv2_kernel_family_compare.csv"))
readr::write_csv(cvloo_kernel_family_compare, file.path(tables_dir, "prediction_cv0_cv00_kernel_family_compare.csv"))
```

## 5. Top-3 tables

The next tables show the three best-performing models in each validation setting. The complete model effects are reported through `FullEffects`, while `Model` gives the manuscript label used in figures.

```{r table-cv12-top3}
cv12_top3 %>%
  knitr::kable(
    caption = "Top 3 models for each trait and validation scheme in CV1 and CV2.",
    align = "llllcc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  )
```

```{r table-cvloo-top3}
cvloo_top3 %>%
  knitr::kable(
    caption = "Top 3 models for each trait, validation scheme, and left-out environment in CV0 and CV00.",
    align = "lllllcc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  )
```

## 6. W-effect tables

The next tables summarize paired contrasts between models with and without the weather kernel.

```{r table-cv12-w-effect}
cv12_w_effect %>%
  dplyr::mutate(
    MeanWithoutW = round(MeanWithoutW, 3),
    MeanWithW = round(MeanWithW, 3),
    DeltaWithW = round(DeltaWithW, 3)
  ) %>%
  knitr::kable(
    caption = "Effect of adding W in CV1 and CV2.",
    align = "lllccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

```{r table-cvloo-w-effect}
cvloo_w_effect %>%
  dplyr::mutate(
    MeanWithoutW = round(MeanWithoutW, 3),
    MeanWithW = round(MeanWithW, 3),
    DeltaWithW = round(DeltaWithW, 3)
  ) %>%
  knitr::kable(
    caption = "Effect of adding W in CV0 and CV00.",
    align = "llllccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

## 7. Kernel-family comparison tables

The next tables compare linear, Gaussian, and arccosine kernels within each analytical pathway and weather status.

```{r table-cv12-kernel-family}
cv12_kernel_family_compare %>%
  dplyr::mutate(
    Linear = round(Linear, 3),
    Gaussian = round(Gaussian, 3),
    Arccosine = round(Arccosine, 3),
    Delta_Gaussian_vs_Linear = round(Delta_Gaussian_vs_Linear, 3),
    Delta_Arccosine_vs_Linear = round(Delta_Arccosine_vs_Linear, 3)
  ) %>%
  knitr::kable(
    caption = "Comparison among kernel families in CV1 and CV2.",
    align = "llllccccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

```{r table-cvloo-kernel-family}
cvloo_kernel_family_compare %>%
  dplyr::mutate(
    Linear = round(Linear, 3),
    Gaussian = round(Gaussian, 3),
    Arccosine = round(Arccosine, 3),
    Delta_Gaussian_vs_Linear = round(Delta_Gaussian_vs_Linear, 3),
    Delta_Arccosine_vs_Linear = round(Delta_Arccosine_vs_Linear, 3)
  ) %>%
  knitr::kable(
    caption = "Comparison among kernel families in CV0 and CV00.",
    align = "lllllccccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

## 8. Automated result-oriented summaries

The tables below provide compact automated summaries that can be inspected before writing the final narrative results.

```{r automated-results-text, results='asis'}
cv12_best_models <- cv12_summary %>%
  dplyr::group_by(Trait, CV) %>%
  dplyr::slice_max(order_by = MeanCorrelation, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(MeanCorrelation = round(MeanCorrelation, 3)) %>%
  dplyr::select(Trait, CV, Model, FullEffects, MeanCorrelation)

cvloo_best_models <- cvloo_summary %>%
  dplyr::group_by(Trait, CV, Env_leave) %>%
  dplyr::slice_max(order_by = MeanCorrelation, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(MeanCorrelation = round(MeanCorrelation, 3)) %>%
  dplyr::select(Trait, CV, Env_leave, Model, FullEffects, MeanCorrelation)

cv12_w_summary <- cv12_w_effect %>%
  dplyr::group_by(Trait, CV) %>%
  dplyr::summarise(MeanDeltaWithW = round(mean(DeltaWithW, na.rm = TRUE), 3), .groups = "drop")

cvloo_w_summary <- cvloo_w_effect %>%
  dplyr::group_by(Trait, CV) %>%
  dplyr::summarise(MeanDeltaWithW = round(mean(DeltaWithW, na.rm = TRUE), 3), .groups = "drop")

cat("### Best models in CV1 and CV2\n\n")
cv12_best_models %>%
  knitr::kable(align = "llllc") %>%
  kableExtra::kable_styling(full_width = FALSE, position = "left")

cat("\n\n### Best models in CV0 and CV00\n\n")
cvloo_best_models %>%
  knitr::kable(align = "lllllc") %>%
  kableExtra::kable_styling(full_width = FALSE, position = "left")

cat("\n\n### Mean effect of W\n\n")
dplyr::bind_rows(
  cv12_w_summary,
  cvloo_w_summary
) %>%
  dplyr::arrange(Trait, CV) %>%
  knitr::kable(align = "llc") %>%
  kableExtra::kable_styling(full_width = FALSE, position = "left")
```

## 9. Figures

The final figures summarize predictive ability for the model catalog in `CV1/CV2` and `CV0/CV00`. The x-axis uses `M1`--`M18`; complete model compositions are provided in Supplementary Table S1.

```{r figure-cv12}
group_levels_en <- c("Genomic", "Phenomic", "Combined")
model_levels_en <- model_catalog$Model

cv12_plot_data <- cv12_summary %>%
  dplyr::mutate(
    Group = factor(Group, levels = group_levels_en),
    Model = factor(Model, levels = model_levels_en),
    CV = factor(CV, levels = c("CV1", "CV2"))
  )

plot_gy_cv1 <- cv12_plot_data %>%
  dplyr::filter(Trait == "GY", CV == "CV1") %>%
  ggplot(aes(x = Model, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30", linewidth = 0.25, na.rm = TRUE) +
  geom_errorbar(aes(ymin = MeanCorrelation - SDCorrelation, ymax = MeanCorrelation + SDCorrelation), width = 0.20, linewidth = 0.35, na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.35) +
  scale_x_discrete(name = "Model") +
  scale_y_continuous(name = "Predictive correlation", breaks = seq(0, 1, by = 0.2), expand = expansion(mult = c(0, 0.02))) +
  coord_cartesian(ylim = c(0, 1)) +
  ggthemes::scale_fill_gdocs(
    name = "Comparison group",
    limits = group_levels_en,
    drop = FALSE,
    guide = guide_legend(title.position = "top", nrow = 1, byrow = TRUE)
  ) +
  labs(title = "CV1") +
  theme_bw(base_size = 16) +
  theme(plot.title = element_text(face = "bold", size = 17, hjust = 0.5), axis.text.x = element_text(size = 13, color = "black"), axis.text.y = element_text(size = 12, color = "black"), axis.title = element_text(face = "bold", size = 14), legend.position = "top", legend.title = element_text(face = "bold", size = 14), legend.text = element_text(size = 12), legend.key.width = grid::unit(1.2, "cm"), panel.grid.major.x = element_blank(), panel.grid.minor = element_blank(), plot.margin = margin(8, 8, 8, 8))

plot_gy_cv2 <- cv12_plot_data %>%
  dplyr::filter(Trait == "GY", CV == "CV2") %>%
  ggplot(aes(x = Model, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30", linewidth = 0.25, na.rm = TRUE) +
  geom_errorbar(aes(ymin = MeanCorrelation - SDCorrelation, ymax = MeanCorrelation + SDCorrelation), width = 0.20, linewidth = 0.35, na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.35) +
  scale_x_discrete(name = "Model") +
  scale_y_continuous(name = "Predictive correlation", breaks = seq(0, 1, by = 0.2), expand = expansion(mult = c(0, 0.02))) +
  coord_cartesian(ylim = c(0, 1)) +
  ggthemes::scale_fill_gdocs(
    name = "Comparison group",
    limits = group_levels_en,
    drop = FALSE,
    guide = guide_legend(title.position = "top", nrow = 1, byrow = TRUE)
  ) +
  labs(title = "CV2") +
  theme_bw(base_size = 16) +
  theme(plot.title = element_text(face = "bold", size = 17, hjust = 0.5), axis.text.x = element_text(size = 13, color = "black"), axis.text.y = element_text(size = 12, color = "black"), axis.title = element_text(face = "bold", size = 14), legend.position = "top", legend.title = element_text(face = "bold", size = 14), legend.text = element_text(size = 12), legend.key.width = grid::unit(1.2, "cm"), panel.grid.major.x = element_blank(), panel.grid.minor = element_blank(), plot.margin = margin(8, 8, 8, 8))

plot_kw_cv1 <- cv12_plot_data %>%
  dplyr::filter(Trait == "KW", CV == "CV1") %>%
  ggplot(aes(x = Model, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30", linewidth = 0.25, na.rm = TRUE) +
  geom_errorbar(aes(ymin = MeanCorrelation - SDCorrelation, ymax = MeanCorrelation + SDCorrelation), width = 0.20, linewidth = 0.35, na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.35) +
  scale_x_discrete(name = "Model") +
  scale_y_continuous(name = "Predictive correlation", breaks = seq(0, 1, by = 0.2), expand = expansion(mult = c(0, 0.02))) +
  coord_cartesian(ylim = c(0, 1)) +
  ggthemes::scale_fill_gdocs(
    name = "Comparison group",
    limits = group_levels_en,
    drop = FALSE,
    guide = guide_legend(title.position = "top", nrow = 1, byrow = TRUE)
  ) +
  labs(title = "CV1") +
  theme_bw(base_size = 16) +
  theme(plot.title = element_text(face = "bold", size = 17, hjust = 0.5), axis.text.x = element_text(size = 13, color = "black"), axis.text.y = element_text(size = 12, color = "black"), axis.title = element_text(face = "bold", size = 14), legend.position = "top", legend.title = element_text(face = "bold", size = 14), legend.text = element_text(size = 12), legend.key.width = grid::unit(1.2, "cm"), panel.grid.major.x = element_blank(), panel.grid.minor = element_blank(), plot.margin = margin(8, 8, 8, 8))

plot_kw_cv2 <- cv12_plot_data %>%
  dplyr::filter(Trait == "KW", CV == "CV2") %>%
  ggplot(aes(x = Model, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30", linewidth = 0.25, na.rm = TRUE) +
  geom_errorbar(aes(ymin = MeanCorrelation - SDCorrelation, ymax = MeanCorrelation + SDCorrelation), width = 0.20, linewidth = 0.35, na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.35) +
  scale_x_discrete(name = "Model") +
  scale_y_continuous(name = "Predictive correlation", breaks = seq(0, 1, by = 0.2), expand = expansion(mult = c(0, 0.02))) +
  coord_cartesian(ylim = c(0, 1)) +
  ggthemes::scale_fill_gdocs(
    name = "Comparison group",
    limits = group_levels_en,
    drop = FALSE,
    guide = guide_legend(title.position = "top", nrow = 1, byrow = TRUE)
  ) +
  labs(title = "CV2") +
  theme_bw(base_size = 16) +
  theme(plot.title = element_text(face = "bold", size = 17, hjust = 0.5), axis.text.x = element_text(size = 13, color = "black"), axis.text.y = element_text(size = 12, color = "black"), axis.title = element_text(face = "bold", size = 14), legend.position = "top", legend.title = element_text(face = "bold", size = 14), legend.text = element_text(size = 12), legend.key.width = grid::unit(1.2, "cm"), panel.grid.major.x = element_blank(), panel.grid.minor = element_blank(), plot.margin = margin(8, 8, 8, 8))

prediction_GY_CV1_CV2_combined <- (plot_gy_cv1 + plot_gy_cv2) +
  patchwork::plot_layout(ncol = 2, guides = "collect") +
  patchwork::plot_annotation(title = "Predictive ability for grain yield (GY)", tag_levels = "A") &
  theme(legend.position = "top", plot.title = element_text(face = "bold", size = 18, hjust = 0.5), plot.tag = element_text(face = "bold", size = 18, color = "black"))

prediction_KW_CV1_CV2_combined <- (plot_kw_cv1 + plot_kw_cv2) +
  patchwork::plot_layout(ncol = 2, guides = "collect") +
  patchwork::plot_annotation(title = "Predictive ability for kernel weight (KW)", tag_levels = "A") &
  theme(legend.position = "top", plot.title = element_text(face = "bold", size = 18, hjust = 0.5), plot.tag = element_text(face = "bold", size = 18, color = "black"))

prediction_GY_CV1_CV2_combined
prediction_KW_CV1_CV2_combined
```

```{r save-figure-cv12, include=FALSE}
ggsave(file.path(figures_dir, "prediction_GY_CV1_CV2_combined.tiff"), plot = prediction_GY_CV1_CV2_combined, width = 18, height = 12, dpi = 600, compression = "lzw", bg = "white")
ggsave(file.path(figures_dir, "prediction_KW_CV1_CV2_combined.tiff"), plot = prediction_KW_CV1_CV2_combined, width = 18, height = 12, dpi = 600, compression = "lzw", bg = "white")
```

```{r figure-cvloo}
cvloo_plot_data <- cvloo_summary %>%
  dplyr::mutate(
    Group = factor(Group, levels = group_levels_en),
    Model = factor(Model, levels = model_levels_en),
    Env_leave = factor(Env_leave, levels = c("CS11_WS", "CS11_WW", "CS12_WS", "CS12_WW")),
    CV = factor(CV, levels = c("CV00", "CV0"))
  )

plot_gy_cv00 <- cvloo_plot_data %>%
  dplyr::filter(Trait == "GY", CV == "CV00") %>%
  ggplot(aes(x = Model, y = MeanCorrelation, fill = Group)) +
  geom_col(
    width = 0.75,
    color = "grey30",
    na.rm = TRUE
  ) +
  geom_errorbar(
    aes(
      ymin = MeanCorrelation - SDCorrelation,
      ymax = MeanCorrelation + SDCorrelation
    ),
    width = 0.20,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  facet_wrap(~ Env_leave, ncol = 2) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey40"
  ) +
  scale_x_discrete(name = "Model") +
  scale_y_continuous(
    name = "Predictive correlation",
    breaks = seq(0, 1, by = 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  ggthemes::scale_fill_gdocs(
    name = "Comparison group",
    limits = group_levels_en,
    drop = FALSE,
    guide = guide_legend(title.position = "top", nrow = 1, byrow = TRUE)
  ) +
  labs(title = "Predictive ability in CV00 for grain yield (GY)") +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.text.x = element_text(size = 12, color = "black"),
    axis.text.y = element_text(size = 12, color = "black"),
    axis.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "top",
    legend.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_gy_cv0 <- cvloo_plot_data %>%
  dplyr::filter(Trait == "GY", CV == "CV0") %>%
  ggplot(aes(x = Model, y = MeanCorrelation, fill = Group)) +
  geom_col(
    width = 0.75,
    color = "grey30",
    na.rm = TRUE
  ) +
  geom_errorbar(
    aes(
      ymin = MeanCorrelation - SDCorrelation,
      ymax = MeanCorrelation + SDCorrelation
    ),
    width = 0.20,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  facet_wrap(~ Env_leave, ncol = 2) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey40"
  ) +
  scale_x_discrete(name = "Model") +
  scale_y_continuous(
    name = "Predictive correlation",
    breaks = seq(0, 1, by = 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  ggthemes::scale_fill_gdocs(
    name = "Comparison group",
    limits = group_levels_en,
    drop = FALSE,
    guide = guide_legend(title.position = "top", nrow = 1, byrow = TRUE)
  ) +
  labs(title = "Predictive ability in CV0 for grain yield (GY)") +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.text.x = element_text(size = 12, color = "black"),
    axis.text.y = element_text(size = 12, color = "black"),
    axis.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "top",
    legend.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_kw_cv00 <- cvloo_plot_data %>%
  dplyr::filter(Trait == "KW", CV == "CV00") %>%
  ggplot(aes(x = Model, y = MeanCorrelation, fill = Group)) +
  geom_col(
    width = 0.75,
    color = "grey30",
    na.rm = TRUE
  ) +
  geom_errorbar(
    aes(
      ymin = MeanCorrelation - SDCorrelation,
      ymax = MeanCorrelation + SDCorrelation
    ),
    width = 0.20,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  facet_wrap(~ Env_leave, ncol = 2) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey40"
  ) +
  scale_x_discrete(name = "Model") +
  scale_y_continuous(
    name = "Predictive correlation",
    breaks = seq(0, 1, by = 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  ggthemes::scale_fill_gdocs(
    name = "Comparison group",
    limits = group_levels_en,
    drop = FALSE,
    guide = guide_legend(title.position = "top", nrow = 1, byrow = TRUE)
  ) +
  labs(title = "Predictive ability in CV00 for kernel weight (KW)") +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.text.x = element_text(size = 12, color = "black"),
    axis.text.y = element_text(size = 12, color = "black"),
    axis.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "top",
    legend.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_kw_cv0 <- cvloo_plot_data %>%
  dplyr::filter(Trait == "KW", CV == "CV0") %>%
  ggplot(aes(x = Model, y = MeanCorrelation, fill = Group)) +
  geom_col(
    width = 0.75,
    color = "grey30",
    na.rm = TRUE
  ) +
  geom_errorbar(
    aes(
      ymin = MeanCorrelation - SDCorrelation,
      ymax = MeanCorrelation + SDCorrelation
    ),
    width = 0.20,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  facet_wrap(~ Env_leave, ncol = 2) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey40"
  ) +
  scale_x_discrete(name = "Model") +
  scale_y_continuous(
    name = "Predictive correlation",
    breaks = seq(0, 1, by = 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  ggthemes::scale_fill_gdocs(
    name = "Comparison group",
    limits = group_levels_en,
    drop = FALSE,
    guide = guide_legend(title.position = "top", nrow = 1, byrow = TRUE)
  ) +
  labs(title = "Predictive ability in CV0 for kernel weight (KW)") +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.text.x = element_text(size = 12, color = "black"),
    axis.text.y = element_text(size = 12, color = "black"),
    axis.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "top",
    legend.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_gy_cv00
plot_gy_cv0
plot_kw_cv00
plot_kw_cv0
```

```{r save-figure-cvloo, include=FALSE}
ggsave(file.path(figures_dir, "prediction_GY_CV00.tiff"), plot = plot_gy_cv00, width = 18, height = 12, dpi = 600, compression = "lzw", bg = "white")
ggsave(file.path(figures_dir, "prediction_GY_CV0.tiff"), plot = plot_gy_cv0, width = 18, height = 12, dpi = 600, compression = "lzw", bg = "white")
ggsave(file.path(figures_dir, "prediction_KW_CV00.tiff"), plot = plot_kw_cv00, width = 18, height = 12, dpi = 600, compression = "lzw", bg = "white")
ggsave(file.path(figures_dir, "prediction_KW_CV0.tiff"), plot = plot_kw_cv0, width = 18, height = 12, dpi = 600, compression = "lzw", bg = "white")
```

## 10. Final note

The visualization step provides the final tables and figures used to interpret the predictive behavior of genomic, phenomic, integrated, weather-assisted, Gaussian, and arccosine models. The model codes shown in the figures correspond to Supplementary Table S1.

## Step navigation

Previous step: [Prediction execution and output organization](analysis_prediction_run_outputs.html)  
Next step: manuscript-oriented result interpretation.
