---
title: "Prediction summaries and figures - Didactic Tutorial"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
---

This file is the **didactic version** of the final visualization script.

Its job is to read the prediction outputs generated by `analysis_prediction.Rmd`, reorganize them into tidy summary tables, and then build the figures and compact displays used in the downstream manuscript workflow.

The current prediction workflow saves **aggregated results by repetition** in `output/results/`:

- `CV1_<trait>_<model>_rep<rep>.csv`
- `CV2_<trait>_<model>_rep<rep>.csv`
- `CV0_<trait>_<env1>_<env2>_<model>_rep<rep>.csv`
- `CV00_<trait>_<env1>_<env2>_<model>_rep<rep>.csv`

Each CSV already contains the environment-level correlations accumulated across folds inside the repetition. Therefore, this script does **not** expect filenames with `fold` in the name, and it does **not** expect raw observation-level columns such as `ObsID`, `observed`, and `predicted`.

### What this script does

The workflow is divided into four practical tasks:

1. locate and read the prediction CSV files;
2. parse the filenames and recover trait, model, repetition, and leave-one-environment-out information;
3. summarize the results for `CV1`, `CV2`, `CV0`, and `CV00`;
4. export cleaned tables and figures for the next manuscript stage.

This didactic version keeps the same logic as the final operational file, but presents the code in a more guided format for study and maintenance.

## 1. Setup

We start by defining chunk defaults for the knitted report. The goal is to make code, tables, and figures readable when the tutorial is rendered as HTML.

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 13,
  fig.height = 8,
  out.width = "100%"
)
```

## 2. Load packages and locate project folders

This section loads the packages used for data manipulation, plotting, and table formatting. It also checks whether the script is being run from the project root or from the `analysis/` folder and then standardizes the output paths.

```{r packages-and-paths}
rm(list = ls())
options(scipen = 999)

packages_needed <- c(
  "dplyr", "tidyr", "readr", "tibble", "ggplot2",
  "ggthemes", "knitr", "kableExtra", "stringr"
)

packages_missing <- packages_needed[!(packages_needed %in% installed.packages()[, "Package"])]
if (length(packages_missing) > 0) {
  install.packages(packages_missing)
}

library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(ggplot2)
library(ggthemes)
library(knitr)
library(kableExtra)
library(stringr)

project_root <- getwd()

if (!dir.exists(file.path(project_root, "output")) &&
    dir.exists(file.path(project_root, "..", "output"))) {
  project_root <- normalizePath(file.path(project_root, ".."))
}

if (!dir.exists(file.path(project_root, "output"))) {
  stop("Could not find the folder 'output/'. Run this script from the project root or from analysis/.")
}

results_dir <- file.path(project_root, "output", "results")
figures_dir <- file.path(project_root, "output", "figures")
tables_dir <- file.path(project_root, "output", "tables")

dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tables_dir, recursive = TRUE, showWarnings = FALSE)
```

## 3. Model catalog used in the reduced pipeline

The visualization stage uses the same reduced 18-model catalog adopted in the final prediction script. Here the catalog is reorganized to emphasize pathway, kernel family, and the presence or absence of weather information.

```{r model-catalog}
model_catalog <- tibble(
  ModelNumber = c(1, 10, 3, 13, 5, 16,
                  2, 11, 4, 14, 6, 17,
                  7, 12, 8, 15, 9, 18),
  Model = sprintf("M%02d", c(1, 10, 3, 13, 5, 16,
                             2, 11, 4, 14, 6, 17,
                             7, 12, 8, 15, 9, 18)),
  Pathway = c(
    rep("Genomic pathway", 6),
    rep("Phenomic pathway", 6),
    rep("Combined G + P pathway", 6)
  ),
  KernelFamily = c(
    "Linear", "Linear",
    "Gaussian", "Gaussian",
    "Arc-cosine", "Arc-cosine",
    "Linear", "Linear",
    "Gaussian", "Gaussian",
    "Arc-cosine", "Arc-cosine",
    "Linear", "Linear",
    "Gaussian", "Gaussian",
    "Arc-cosine", "Arc-cosine"
  ),
  WeatherStatus = c(
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W",
    "Without W", "With W"
  ),
  Group = c(
    "Genomic baseline",
    "Weather single-source",
    "Genomic baseline",
    "Weather single-source",
    "Genomic baseline",
    "Weather single-source",
    "Phenomic baseline",
    "Weather single-source",
    "Phenomic baseline",
    "Weather single-source",
    "Phenomic baseline",
    "Weather single-source",
    "Combined baseline",
    "Weather combined",
    "Combined baseline",
    "Weather combined",
    "Combined baseline",
    "Weather combined"
  ),
  Description = c(
    "Genomic linear + E",
    "Genomic linear + W",
    "Genomic Gaussian + E",
    "Genomic Gaussian + W",
    "Genomic arc-cosine + E",
    "Genomic arc-cosine + W",
    "Phenomic linear + E",
    "Phenomic linear + W",
    "Phenomic Gaussian + E",
    "Phenomic Gaussian + W",
    "Phenomic arc-cosine + E",
    "Phenomic arc-cosine + W",
    "Combined linear + E",
    "Combined linear + W",
    "Combined Gaussian + E",
    "Combined Gaussian + W",
    "Combined arc-cosine + E",
    "Combined arc-cosine + W"
  ),
  ShortDescription = c(
    "G + E",
    "G + W",
    "GGK + E",
    "GGK + W",
    "GAK + E",
    "GAK + W",
    "P + E",
    "P + W",
    "PGK + E",
    "PGK + W",
    "PAK + E",
    "PAK + W",
    "G+P + E",
    "G+P + W",
    "GGK+PGK + E",
    "GGK+PGK + W",
    "GAK+PAK + E",
    "GAK+PAK + W"
  )
)

model_catalog
```

## 4. Comparison pairs used for W vs no-W comparisons

These pairs are used later to quantify the gain or loss associated with adding weather information (`W`) within the same pathway and kernel family.

```{r comparison-pairs}
comparison_pairs <- tibble(
  Comparison = c(
    "Genomic linear",
    "Genomic Gaussian",
    "Genomic arc-cosine",
    "Phenomic linear",
    "Phenomic Gaussian",
    "Phenomic arc-cosine",
    "Combined linear",
    "Combined Gaussian",
    "Combined arc-cosine"
  ),
  WithoutW = c("M01", "M03", "M05", "M02", "M04", "M06", "M07", "M08", "M09"),
  WithW = c("M10", "M13", "M16", "M11", "M14", "M17", "M12", "M15", "M18")
)

comparison_pairs
```

## 5. Locate prediction files

Before reading any data, the script scans `output/results/` and separates files into two groups: `CV1/CV2` and `CV0/CV00`.

```{r locate-files}
all_result_files <- list.files(results_dir, pattern = "\\.csv$", full.names = TRUE)
all_result_names <- basename(all_result_files)

cv12_files <- all_result_files[grepl("^CV1_|^CV2_", all_result_names)]
cvloo_files <- all_result_files[grepl("^CV0_|^CV00_", all_result_names)]

cat("Number of CV1/CV2 files:", length(cv12_files), "\n")
cat("Number of CV0/CV00 files:", length(cvloo_files), "\n")

if (length(cv12_files) == 0 && length(cvloo_files) == 0) {
  stop("No prediction CSV files were found in output/results/. Run analysis_prediction first.")
}
```

## 6. Read CV1 and CV2 files generated by the final prediction script

In this step, each filename is parsed to recover the validation scheme, trait, model, and repetition. The fold information is not in the filename anymore; it is recovered from the `.id` column stored inside each CSV.

```{r read-cv12}
cv12_list <- vector("list", length(cv12_files))

for (i in seq_along(cv12_files)) {
  current_file <- cv12_files[i]
  file_name <- basename(current_file)
  file_name <- sub("\\.csv$", "", file_name)
  parts <- strsplit(file_name, "_")[[1]]

  if (length(parts) != 4) {
    stop(
      "Unexpected filename for CV1/CV2 file: ", file_name,
      ". Expected pattern CV1_<trait>_<model>_rep<rep>.csv or CV2_<trait>_<model>_rep<rep>.csv"
    )
  }

  current_cv <- parts[1]
  current_trait <- parts[2]
  current_model_label <- parts[3]
  current_model_number <- as.numeric(gsub("M", "", current_model_label))
  current_rep <- as.numeric(gsub("rep", "", parts[4]))

  current_data <- readr::read_csv(current_file, show_col_types = FALSE)

  if (!(".id" %in% names(current_data)) || !("Env" %in% names(current_data)) || !("cor" %in% names(current_data))) {
    stop(
      "Unexpected column structure in file: ", file_name,
      ". The final prediction script should export columns including .id, Env, and cor."
    )
  }

  current_data <- current_data %>%
    rename(Fold = .id) %>%
    mutate(
      Fold = suppressWarnings(as.integer(Fold)),
      CV = current_cv,
      Trait = current_trait,
      ModelNumber = current_model_number,
      Model = current_model_label,
      Rep = current_rep
    )

  cv12_list[[i]] <- current_data
}

cv12_raw <- bind_rows(cv12_list) %>%
  left_join(model_catalog, by = c("ModelNumber", "Model"))

readr::write_csv(cv12_raw, file.path(tables_dir, "prediction_cv1_cv2_raw.csv"))
```

## 7. Read CV0 and CV00 files generated by the final prediction script

The logic is analogous to the previous step, but now the filename also carries the held-out environment label. Because the environment names themselves contain an underscore (for example, `CS11_WS`), the parser reads this label as two filename tokens and then recombines them into a single `Env_leave` value.

```{r read-cvloo}
cvloo_list <- vector("list", length(cvloo_files))

for (i in seq_along(cvloo_files)) {
  current_file <- cvloo_files[i]
  file_name <- basename(current_file)
  file_name <- sub("\\.csv$", "", file_name)
  parts <- strsplit(file_name, "_")[[1]]

  if (length(parts) != 6) {
    stop(
      "Unexpected filename for CV0/CV00 file: ", file_name,
      ". Expected pattern CV0_<trait>_<env_part1>_<env_part2>_<model>_rep<rep>.csv or CV00_<trait>_<env_part1>_<env_part2>_<model>_rep<rep>.csv, where <env_part1>_<env_part2> forms one held-out environment label such as CS11_WS."
    )
  }

  current_cv <- parts[1]
  current_trait <- parts[2]
  current_env_part1 <- parts[3]
  current_env_part2 <- parts[4]
  current_env_leave <- paste(current_env_part1, current_env_part2, sep = "_")
  current_model_label <- parts[5]
  current_model_number <- as.numeric(gsub("M", "", current_model_label))
  current_rep <- as.numeric(gsub("rep", "", parts[6]))

  current_data <- readr::read_csv(current_file, show_col_types = FALSE)

  if (!(".id" %in% names(current_data)) || !("Env" %in% names(current_data)) || !("cor" %in% names(current_data))) {
    stop(
      "Unexpected column structure in file: ", file_name,
      ". The final prediction script should export columns including .id, Env, and cor."
    )
  }

  current_data <- current_data %>%
    rename(Fold = .id) %>%
    mutate(
      Fold = suppressWarnings(as.integer(Fold)),
      CV = current_cv,
      Trait = current_trait,
      Env_leave = current_env_leave,
      ModelNumber = current_model_number,
      Model = current_model_label,
      Rep = current_rep
    )

  cvloo_list[[i]] <- current_data
}

cvloo_raw <- bind_rows(cvloo_list) %>%
  left_join(model_catalog, by = c("ModelNumber", "Model"))

readr::write_csv(cvloo_raw, file.path(tables_dir, "prediction_cv0_cv00_raw.csv"))
readr::write_csv(model_catalog, file.path(tables_dir, "prediction_model_catalog.csv"))
```

## 8. Summarize CV1 and CV2

The summary is built progressively:

1. first by environment and fold;
2. then by environment within repetition;
3. then by repetition;
4. and finally by model-level means and standard deviations across repetitions.

### 8.1 Summarize CV1 and CV2 by environment and fold

This first summary keeps the most detailed analytical level used later in the script: each environment inside each fold of each repetition.

```{r summarize-cv12-env-fold}
cv12_env_fold <- cv12_raw %>%
  group_by(
    ModelNumber, Model, Group, Pathway, KernelFamily, WeatherStatus,
    Description, ShortDescription, Trait, CV, Rep, Fold, Env
  ) %>%
  summarise(
    EnvCorrelation = mean(cor, na.rm = TRUE),
    ValidationCount = sum(!is.na(cor)),
    MeanRuntimeSec = mean(Runtime_sec, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    EnvCorrelation = ifelse(is.nan(EnvCorrelation), NA_real_, EnvCorrelation),
    MeanRuntimeSec = ifelse(is.nan(MeanRuntimeSec), NA_real_, MeanRuntimeSec)
  ) %>%
  filter(!is.na(EnvCorrelation), ValidationCount > 0)
```

### 8.2 Summarize CV1 and CV2 by environment within repetition

Now the script combines all folds of the same repetition for each environment using the validation counts as weights.

```{r summarize-cv12-env-rep}
cv12_env_rep <- cv12_env_fold %>%
  group_by(
    ModelNumber, Model, Group, Pathway, KernelFamily, WeatherStatus,
    Description, ShortDescription, Trait, CV, Rep, Env
  ) %>%
  summarise(
    EnvCorrelation = weighted.mean(EnvCorrelation, w = ValidationCount, na.rm = TRUE),
    ValidationCount = sum(ValidationCount, na.rm = TRUE),
    MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop"
  )
```

### 8.3 Summarize CV1 and CV2 by repetition

This step collapses the environment-level information into one weighted correlation per repetition.

```{r summarize-cv12-rep}
cv12_rep_summary <- cv12_env_rep %>%
  group_by(
    ModelNumber, Model, Group, Pathway, KernelFamily, WeatherStatus,
    Description, ShortDescription, Trait, CV, Rep
  ) %>%
  summarise(
    WeightedCorrelation = weighted.mean(EnvCorrelation, w = ValidationCount, na.rm = TRUE),
    MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop"
  )
```

### 8.4 Build the final CV1 and CV2 summary table

The final table reports the average correlation and its standard deviation across repetitions for each model.

```{r summarize-cv12-final}
cv12_summary <- cv12_rep_summary %>%
  group_by(
    ModelNumber, Model, Group, Pathway, KernelFamily, WeatherStatus,
    Description, ShortDescription, Trait, CV
  ) %>%
  summarise(
    MeanCorrelation = mean(WeightedCorrelation, na.rm = TRUE),
    SDCorrelation = sd(WeightedCorrelation, na.rm = TRUE),
    NumberOfRepetitions = n(),
    MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    SDCorrelation = ifelse(is.na(SDCorrelation), 0, SDCorrelation)
  )
```

### 8.5 Export the CV1 and CV2 summary tables

These files are later reused by the manuscript script and also serve as intermediate checkpoints for debugging.

```{r export-cv12-summaries}
readr::write_csv(cv12_env_fold, file.path(tables_dir, "prediction_cv1_cv2_by_rep_fold_env.csv"))
readr::write_csv(cv12_env_rep, file.path(tables_dir, "prediction_cv1_cv2_by_rep_env.csv"))
readr::write_csv(cv12_rep_summary, file.path(tables_dir, "prediction_cv1_cv2_by_rep.csv"))
readr::write_csv(cv12_summary, file.path(tables_dir, "prediction_cv1_cv2_summary.csv"))
```

## 9. Summarize CV0 and CV00

For `CV0` and `CV00`, the script keeps only the rows where the environment in the data matches the environment that was left out. This ensures that the reported correlation truly refers to the leave-one-environment-out target.

### 9.1 Keep only the leave-one-environment-out rows

This filter is conceptually important because the raw files may contain rows from multiple environments, but the reported result must correspond only to the target environment that was masked during prediction.

```{r filter-cvloo-target-env}
cvloo_env_filtered <- cvloo_raw %>%
  mutate(
    Env = as.character(Env),
    Env_leave = as.character(Env_leave)
  ) %>%
  filter(Env == Env_leave)
```

### 9.2 Summarize CV0 and CV00 by environment and fold

At this level the script keeps each fold separate before combining repetitions.

```{r summarize-cvloo-env-fold}
cvloo_env_fold <- cvloo_env_filtered %>%
  group_by(
    ModelNumber, Model, Group, Pathway, KernelFamily, WeatherStatus,
    Description, ShortDescription, Trait, CV, Env_leave, Rep, Fold, Env
  ) %>%
  summarise(
    LOEOCorrelation = mean(cor, na.rm = TRUE),
    ValidationCount = sum(!is.na(cor)),
    MeanRuntimeSec = mean(Runtime_sec, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    LOEOCorrelation = ifelse(is.nan(LOEOCorrelation), NA_real_, LOEOCorrelation),
    MeanRuntimeSec = ifelse(is.nan(MeanRuntimeSec), NA_real_, MeanRuntimeSec)
  ) %>%
  filter(!is.na(LOEOCorrelation), ValidationCount > 0)
```

### 9.3 Summarize CV0 and CV00 by repetition

Now the fold-level results are combined inside each repetition using the validation counts as weights.

```{r summarize-cvloo-rep}
cvloo_rep_summary <- cvloo_env_fold %>%
  group_by(
    ModelNumber, Model, Group, Pathway, KernelFamily, WeatherStatus,
    Description, ShortDescription, Trait, CV, Env_leave, Rep
  ) %>%
  summarise(
    LOEOCorrelation = weighted.mean(LOEOCorrelation, w = ValidationCount, na.rm = TRUE),
    ValidationCount = sum(ValidationCount, na.rm = TRUE),
    MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop"
  )
```

### 9.4 Build the final CV0 and CV00 summary table

This table is the main source for the leave-one-environment-out summaries used later in the manuscript and the figures.

```{r summarize-cvloo-final}
cvloo_summary <- cvloo_rep_summary %>%
  group_by(
    ModelNumber, Model, Group, Pathway, KernelFamily, WeatherStatus,
    Description, ShortDescription, Trait, CV, Env_leave
  ) %>%
  summarise(
    MeanCorrelation = mean(LOEOCorrelation, na.rm = TRUE),
    SDCorrelation = sd(LOEOCorrelation, na.rm = TRUE),
    NumberOfRuns = n(),
    MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    SDCorrelation = ifelse(is.na(SDCorrelation), 0, SDCorrelation)
  )
```

### 9.5 Export the CV0 and CV00 summary tables

These files mirror the CV1/CV2 exports, but now for the leave-one-environment-out scenarios.

```{r export-cvloo-summaries}
readr::write_csv(cvloo_env_fold, file.path(tables_dir, "prediction_cv0_cv00_by_rep_fold_env.csv"))
readr::write_csv(cvloo_rep_summary, file.path(tables_dir, "prediction_cv0_cv00_by_run.csv"))
readr::write_csv(cvloo_summary, file.path(tables_dir, "prediction_cv0_cv00_summary.csv"))
```

## 10. Comparison tables

This section derives secondary analytical tables, especially the contrast between models with and without `W` and the comparison among kernel families inside each pathway.

### 10.1 Compare models with and without W for CV1 and CV2

The comparison pairs were defined earlier in the script. Here the summary table is joined twice: once for the model without weather information and once for the corresponding model with weather information.

```{r comparison-cv12-w-effect}
cv12_w_effect <- comparison_pairs %>%
  left_join(
    cv12_summary %>%
      select(Trait, CV, Model, MeanCorrelation, SDCorrelation),
    by = c("WithoutW" = "Model")
  ) %>%
  rename(
    MeanWithoutW = MeanCorrelation,
    SDWithoutW = SDCorrelation
  ) %>%
  left_join(
    cv12_summary %>%
      select(Trait, CV, Model, MeanCorrelation, SDCorrelation),
    by = c("WithW" = "Model", "Trait", "CV")
  ) %>%
  rename(
    MeanWithW = MeanCorrelation,
    SDWithW = SDCorrelation
  ) %>%
  mutate(
    DeltaWithW = MeanWithW - MeanWithoutW
  ) %>%
  arrange(Trait, CV, Comparison)
```

### 10.2 Compare models with and without W for CV0 and CV00

This is the same analytical idea as above, but now conditioned on the environment that was left out.

```{r comparison-cvloo-w-effect}
cvloo_w_effect <- comparison_pairs %>%
  left_join(
    cvloo_summary %>%
      select(Trait, CV, Env_leave, Model, MeanCorrelation, SDCorrelation),
    by = c("WithoutW" = "Model")
  ) %>%
  rename(
    MeanWithoutW = MeanCorrelation,
    SDWithoutW = SDCorrelation
  ) %>%
  left_join(
    cvloo_summary %>%
      select(Trait, CV, Env_leave, Model, MeanCorrelation, SDCorrelation),
    by = c("WithW" = "Model", "Trait", "CV", "Env_leave")
  ) %>%
  rename(
    MeanWithW = MeanCorrelation,
    SDWithW = SDCorrelation
  ) %>%
  mutate(
    DeltaWithW = MeanWithW - MeanWithoutW
  ) %>%
  arrange(Trait, CV, Env_leave, Comparison)
```

### 10.3 Compare kernel families for CV1 and CV2

This table places the linear, Gaussian, and arc-cosine kernels side by side within each pathway and weather configuration.

```{r comparison-cv12-kernel-family}
cv12_kernel_family_compare <- cv12_summary %>%
  select(Trait, CV, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  mutate(
    Delta_Gaussian_vs_Linear = Gaussian - Linear,
    Delta_ArcCos_vs_Linear = `Arc-cosine` - Linear,
    Delta_ArcCos_vs_Gaussian = `Arc-cosine` - Gaussian
  ) %>%
  arrange(Trait, CV, Pathway, WeatherStatus)
```

### 10.4 Compare kernel families for CV0 and CV00

The same kernel comparison is repeated here for each leave-one-environment-out target.

```{r comparison-cvloo-kernel-family}
cvloo_kernel_family_compare <- cvloo_summary %>%
  select(Trait, CV, Env_leave, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  mutate(
    Delta_Gaussian_vs_Linear = Gaussian - Linear,
    Delta_ArcCos_vs_Linear = `Arc-cosine` - Linear,
    Delta_ArcCos_vs_Gaussian = `Arc-cosine` - Gaussian
  ) %>%
  arrange(Trait, CV, Env_leave, Pathway, WeatherStatus)
```

### 10.5 Export the comparison tables

These outputs are later reused directly in manuscript reporting and result interpretation.

```{r export-comparison-tables}
readr::write_csv(cv12_w_effect, file.path(tables_dir, "prediction_cv1_cv2_w_effect_pairs.csv"))
readr::write_csv(cvloo_w_effect, file.path(tables_dir, "prediction_cv0_cv00_w_effect_pairs.csv"))
readr::write_csv(cv12_kernel_family_compare, file.path(tables_dir, "prediction_cv1_cv2_kernel_family_compare.csv"))
readr::write_csv(cvloo_kernel_family_compare, file.path(tables_dir, "prediction_cv0_cv00_kernel_family_compare.csv"))
```

## 11. Compact display tables

These objects are simplified versions of the full summaries. They are useful for manuscript tables, supplementary material, and quick inspection of the best-performing models.

### 11.1 Build the compact CV1 and CV2 display table

This table keeps the main summary columns but rounds the numerical fields for easier reading.

```{r compact-cv12-table}
cv12_table_display <- cv12_summary %>%
  arrange(Trait, CV, Pathway, KernelFamily, WeatherStatus) %>%
  mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3),
    MeanRuntimeSec = round(MeanRuntimeSec, 2)
  ) %>%
  select(
    Trait, CV, Pathway, KernelFamily, WeatherStatus,
    ShortDescription, MeanCorrelation, SDCorrelation,
    NumberOfRepetitions, MeanRuntimeSec
  )
```

### 11.2 Build the compact CV0 and CV00 display table

The leave-one-environment-out display table keeps the environment column because it is part of the interpretation.

```{r compact-cvloo-table}
cvloo_table_display <- cvloo_summary %>%
  arrange(Trait, CV, Env_leave, Pathway, KernelFamily, WeatherStatus) %>%
  mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3),
    MeanRuntimeSec = round(MeanRuntimeSec, 2)
  ) %>%
  select(
    Trait, CV, Env_leave, Pathway, KernelFamily, WeatherStatus,
    ShortDescription, MeanCorrelation, SDCorrelation,
    NumberOfRuns, MeanRuntimeSec
  )
```

### 11.3 Build the top-3 objects for CV1 and CV2

These objects are displayed in the next section and are also exported for downstream use.

```{r build-cv12-top3}
cv12_top3 <- cv12_summary %>%
  group_by(Trait, CV) %>%
  slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3)
  ) %>%
  select(Trait, CV, ShortDescription, MeanCorrelation, SDCorrelation)
```

### 11.4 Build the top-3 objects for CV0 and CV00

The leave-one-environment-out top-3 tables are stratified by target environment.

```{r build-cvloo-top3}
cvloo_top3 <- cvloo_summary %>%
  group_by(Trait, CV, Env_leave) %>%
  slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    MeanCorrelation = round(MeanCorrelation, 3),
    SDCorrelation = round(SDCorrelation, 3)
  ) %>%
  select(Trait, CV, Env_leave, ShortDescription, MeanCorrelation, SDCorrelation)
```

### 11.5 Export the compact and top-3 tables

All derived display-oriented tables are written to disk here so that later scripts can reuse them directly.

```{r export-compact-and-top3}
readr::write_csv(cv12_table_display, file.path(tables_dir, "prediction_cv1_cv2_display_table.csv"))
readr::write_csv(cvloo_table_display, file.path(tables_dir, "prediction_cv0_cv00_display_table.csv"))
readr::write_csv(cv12_top3, file.path(tables_dir, "prediction_cv1_cv2_top3.csv"))
readr::write_csv(cvloo_top3, file.path(tables_dir, "prediction_cv0_cv00_top3.csv"))
```

## 12. Top-3 tables

The next two chunks display the three best-performing models within each validation setting.

```{r table-cv12-top3}
cv12_top3 %>%
  knitr::kable(
    caption = "Top 3 models for each trait and validation scheme in CV1 and CV2.",
    align = "llccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  )
```

```{r table-cvloo-top3}
cvloo_top3 %>%
  knitr::kable(
    caption = "Top 3 models for each trait, validation scheme, and held-out environment in CV0 and CV00.",
    align = "lllccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  )
```

## 13. Weather-effect tables

These tables present the pairwise comparisons between models with and without weather information.

```{r table-cv12-w-effect}
cv12_w_effect %>%
  mutate(
    MeanWithoutW = round(MeanWithoutW, 3),
    MeanWithW = round(MeanWithW, 3),
    DeltaWithW = round(DeltaWithW, 3)
  ) %>%
  knitr::kable(
    caption = "Pairwise comparison of models without and with the weather kernel (W) in CV1 and CV2.",
    align = "lllccccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

```{r table-cvloo-w-effect}
cvloo_w_effect %>%
  mutate(
    MeanWithoutW = round(MeanWithoutW, 3),
    MeanWithW = round(MeanWithW, 3),
    DeltaWithW = round(DeltaWithW, 3)
  ) %>%
  knitr::kable(
    caption = "Pairwise comparison of models without and with the weather kernel (W) in CV0 and CV00.",
    align = "llllccccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

## 14. Kernel-family comparison tables

```{r table-cv12-kernel-family}
cv12_kernel_family_compare %>%
  mutate(
    Linear = round(Linear, 3),
    Gaussian = round(Gaussian, 3),
    `Arc-cosine` = round(`Arc-cosine`, 3),
    Delta_Gaussian_vs_Linear = round(Delta_Gaussian_vs_Linear, 3),
    Delta_ArcCos_vs_Linear = round(Delta_ArcCos_vs_Linear, 3),
    Delta_ArcCos_vs_Gaussian = round(Delta_ArcCos_vs_Gaussian, 3)
  ) %>%
  knitr::kable(
    caption = "Comparison of kernel families within each analytical pathway in CV1 and CV2.",
    align = "llllcccccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

```{r table-cvloo-kernel-family}
cvloo_kernel_family_compare %>%
  mutate(
    Linear = round(Linear, 3),
    Gaussian = round(Gaussian, 3),
    `Arc-cosine` = round(`Arc-cosine`, 3),
    Delta_Gaussian_vs_Linear = round(Delta_Gaussian_vs_Linear, 3),
    Delta_ArcCos_vs_Linear = round(Delta_ArcCos_vs_Linear, 3),
    Delta_ArcCos_vs_Gaussian = round(Delta_ArcCos_vs_Gaussian, 3)
  ) %>%
  knitr::kable(
    caption = "Comparison of kernel families within each analytical pathway in CV0 and CV00.",
    align = "lllllcccccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

## 15. Figures

```{r figure-cv12}
cv12_plot_data <- cv12_summary %>%
  mutate(
    ShortDescription = factor(ShortDescription, levels = model_catalog$ShortDescription)
  )

lower_limit_cv12 <- min(cv12_plot_data$MeanCorrelation - cv12_plot_data$SDCorrelation, na.rm = TRUE)
upper_limit_cv12 <- max(cv12_plot_data$MeanCorrelation + cv12_plot_data$SDCorrelation, na.rm = TRUE)

p_cv12 <- ggplot(cv12_plot_data, aes(x = ShortDescription, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30") +
  geom_errorbar(
    aes(
      ymin = MeanCorrelation - SDCorrelation,
      ymax = MeanCorrelation + SDCorrelation
    ),
    width = 0.20,
    linewidth = 0.35
  ) +
  facet_grid(Trait ~ CV) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  coord_cartesian(ylim = c(lower_limit_cv12 - 0.03, upper_limit_cv12 + 0.03)) +
  scale_x_discrete(name = "Model description") +
  scale_y_continuous(name = "Predictive correlation") +
  ggthemes::scale_fill_gdocs(name = "Comparison group") +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
    strip.text = element_text(face = "bold", size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7),
    axis.text.y = element_text(size = 9),
    axis.title = element_text(size = 10),
    legend.position = "top",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  ) +
  labs(title = "Predictive ability under CV1 and CV2")

p_cv12

ggsave(file.path(figures_dir, "prediction_cv1_cv2.tiff"), plot = p_cv12, width = 14, height = 8, dpi = 300, bg = "white")
ggsave(file.path(figures_dir, "prediction_cv1_cv2.pdf"), plot = p_cv12, width = 14, height = 8, bg = "white")
```

```{r figure-cvloo}
cvloo_plot_data <- cvloo_summary %>%
  mutate(
    ShortDescription = factor(ShortDescription, levels = model_catalog$ShortDescription),
    Env_leave = factor(Env_leave, levels = c("CS11_WS", "CS11_WW", "CS12_WS", "CS12_WW"))
  )

lower_limit_cvloo <- min(cvloo_plot_data$MeanCorrelation - cvloo_plot_data$SDCorrelation, na.rm = TRUE)
upper_limit_cvloo <- max(cvloo_plot_data$MeanCorrelation + cvloo_plot_data$SDCorrelation, na.rm = TRUE)

p_cvloo <- ggplot(cvloo_plot_data, aes(x = ShortDescription, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30") +
  geom_errorbar(
    aes(
      ymin = MeanCorrelation - SDCorrelation,
      ymax = MeanCorrelation + SDCorrelation
    ),
    width = 0.20,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  facet_grid(Env_leave ~ Trait + CV) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  coord_cartesian(ylim = c(lower_limit_cvloo - 0.03, upper_limit_cvloo + 0.03)) +
  scale_x_discrete(name = "Model description") +
  scale_y_continuous(name = "Predictive correlation") +
  ggthemes::scale_fill_gdocs(name = "Comparison group") +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
    axis.text.y = element_text(size = 8),
    axis.title = element_text(size = 10),
    legend.position = "top",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  ) +
  labs(title = "Predictive ability under CV0 and CV00")

p_cvloo

ggsave(file.path(figures_dir, "prediction_cv0_cv00.tiff"), plot = p_cvloo, width = 16, height = 12, dpi = 300, bg = "white")
ggsave(file.path(figures_dir, "prediction_cv0_cv00.pdf"), plot = p_cvloo, width = 16, height = 12, bg = "white")
```

## 16. Final notes

This tutorialized version is aligned with the final prediction workflow.

It expects the files exported by `analysis_prediction.Rmd`, where:

- each file is already aggregated inside the repetition;
- the fold information appears in the internal `.id` column;
- `CV0` and `CV00` still need `Env == Env_leave` filtering for the held-out-environment summaries.

After running this file, the next step in the repository is to execute the manuscript-oriented script that consumes the summary tables written to `output/tables/`.
