---
title: "Prediction summaries and figures"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Integrating-nir-genomic-kernel/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Integrating-nir-genomic-kernel
---

**Language / Idioma:** [English](visualization.html) | [Português](visualization_pt.html)

The purpose of this step is to take the raw prediction outputs produced in `analysis_prediction_run_outputs.Rmd`, reorganize them into interpretable summary tables, and then translate these summaries into compact comparisons and final figures.

## 1. Setup

The document starts by defining a consistent rendering style for code, tables, and figures. This helps the tutorial remain readable when inspected in HTML and keeps the visual presentation stable across the outputs produced in this step.

```{r setup}
knitr::opts_chunk$set( echo = TRUE,
 warning = FALSE,
 message = FALSE,
 fig.width = 13,
 fig.height = 8,
  out.width = "100%" )
```

## 2. Load packages and locate project folders

The visualization step depends on a small set of packages for data manipulation, figure construction, and table formatting. It also needs to locate the project folders correctly, because the summaries generated here must be written to stable output paths that can later be reused by the site and by manuscript-oriented materials.

This section standardizes the project root, locates the folders that contain the prediction outputs, and prepares the directories where the final tables and figures will be written.

```{r packages-and-paths}
rm(list = ls())
options(scipen = 999)
packages_needed <- c( "dplyr",
 "tidyr",
 "readr",
 "tibble",
 "ggplot2",
 "ggthemes",
 "knitr",
 "kableExtra",
  "stringr" )
packages_missing <- packages_needed[!(packages_needed %in% installed.packages()[, "Package"])]
if (length(packages_missing) > 0) {
  stop(
    paste0(
      "The following packages are required for this module, but are not installed: ",
      paste(packages_missing, collapse = ", "),
      ". Install them before rendering visualization.Rmd."
    )
  )
}
library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(ggplot2)
library(ggthemes)
library(knitr)
library(kableExtra)
library(stringr)
project_root <- getwd()
if (!dir.exists(file.path(project_root, "output")) && dir.exists(file.path(project_root, "..", "output"))) {
  project_root <- normalizePath(file.path(project_root, ".."))
}
if (!dir.exists(file.path(project_root, "output"))) {
  stop("Could not locate the 'output/' directory.\nRun this script from the project root or from analysis/.")
}
results_dir <- file.path(project_root, "output", "results")
figures_dir <- file.path(project_root, "output", "figures")
tables_dir <- file.path(project_root, "output", "tables")
dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tables_dir, recursive = TRUE, showWarnings = FALSE)
```

## 3. Model catalog used in the reduced pipeline

The visualization step uses the same reduced catalog of 18 models adopted in the prediction workflow. Here, this catalog is reorganized so that the summaries become easier to interpret analytically. Instead of showing only model numbers, the table emphasizes the represented pathway, the kernel family used, and whether climatic information is present.

This organization is important because the comparisons shown later in the document are not arbitrary. They were designed to help the reader understand where predictive gains come from and whether these gains are associated with climatic information, kernel choice, or integration between genomic and phenomic sources.

```{r model-catalog}
model_catalog <- tibble(
  ModelNumber = c(1, 10, 3, 13, 5, 16, 2, 11, 4, 14, 6, 17, 7, 12, 8, 15, 9, 18),
  Model = paste0("Eta", c(1, 10, 3, 13, 5, 16, 2, 11, 4, 14, 6, 17, 7, 12, 8, 15, 9, 18)),
  Pathway = c(
    rep("Genomic pathway", 6),
    rep("Phenomic pathway", 6),
    rep("Combined G + P pathway", 6)
  ),
  KernelFamily = c(
    "Linear", "Linear", "Gaussian", "Gaussian", "Arc-cosine", "Arc-cosine",
    "Linear", "Linear", "Gaussian", "Gaussian", "Arc-cosine", "Arc-cosine",
    "Linear", "Linear", "Gaussian", "Gaussian", "Arc-cosine", "Arc-cosine"
  ),
  WeatherStatus = c(
    "Without W", "With W", "Without W", "With W", "Without W", "With W",
    "Without W", "With W", "Without W", "With W", "Without W", "With W",
    "Without W", "With W", "Without W", "With W", "Without W", "With W"
  ),
  Group = c(
    rep("Genomic", 6),
    rep("Phenomic", 6),
    rep("Combined", 6)
  ),
  Description = c(
    "Genomic linear + E", "Genomic linear + W", "Genomic Gaussian + E", "Genomic Gaussian + W", "Genomic arc-cosine + E", "Genomic arc-cosine + W",
    "Phenomic linear + E", "Phenomic linear + W", "Phenomic Gaussian + E", "Phenomic Gaussian + W", "Phenomic arc-cosine + E", "Phenomic arc-cosine + W",
    "Combined linear + E", "Combined linear + W", "Combined Gaussian + E", "Combined Gaussian + W", "Combined arc-cosine + E", "Combined arc-cosine + W"
  ),
  ShortDescription = c(
    "G + E", "G + W", "GGK + E", "GGK + W", "GAK + E", "GAK + W",
    "P + E", "P + W", "PGK + E", "PGK + W", "PAK + E", "PAK + W",
    "G+P + E", "G+P + W", "GGK+PGK + E", "GGK+PGK + W", "GAK+PAK + E", "GAK+PAK + W"
  ),
  FullEffects = c(
    "E + G + GE",
    "E + W + G + GE + GW",
    "E + GGK + GGKE",
    "E + W + GGK + GGKE + GGKW",
    "E + GAK + GAKE",
    "E + W + GAK + GAKE + GAKW",
    "E + P + PE",
    "E + W + P + PE + PW",
    "E + PGK + PGKE",
    "E + W + PGK + PGKE + PGKW",
    "E + PAK + PAKE",
    "E + W + PAK + PAKE + PAKW",
    "E + G + P + GE + PE",
    "E + W + G + P + GE + PE + GW + PW",
    "E + GGK + PGK + GGKE + PGKE",
    "E + W + GGK + PGK + GGKE + PGKE + GGKW + PGKW",
    "E + GAK + PAK + GAKE + PAKE",
    "E + W + GAK + PAK + GAKE + PAKE + GAKW + PAKW"
  ),
  InterpretiveDescription = c(
    "Linear genomic model with environment and G×E interaction.",
    "Linear genomic-weather model with environment, weather, G×E, and G×W interaction.",
    "Gaussian genomic model with environment and genomic-by-environment interaction.",
    "Gaussian genomic-weather model with environment, weather, genomic-by-environment, and genomic-by-weather interaction.",
    "Arc-cosine genomic model with environment and genomic-by-environment interaction.",
    "Arc-cosine genomic-weather model with environment, weather, genomic-by-environment, and genomic-by-weather interaction.",
    "Linear phenomic model with environment and P×E interaction.",
    "Linear phenomic-weather model with environment, weather, P×E, and P×W interaction.",
    "Gaussian phenomic model with environment and phenomic-by-environment interaction.",
    "Gaussian phenomic-weather model with environment, weather, phenomic-by-environment, and phenomic-by-weather interaction.",
    "Arc-cosine phenomic model with environment and phenomic-by-environment interaction.",
    "Arc-cosine phenomic-weather model with environment, weather, phenomic-by-environment, and phenomic-by-weather interaction.",
    "Linear combined genomic-phenomic model with environment and both G×E and P×E interaction terms.",
    "Linear combined genomic-phenomic-weather model with environment, weather, G×E, P×E, G×W, and P×W interaction terms.",
    "Gaussian combined genomic-phenomic model with environment and both genomic and phenomic interaction terms.",
    "Gaussian combined genomic-phenomic-weather model with environment, weather, and the full set of genomic and phenomic interaction terms.",
    "Arc-cosine combined genomic-phenomic model with environment and both genomic and phenomic interaction terms.",
    "Arc-cosine combined genomic-phenomic-weather model with environment, weather, and the full set of genomic and phenomic interaction terms."
  )
)

model_catalog
```

## 4. Comparison pairs used for contrasts between without W and with W

A central question in this project is whether adding the climatic kernel (`W`) improves predictive ability. To answer this question fairly, the comparison must always be made between structurally similar models that differ only by the inclusion of climatic information.

The comparison pairs defined below provide exactly this structure. They are used later to build tables that quantify how predictive performance changes when `W` is introduced within the same pathway and the same kernel family.

```{r comparison-pairs}
comparison_pairs <- tibble( Comparison = c( "Genomic linear", "Genomic Gaussian", "Genomic arc-cosine", "Phenomic linear", "Phenomic Gaussian", "Phenomic arc-cosine", "Combined linear", "Combined Gaussian", "Combined arc-cosine" ),
 WithoutW = c("Eta1", "Eta3", "Eta5", "Eta2", "Eta4", "Eta6", "Eta7", "Eta8", "Eta9"),
  WithW = c("Eta10", "Eta13", "Eta16", "Eta11", "Eta14", "Eta17", "Eta12", "Eta15", "Eta18") )
comparison_pairs
```

## 5. Locate prediction files

The first operational step of the visualization stage is to locate the raw prediction files produced by the previous module. In the current pipeline standard, these files are written recursively to `output/results/`, with an explicit hierarchy by validation scheme, trait, repetition, and fold.

At this stage, the script reads the results folder recursively, keeps only prediction CSVs stored inside the structured `CV1`, `CV2`, `CV0`, and `CV00` branches, and ignores other artifacts such as `bglr_runs`. This makes the later summaries compatible with the new `Rep/Fold` contract and preserves the separation between genotype-based validation problems and leave-one-environment-out problems.

```{r load-generated-visualization-outputs, include=FALSE}
required_tables <- c(
  "prediction_cv1_cv2_summary.csv",
  "prediction_cv0_cv00_summary.csv",
  "prediction_cv1_cv2_w_effect_pairs.csv",
  "prediction_cv0_cv00_w_effect_pairs.csv",
  "prediction_cv1_cv2_kernel_family_compare.csv",
  "prediction_cv0_cv00_kernel_family_compare.csv",
  "prediction_cv1_cv2_display_table.csv",
  "prediction_cv0_cv00_display_table.csv",
  "prediction_cv1_cv2_top3.csv",
  "prediction_cv0_cv00_top3.csv"
)

missing_tables <- required_tables[
  !file.exists(file.path(tables_dir, required_tables))
]

if (length(missing_tables) > 0) {
  stop(
    "This visualization file depends on tables already generated in output/tables/.\n",
    "Missing files: ",
    paste(missing_tables, collapse = ", ")
  )
}

model_catalog <- model_catalog %>%
  dplyr::select(
    ModelNumber,
    Model,
    Pathway,
    KernelFamily,
    WeatherStatus,
    Group,
    Description,
    ShortDescription,
    FullEffects,
    InterpretiveDescription
  )

required_model_catalog_cols <- c(
  "ModelNumber",
  "Model",
  "Pathway",
  "KernelFamily",
  "WeatherStatus",
  "Group",
  "Description",
  "ShortDescription",
  "FullEffects",
  "InterpretiveDescription"
)

missing_model_catalog_cols <- setdiff(required_model_catalog_cols, names(model_catalog))
if (length(missing_model_catalog_cols) > 0) {
  stop(
    "O objeto model_catalog foi criado sem as colunas obrigatórias: ",
    paste(missing_model_catalog_cols, collapse = ", "),
    "."
  )
}

cv12_summary <- readr::read_csv(
  file.path(tables_dir, "prediction_cv1_cv2_summary.csv"),
  show_col_types = FALSE
)

if (!all(c("ModelNumber", "Model") %in% names(cv12_summary))) {
  stop(
    "prediction_cv1_cv2_summary.csv must contain ModelNumber and Model to reattach FullEffects and InterpretiveDescription."
  )
}

cv12_summary <- cv12_summary %>%
  dplyr::select(
    -dplyr::any_of(
      c(
        "FullEffects",
        "InterpretiveDescription",
        "FullEffects.x",
        "FullEffects.y",
        "InterpretiveDescription.x",
        "InterpretiveDescription.y"
      )
    )
  ) %>%
  left_join(
    model_catalog %>% dplyr::select(ModelNumber, Model, FullEffects, InterpretiveDescription),
    by = c("ModelNumber", "Model")
  )

required_cv12_summary_cols <- c(
  "Trait",
  "CV",
  "Group",
  "MeanCorrelation",
  "SDCorrelation",
  "FullEffects"
)

missing_cv12_summary_cols <- setdiff(required_cv12_summary_cols, names(cv12_summary))
if (length(missing_cv12_summary_cols) > 0) {
  stop(
    "cv12_summary is missing required columns after loading: ",
    paste(missing_cv12_summary_cols, collapse = ", "),
    "."
  )
}

cvloo_summary <- readr::read_csv(
  file.path(tables_dir, "prediction_cv0_cv00_summary.csv"),
  show_col_types = FALSE
)

if (!all(c("ModelNumber", "Model") %in% names(cvloo_summary))) {
  stop(
    "prediction_cv0_cv00_summary.csv must contain ModelNumber and Model to reattach FullEffects and InterpretiveDescription."
  )
}

cvloo_summary <- cvloo_summary %>%
  dplyr::select(
    -dplyr::any_of(
      c(
        "FullEffects",
        "InterpretiveDescription",
        "FullEffects.x",
        "FullEffects.y",
        "InterpretiveDescription.x",
        "InterpretiveDescription.y"
      )
    )
  ) %>%
  left_join(
    model_catalog %>% dplyr::select(ModelNumber, Model, FullEffects, InterpretiveDescription),
    by = c("ModelNumber", "Model")
  )

required_cvloo_summary_cols <- c(
  "Trait",
  "CV",
  "Env_leave",
  "Group",
  "MeanCorrelation",
  "SDCorrelation",
  "FullEffects"
)

missing_cvloo_summary_cols <- setdiff(required_cvloo_summary_cols, names(cvloo_summary))
if (length(missing_cvloo_summary_cols) > 0) {
  stop(
    "cvloo_summary is missing required columns after loading: ",
    paste(missing_cvloo_summary_cols, collapse = ", "),
    "."
  )
}

cv12_w_effect <- readr::read_csv(
  file.path(tables_dir, "prediction_cv1_cv2_w_effect_pairs.csv"),
  show_col_types = FALSE
)

required_cv12_w_effect_cols <- c(
  "Comparison",
  "Trait",
  "CV",
  "MeanWithoutW",
  "MeanWithW",
  "DeltaWithW"
)

missing_cv12_w_effect_cols <- setdiff(required_cv12_w_effect_cols, names(cv12_w_effect))
if (length(missing_cv12_w_effect_cols) > 0) {
  stop(
    "prediction_cv1_cv2_w_effect_pairs.csv is missing required columns: ",
    paste(missing_cv12_w_effect_cols, collapse = ", "),
    "."
  )
}

cvloo_w_effect <- readr::read_csv(
  file.path(tables_dir, "prediction_cv0_cv00_w_effect_pairs.csv"),
  show_col_types = FALSE
)

required_cvloo_w_effect_cols <- c(
  "Comparison",
  "Trait",
  "CV",
  "Env_leave",
  "MeanWithoutW",
  "MeanWithW",
  "DeltaWithW"
)

missing_cvloo_w_effect_cols <- setdiff(required_cvloo_w_effect_cols, names(cvloo_w_effect))
if (length(missing_cvloo_w_effect_cols) > 0) {
  stop(
    "prediction_cv0_cv00_w_effect_pairs.csv is missing required columns: ",
    paste(missing_cvloo_w_effect_cols, collapse = ", "),
    "."
  )
}

cv12_kernel_family_compare <- readr::read_csv(
  file.path(tables_dir, "prediction_cv1_cv2_kernel_family_compare.csv"),
  show_col_types = FALSE
)

required_cv12_kernel_cols <- c(
  "Trait",
  "CV",
  "Pathway",
  "WeatherStatus",
  "Linear",
  "Gaussian",
  "Arc-cosine",
  "Delta_Gaussian_vs_Linear",
  "Delta_ArcCos_vs_Linear",
  "Delta_ArcCos_vs_Gaussian"
)

missing_cv12_kernel_cols <- setdiff(required_cv12_kernel_cols, names(cv12_kernel_family_compare))
if (length(missing_cv12_kernel_cols) > 0) {
  stop(
    "prediction_cv1_cv2_kernel_family_compare.csv is missing required columns: ",
    paste(missing_cv12_kernel_cols, collapse = ", "),
    "."
  )
}

cvloo_kernel_family_compare <- readr::read_csv(
  file.path(tables_dir, "prediction_cv0_cv00_kernel_family_compare.csv"),
  show_col_types = FALSE
)

required_cvloo_kernel_cols <- c(
  "Trait",
  "CV",
  "Env_leave",
  "Pathway",
  "WeatherStatus",
  "Linear",
  "Gaussian",
  "Arc-cosine",
  "Delta_Gaussian_vs_Linear",
  "Delta_ArcCos_vs_Linear",
  "Delta_ArcCos_vs_Gaussian"
)

missing_cvloo_kernel_cols <- setdiff(required_cvloo_kernel_cols, names(cvloo_kernel_family_compare))
if (length(missing_cvloo_kernel_cols) > 0) {
  stop(
    "prediction_cv0_cv00_kernel_family_compare.csv is missing required columns: ",
    paste(missing_cvloo_kernel_cols, collapse = ", "),
    "."
  )
}

cv12_table_display <- readr::read_csv(
  file.path(tables_dir, "prediction_cv1_cv2_display_table.csv"),
  show_col_types = FALSE
)

required_cv12_display_cols <- c(
  "Trait",
  "CV",
  "Pathway",
  "KernelFamily",
  "WeatherStatus",
  "ShortDescription",
  "MeanCorrelation",
  "SDCorrelation",
  "NumberOfRepetitions",
  "MeanRuntimeSec"
)

missing_cv12_display_cols <- setdiff(required_cv12_display_cols, names(cv12_table_display))
if (length(missing_cv12_display_cols) > 0) {
  stop(
    "prediction_cv1_cv2_display_table.csv is missing required columns: ",
    paste(missing_cv12_display_cols, collapse = ", "),
    "."
  )
}

cvloo_table_display <- readr::read_csv(
  file.path(tables_dir, "prediction_cv0_cv00_display_table.csv"),
  show_col_types = FALSE
)

required_cvloo_display_cols <- c(
  "Trait",
  "CV",
  "Env_leave",
  "Pathway",
  "KernelFamily",
  "WeatherStatus",
  "ShortDescription",
  "MeanCorrelation",
  "SDCorrelation",
  "NumberOfRuns",
  "MeanRuntimeSec"
)

missing_cvloo_display_cols <- setdiff(required_cvloo_display_cols, names(cvloo_table_display))
if (length(missing_cvloo_display_cols) > 0) {
  stop(
    "prediction_cv0_cv00_display_table.csv is missing required columns: ",
    paste(missing_cvloo_display_cols, collapse = ", "),
    "."
  )
}

cv12_top3 <- readr::read_csv(
  file.path(tables_dir, "prediction_cv1_cv2_top3.csv"),
  show_col_types = FALSE
)

required_cv12_top3_cols <- c(
  "Trait",
  "CV",
  "ShortDescription",
  "MeanCorrelation",
  "SDCorrelation"
)

missing_cv12_top3_cols <- setdiff(required_cv12_top3_cols, names(cv12_top3))
if (length(missing_cv12_top3_cols) > 0) {
  stop(
    "prediction_cv1_cv2_top3.csv is missing required columns: ",
    paste(missing_cv12_top3_cols, collapse = ", "),
    "."
  )
}

cvloo_top3 <- readr::read_csv(
  file.path(tables_dir, "prediction_cv0_cv00_top3.csv"),
  show_col_types = FALSE
)

required_cvloo_top3_cols <- c(
  "Trait",
  "CV",
  "Env_leave",
  "ShortDescription",
  "MeanCorrelation",
  "SDCorrelation"
)

missing_cvloo_top3_cols <- setdiff(required_cvloo_top3_cols, names(cvloo_top3))
if (length(missing_cvloo_top3_cols) > 0) {
  stop(
    "prediction_cv0_cv00_top3.csv is missing required columns: ",
    paste(missing_cvloo_top3_cols, collapse = ", "),
    "."
  )
}

model_catalog <- model_catalog %>%
  dplyr::mutate(
    Pathway = dplyr::recode(
      Pathway,
      "Genomic pathway" = "Genomic pathway",
      "Phenomic pathway" = "Phenomic pathway",
      "Combined G + P pathway" = "Combined G + P pathway"
    ),
    KernelFamily = dplyr::recode(
      KernelFamily,
      "Gaussian" = "Gaussian"
    )
  )

cv12_summary <- cv12_summary %>%
  dplyr::mutate(
    Pathway = dplyr::recode(
      Pathway,
      "Genomic pathway" = "Genomic pathway",
      "Phenomic pathway" = "Phenomic pathway",
      "Combined G + P pathway" = "Combined G + P pathway"
    ),
    KernelFamily = dplyr::recode(
      KernelFamily,
      "Gaussian" = "Gaussian"
    ),
    WeatherStatus = dplyr::recode(
      WeatherStatus,
      "Without W" = "Without W",
      "With W" = "With W"
    ),
    Group = dplyr::case_when(
      Pathway == "Genomic pathway" ~ "Genomic",
      Pathway == "Phenomic pathway" ~ "Phenomic",
      Pathway == "Combined G + P pathway" ~ "Combined",
      TRUE ~ as.character(Group)
    ),
    Description = dplyr::recode(
      Description,
      "Genomic linear + E" = "Genomic linear + E",
      "Genomic linear + W" = "Genomic linear + W",
      "Genomic Gaussian + E" = "Genomic Gaussian + E",
      "Genomic Gaussian + W" = "Genomic Gaussian + W",
      "Genomic arc-cosine + E" = "Genomic arc-cosine + E",
      "Genomic arc-cosine + W" = "Genomic arc-cosine + W",
      "Phenomic linear + E" = "Phenomic linear + E",
      "Phenomic linear + W" = "Phenomic linear + W",
      "Phenomic Gaussian + E" = "Phenomic Gaussian + E",
      "Phenomic Gaussian + W" = "Phenomic Gaussian + W",
      "Phenomic arc-cosine + E" = "Phenomic arc-cosine + E",
      "Phenomic arc-cosine + W" = "Phenomic arc-cosine + W",
      "Combined linear + E" = "Combined linear + E",
      "Combined linear + W" = "Combined linear + W",
      "Combined Gaussian + E" = "Combined Gaussian + E",
      "Combined Gaussian + W" = "Combined Gaussian + W",
      "Combined arc-cosine + E" = "Combined arc-cosine + E",
      "Combined arc-cosine + W" = "Combined arc-cosine + W"
    )
  )

cvloo_summary <- cvloo_summary %>%
  dplyr::mutate(
    Pathway = dplyr::recode(
      Pathway,
      "Genomic pathway" = "Genomic pathway",
      "Phenomic pathway" = "Phenomic pathway",
      "Combined G + P pathway" = "Combined G + P pathway"
    ),
    KernelFamily = dplyr::recode(
      KernelFamily,
      "Gaussian" = "Gaussian"
    ),
    WeatherStatus = dplyr::recode(
      WeatherStatus,
      "Without W" = "Without W",
      "With W" = "With W"
    ),
    Group = dplyr::case_when(
      Pathway == "Genomic pathway" ~ "Genomic",
      Pathway == "Phenomic pathway" ~ "Phenomic",
      Pathway == "Combined G + P pathway" ~ "Combined",
      TRUE ~ as.character(Group)
    ),
    Description = dplyr::recode(
      Description,
      "Genomic linear + E" = "Genomic linear + E",
      "Genomic linear + W" = "Genomic linear + W",
      "Genomic Gaussian + E" = "Genomic Gaussian + E",
      "Genomic Gaussian + W" = "Genomic Gaussian + W",
      "Genomic arc-cosine + E" = "Genomic arc-cosine + E",
      "Genomic arc-cosine + W" = "Genomic arc-cosine + W",
      "Phenomic linear + E" = "Phenomic linear + E",
      "Phenomic linear + W" = "Phenomic linear + W",
      "Phenomic Gaussian + E" = "Phenomic Gaussian + E",
      "Phenomic Gaussian + W" = "Phenomic Gaussian + W",
      "Phenomic arc-cosine + E" = "Phenomic arc-cosine + E",
      "Phenomic arc-cosine + W" = "Phenomic arc-cosine + W",
      "Combined linear + E" = "Combined linear + E",
      "Combined linear + W" = "Combined linear + W",
      "Combined Gaussian + E" = "Combined Gaussian + E",
      "Combined Gaussian + W" = "Combined Gaussian + W",
      "Combined arc-cosine + E" = "Combined arc-cosine + E",
      "Combined arc-cosine + W" = "Combined arc-cosine + W"
    )
  )

cv12_w_effect <- cv12_w_effect %>%
  dplyr::mutate(
    Comparison = dplyr::recode(
      Comparison,
      "Genomic linear" = "Genomic linear",
      "Genomic Gaussian" = "Genomic Gaussian",
      "Genomic arc-cosine" = "Genomic arc-cosine",
      "Phenomic linear" = "Phenomic linear",
      "Phenomic Gaussian" = "Phenomic Gaussian",
      "Phenomic arc-cosine" = "Phenomic arc-cosine",
      "Combined linear" = "Combined linear",
      "Combined Gaussian" = "Combined Gaussian",
      "Combined arc-cosine" = "Combined arc-cosine"
    )
  )

cvloo_w_effect <- cvloo_w_effect %>%
  dplyr::mutate(
    Comparison = dplyr::recode(
      Comparison,
      "Genomic linear" = "Genomic linear",
      "Genomic Gaussian" = "Genomic Gaussian",
      "Genomic arc-cosine" = "Genomic arc-cosine",
      "Phenomic linear" = "Phenomic linear",
      "Phenomic Gaussian" = "Phenomic Gaussian",
      "Phenomic arc-cosine" = "Phenomic arc-cosine",
      "Combined linear" = "Combined linear",
      "Combined Gaussian" = "Combined Gaussian",
      "Combined arc-cosine" = "Combined arc-cosine"
    )
  )

cv12_kernel_family_compare <- cv12_kernel_family_compare %>%
  dplyr::mutate(
    Pathway = dplyr::recode(
      Pathway,
      "Genomic pathway" = "Genomic pathway",
      "Phenomic pathway" = "Phenomic pathway",
      "Combined G + P pathway" = "Combined G + P pathway"
    ),
    WeatherStatus = dplyr::recode(
      WeatherStatus,
      "Without W" = "Without W",
      "With W" = "With W"
    )
  )

cvloo_kernel_family_compare <- cvloo_kernel_family_compare %>%
  dplyr::mutate(
    Pathway = dplyr::recode(
      Pathway,
      "Genomic pathway" = "Genomic pathway",
      "Phenomic pathway" = "Phenomic pathway",
      "Combined G + P pathway" = "Combined G + P pathway"
    ),
    WeatherStatus = dplyr::recode(
      WeatherStatus,
      "Without W" = "Without W",
      "With W" = "With W"
    )
  )

cv12_table_display <- cv12_table_display %>%
  dplyr::mutate(
    Pathway = dplyr::recode(
      Pathway,
      "Genomic pathway" = "Genomic pathway",
      "Phenomic pathway" = "Phenomic pathway",
      "Combined G + P pathway" = "Combined G + P pathway"
    ),
    KernelFamily = dplyr::recode(
      KernelFamily,
      "Gaussian" = "Gaussian"
    ),
    WeatherStatus = dplyr::recode(
      WeatherStatus,
      "Without W" = "Without W",
      "With W" = "With W"
    )
  )

cvloo_table_display <- cvloo_table_display %>%
  dplyr::mutate(
    Pathway = dplyr::recode(
      Pathway,
      "Genomic pathway" = "Genomic pathway",
      "Phenomic pathway" = "Phenomic pathway",
      "Combined G + P pathway" = "Combined G + P pathway"
    ),
    KernelFamily = dplyr::recode(
      KernelFamily,
      "Gaussian" = "Gaussian"
    ),
    WeatherStatus = dplyr::recode(
      WeatherStatus,
      "Without W" = "Without W",
      "With W" = "With W"
    )
  )
```

```{r locate-files, eval=FALSE}
all_result_rel_paths <- list.files(results_dir, pattern = "\\.csv$", full.names = FALSE, recursive = TRUE)
all_result_rel_paths <- gsub("\\\\", "/", all_result_rel_paths)
structured_idx <- grepl("^CV1/|^CV2/|^CV0/|^CV00/", all_result_rel_paths)
all_result_rel_paths <- all_result_rel_paths[structured_idx]
all_result_files <- file.path(results_dir, all_result_rel_paths)
cv12_idx <- grepl("^CV1/|^CV2/", all_result_rel_paths)
cvloo_idx <- grepl("^CV0/|^CV00/", all_result_rel_paths)
cv12_files <- all_result_files[cv12_idx]
cv12_rel_paths <- all_result_rel_paths[cv12_idx]
cvloo_files <- all_result_files[cvloo_idx]
cvloo_rel_paths <- all_result_rel_paths[cvloo_idx]
cat("Número de arquivos CV1/CV2:", length(cv12_files), "
")
cat("Número de arquivos CV0/CV00:", length(cvloo_files), "
")
if (length(cv12_files) == 0 && length(cvloo_files) == 0) {
  stop("No prediction CSV file was found in output/results/.
Execute analysis_prediction_run_outputs primeiro.")
}
```

## 6. Read CV1 and CV2 files generated by the final prediction script

The `CV1` and `CV2` files contain the raw outputs needed to summarize predictive performance when the challenge is defined along the genotype axis. In the current workflow, repetition and fold are encoded both in the folder hierarchy and in the file name, so the script reconstructs these metadata explicitly before building the summaries.

This design is useful because it keeps the file system organized for quick inspection and, at the same time, allows the visualization step to rebuild unified tables across repetitions and folds.

```{r read-cv12, eval=FALSE}
cv12_list <- vector("list", length(cv12_files))
for (i in seq_along(cv12_files)) {
  current_file <- cv12_files[i]
  current_rel_path <- cv12_rel_paths[i]
  path_parts <- strsplit(current_rel_path, "/")[[1]]

  if (length(path_parts) != 5) {
    stop(
      "Estrutura de diretórios inesperada para arquivo CV1/CV2: ",
      current_rel_path,
      ".\nPadrão esperado: CV1/<trait>/repXX/foldXX/<file>.csv ou CV2/<trait>/repXX/foldXX/<file>.csv"
    )
  }

  current_cv <- path_parts[1]
  trait_from_path <- path_parts[2]
  rep_from_path <- as.numeric(gsub("^rep", "", path_parts[3]))
  fold_from_path <- as.numeric(gsub("^fold", "", path_parts[4]))

  file_name <- sub("\\.csv$", "", basename(current_file))
  file_parts <- strsplit(file_name, "_")[[1]]

  if (length(file_parts) != 5) {
    stop(
      "Nome de arquivo inesperado para CV1/CV2: ",
      file_name,
      ".\nPadrão esperado: CV1_<trait>_Eta<model>_repXX_foldXX.csv ou CV2_<trait>_Eta<model>_repXX_foldXX.csv"
    )
  }

  current_trait <- file_parts[2]
  current_model_label <- file_parts[3]
  current_model_number <- as.numeric(gsub("^Eta", "", current_model_label))
  current_rep <- as.numeric(gsub("^rep", "", file_parts[4]))
  current_fold <- as.numeric(gsub("^fold", "", file_parts[5]))

  if (!identical(current_cv, file_parts[1]) || !identical(trait_from_path, current_trait) || !identical(rep_from_path, current_rep) || !identical(fold_from_path, current_fold)) {
    stop("Mismatch between path metadata and file name metadata in: ", current_rel_path)
  }

  current_data <- readr::read_csv(current_file, show_col_types = FALSE)

  if (!("Env" %in% names(current_data)) || !("cor" %in% names(current_data))) {
    stop(
      "Estrutura de colunas inesperada no arquivo: ",
      file_name,
      ".\nO script final de predição deve exportar colunas incluindo Env e cor."
    )
  }

  current_data <- current_data %>%
    mutate(
      RelativePath = current_rel_path,
      Fold = current_fold,
      CV = current_cv,
      Trait = current_trait,
      ModelNumber = current_model_number,
      Model = current_model_label,
      Rep = current_rep
    )

  cv12_list[[i]] <- current_data
}

cv12_raw <- bind_rows(cv12_list) %>%
  left_join(model_catalog, by = c("ModelNumber", "Model"))

if (!("Runtime_sec" %in% names(cv12_raw))) cv12_raw$Runtime_sec <- NA_real_
readr::write_csv(cv12_raw, file.path(tables_dir, "prediction_cv1_cv2_raw.csv"))
```

## 7. Read CV0 and CV00 files generated by the final prediction script

The `CV0` and `CV00` files represent the most demanding prediction problems in this study because they involve an environment left out from model fitting. In the current pipeline standard, the leave-one-environment-out label is encoded both in the folder hierarchy and in the file name, so the script reconstructs this information explicitly before building the leave-one-environment-out summaries.

This section reads these files, checks the consistency of the structured paths, and prepares the metadata needed to keep `Env_leave` explicit throughout the summaries.

```{r read-cvloo, eval=FALSE}
cvloo_list <- vector("list", length(cvloo_files))
for (i in seq_along(cvloo_files)) {
  current_file <- cvloo_files[i]
  current_rel_path <- cvloo_rel_paths[i]
  path_parts <- strsplit(current_rel_path, "/")[[1]]

  if (length(path_parts) != 6) {
    stop(
      "Estrutura de diretórios inesperada para arquivo CV0/CV00: ",
      current_rel_path,
      ".\nPadrão esperado: CV0/<Env_leave>/<trait>/repXX/foldXX/<file>.csv ou CV00/<Env_leave>/<trait>/repXX/foldXX/<file>.csv"
    )
  }

  current_cv <- path_parts[1]
  env_from_path <- path_parts[2]
  trait_from_path <- path_parts[3]
  rep_from_path <- as.numeric(gsub("^rep", "", path_parts[4]))
  fold_from_path <- as.numeric(gsub("^fold", "", path_parts[5]))

  file_name <- sub("\\.csv$", "", basename(current_file))
  file_parts <- strsplit(file_name, "_")[[1]]

  if (length(file_parts) < 6) {
    stop(
      "Nome de arquivo inesperado para CV0/CV00: ",
      file_name,
      ".\nPadrão esperado: CV0_<trait>_<Env_leave>_Eta<model>_repXX_foldXX.csv ou CV00_<trait>_<Env_leave>_Eta<model>_repXX_foldXX.csv"
    )
  }

  current_trait <- file_parts[2]
  current_model_label <- file_parts[length(file_parts) - 2]
  current_model_number <- as.numeric(gsub("^Eta", "", current_model_label))
  current_rep <- as.numeric(gsub("^rep", "", file_parts[length(file_parts) - 1]))
  current_fold <- as.numeric(gsub("^fold", "", file_parts[length(file_parts)]))
  current_env_leave <- paste(file_parts[3:(length(file_parts) - 3)], collapse = "_")

  if (!identical(current_cv, file_parts[1]) || !identical(env_from_path, current_env_leave) || !identical(trait_from_path, current_trait) || !identical(rep_from_path, current_rep) || !identical(fold_from_path, current_fold)) {
    stop("Mismatch between path metadata and file name metadata in: ", current_rel_path)
  }

  current_data <- readr::read_csv(current_file, show_col_types = FALSE)

  if (!("Env" %in% names(current_data)) || !("cor" %in% names(current_data))) {
    stop(
      "Estrutura de colunas inesperada no arquivo: ",
      file_name,
      ".\nO script final de predição deve exportar colunas incluindo Env e cor."
    )
  }

  current_data <- current_data %>%
    mutate(
      RelativePath = current_rel_path,
      Fold = current_fold,
      CV = current_cv,
      Trait = current_trait,
      Env_leave = current_env_leave,
      ModelNumber = current_model_number,
      Model = current_model_label,
      Rep = current_rep
    )

  cvloo_list[[i]] <- current_data
}
cvloo_raw <- bind_rows(cvloo_list) %>%
  left_join(model_catalog, by = c("ModelNumber", "Model"))
if (!("Runtime_sec" %in% names(cvloo_raw))) cvloo_raw$Runtime_sec <- NA_real_
readr::write_csv(cvloo_raw, file.path(tables_dir, "prediction_cv0_cv00_raw.csv"))
readr::write_csv(model_catalog, file.path(tables_dir, "prediction_model_catalog.csv"))
```

## 8. Summarize CV1 and CV2

This step builds the summaries for `CV1` and `CV2` sequentially: first by environment and fold, then by environment within repetition, then by repetition, and finally by model-level means and standard deviations across repetitions.

```{r summarize-cv12, eval=FALSE}
cv12_env_fold <- cv12_raw %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Rep,
   Fold,
    Env ) %>%
  summarise( EnvCorrelation = mean(cor, na.rm = TRUE),
   ValidationCount = sum(!is.na(cor)),
   MeanRuntimeSec = mean(Runtime_sec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( EnvCorrelation = ifelse(is.nan(EnvCorrelation), NA_real_, EnvCorrelation),
    MeanRuntimeSec = ifelse(is.nan(MeanRuntimeSec), NA_real_, MeanRuntimeSec) ) %>%
  filter(!is.na(EnvCorrelation), ValidationCount > 0)
cv12_env_rep <- cv12_env_fold %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Rep,
    Env ) %>%
  summarise( EnvCorrelation = weighted.mean(EnvCorrelation, w = ValidationCount, na.rm = TRUE),
   ValidationCount = sum(ValidationCount, na.rm = TRUE),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" )
cv12_rep_summary <- cv12_env_rep %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
    Rep ) %>%
  summarise( WeightedCorrelation = weighted.mean(EnvCorrelation, w = ValidationCount, na.rm = TRUE),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" )
cv12_summary <- cv12_rep_summary %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
    CV ) %>%
  summarise( MeanCorrelation = mean(WeightedCorrelation, na.rm = TRUE),
   SDCorrelation = sd(WeightedCorrelation, na.rm = TRUE),
   NumberOfRepetitions = n(),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( SDCorrelation = ifelse(is.na(SDCorrelation), 0, SDCorrelation) ) %>%
  left_join(
    model_catalog %>% select(ModelNumber, Model, FullEffects, InterpretiveDescription),
    by = c("ModelNumber", "Model")
  )
readr::write_csv(cv12_env_fold, file.path(tables_dir, "prediction_cv1_cv2_by_rep_fold_env.csv"))
readr::write_csv(cv12_env_rep, file.path(tables_dir, "prediction_cv1_cv2_by_rep_env.csv"))
readr::write_csv(cv12_rep_summary, file.path(tables_dir, "prediction_cv1_cv2_by_rep.csv"))
readr::write_csv(cv12_summary, file.path(tables_dir, "prediction_cv1_cv2_summary.csv"))
```

## 9. Summarize CV0 and CV00

For leave-one-environment-out scenarios, the script first restricts the summaries to rows in which `Env == Env_leave`. This keeps only the component of the left-out environment that is conceptually relevant for `CV0` and `CV00`.

```{r summarize-cvloo, eval=FALSE}
cvloo_env_filtered <- cvloo_raw %>%
  filter(Env == Env_leave)
cvloo_env_fold <- cvloo_env_filtered %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Env_leave,
   Rep,
   Fold,
    Env ) %>%
  summarise( LOEOCorrelation = mean(cor, na.rm = TRUE),
   ValidationCount = sum(!is.na(cor)),
   MeanRuntimeSec = mean(Runtime_sec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( LOEOCorrelation = ifelse(is.nan(LOEOCorrelation), NA_real_, LOEOCorrelation),
    MeanRuntimeSec = ifelse(is.nan(MeanRuntimeSec), NA_real_, MeanRuntimeSec) ) %>%
  filter(!is.na(LOEOCorrelation), ValidationCount > 0)
cvloo_rep_summary <- cvloo_env_fold %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Env_leave,
    Rep ) %>%
  summarise( LOEOCorrelation = weighted.mean(LOEOCorrelation, w = ValidationCount, na.rm = TRUE),
   ValidationCount = sum(ValidationCount, na.rm = TRUE),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" )
cvloo_summary <- cvloo_rep_summary %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
    Env_leave ) %>%
  summarise( MeanCorrelation = mean(LOEOCorrelation, na.rm = TRUE),
   SDCorrelation = sd(LOEOCorrelation, na.rm = TRUE),
   NumberOfRuns = n(),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( SDCorrelation = ifelse(is.na(SDCorrelation), 0, SDCorrelation) ) %>%
  left_join(
    model_catalog %>% select(ModelNumber, Model, FullEffects, InterpretiveDescription),
    by = c("ModelNumber", "Model")
  )
readr::write_csv(cvloo_env_fold, file.path(tables_dir, "prediction_cv0_cv00_by_rep_fold_env.csv"))
readr::write_csv(cvloo_rep_summary, file.path(tables_dir, "prediction_cv0_cv00_by_run.csv"))
readr::write_csv(cvloo_summary, file.path(tables_dir, "prediction_cv0_cv00_summary.csv"))
```

## 10. Comparison tables

This section derives secondary analytical tables, especially the contrast between models with and without `W` and the comparison among kernel families within each pathway.

```{r comparison-tables, eval=FALSE}
cv12_w_effect <- comparison_pairs %>%
  left_join( cv12_summary %>% select(Trait, CV, Model, MeanCorrelation, SDCorrelation),
    by = c("WithoutW" = "Model") ) %>%
  rename( MeanWithoutW = MeanCorrelation, SDWithoutW = SDCorrelation ) %>%
  left_join( cv12_summary %>% select(Trait, CV, Model, MeanCorrelation, SDCorrelation),
    by = c("WithW" = "Model", "Trait", "CV") ) %>%
  rename( MeanWithW = MeanCorrelation, SDWithW = SDCorrelation ) %>%
  mutate( DeltaWithW = MeanWithW - MeanWithoutW ) %>%
  arrange(Trait, CV, Comparison)
cvloo_w_effect <- comparison_pairs %>%
  left_join( cvloo_summary %>% select(Trait, CV, Env_leave, Model, MeanCorrelation, SDCorrelation),
    by = c("WithoutW" = "Model") ) %>%
  rename( MeanWithoutW = MeanCorrelation, SDWithoutW = SDCorrelation ) %>%
  left_join( cvloo_summary %>% select(Trait, CV, Env_leave, Model, MeanCorrelation, SDCorrelation),
    by = c("WithW" = "Model", "Trait", "CV", "Env_leave") ) %>%
  rename( MeanWithW = MeanCorrelation, SDWithW = SDCorrelation ) %>%
  mutate( DeltaWithW = MeanWithW - MeanWithoutW ) %>%
  arrange(Trait, CV, Env_leave, Comparison)
cv12_kernel_family_compare <- cv12_summary %>%
  select(Trait, CV, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  mutate( Delta_Gaussian_vs_Linear = Gaussian - Linear,
   Delta_ArcCos_vs_Linear = `Arc-cosine` - Linear,
    Delta_ArcCos_vs_Gaussian = `Arc-cosine` - Gaussian ) %>%
  arrange(Trait, CV, Pathway, WeatherStatus)
cvloo_kernel_family_compare <- cvloo_summary %>%
  select(Trait, CV, Env_leave, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  mutate( Delta_Gaussian_vs_Linear = Gaussian - Linear,
   Delta_ArcCos_vs_Linear = `Arc-cosine` - Linear,
    Delta_ArcCos_vs_Gaussian = `Arc-cosine` - Gaussian ) %>%
  arrange(Trait, CV, Env_leave, Pathway, WeatherStatus)
readr::write_csv(cv12_w_effect, file.path(tables_dir, "prediction_cv1_cv2_w_effect_pairs.csv"))
readr::write_csv(cvloo_w_effect, file.path(tables_dir, "prediction_cv0_cv00_w_effect_pairs.csv"))
readr::write_csv(cv12_kernel_family_compare,
  file.path(tables_dir, "prediction_cv1_cv2_kernel_family_compare.csv"))
readr::write_csv(cvloo_kernel_family_compare,
  file.path(tables_dir, "prediction_cv0_cv00_kernel_family_compare.csv"))
```

## 11. Compact display tables

These objects are simplified versions of the full summaries. They are useful for manuscript tables, supplementary material, and quick inspection of the best-performing models.

```{r display-tables, eval=FALSE}
cv12_table_display <- cv12_summary %>%
  arrange(Trait, CV, Pathway, KernelFamily, WeatherStatus) %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3),
   SDCorrelation = round(SDCorrelation, 3),
    MeanRuntimeSec = round(MeanRuntimeSec, 2) ) %>%
  select( Trait,
   CV,
   Pathway,
   KernelFamily,
   WeatherStatus,
   ShortDescription,
   MeanCorrelation,
   SDCorrelation,
   NumberOfRepetitions,
    MeanRuntimeSec )
cvloo_table_display <- cvloo_summary %>%
  arrange(Trait, CV, Env_leave, Pathway, KernelFamily, WeatherStatus) %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3),
   SDCorrelation = round(SDCorrelation, 3),
    MeanRuntimeSec = round(MeanRuntimeSec, 2) ) %>%
  select( Trait,
   CV,
   Env_leave,
   Pathway,
   KernelFamily,
   WeatherStatus,
   ShortDescription,
   MeanCorrelation,
   SDCorrelation,
   NumberOfRuns,
    MeanRuntimeSec )
cv12_top3 <- cv12_summary %>%
  group_by(Trait, CV) %>%
  slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3), SDCorrelation = round(SDCorrelation, 3) ) %>%
  select(Trait, CV, ShortDescription, MeanCorrelation, SDCorrelation)
cvloo_top3 <- cvloo_summary %>%
  group_by(Trait, CV, Env_leave) %>%
  slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3), SDCorrelation = round(SDCorrelation, 3) ) %>%
  select(Trait, CV, Env_leave, ShortDescription, MeanCorrelation, SDCorrelation)
readr::write_csv(cv12_table_display, file.path(tables_dir, "prediction_cv1_cv2_display_table.csv"))
readr::write_csv(cvloo_table_display, file.path(tables_dir, "prediction_cv0_cv00_display_table.csv"))
readr::write_csv(cv12_top3, file.path(tables_dir, "prediction_cv1_cv2_top3.csv"))
readr::write_csv(cvloo_top3, file.path(tables_dir, "prediction_cv0_cv00_top3.csv"))
```

## 12. Top-3 tables

The next two tables display the three best-performing models in each validation scheme.

```{r table-cv12-top3}
if (!exists("cv12_top3")) {
  stop("The object cv12_top3 was not loaded before the table-cv12-top3 chunk.")
}
if (!all(c("Trait", "CV", "ShortDescription", "MeanCorrelation", "SDCorrelation") %in% names(cv12_top3))) {
  stop("cv12_top3 does not contain the columns required by the table-cv12-top3 chunk.")
}

cv12_top3 %>%
  knitr::kable( caption = "Top 3 models for each trait and validation scheme in CV1 and CV2.",
    align = "llccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" )
```

```{r table-cvloo-top3}
if (!exists("cvloo_top3")) {
  stop("The object cvloo_top3 was not loaded before the table-cvloo-top3 chunk.")
}
if (!all(c("Trait", "CV", "Env_leave", "ShortDescription", "MeanCorrelation", "SDCorrelation") %in% names(cvloo_top3))) {
  stop("cvloo_top3 does not contain the columns required by the table-cvloo-top3 chunk.")
}

cvloo_top3 %>%
  knitr::kable( caption = "Top 3 models for each trait, validation scheme, and left-out environment in CV0 and CV00.",
    align = "lllccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" )
```

## 13. W-effect tables

These tables present the paired comparisons between models without and with climatic information.

```{r table-cv12-w-effect}
if (!exists("cv12_w_effect")) {
  stop("The object cv12_w_effect was not loaded before the table-cv12-w-effect chunk.")
}
if (!all(c("Comparison", "Trait", "CV", "MeanWithoutW", "MeanWithW", "DeltaWithW") %in% names(cv12_w_effect))) {
  stop("cv12_w_effect does not contain the columns required by the table-cv12-w-effect chunk.")
}

cv12_w_effect %>%
  mutate( MeanWithoutW = round(MeanWithoutW, 3),
   MeanWithW = round(MeanWithW, 3),
    DeltaWithW = round(DeltaWithW, 3) ) %>%
  knitr::kable( caption = "Paired comparison between models without and with the climatic kernel (W) in CV1 and CV2.",
    align = "lllccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

```{r table-cvloo-w-effect}
if (!exists("cvloo_w_effect")) {
  stop("The object cvloo_w_effect was not loaded before the table-cvloo-w-effect chunk.")
}
if (!all(c("Comparison", "Trait", "CV", "Env_leave", "MeanWithoutW", "MeanWithW", "DeltaWithW") %in% names(cvloo_w_effect))) {
  stop("cvloo_w_effect does not contain the columns required by the table-cvloo-w-effect chunk.")
}

cvloo_w_effect %>%
  mutate( MeanWithoutW = round(MeanWithoutW, 3),
   MeanWithW = round(MeanWithW, 3),
    DeltaWithW = round(DeltaWithW, 3) ) %>%
  knitr::kable( caption = "Paired comparison between models without and with the climatic kernel (W) in CV0 and CV00.",
    align = "llllccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

## 14. Kernel-family comparison tables

Here the script contrasts linear, Gaussian, and arc-cosine kernels within each analytical pathway.

```{r table-cv12-kernel-family}
if (!exists("cv12_kernel_family_compare")) {
  stop("The object cv12_kernel_family_compare was not loaded before the table-cv12-kernel-family chunk.")
}
if (!all(c("Trait", "CV", "Pathway", "WeatherStatus", "Linear", "Gaussian", "Arc-cosine") %in% names(cv12_kernel_family_compare))) {
  stop("cv12_kernel_family_compare does not contain the columns required by the table-cv12-kernel-family chunk.")
}

cv12_kernel_family_compare %>%
  mutate( Linear = round(Linear, 3),
   Gaussian = round(Gaussian, 3),
   `Arc-cosine` = round(`Arc-cosine`, 3),
   Delta_Gaussian_vs_Linear = round(Delta_Gaussian_vs_Linear, 3),
   Delta_ArcCos_vs_Linear = round(Delta_ArcCos_vs_Linear, 3),
    Delta_ArcCos_vs_Gaussian = round(Delta_ArcCos_vs_Gaussian, 3) ) %>%
  knitr::kable( caption = "Comparison among kernel families within each analytical pathway in CV1 and CV2.",
    align = "llllcccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

```{r table-cvloo-kernel-family}
if (!exists("cvloo_kernel_family_compare")) {
  stop("The object cvloo_kernel_family_compare was not loaded before the table-cvloo-kernel-family chunk.")
}
if (!all(c("Trait", "CV", "Env_leave", "Pathway", "WeatherStatus", "Linear", "Gaussian", "Arc-cosine") %in% names(cvloo_kernel_family_compare))) {
  stop("cvloo_kernel_family_compare does not contain the columns required by the table-cvloo-kernel-family chunk.")
}

cvloo_kernel_family_compare %>%
  mutate( Linear = round(Linear, 3),
   Gaussian = round(Gaussian, 3),
   `Arc-cosine` = round(`Arc-cosine`, 3),
   Delta_Gaussian_vs_Linear = round(Delta_Gaussian_vs_Linear, 3),
   Delta_ArcCos_vs_Linear = round(Delta_ArcCos_vs_Linear, 3),
    Delta_ArcCos_vs_Gaussian = round(Delta_ArcCos_vs_Gaussian, 3) ) %>%
  knitr::kable( caption = "Comparison among kernel families within each analytical pathway in CV0 and CV00.",
    align = "lllllcccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```


## 15. Automated result-oriented text

The following paragraphs are generated directly from the summary objects created in this script. Whenever the prediction outputs are updated and the file is rendered again, this text is refreshed automatically from the current results.

```{r automated-results-text, results='asis'}
fmt3 <- function(x) sprintf("%.3f", round(x, 3))

trait_labels <- c(
  "GY" = "Grain yield (GY)",
  "KW" = "Kernel weight (KW)"
)

traits_to_report <- intersect(
  c("GY", "KW"),
  unique(c(cv12_summary$Trait, cvloo_summary$Trait))
)

for (current_trait in traits_to_report) {
  cat("### ", trait_labels[[current_trait]], "\n\n", sep = "")

  for (current_cv in c("CV1", "CV2")) {
    current_data <- cv12_summary %>%
      dplyr::filter(Trait == current_trait, CV == current_cv) %>%
      dplyr::arrange(dplyr::desc(MeanCorrelation))

    if (nrow(current_data) == 0) next

    best_models <- current_data %>%
      dplyr::slice_head(n = min(3, nrow(current_data)))

    worst_models <- current_data %>%
      dplyr::arrange(MeanCorrelation) %>%
      dplyr::slice_head(n = min(3, nrow(current_data)))

    cat(
      "For **", current_trait, "** in **", current_cv, "**, the best-performing model was **",
      best_models$FullEffects[1], "** (", fmt3(best_models$MeanCorrelation[1]), ")",
      sep = ""
    )

    if (nrow(best_models) >= 2) {
      cat(
        ", followed by **", best_models$FullEffects[2], "** (",
        fmt3(best_models$MeanCorrelation[2]), ")",
        sep = ""
      )
    }

    if (nrow(best_models) >= 3) {
      cat(
        " and **", best_models$FullEffects[3], "** (",
        fmt3(best_models$MeanCorrelation[3]), ")",
        sep = ""
      )
    }

    cat(
      ". The lowest values in this scenario were observed for **",
      paste0(worst_models$FullEffects, collapse = "**, **"),
      "**, ranging from ",
      fmt3(min(worst_models$MeanCorrelation, na.rm = TRUE)),
      " to ",
      fmt3(max(worst_models$MeanCorrelation, na.rm = TRUE)),
      ".\n\n",
      sep = ""
    )
  }

  for (current_cv in c("CV0", "CV00")) {
    current_data <- cvloo_summary %>%
      dplyr::filter(Trait == current_trait, CV == current_cv)

    if (nrow(current_data) == 0) next

    best_by_env <- current_data %>%
      dplyr::group_by(Env_leave) %>%
      dplyr::slice_max(MeanCorrelation, n = 1, with_ties = FALSE) %>%
      dplyr::ungroup() %>%
      dplyr::arrange(Env_leave)

    worst_by_env <- current_data %>%
      dplyr::group_by(Env_leave) %>%
      dplyr::slice_min(MeanCorrelation, n = 1, with_ties = FALSE) %>%
      dplyr::ungroup() %>%
      dplyr::arrange(Env_leave)

    cat(
      "In **", current_cv, "**, the best-performing models by left-out environment were ",
      paste0(
        "**", best_by_env$FullEffects, "** in **", best_by_env$Env_leave,
        "** (", fmt3(best_by_env$MeanCorrelation), ")",
        collapse = "; "
      ),
      ". Across environments, the lowest values ranged from ",
      fmt3(min(worst_by_env$MeanCorrelation, na.rm = TRUE)),
      " to ",
      fmt3(max(worst_by_env$MeanCorrelation, na.rm = TRUE)),
      ".\n\n",
      sep = ""
    )
  }

  current_w <- dplyr::bind_rows(
    cv12_w_effect %>%
      dplyr::filter(Trait == current_trait) %>%
      dplyr::group_by(CV) %>%
      dplyr::summarise(
        MeanDeltaWithW = mean(DeltaWithW, na.rm = TRUE),
        .groups = "drop"
      ),
    cvloo_w_effect %>%
      dplyr::filter(Trait == current_trait) %>%
      dplyr::group_by(CV) %>%
      dplyr::summarise(
        MeanDeltaWithW = mean(DeltaWithW, na.rm = TRUE),
        .groups = "drop"
      )
  ) %>%
    dplyr::mutate(
      CV = factor(CV, levels = c("CV1", "CV2", "CV0", "CV00"))
    ) %>%
    dplyr::arrange(CV)

  if (nrow(current_w) > 0) {
    cat(
      "Across paired contrasts, the mean effect of adding **W** was ",
      paste0(
        "**", current_w$CV, "**: ",
        fmt3(current_w$MeanDeltaWithW),
        collapse = "; "
      ),
      ".\n\n",
      sep = ""
    )
  }

  current_kernel <- dplyr::bind_rows(
    cv12_kernel_family_compare %>%
      dplyr::filter(Trait == current_trait) %>%
      dplyr::group_by(CV, Pathway, WeatherStatus) %>%
      dplyr::summarise(
        Delta_Gaussian_vs_Linear = mean(Delta_Gaussian_vs_Linear, na.rm = TRUE),
        Delta_ArcCos_vs_Linear = mean(Delta_ArcCos_vs_Linear, na.rm = TRUE),
        .groups = "drop"
      ),
    cvloo_kernel_family_compare %>%
      dplyr::filter(Trait == current_trait) %>%
      dplyr::group_by(CV, Pathway, WeatherStatus) %>%
      dplyr::summarise(
        Delta_Gaussian_vs_Linear = mean(Delta_Gaussian_vs_Linear, na.rm = TRUE),
        Delta_ArcCos_vs_Linear = mean(Delta_ArcCos_vs_Linear, na.rm = TRUE),
        .groups = "drop"
      )
  )

  if (nrow(current_kernel) > 0) {
    cat(
      "Across pathway-specific kernel contrasts, the mean Gaussian-minus-linear difference was ",
      fmt3(mean(current_kernel$Delta_Gaussian_vs_Linear, na.rm = TRUE)),
      " and the mean arc-cosine-minus-linear difference was ",
      fmt3(mean(current_kernel$Delta_ArcCos_vs_Linear, na.rm = TRUE)),
      ".\n\n",
      sep = ""
    )
  }
}
```


## 16. Figures

The final figures summarize predictive ability for the full model catalog in `CV1/CV2` and `CV0/CV00`.

```{r figure-cv12}
if (!exists("cv12_summary")) {
  stop("The object cv12_summary was not loaded before the figure-cv12 chunk.")
}
if (!all(c("Trait", "CV", "MeanCorrelation", "SDCorrelation", "FullEffects", "Group") %in% names(cv12_summary))) {
  stop("cv12_summary does not contain the columns required by the figure-cv12 chunk.")
}

group_levels_en <- c("Genomic", "Phenomic", "Combined")
model_order_wrapped_en <- stringr::str_wrap(model_catalog$FullEffects, width = 22)

cv12_plot_data <- cv12_summary %>%
  mutate(
    FullEffectsWrapped = stringr::str_wrap(FullEffects, width = 22),
    PlotOrder = match(FullEffects, model_catalog$FullEffects),
    Group = dplyr::case_when(
      Pathway == "Genomic pathway" ~ "Genomic",
      Pathway == "Phenomic pathway" ~ "Phenomic",
      Pathway == "Combined G + P pathway" ~ "Combined",
      TRUE ~ as.character(Group)
    ),
    Group = factor(Group, levels = group_levels_en)
  )

cv12_traits <- unique(cv12_plot_data$Trait)
cv12_schemes <- unique(cv12_plot_data$CV)
cv12_plot_objects <- list()

for (current_trait in cv12_traits) {
  for (current_cv in cv12_schemes) {
    trait_cv_data <- cv12_plot_data %>%
      filter(Trait == current_trait, CV == current_cv) %>%
      arrange(PlotOrder) %>%
      mutate(
        FullEffectsWrapped = factor(
          FullEffectsWrapped,
          levels = model_order_wrapped_en
        )
      )

    if (nrow(trait_cv_data) == 0) next

    lower_limit_trait_cv <- min(trait_cv_data$MeanCorrelation - trait_cv_data$SDCorrelation, na.rm = TRUE)
    upper_limit_trait_cv <- max(trait_cv_data$MeanCorrelation + trait_cv_data$SDCorrelation, na.rm = TRUE)
    plot_lower_limit_trait_cv <- min(0, lower_limit_trait_cv - 0.03)
    plot_upper_limit_trait_cv <- upper_limit_trait_cv + 0.03

    plot_key <- paste(current_trait, tolower(current_cv), sep = "__")

    cv12_plot_objects[[plot_key]] <- ggplot(
      trait_cv_data,
      aes(x = FullEffectsWrapped, y = MeanCorrelation, fill = Group)
    ) +
      geom_col(width = 0.75, color = "grey30", na.rm = TRUE) +
      geom_errorbar(
        aes(
          ymin = MeanCorrelation - SDCorrelation,
          ymax = MeanCorrelation + SDCorrelation
        ),
        width = 0.20,
        linewidth = 0.35,
        na.rm = TRUE
      ) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
      scale_x_discrete(name = "Full model effects") +
      scale_y_continuous(
        name = "Predictive correlation",
        expand = expansion(mult = c(0, 0.02))
      ) +
      coord_cartesian(ylim = c(plot_lower_limit_trait_cv, plot_upper_limit_trait_cv)) +
      ggthemes::scale_fill_gdocs(name = "Comparison group") +
      theme_bw() +
      theme(
        plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
        axis.text.x = element_text(size = 7, angle = 60, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 10),
        legend.position = "top",
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)
      ) +
      labs(
        title = paste("Predictive ability in", current_cv, "for", current_trait)
      )

    print(cv12_plot_objects[[plot_key]])
  }
}
```

```{r save-figure-cv12, include=FALSE}
if (!exists("cv12_plot_objects")) {
  stop("The object cv12_plot_objects was not created before the save-figure-cv12 chunk.")
}

for (plot_key in names(cv12_plot_objects)) {
  plot_parts <- strsplit(plot_key, "__", fixed = TRUE)[[1]]
  current_trait <- plot_parts[1]
  current_cv <- plot_parts[2]

  ggsave(
    file.path(figures_dir, paste0("prediction_", current_trait, "_", current_cv, ".tiff")),
    plot = cv12_plot_objects[[plot_key]],
    width = 18,
    height = 11,
    dpi = 300,
    bg = "white"
  )
}
```

```{r figure-cvloo}
if (!exists("cvloo_summary")) {
  stop("The object cvloo_summary was not loaded before the figure-cvloo chunk.")
}
if (!all(c("Trait", "CV", "Env_leave", "MeanCorrelation", "SDCorrelation", "FullEffects", "Group") %in% names(cvloo_summary))) {
  stop("cvloo_summary does not contain the columns required by the figure-cvloo chunk.")
}

group_levels_en <- c("Genomic", "Phenomic", "Combined")
model_order_wrapped_en <- stringr::str_wrap(model_catalog$FullEffects, width = 22)

cvloo_plot_data <- cvloo_summary %>%
  mutate(
    FullEffectsWrapped = stringr::str_wrap(FullEffects, width = 22),
    Env_leave = factor(Env_leave, levels = c("CS11_WS", "CS11_WW", "CS12_WS", "CS12_WW")),
    PlotOrder = match(FullEffects, model_catalog$FullEffects),
    Group = dplyr::case_when(
      Pathway == "Genomic pathway" ~ "Genomic",
      Pathway == "Phenomic pathway" ~ "Phenomic",
      Pathway == "Combined G + P pathway" ~ "Combined",
      TRUE ~ as.character(Group)
    ),
    Group = factor(Group, levels = group_levels_en)
  )

cvloo_traits <- unique(cvloo_plot_data$Trait)
cvloo_schemes <- unique(cvloo_plot_data$CV)
cvloo_plot_objects <- list()

for (current_trait in cvloo_traits) {
  for (current_cv in cvloo_schemes) {
    trait_cv_data <- cvloo_plot_data %>%
      filter(Trait == current_trait, CV == current_cv)

    if (nrow(trait_cv_data) == 0) next

    trait_cv_data <- trait_cv_data %>%
      arrange(Env_leave, PlotOrder) %>%
      mutate(
        FullEffectsWrapped = factor(
          FullEffectsWrapped,
          levels = model_order_wrapped_en
        )
      )

    lower_limit_trait_cv <- min(trait_cv_data$MeanCorrelation - trait_cv_data$SDCorrelation, na.rm = TRUE)
    upper_limit_trait_cv <- max(trait_cv_data$MeanCorrelation + trait_cv_data$SDCorrelation, na.rm = TRUE)
    plot_lower_limit_trait_cv <- min(0, lower_limit_trait_cv - 0.03)
    plot_upper_limit_trait_cv <- upper_limit_trait_cv + 0.03

    plot_key <- paste(current_trait, current_cv, sep = "__")

    cvloo_plot_objects[[plot_key]] <- ggplot(
      trait_cv_data,
      aes(x = FullEffectsWrapped, y = MeanCorrelation, fill = Group)
    ) +
      geom_col(width = 0.75, color = "grey30", na.rm = TRUE) +
      geom_errorbar(
        aes(
          ymin = MeanCorrelation - SDCorrelation,
          ymax = MeanCorrelation + SDCorrelation
        ),
        width = 0.20,
        linewidth = 0.35,
        na.rm = TRUE
      ) +
      facet_wrap(~ Env_leave, ncol = 2) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
      scale_x_discrete(name = "Full model effects") +
      scale_y_continuous(
        name = "Predictive correlation",
        expand = expansion(mult = c(0, 0.02))
      ) +
      coord_cartesian(ylim = c(plot_lower_limit_trait_cv, plot_upper_limit_trait_cv)) +
      ggthemes::scale_fill_gdocs(name = "Comparison group") +
      theme_bw() +
      theme(
        plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
        strip.text = element_text(face = "bold", size = 9),
        axis.text.x = element_text(size = 6.5, angle = 60, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 10),
        legend.position = "top",
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)
      ) +
      labs(
        title = paste("Predictive ability in", current_cv, "for", current_trait)
      )

    print(cvloo_plot_objects[[plot_key]])
  }
}
```

```{r save-figure-cvloo, include=FALSE}
if (!exists("cvloo_plot_objects")) {
  stop("The object cvloo_plot_objects was not created before the save-figure-cvloo chunk.")
}

for (plot_key in names(cvloo_plot_objects)) {
  plot_parts <- strsplit(plot_key, "__", fixed = TRUE)[[1]]
  current_trait <- plot_parts[1]
  current_cv <- plot_parts[2]

  ggsave(
    file.path(figures_dir, paste0("prediction_", current_trait, "_", current_cv, ".tiff")),
    plot = cvloo_plot_objects[[plot_key]],
    width = 18,
    height = 12,
    dpi = 300,
    bg = "white"
  )
}
```

## 17. Final note

This didactic version is aligned with the current prediction workflow based on `analysis_prediction_run_outputs.Rmd`. It expects files in which:

- repetition and fold are encoded both in the folder hierarchy and in the file name;
- `CV0` and `CV00` carry the left-out environment label both in the folder hierarchy and in the file name;
- prediction CSVs are read recursively from `output/results/`;
- `CV0` and `CV00` still require the filter `Env == Env_leave` for the left-out-environment summaries.

After running this file, the next step in the repository is to run the manuscript-oriented script that consumes the summary tables written to `output/tables/`.

## Step navigation

**Previous step:** [Prediction — Part II](analysis_prediction_run_outputs.html)  
**Next site reference:** [Home](index.html)

Outside the site menu, the next analytical step is to generate manuscript-oriented summaries and review tables from the outputs created here.

