---
title: "Prediction summaries and figures - didactic tutorial"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
---

**Language / Idioma:** [English](visualization.html) | [Português](visualization_pt.html)

This file is the **didactic version** of the final visualization workflow. Its role is to read the prediction outputs generated by `analysis_prediction_run_outputs.Rmd`, reorganize them into tidy summary tables, and then build the figures and compact displays used later in the manuscript.

The current prediction workflow saves raw prediction output files in `output/results/` using the following patterns:

- `CV1_<trait>_<model>_rep<rep>_fold<fold>.csv`
- `CV2_<trait>_<model>_rep<rep>_fold<fold>.csv`
- `CV0_<trait>_<Env_leave>_<model>_rep<rep>_fold<fold>.csv`
- `CV00_<trait>_<Env_leave>_<model>_rep<rep>_fold<fold>.csv`

In this workflow, repetition and fold are recovered directly from the filename. For `CV0` and `CV00`, the held-out environment label is also recovered from the filename and later matched to `Env` through the filter `Env == Env_leave`.

In line with the current analytical interpretation of the repository, the **weather kernel (`W`)** should be read as a weather-derived source of similarity that complements, but does not replace, the categorical environment effect (`E`). The phenomic component (`P`) should also be interpreted with caution, because NIRS-based similarity may capture strong physiological response to the environment in addition to potentially useful predictive structure.

### What this script does

The workflow is organized around four practical tasks:

1. locate and read the prediction CSV files;
2. parse the filenames and recover trait, model, repetition, and leave-one-environment-out information;
3. summarize results for `CV1`, `CV2`, `CV0`, and `CV00`;
4. export clean tables and figures for the next manuscript-oriented stage.

To make interpretation easier, the models are ordered to help answer three practical questions:

1. does moving from linear to Gaussian or arc-cosine kernels improve predictive ability?
2. does adding the weather kernel (`W`) improve predictive ability?
3. does combining genomic and phenomic information improve predictive ability?

## Inputs, outputs and next step

**Inputs of this step**
- raw prediction outputs written by `analysis_prediction_run_outputs.Rmd`
- light prediction metadata tables generated in the previous stage

**Outputs of this step**
- final summary tables
- model rankings
- W-effect comparisons
- kernel-family comparisons
- final figures for downstream interpretation

**Next step**
- the outputs generated here feed the manuscript-oriented summaries and the final interpretation of the project results.

## 1. Setup

We start by defining the default chunk options for the rendered report. The goal is to keep code, tables, and figures readable when the tutorial is rendered in HTML.

This chunk sets the global knitting options for the document so that the rendered tables and figures follow a consistent presentation style.

```{r setup}
knitr::opts_chunk$set( echo = TRUE,
 warning = FALSE,
 message = FALSE,
 fig.width = 13,
 fig.height = 8,
  out.width = "100%" )
```

## 2. Load packages and locate project folders

This section loads the packages used for data manipulation, figure construction, and table formatting. It also checks whether the script is being executed from the project root or from `analysis/`, and then standardizes the output paths.

This chunk loads the plotting and data-wrangling packages, locates the project folders, and prepares the output directories used by the visualization module.

```{r packages-and-paths}
rm(list = ls())
options(scipen = 999)
packages_needed <- c( "dplyr",
 "tidyr",
 "readr",
 "tibble",
 "ggplot2",
 "ggthemes",
 "knitr",
 "kableExtra",
  "stringr" )
packages_missing <- packages_needed[!(packages_needed %in% installed.packages()[, "Package"])]
if (length(packages_missing) > 0) {
  install.packages(packages_missing)
}
library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(ggplot2)
library(ggthemes)
library(knitr)
library(kableExtra)
library(stringr)
project_root <- getwd()
if (!dir.exists(file.path(project_root, "output")) && dir.exists(file.path(project_root, "..", "output"))) {
  project_root <- normalizePath(file.path(project_root, ".."))
}
if (!dir.exists(file.path(project_root, "output"))) {
  stop("Could not find the folder 'output/'.\nRun this script from the project root or from analysis/.")
}
results_dir <- file.path(project_root, "output", "results")
figures_dir <- file.path(project_root, "output", "figures")
tables_dir <- file.path(project_root, "output", "tables")
dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tables_dir, recursive = TRUE, showWarnings = FALSE)
```

## 3. Model catalog used in the reduced pipeline

The visualization stage uses the same reduced catalog of 18 models adopted in the final prediction workflow. Here the catalog is reorganized to emphasize analytical pathway, kernel family, and the presence or absence of weather information.

This chunk defines the reduced model catalog used to annotate summaries, figures, and comparison tables throughout the document.

```{r model-catalog}
model_catalog <- tibble( ModelNumber = c(1, 10, 3, 13, 5, 16, 2, 11, 4, 14, 6, 17, 7, 12, 8, 15, 9, 18),
 Model = sprintf("M%02d", c(1, 10, 3, 13, 5, 16, 2, 11, 4, 14, 6, 17, 7, 12, 8, 15, 9, 18)),
 Pathway = c( rep("Genomic pathway", 6), rep("Phenomic pathway", 6), rep("Combined G + P pathway", 6) ),
 KernelFamily = c( "Linear", "Linear", "Gaussian", "Gaussian", "Arc-cosine", "Arc-cosine", "Linear", "Linear", "Gaussian", "Gaussian", "Arc-cosine", "Arc-cosine", "Linear", "Linear", "Gaussian", "Gaussian", "Arc-cosine", "Arc-cosine" ),
 WeatherStatus = c( "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W", "Without W", "With W" ),
 Group = c( "Genomic baseline", "Weather single-source", "Genomic baseline", "Weather single-source", "Genomic baseline", "Weather single-source", "Phenomic baseline", "Weather single-source", "Phenomic baseline", "Weather single-source", "Phenomic baseline", "Weather single-source", "Combined baseline", "Weather combined", "Combined baseline", "Weather combined", "Combined baseline", "Weather combined" ),
 Description = c( "Genomic linear + E", "Genomic linear + W", "Genomic Gaussian + E", "Genomic Gaussian + W", "Genomic arc-cosine + E", "Genomic arc-cosine + W", "Phenomic linear + E", "Phenomic linear + W", "Phenomic Gaussian + E", "Phenomic Gaussian + W", "Phenomic arc-cosine + E", "Phenomic arc-cosine + W", "Combined linear + E", "Combined linear + W", "Combined Gaussian + E", "Combined Gaussian + W", "Combined arc-cosine + E", "Combined arc-cosine + W" ),
  ShortDescription = c( "G + E", "G + W", "GGK + E", "GGK + W", "GAK + E", "GAK + W", "P + E", "P + W", "PGK + E", "PGK + W", "PAK + E", "PAK + W", "G+P + E", "G+P + W", "GGK+PGK + E", "GGK+PGK + W", "GAK+PAK + E", "GAK+PAK + W" ) )
model_catalog
```

## 4. Comparison pairs used for W versus no-W contrasts

This auxiliary table defines the pairwise contrasts between models that differ only by the inclusion of the weather kernel (`W`). These comparisons are later exported as manuscript-ready analytical tables.

This chunk specifies the model pairs compared in the weather-versus-no-weather contrasts used later in the tutorial.

```{r comparison-pairs}
comparison_pairs <- tibble( Comparison = c( "Genomic linear", "Genomic Gaussian", "Genomic arc-cosine", "Phenomic linear", "Phenomic Gaussian", "Phenomic arc-cosine", "Combined linear", "Combined Gaussian", "Combined arc-cosine" ),
 WithoutW = c("M01", "M03", "M05", "M02", "M04", "M06", "M07", "M08", "M09"),
  WithW = c("M10", "M13", "M16", "M11", "M14", "M17", "M12", "M15", "M18") )
comparison_pairs
```

## 5. Locate prediction files

At this point the script scans `output/results/`, separates `CV1/CV2` files from `CV0/CV00` files, and stops early if no prediction outputs are found.

This chunk finds the prediction result files generated by the previous module and separates them according to the validation scenario they belong to.

```{r locate-files}
all_result_files <- list.files(results_dir, pattern = "\\.csv$", full.names = TRUE)
all_result_names <- basename(all_result_files)
cv12_files <- all_result_files[grepl("^CV1_|^CV2_", all_result_names)]
cvloo_files <- all_result_files[grepl("^CV0_|^CV00_", all_result_names)]
cat("Number of CV1/CV2 files:", length(cv12_files), "\n")
cat("Number of CV0/CV00 files:", length(cvloo_files), "\n")
if (length(cv12_files) == 0 && length(cvloo_files) == 0) {
  stop("No prediction CSV files were found in output/results/.\nRun analysis_prediction_run_outputs first.")
}
```

## 6. Read CV1 and CV2 files generated by the final prediction script

The next step reads the `CV1` and `CV2` files, parses the filename, and reconstructs the metadata needed for downstream summaries. In this workflow, repetition and fold are recovered directly from the filename.

This chunk reads the CV1 and CV2 files, parses their names, and reconstructs the metadata needed for downstream summaries.

```{r read-cv12}
cv12_list <- vector("list", length(cv12_files))
for (i in seq_along(cv12_files)) {
  current_file <- cv12_files[i]
  file_name <- basename(current_file)
  file_name <- sub("\.csv$", "", file_name)
  parts <- strsplit(file_name, "_")[[1]]

  if (length(parts) != 5) {
    stop(
      "Unexpected filename for CV1/CV2 file: ",
      file_name,
      ".
Expected pattern CV1_<trait>_<model>_rep<rep>_fold<fold>.csv or CV2_<trait>_<model>_rep<rep>_fold<fold>.csv"
    )
  }

  current_cv <- parts[1]
  current_trait <- parts[2]
  current_model_label <- parts[3]
  current_model_number <- as.numeric(gsub("M", "", current_model_label))
  current_rep <- as.numeric(gsub("rep", "", parts[4]))
  current_fold <- as.numeric(gsub("fold", "", parts[5]))

  current_data <- readr::read_csv(current_file, show_col_types = FALSE)

  if (!("Env" %in% names(current_data)) || !("cor" %in% names(current_data))) {
    stop(
      "Unexpected column structure in file: ",
      file_name,
      ".
The final prediction script should export columns including Env and cor."
    )
  }

  current_data <- current_data %>%
    mutate(
      Fold = current_fold,
      CV = current_cv,
      Trait = current_trait,
      ModelNumber = current_model_number,
      Model = current_model_label,
      Rep = current_rep
    )

  cv12_list[[i]] <- current_data
}

cv12_raw <- bind_rows(cv12_list) %>%
  left_join(model_catalog, by = c("ModelNumber", "Model"))

readr::write_csv(cv12_raw, file.path(tables_dir, "prediction_cv1_cv2_raw.csv"))
```

## 7. Read CV0 and CV00 files generated by the final prediction script

The logic is analogous to the previous section, but now the filename also carries the label of the held-out environment. In the final workflow, that label may contain multiple underscores, so the parser reconstructs `Env_leave` from the filename by reading from right to left.

This chunk reads the CV0 and CV00 files, parses the held-out environment labels, and prepares the data for leave-one-environment-out summaries.

```{r read-cvloo}
cvloo_list <- vector("list", length(cvloo_files))
for (i in seq_along(cvloo_files)) {
  current_file <- cvloo_files[i]
  file_name <- basename(current_file)
  file_name <- sub("\.csv$", "", file_name)
  parts <- strsplit(file_name, "_")[[1]]

  if (length(parts) < 6) {
    stop(
      "Unexpected filename for CV0/CV00 file: ",
      file_name,
      ".
Expected pattern CV0_<trait>_<Env_leave>_<model>_rep<rep>_fold<fold>.csv or CV00_<trait>_<Env_leave>_<model>_rep<rep>_fold<fold>.csv"
    )
  }

  current_cv <- parts[1]
  current_trait <- parts[2]
  current_model_label <- parts[length(parts) - 2]
  current_model_number <- as.numeric(gsub("M", "", current_model_label))
  current_rep <- as.numeric(gsub("rep", "", parts[length(parts) - 1]))
  current_fold <- as.numeric(gsub("fold", "", parts[length(parts)]))
  current_env_leave <- paste(parts[3:(length(parts) - 3)], collapse = "_")

  current_data <- readr::read_csv(current_file, show_col_types = FALSE)

  if (!("Env" %in% names(current_data)) || !("cor" %in% names(current_data))) {
    stop(
      "Unexpected column structure in file: ",
      file_name,
      ".
The final prediction script should export columns including Env and cor."
    )
  }

  current_data <- current_data %>%
    mutate(
      Fold = current_fold,
      CV = current_cv,
      Trait = current_trait,
      Env_leave = current_env_leave,
      ModelNumber = current_model_number,
      Model = current_model_label,
      Rep = current_rep
    )

  cvloo_list[[i]] <- current_data
}
cvloo_raw <- bind_rows(cvloo_list) %>%
  left_join(model_catalog, by = c("ModelNumber", "Model"))
readr::write_csv(cvloo_raw, file.path(tables_dir, "prediction_cv0_cv00_raw.csv"))
readr::write_csv(model_catalog, file.path(tables_dir, "prediction_model_catalog.csv"))
```

## 8. Summarize CV1 and CV2

This stage builds the summaries for `CV1` and `CV2` in a stepwise way: first by environment and fold, then by environment within repetition, then by repetition, and finally by model-level means and standard deviations across repetitions.

This chunk builds the hierarchical summaries for CV1 and CV2, moving from raw files to environment-level, repetition-level, and model-level results.

```{r summarize-cv12}
cv12_env_fold <- cv12_raw %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Rep,
   Fold,
    Env ) %>%
  summarise( EnvCorrelation = mean(cor, na.rm = TRUE),
   ValidationCount = sum(!is.na(cor)),
   MeanRuntimeSec = mean(Runtime_sec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( EnvCorrelation = ifelse(is.nan(EnvCorrelation), NA_real_, EnvCorrelation),
    MeanRuntimeSec = ifelse(is.nan(MeanRuntimeSec), NA_real_, MeanRuntimeSec) ) %>%
  filter(!is.na(EnvCorrelation), ValidationCount > 0)
cv12_env_rep <- cv12_env_fold %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Rep,
    Env ) %>%
  summarise( EnvCorrelation = weighted.mean(EnvCorrelation, w = ValidationCount, na.rm = TRUE),
   ValidationCount = sum(ValidationCount, na.rm = TRUE),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" )
cv12_rep_summary <- cv12_env_rep %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
    Rep ) %>%
  summarise( WeightedCorrelation = weighted.mean(EnvCorrelation, w = ValidationCount, na.rm = TRUE),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" )
cv12_summary <- cv12_rep_summary %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
    CV ) %>%
  summarise( MeanCorrelation = mean(WeightedCorrelation, na.rm = TRUE),
   SDCorrelation = sd(WeightedCorrelation, na.rm = TRUE),
   NumberOfRepetitions = n(),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( SDCorrelation = ifelse(is.na(SDCorrelation), 0, SDCorrelation) )
readr::write_csv(cv12_env_fold, file.path(tables_dir, "prediction_cv1_cv2_by_rep_fold_env.csv"))
readr::write_csv(cv12_env_rep, file.path(tables_dir, "prediction_cv1_cv2_by_rep_env.csv"))
readr::write_csv(cv12_rep_summary, file.path(tables_dir, "prediction_cv1_cv2_by_rep.csv"))
readr::write_csv(cv12_summary, file.path(tables_dir, "prediction_cv1_cv2_summary.csv"))
```

## 9. Summarize CV0 and CV00

For leave-one-environment-out scenarios, the script first restricts the summaries to rows in which `Env == Env_leave`. This keeps only the held-out-environment component that is conceptually relevant for `CV0` and `CV00`.

This chunk builds the corresponding summaries for CV0 and CV00, keeping the held-out environment explicit in the final tables.

```{r summarize-cvloo}
cvloo_env_filtered <- cvloo_raw %>%
  filter(Env == Env_leave)
cvloo_env_fold <- cvloo_env_filtered %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Env_leave,
   Rep,
   Fold,
    Env ) %>%
  summarise( LOEOCorrelation = mean(cor, na.rm = TRUE),
   ValidationCount = sum(!is.na(cor)),
   MeanRuntimeSec = mean(Runtime_sec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( LOEOCorrelation = ifelse(is.nan(LOEOCorrelation), NA_real_, LOEOCorrelation),
    MeanRuntimeSec = ifelse(is.nan(MeanRuntimeSec), NA_real_, MeanRuntimeSec) ) %>%
  filter(!is.na(LOEOCorrelation), ValidationCount > 0)
cvloo_rep_summary <- cvloo_env_fold %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
   Env_leave,
    Rep ) %>%
  summarise( LOEOCorrelation = weighted.mean(LOEOCorrelation, w = ValidationCount, na.rm = TRUE),
   ValidationCount = sum(ValidationCount, na.rm = TRUE),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" )
cvloo_summary <- cvloo_rep_summary %>%
  group_by( ModelNumber,
   Model,
   Group,
   Pathway,
   KernelFamily,
   WeatherStatus,
   Description,
   ShortDescription,
   Trait,
   CV,
    Env_leave ) %>%
  summarise( MeanCorrelation = mean(LOEOCorrelation, na.rm = TRUE),
   SDCorrelation = sd(LOEOCorrelation, na.rm = TRUE),
   NumberOfRuns = n(),
   MeanRuntimeSec = mean(MeanRuntimeSec, na.rm = TRUE),
    .groups = "drop" ) %>%
  mutate( SDCorrelation = ifelse(is.na(SDCorrelation), 0, SDCorrelation) )
readr::write_csv(cvloo_env_fold, file.path(tables_dir, "prediction_cv0_cv00_by_rep_fold_env.csv"))
readr::write_csv(cvloo_rep_summary, file.path(tables_dir, "prediction_cv0_cv00_by_run.csv"))
readr::write_csv(cvloo_summary, file.path(tables_dir, "prediction_cv0_cv00_summary.csv"))
```

## 10. Comparison tables

This section derives secondary analytical tables, especially the contrast between models with and without `W`, and the comparison between kernel families within each pathway.

This chunk constructs the derived comparison tables used to evaluate the effect of `W` and to compare kernel families within each pathway.

```{r comparison-tables}
cv12_w_effect <- comparison_pairs %>%
  left_join( cv12_summary %>% select(Trait, CV, Model, MeanCorrelation, SDCorrelation),
    by = c("WithoutW" = "Model") ) %>%
  rename( MeanWithoutW = MeanCorrelation, SDWithoutW = SDCorrelation ) %>%
  left_join( cv12_summary %>% select(Trait, CV, Model, MeanCorrelation, SDCorrelation),
    by = c("WithW" = "Model", "Trait", "CV") ) %>%
  rename( MeanWithW = MeanCorrelation, SDWithW = SDCorrelation ) %>%
  mutate( DeltaWithW = MeanWithW - MeanWithoutW ) %>%
  arrange(Trait, CV, Comparison)
cvloo_w_effect <- comparison_pairs %>%
  left_join( cvloo_summary %>% select(Trait, CV, Env_leave, Model, MeanCorrelation, SDCorrelation),
    by = c("WithoutW" = "Model") ) %>%
  rename( MeanWithoutW = MeanCorrelation, SDWithoutW = SDCorrelation ) %>%
  left_join( cvloo_summary %>% select(Trait, CV, Env_leave, Model, MeanCorrelation, SDCorrelation),
    by = c("WithW" = "Model", "Trait", "CV", "Env_leave") ) %>%
  rename( MeanWithW = MeanCorrelation, SDWithW = SDCorrelation ) %>%
  mutate( DeltaWithW = MeanWithW - MeanWithoutW ) %>%
  arrange(Trait, CV, Env_leave, Comparison)
cv12_kernel_family_compare <- cv12_summary %>%
  select(Trait, CV, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  mutate( Delta_Gaussian_vs_Linear = Gaussian - Linear,
   Delta_ArcCos_vs_Linear = `Arc-cosine` - Linear,
    Delta_ArcCos_vs_Gaussian = `Arc-cosine` - Gaussian ) %>%
  arrange(Trait, CV, Pathway, WeatherStatus)
cvloo_kernel_family_compare <- cvloo_summary %>%
  select(Trait, CV, Env_leave, Pathway, WeatherStatus, KernelFamily, MeanCorrelation) %>%
  pivot_wider(names_from = KernelFamily, values_from = MeanCorrelation) %>%
  mutate( Delta_Gaussian_vs_Linear = Gaussian - Linear,
   Delta_ArcCos_vs_Linear = `Arc-cosine` - Linear,
    Delta_ArcCos_vs_Gaussian = `Arc-cosine` - Gaussian ) %>%
  arrange(Trait, CV, Env_leave, Pathway, WeatherStatus)
readr::write_csv(cv12_w_effect, file.path(tables_dir, "prediction_cv1_cv2_w_effect_pairs.csv"))
readr::write_csv(cvloo_w_effect, file.path(tables_dir, "prediction_cv0_cv00_w_effect_pairs.csv"))
readr::write_csv(cv12_kernel_family_compare,
  file.path(tables_dir, "prediction_cv1_cv2_kernel_family_compare.csv"))
readr::write_csv(cvloo_kernel_family_compare,
  file.path(tables_dir, "prediction_cv0_cv00_kernel_family_compare.csv"))
```

## 11. Compact display tables

These objects are simplified versions of the full summaries. They are useful for manuscript tables, supplementary material, and quick inspection of the best-performing models.

This chunk prepares the compact display tables and Top-3 rankings that are easier to inspect than the full summary outputs.

```{r display-tables}
cv12_table_display <- cv12_summary %>%
  arrange(Trait, CV, Pathway, KernelFamily, WeatherStatus) %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3),
   SDCorrelation = round(SDCorrelation, 3),
    MeanRuntimeSec = round(MeanRuntimeSec, 2) ) %>%
  select( Trait,
   CV,
   Pathway,
   KernelFamily,
   WeatherStatus,
   ShortDescription,
   MeanCorrelation,
   SDCorrelation,
   NumberOfRepetitions,
    MeanRuntimeSec )
cvloo_table_display <- cvloo_summary %>%
  arrange(Trait, CV, Env_leave, Pathway, KernelFamily, WeatherStatus) %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3),
   SDCorrelation = round(SDCorrelation, 3),
    MeanRuntimeSec = round(MeanRuntimeSec, 2) ) %>%
  select( Trait,
   CV,
   Env_leave,
   Pathway,
   KernelFamily,
   WeatherStatus,
   ShortDescription,
   MeanCorrelation,
   SDCorrelation,
   NumberOfRuns,
    MeanRuntimeSec )
cv12_top3 <- cv12_summary %>%
  group_by(Trait, CV) %>%
  slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3), SDCorrelation = round(SDCorrelation, 3) ) %>%
  select(Trait, CV, ShortDescription, MeanCorrelation, SDCorrelation)
cvloo_top3 <- cvloo_summary %>%
  group_by(Trait, CV, Env_leave) %>%
  slice_max(order_by = MeanCorrelation, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  mutate( MeanCorrelation = round(MeanCorrelation, 3), SDCorrelation = round(SDCorrelation, 3) ) %>%
  select(Trait, CV, Env_leave, ShortDescription, MeanCorrelation, SDCorrelation)
readr::write_csv(cv12_table_display, file.path(tables_dir, "prediction_cv1_cv2_display_table.csv"))
readr::write_csv(cvloo_table_display, file.path(tables_dir, "prediction_cv0_cv00_display_table.csv"))
readr::write_csv(cv12_top3, file.path(tables_dir, "prediction_cv1_cv2_top3.csv"))
readr::write_csv(cvloo_top3, file.path(tables_dir, "prediction_cv0_cv00_top3.csv"))
```

## 12. Top-3 tables

The next two chunks display the three best-performing models in each validation scheme.

This chunk renders the Top-3 table for CV1 and CV2 so that the best-performing models can be inspected directly in the document.

```{r table-cv12-top3}
cv12_top3 %>%
  knitr::kable( caption = "Top 3 models for each trait and validation scheme in CV1 and CV2.",
    align = "llccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" )
```

This chunk renders the Top-3 table for CV0 and CV00, stratified by held-out environment.

```{r table-cvloo-top3}
cvloo_top3 %>%
  knitr::kable( caption = "Top 3 models for each trait, validation scheme, and held-out environment in CV0 and CV00.",
    align = "lllccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" )
```

## 13. W-effect tables

These tables present the pairwise comparisons between models without and with weather information.

This chunk renders the pairwise comparison table that contrasts models without and with the weather kernel in CV1 and CV2.

```{r table-cv12-w-effect}
cv12_w_effect %>%
  mutate( MeanWithoutW = round(MeanWithoutW, 3),
   MeanWithW = round(MeanWithW, 3),
    DeltaWithW = round(DeltaWithW, 3) ) %>%
  knitr::kable( caption = "Pairwise comparison of models without and with the weather kernel (W) in CV1 and CV2.",
    align = "lllccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

This chunk renders the analogous weather-effect comparison table for CV0 and CV00.

```{r table-cvloo-w-effect}
cvloo_w_effect %>%
  mutate( MeanWithoutW = round(MeanWithoutW, 3),
   MeanWithW = round(MeanWithW, 3),
    DeltaWithW = round(DeltaWithW, 3) ) %>%
  knitr::kable( caption = "Pairwise comparison of models without and with the weather kernel (W) in CV0 and CV00.",
    align = "llllccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

## 14. Kernel-family comparison tables

Here the script contrasts linear, Gaussian, and arc-cosine kernels within each analytical pathway.

This chunk renders the table that compares linear, Gaussian, and arc-cosine kernels within CV1 and CV2.

```{r table-cv12-kernel-family}
cv12_kernel_family_compare %>%
  mutate( Linear = round(Linear, 3),
   Gaussian = round(Gaussian, 3),
   `Arc-cosine` = round(`Arc-cosine`, 3),
   Delta_Gaussian_vs_Linear = round(Delta_Gaussian_vs_Linear, 3),
   Delta_ArcCos_vs_Linear = round(Delta_ArcCos_vs_Linear, 3),
    Delta_ArcCos_vs_Gaussian = round(Delta_ArcCos_vs_Gaussian, 3) ) %>%
  knitr::kable( caption = "Comparison of kernel families within each analytical pathway in CV1 and CV2.",
    align = "llllcccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

This chunk renders the corresponding kernel-family comparison table for CV0 and CV00.

```{r table-cvloo-kernel-family}
cvloo_kernel_family_compare %>%
  mutate( Linear = round(Linear, 3),
   Gaussian = round(Gaussian, 3),
   `Arc-cosine` = round(`Arc-cosine`, 3),
   Delta_Gaussian_vs_Linear = round(Delta_Gaussian_vs_Linear, 3),
   Delta_ArcCos_vs_Linear = round(Delta_ArcCos_vs_Linear, 3),
    Delta_ArcCos_vs_Gaussian = round(Delta_ArcCos_vs_Gaussian, 3) ) %>%
  knitr::kable( caption = "Comparison of kernel families within each analytical pathway in CV0 and CV00.",
    align = "lllllcccccc" ) %>%
  kableExtra::kable_styling( bootstrap_options = c("striped", "hover", "condensed", "responsive"),
   full_width = FALSE,
    position = "left" ) %>%
  kableExtra::scroll_box(width = "100%", height = "320px")
```

## 15. Figures

The final figures summarize predictive ability for the full model catalog under `CV1/CV2` and `CV0/CV00`.

This chunk builds the main figure for CV1 and CV2, combining mean predictive correlations and uncertainty bars across models.

```{r figure-cv12}
cv12_plot_data <- cv12_summary %>%
  mutate( ShortDescription = factor(ShortDescription, levels = model_catalog$ShortDescription) )
lower_limit_cv12 <- min(cv12_plot_data$MeanCorrelation - cv12_plot_data$SDCorrelation, na.rm = TRUE)
upper_limit_cv12 <- max(cv12_plot_data$MeanCorrelation + cv12_plot_data$SDCorrelation, na.rm = TRUE)
p_cv12 <- ggplot(cv12_plot_data, aes(x = ShortDescription, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30") +
  geom_errorbar( aes( ymin = MeanCorrelation - SDCorrelation, ymax = MeanCorrelation + SDCorrelation ),
   width = 0.20,
    linewidth = 0.35 ) +
  facet_grid(Trait ~ CV) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  coord_cartesian(ylim = c(lower_limit_cv12 - 0.03, upper_limit_cv12 + 0.03)) +
  scale_x_discrete(name = "Model description") +
  scale_y_continuous(name = "Predictive correlation") +
  ggthemes::scale_fill_gdocs(name = "Comparison group") +
  theme_bw() +
  theme( plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
   strip.text = element_text(face = "bold", size = 10),
   axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7),
   axis.text.y = element_text(size = 9),
   axis.title = element_text(size = 10),
   legend.position = "top",
   legend.title = element_text(size = 9),
    legend.text = element_text(size = 8) ) +
  labs(title = "Predictive ability under CV1 and CV2")
p_cv12
ggsave(file.path(figures_dir, "prediction_cv1_cv2.tiff"),
 plot = p_cv12,
 width = 14,
 height = 8,
 dpi = 300,
  bg = "white")
ggsave(file.path(figures_dir, "prediction_cv1_cv2.pdf"), plot = p_cv12, width = 14, height = 8, bg = "white")
```

This chunk builds the final figure for CV0 and CV00, keeping the held-out environments explicit in the layout.

```{r figure-cvloo}
cvloo_plot_data <- cvloo_summary %>%
  mutate( ShortDescription = factor(ShortDescription, levels = model_catalog$ShortDescription),
    Env_leave = factor(Env_leave, levels = c("CS11_WS", "CS11_WW", "CS12_WS", "CS12_WW")) )
lower_limit_cvloo <- min(cvloo_plot_data$MeanCorrelation - cvloo_plot_data$SDCorrelation, na.rm = TRUE)
upper_limit_cvloo <- max(cvloo_plot_data$MeanCorrelation + cvloo_plot_data$SDCorrelation, na.rm = TRUE)
p_cvloo <- ggplot(cvloo_plot_data, aes(x = ShortDescription, y = MeanCorrelation, fill = Group)) +
  geom_col(width = 0.75, color = "grey30") +
  geom_errorbar( aes( ymin = MeanCorrelation - SDCorrelation, ymax = MeanCorrelation + SDCorrelation ),
   width = 0.20,
   linewidth = 0.35,
    na.rm = TRUE ) +
  facet_grid(Env_leave ~ Trait + CV) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  coord_cartesian(ylim = c(lower_limit_cvloo - 0.03, upper_limit_cvloo + 0.03)) +
  scale_x_discrete(name = "Model description") +
  scale_y_continuous(name = "Predictive correlation") +
  ggthemes::scale_fill_gdocs(name = "Comparison group") +
  theme_bw() +
  theme( plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
   strip.text = element_text(face = "bold", size = 9),
   axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
   axis.text.y = element_text(size = 8),
   axis.title = element_text(size = 10),
   legend.position = "top",
   legend.title = element_text(size = 9),
    legend.text = element_text(size = 8) ) +
  labs(title = "Predictive ability under CV0 and CV00")
p_cvloo
ggsave(file.path(figures_dir, "prediction_cv0_cv00.tiff"),
 plot = p_cvloo,
 width = 16,
 height = 12,
 dpi = 300,
  bg = "white")
ggsave(file.path(figures_dir, "prediction_cv0_cv00.pdf"),
 plot = p_cvloo,
 width = 16,
 height = 12,
  bg = "white")
```

## 16. Final notes

This tutorialized version is aligned with the current prediction workflow based on `analysis_prediction_run_outputs.Rmd`. It expects files in which:

- repetition and fold are encoded in the filename;
- `CV0` and `CV00` carry the held-out environment label in the filename;
- `CV0` and `CV00` still require the `Env == Env_leave` filter for the held-out-environment summaries.

After running this file, the next repository step is to run the manuscript-oriented script that consumes the summary tables written to `output/tables/`.

## Stage navigation

**Previous stage:** [Prediction — Part II](analysis_prediction_run_outputs.html)  
**Next reference in the site:** [Home](index.html)

Outside the site menu, the next analytical step is to generate manuscript-oriented summaries and review tables from the outputs created here.
